### 2023-02-04 Analysis.rmd
initial commit with setup to load forecasts

---
title: "Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
library(here)
library(magrittr)
library(scoringutils)
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(ggridges)
library(RColorBrewer)
library(patchwork)
library(stringr)
library(covid.german.forecasts)
library(grates)
library(ggdist)
library(ggthemes)
library(scales)
library(dplyr)
library(tidyr)
library(forcats)

```

### 2023-02-04 get-paper-data.R

@@ -1,0 +1,228 @@
+## ========================================================================== ##
+##               Code to refetch data from                   ##
+## ========================================================================== ##
+
+library(covidHubUtils)
+library(dplyr)
+library(data.table)
+library(purrr)
+library(scoringutils)
+library(here)
+library(stringr)
+
+analysis_date <- "2022-12-12"
+
+## -------------------------------------------------------------------------- ##  
+##                  Download and process Forecast Hub data                    ##
+## -------------------------------------------------------------------------- ##
+
+# -------------------- Download Forecast Hub folder using SVN ---------------- #
+# svn checkout https://github.com/epiforecasts/covid19-forecast-hub-europe/trunk/data-processed
+
+# svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/submissions
+
+
+# ---------------load truth data using the covidHubutils package ------------- #
+if (file.exists("data/weekly-truth-Europe.csv")) {
+  truth <- fread("data/weekly-truth-Europe.csv")
+} else {
+  truth <- covidHubUtils::load_truth(hub = "ECDC") |>
+    filter(target_variable %in% c("inc case", "inc death")) |>
+    mutate(target_variable = ifelse(target_variable == "inc case", 
+                                    "Cases", "Deaths")) |>
+    rename(target_type = target_variable, 
+           true_value = value) |>
+    select(-model) |>
+    filter(location == "GB")
+  
+  fwrite(truth, "data/weekly-truth-Europe.csv")
+}
+
+
+
+load_single_model <- function(root_dir,
+                              model_name) {
+  
+  locations <- select(truth, location, location_name) |>
+    filter(location == "GB") |>
+    unique()
+  
+  folders <- list.files(root_dir)
+  files <- map(folders,
+               .f = function(folder_name) {
+                 files <- list.files(here(root_dir, folder_name))
+                 paste(here(root_dir, folder_name, files))
+               }) %>%
+    unlist()
+  
+  forecasts <- suppressMessages(map_dfr(files, 
+                                        function(file) {
+                                          fread(file) |>
+                                            mutate_all(as.character) |>
+                                            mutate(model = model_name,
+                                                   quantile = as.numeric(quantile),
[…]

### 2023-02-05 Analysis.rmd
update code to fetch prediction data from github

@@
 library(stringr)
-library(covid.german.forecasts)
 library(grates)
@@
 ```
+
+```{r}
+forecasts <- fread("data/forecast-data.csv")
+scores <- fread("data/all-scores-crowd-forecasts.csv")
+weekly_truth <- fread("data/weekly-truth-Europe.csv")
+daily_truth <- fread("data/daily-truth-Europe.csv")
+
+time_start <- as.Date("2021-05-24")
+time_stop <- as.Date("2021-08-16")
+
+label_fn <- function(x) {
+  x <- ifelse(x%%1 == 0, 
+              as.integer(x), x)
+  ifelse(x < 1000, 
+         paste(x), 
+         ifelse(x < 1e6, 
+                paste0(x / 1000, "k"),
+                ifelse(x < 1e9, 
+                       paste0(x / 1e6, "m"), 
+                       paste0(x / 1e9, "b"))
+         )
+  )
+}
+
+```
+
+```{r}
+weekly_truth |>
+  ggplot(aes(x = target_end_date, y = true_value)) +
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = 0, ymax = Inf,
+           fill = "darkolivegreen3", alpha = .1) + 
+  geom_line() + 
+  geom_point(size = 0.8, aes(color = status)) + 
+  geom_bar(data = daily_truth |>
+             mutate(true_value = true_value * 7), 
+           stat = "identity", 
+           fill = "skyblue3", alpha = 0.4) + 
+  facet_wrap(~ target_type, 
+             nrow = 2, scale = "free_y") + 
+  scale_y_continuous(labels = label_fn) +
+  theme_scoringutils() + 
+  scale_colour_manual(values = c("red", "black")) + 
+  labs(y = "Observed value", x = "Date", color = "Data status")
+```
+
+```{r}
+forecasts |>
+  filter(target_end_date >= "2021-04-01", 
+         target_end_date <= "2021-10-01") |>
+  filter(model == "direct crowd") |>
+
+```

### 2023-02-05 fetch-data.sh

@@ -1,0 +1,10 @@
+#! bin/bash
+
+cd data
+svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/submissions
+
+cd crowd-direct-forecast
+svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/crowd-direct-forecast/processed-forecast-data
+
+cd ../crowd-rt-forecast
+svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/crowd-rt-forecast/processed-forecast-data

### 2023-02-05 fetch-data.sh
add data from the European Forecast Hub ensemble

@@ -8,3 +8,7 @@ svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/crowd-d
 
 cd ../crowd-rt-forecast
 svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/crowd-rt-forecast/processed-forecast-data
+
+cd ../EuroCOVIDhub-ensemble
+svn checkout https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe/trunk/data-processed/EuroCOVIDhub-ensemble
+mv EuroCOVIDhub-ensemble processed-forecast-data

### 2023-02-05 get-paper-data.R

@@ -9,6 +9,7 @@ library(purrr)
 library(scoringutils)
 library(here)
 library(stringr)
+library(tidyr)
 
 analysis_date <- "2022-12-12"
 
@@ -16,120 +17,13 @@ analysis_date <- "2022-12-12"
 ##                  Download and process Forecast Hub data                    ##
 ## -------------------------------------------------------------------------- ##
 
-# -------------------- Download Forecast Hub folder using SVN ---------------- #
-# svn checkout https://github.com/epiforecasts/covid19-forecast-hub-europe/trunk/data-processed
-
-# svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/submissions
+# ----------------------------- Renew Forecast Hub data ---------------------- #
+# bash fetch-data.sh
 
 
-# ---------------load truth data using the covidHubutils package ------------- #
-if (file.exists("data/weekly-truth-Europe.csv")) {
-  truth <- fread("data/weekly-truth-Europe.csv")
-} else {
-  truth <- covidHubUtils::load_truth(hub = "ECDC") |>
-    filter(target_variable %in% c("inc case", "inc death")) |>
-    mutate(target_variable = ifelse(target_variable == "inc case", 
-                                    "Cases", "Deaths")) |>
-    rename(target_type = target_variable, 
-           true_value = value) |>
-    select(-model) |>
-    filter(location == "GB")
-  
-  fwrite(truth, "data/weekly-truth-Europe.csv")
-}
-
-
-
-load_single_model <- function(root_dir,
-                              model_name) {
-  
-  locations <- select(truth, location, location_name) |>
-    filter(location == "GB") |>
-    unique()
-  
-  folders <- list.files(root_dir)
-  files <- map(folders,
-               .f = function(folder_name) {
-                 files <- list.files(here(root_dir, folder_name))
-                 paste(here(root_dir, folder_name, files))
-               }) %>%
-    unlist()
-  
-  forecasts <- suppressMessages(map_dfr(files, 
-                                        function(file) {
-                                          fread(file) |>
-                                            mutate_all(as.character) |>
-                                            mutate(model = model_name,
-                                                   quantile = as.numeric(quantile),
-                                                   target_end_date = as.character(target_end_date),
-                                                   horizon = as.numeric(gsub("([0-9]+).*$", "\\1", target))) %>%
-                                            filter(grepl("inc", target),
[…]

### 2023-02-05 get-paper-data.R
add Hub ensemble to forecasts and scores data

@@ -11,7 +11,9 @@ library(here)
 library(stringr)
 library(tidyr)
 
-analysis_date <- "2022-12-12"
+analysis_date <- Sys.Date()
+time_start <- as.Date("2021-05-24")
+time_stop <- as.Date("2021-08-16")
 
 ## -------------------------------------------------------------------------- ##  
 ##                  Download and process Forecast Hub data                    ##
@@ -180,16 +182,20 @@ forecasts <- load_forecasts(
 # ---------------------- load individual prediction data --------------------- #
 
 root_dirs <- c(here::here("data", "crowd-direct-forecast", "processed-forecast-data"),
-               here::here("data", "crowd-rt-forecast", "processed-forecast-data"))
+               here::here("data", "crowd-rt-forecast", "processed-forecast-data"), 
+               here::here("data", "EuroCOVIDhub-ensemble", "processed-forecast-data"))
 file_paths_forecast <- c(here::here(root_dirs[1], list.files(root_dirs[1])),
-                         here::here(root_dirs[2], list.files(root_dirs[2])))
+                         here::here(root_dirs[2], list.files(root_dirs[2])), 
+                         here::here(root_dirs[3], list.files(root_dirs[3])))
 
 individual_prediction_data <- purrr::map_dfr(file_paths_forecast,
                                              .f = function(x) {
                                                data <- data.table::fread(x)
-                                               
                                                if ("board_name" %in% names(data)) {
                                                  setnames(data, old = "board_name", new = "model")
+                                               }
+                                               if (!("model" %in% names(data))) {
+                                                 data[, model := "EuroCOVIDhub-ensemble"]
                                                }
                                                data[, target_end_date := as.character(target_end_date)]
                                                data[, forecast_date := calc_submission_due_date(forecast_date)]

### 2023-02-05 fetch-data.sh
add data about models included in the Hub ensemble

@@ -12,3 +12,5 @@ svn checkout https://github.com/epiforecasts/europe-covid-forecast/trunk/crowd-r
 cd ../EuroCOVIDhub-ensemble
 svn checkout https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe/trunk/data-processed/EuroCOVIDhub-ensemble
 mv EuroCOVIDhub-ensemble processed-forecast-data
+
+svn checkout https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe/trunk/code/ensemble/EuroCOVIDhub/criteria

### 2023-02-05 get-paper-data.R

@@ -222,6 +222,29 @@ data <- merge_pred_and_obs(combined_forecasts,
 fwrite(data, "data/forecast-data.csv")
 
 
+# --- collect data on how many models were included in the hub-ensemble
+
+root_dirs <- c(here::here("data", "EuroCOVIDhub-ensemble", "criteria"))
+file_paths_inclusion <- c(here::here(root_dirs[1], list.files(root_dirs[1])))
+
+ensemble_inclusions <- purrr::map_dfr(file_paths_inclusion,
+                                      .f = function(x) {
+                                        data <- data.table::fread(x)
+                                        data[location == "GB" & included_in_ensemble]
+                                        
+                                        date <- substr(x,1,nchar(x)-4)
+                                        date <- substr(date, nchar(date)-9, nchar(date))
+                                        
+                                        data[, forecast_date := date]
+                                        return(data)
+                                      }) |>
+  filter(forecast_date >= time_start, 
+         forecast_date <= time_stop)
+
+fwrite(ensemble_inclusions, "data/ensemble-inclusions.csv")
+
+
+
 ## -------------------------------------------------------------------------- ##  
 ##                               Score forecasts                              ##
 ## -------------------------------------------------------------------------- ##

### 2023-02-07 Analysis.rmd
add Figures

@@
 
+colors <- RColorBrewer::brewer.pal(n = 10, name = "Set1")
+
 ```
@@
 forecasts <- fread("data/forecast-data.csv")
-scores <- fread("data/all-scores-crowd-forecasts.csv")
-weekly_truth <- fread("data/weekly-truth-Europe.csv")
-daily_truth <- fread("data/daily-truth-Europe.csv")
+scores <- fread("data/all-scores-crowd-forecasts.csv") |>
+    mutate(type_and_scale = factor(type_and_scale, 
+                                 levels = c("Cases - natural", "Deaths - natural", 
+                                            "Cases - sqrt", "Deaths - sqrt", 
+                                            "Cases - log", "Deaths - log")))
+weekly_truth <- fread("data/weekly-truth.csv")
+daily_truth <- fread("data/daily-truth.csv")
+ensemble_inclusions <- fread("data/ensemble-inclusions.csv")
 
@@
 time_stop <- as.Date("2021-08-16")
+
+models_all <- forecasts$model |> unique() |> sort()
+models_all <- models_all[!models_all %in% c("", "EpiExpert-ensemble")]
+models_non_human <- c("crowd-ensemble", "crowd-direct", "crowd-rt",
+                      "EuroCOVIDhub-ensemble")
+models_human <- models_all[!(models_all %in% models_non_human)]
+models_rt <- models_human[grepl("(Rt)", models_human)]
+models_direct <- models_human[!grepl("(Rt)", models_human)]
 
@@
 ```{r}
-forecasts |>
+
+models <- forecasts$model |> unique()
+
+horizon_0 <- weekly_truth |>
+  filter(target_end_date >= "2021-04-01") |>
+  mutate(prediction = true_value, 
+         forecast_date = target_end_date + 2, 
+         horizon = 0, 
+         quantile = 0.5, 
+         model = list(models)) |>
+  tidyr::unnest(cols = model)
+
+plot_df <- forecasts |>
+  filter(quantile == 0.5) |>
+  rbind(horizon_0) |>
   filter(target_end_date >= "2021-04-01", 
          target_end_date <= "2021-10-01") |>
-  filter(model == "direct crowd") |>
+  filter(!(model %in% c("crowd-rt", "EpiExpert-ensemble"))) |>
+  filter(!(grepl("(Rt)", model))) |>
+  mutate(colouring = ifelse(model == "crowd-ensemble", 
+                            "Crowd ensemble", 
+                            "individual particpant"))
+
+plot_df |>
+  ggplot(aes(x = target_end_date)) +
+  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
+                color = "Individual participant"), 
+            linewidth = .1) + 
+  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
+            aes(y = prediction, group = interaction(forecast_date, model), 
+                color = "Crowd ensemble"),
+            linewidth = .4) + 
+  geom_line(aes(y = true_value),
+            linewidth = .4, color = "black") + 
+  geom_point(aes(y = true_value),
+             size = .5, color = "black") + 
+  scale_color_manual(values = c(colors[1], "grey70")) + 
+  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
+  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
+  theme_scoringutils()
+```
+
+```{r}
+
+forecasts |>
+  filter(quantile == 0.5) |>
+  rbind(horizon_0) |>
+  filter(target_end_date >= "2021-04-01", 
+         target_end_date <= "2021-10-01") |>
+  filter((model %in% c("crowd-ensemble", "EuroCOVIDhub-ensemble"))) |>
+  ggplot(aes(x = target_end_date)) +
+  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
+                color = model), 
+            linewidth = .4) + 
+  geom_line(aes(y = true_value),
+            linewidth = .4, color = "black") + 
+  geom_point(aes(y = true_value),
+             size = .5, color = "black") + 
+  scale_color_manual(values = colors[1:2]) +
+  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
+  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
+  theme_scoringutils()
+```
+
+```{r}
+
+comparison_plot <- function(scores, models = c("crowd-direct", "EuroCOVIDhub-ensemble")) {
+  mean_wis <- scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models) |>
+  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale")) |>
+  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale"),
+                   fun = signif, digits = 3) |>
+  mutate(wis_mean = paste("Mean:", label_fn(interval_score)))
+
+scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models) |>
+  arrange(scale, target_type, model) |>
+  ggplot(aes(y = interval_score, x = model)) + 
+  geom_violin(aes(fill = model), color = NA,
+         alpha = 0.2) + 
+  geom_boxplot(alpha = 0.5) + 
+  scale_fill_brewer(palette = "Set1", name = "Forecast target") +
+
+  geom_text(data = mean_wis, 
+            aes(label = wis_mean, y = -Inf), 
+            vjust   = -1) + 
+  facet_wrap(~ type_and_scale, scale = "free") + 
+  scale_y_continuous(labels = label_fn, 
+                     expand = expansion(mult = c(0.15, 0.05))) + 
+  labs(y = "Weighted interval score", x = "Forecaster") + 
+  theme_scoringutils() + 
+  theme(legend.position = "None")
+}
+
+comparison_plot(scores |>
+                  filter(horizon == 2), 
+                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))
+
+comparison_plot(scores |>
+                  filter(horizon == 2), 
+                models = c("crowd-direct", "crowd-rt", "crowd-ensemble"))
 
 ```
+
+## Weekly participation
+
+```{r}
+
+available <- function(forecasts, filter, name) {
+  forecasts |>
+    filter(model %in% filter) |>
+    avail_forecasts(by = c("forecast_date", "target_type"), 
+                    collapse = c("horizon", "quantile", "target_end_date")) |>
+    mutate(type = name)
+}
+
+## Number of forecasts
+avail <- 
+  available(forecasts, filter = models_direct, name = "Direct") |>
+  rbind(available(forecasts, filter = models_rt, name = "Rt"))
+
+plot_num_forecasters <- avail |>
+  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
+             fill = type)) +
+  geom_bar(stat = "identity", color = "white") + 
+  facet_wrap(~ target_type) + 
+  labs(x = "Forecast date", fill = "Forecast method") + 
+  theme_scoringutils()
+
+ggsave("output/figures/num-forecasters-week.png", plot = plot_num_forecasters, 
+       width = 7, height = 3)
+
+## Number of Hub ensemble members
+plot_ensemble_members <- ensemble_inclusions |>
+  mutate(prediction = 1, true_value = 1) |>
+  avail_forecasts(by = c("forecast_date", "target_type"), 
+                    collapse = c("horizon", "quantile", "target_end_date")) |>
+  ggplot(aes(x = forecast_date, y = `Number forecasts`)) +
+  geom_bar(stat = "identity", fill = "grey60", color = "white") + 
+  facet_wrap(~ target_type) + 
+  labs(x = "Forecast date", fill = "Forecast method") + 
+  theme_scoringutils()
+
+ggsave("output/figures/num-ensemble-members.png", plot = plot_ensemble_members, 
+       width = 7, height = 3)
+
+## number of unique participants
+unique_participants <- gsub("\ \\(Rt\\)", "", models_human) |> unique()
+length(unique_participants)
+
+## Number of unique participants in every week
+forecasts |>
+  filter(model %in% models_human) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  select(model, forecast_date) |>
+  group_by(forecast_date) |>
+  summarise(n = length(unique(model))) |>
+  summary()
+
+## number of times a forecaster submitted
+forecasts |>
+  filter(model %in% models_human) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  select(forecast_date, model) |>
+  group_by(model) |>
+  summarise(n = length(unique(forecast_date))) |>
+  summary()
+```
+
+## Performance
+
+```{r}
+
+df <- scores |>
+  filter(model %in% models_non_human, 
+         scale %in% c("log", "natural")) |>
+    summarise_scores(by = c("horizon", "model", 
+                            "target_type", "scale", "type_and_scale")) |>
+  rename("Dispersion" = dispersion, 
+         "Overprediction" = overprediction, 
+         "Underprediction" = underprediction, 
+         Model = model) |>
+    pivot_longer(names_to = "WIS components",
+                 cols = c(`Dispersion`, `Overprediction`, `Underprediction`)) |>
+    mutate(`WIS components` = fct_relevel(`WIS components`, 
+                                          c("Dispersion", "Underprediction", 
+                                            "Overprediction")))
+
+component_plot <- function(df, scale_filter, type_filter) {
+  df |>
+  filter(scale == scale_filter) |>
+  filter(target_type == type_filter) |>
+  ggplot(aes(x = reorder(Model, (interval_score)), y = value, fill = Model, 
+             group = Model, alpha = `WIS components`)) + 
+  scale_alpha_manual(values = c(1, 0.1, 0.6)) +
+  # scale_fill_manual(values = vals) + 
+  geom_bar(stat = "identity", position = "stack",
+           color = "black",
+           size = 0.1,
+           width = 0.8) +
+  facet_wrap(~ horizon, ncol = 4, 
+             strip.position = "bottom") + 
+  theme_scoringutils() + 
+  # theme(axis.text.x = element_blank()) + 
+  theme(axis.text.x = element_blank(),
+        axis.line.x = element_blank(), 
+        axis.ticks.x = element_blank(),
+        plot.title = element_text(hjust = 0.5,
+                                  size = 8)) + 
+  labs(x = NULL, y = "WIS", title = paste(type_filter, "-", scale_filter)) 
+}
+
+p_nat_case <- component_plot(df, "natural", "Cases")
+p_nat_death <- component_plot(df, "natural", "Deaths")
+p_log_case <- component_plot(df, "log", "Cases")
+p_log_death <- component_plot(df, "log", "Deaths")
+
+cov <- forecasts |>
+  filter(model %in% models_non_human) |>
+  score() |>
+  add_coverage(by = c("model", "target_type", "horizon")) |>
+  summarise_scores(by = c("model", "target_type", "range", "horizon"))
+
+cov_plot <- function(cov_scores, filter_range) {
+  cov_scores |>
+    filter(range == filter_range) |>
+    ggplot(aes(x = horizon, y = coverage, color = model)) + 
+    geom_line(show.legend = FALSE) + 
+    geom_point(show.legend = FALSE) + 
+    geom_hline(yintercept = filter_range / 100, 
+               linetype = "dashed", colour = "grey80") + 
+    facet_wrap(~ target_type, ncol = 2) + 
+    scale_y_continuous(labels = \(x) {paste0(100 * x, "%")}) + 
+    theme_scoringutils() + 
+    labs(x = "Horizon", y = paste("Coverage -", filter_range, "PI"))
+}
+
+p_cov50 <- cov_plot(cov, 50)
+p_cov90 <- cov_plot(cov, 90)
+
+p_performance <- (p_nat_case + p_nat_death) / 
+  (p_log_case + p_log_death) + 
+  (p_cov50 + p_cov90) + 
+  plot_layout(guides = "collect") &
+  plot_annotation(tag_levels = 'A') & 
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin())
+
+p_performance
+
+ggsave("output/figures/performance.png", plot = p_performance, 
+       width = 7, height = 8)
+
+```
+
+Table with performance
+
+```{r, eval=FALSE}
+
+coverage <- cov |>
+  select(model, target_type, horizon, coverage_50, coverage_90) |>
+  unique()
+
+scores |>
+  filter(model %in% models_non_human, 
+         scale %in% c("log", "natural")) |>
+  summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
+  inner_join(coverage) |>
+  filter(horizon == 2) |>
+  select(model, target_type, scale, interval_score, coverage_50, coverage_90) |>
+  pivot_wider(names_from = scale, values_from = interval_score) |>
+  select(model, target_type, natural, log, coverage_50, coverage_90) |>
+  mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
+  kable(format = "latex", booktabs = TRUE, 
+        align = c("l", "l", rep("c", 4)),
+        linesep = c('', '', '', '\\addlinespace'),
+        col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
+                      "Coverage 50%", "Coverage 90%")) |> 
+    kable_styling(latex_options = c("scale_down", 
+                                    "hold_position")) 
+```
+
+```{r}
+
+plot_pred_obs <- function(forecasts, scale = "natural") {
+
+  truth <- weekly_truth |>
+    mutate(target_type = paste(target_type, "-", scale))
+
+  if (scale == "log") {
+    forecasts <- forecasts |>
+      mutate(prediction = log(prediction + 1))
+
+    truth <- truth |>
+      mutate(true_value = log(true_value + 1))
+
+  }
+
+  forecasts |>
+    filter(model %in% models_non_human, 
+           horizon == 2) |>
+    filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
+    dcast(... ~ quantile, value.var = "prediction") |>
+    mutate(target_type = paste(target_type, "-", scale)) |>
+    ggplot(aes(x = target_end_date)) + 
+    geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
+                   size = 1.1,
+                   alpha = 1,
+                   position = position_dodge(width = 4.8), 
+                   show.legend = FALSE) + 
+    geom_line(data = truth |>
+                filter(target_end_date >= "2021-05-01", 
+                       target_end_date <= "2021-09-01"), 
+              aes(y = true_value)) + 
+    geom_point(data = truth |>
+                 filter(target_end_date >= "2021-05-01", 
+                        target_end_date <= "2021-09-01"), 
+               aes(y = true_value), size = 0.7) + 
+    theme_scoringutils() +
+    facet_wrap(~ target_type, scales = "free", ncol = 2) + 
+    scale_y_continuous(labels = label_fn) + 
+    theme(legend.position = "bottom") + 
+    labs(y = "Forecasts", x = "Date", colour = "Model")
+}
+
+p_pred_nat <- plot_pred_obs(forecasts)
+p_pred_log <- plot_pred_obs(forecasts, scale = "log")
+
+plot_wis_components <- function(scores) {
+  scores |>
+    filter(model %in% models_non_human, 
+           horizon == 2) |>
+    summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
+    select(-interval_score) |>
+    # add dispersion value to produce accurate "stacked" barplot
+    mutate(underprediction = underprediction + dispersion, 
+           overprediction = overprediction + dispersion) |>
+    data.table::melt(measure.vars = c(
+      "overprediction",
+      "underprediction",
+      "dispersion"
+    ),
+    variable.name = "wis_component_name",
+    value.name = "component_value"
+    ) |>
+    mutate(wis_component_name = stringr::str_to_title(wis_component_name)) |>
+    mutate(wis_component_name = factor(
+      wis_component_name, 
+      levels = c("Dispersion", "Overprediction", "Underprediction"))) |>
+    mutate(target_end_date = as.Date(target_end_date)) |>
+    ggplot(aes(x = target_end_date, group = model)) +
+    geom_col(
+      position = "dodge",
+      aes(y = component_value, 
+          fill = model, 
+          group = model, alpha = wis_component_name), 
+      color = "black", 
+      size = 0.1,
+      width = 5.5
+    ) +
+    facet_wrap(~ type_and_scale, scales = "free_y") +
+    theme_scoringutils() +
+    scale_alpha_manual(values = c(1, 0.6, 0.1)) +
+    scale_y_continuous(labels = label_fn) + 
+    expand_limits(x = as.Date(c("2021-05-01", "2021-09-01"))) +
+    labs(y = "WIS", x = "Target end date", 
+         fill = "Forecaster", alpha = "WIS component") +
+    theme(legend.position = "bottom", 
+          legend.box="vertical", legend.margin = margin()) 
+}
+
+p_scores_nat <- scores |>
+  filter(scale == "natural") |>
+  plot_wis_components()
+
+p_scores_log <- scores |>
+  filter(scale == "log") |>
+  plot_wis_components()
+
+p_scores_and_pred <- p_pred_nat / p_scores_nat / p_pred_log / p_scores_log + 
+  plot_layout(guides = "collect", heights = c(1, 0.8, 1, 0.8)) &
+  plot_annotation(tag_levels = 'A') & 
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin())
+
+ggsave("output/figures/scores-and-forecasts.png", plot = p_scores_and_pred, 
+       width = 9, height = 9)
+
+```
+
+Interesting questions
+- [x] performance for cases and deaths
+- [x] on the natural scale and log scale
+
+- performance experts vs. non-experts
+- performance Rt vs. non-Rt forecast
+- need a table with number of participants
+- add calibration plot (maybe just copy the plot from the Germany / Poland paper)
+- need to find the numbers of member models
+
+Methods
+- Hub ensemble switched from mean to median on 2021-07-19
+- did we do do mean or median? 
+- how did we end up with the crowd ensemble

### 2023-02-07 get-paper-data.R
Update forecast data and inclusion criteria for ensemble models

@@ -222,22 +222,27 @@ data <- merge_pred_and_obs(combined_forecasts,
 fwrite(data, "data/forecast-data.csv")
 
 
-# --- collect data on how many models were included in the hub-ensemble
+# ---- collect data on how many models were included in the hub-ensemble ----- #
 
 root_dirs <- c(here::here("data", "EuroCOVIDhub-ensemble", "criteria"))
 file_paths_inclusion <- c(here::here(root_dirs[1], list.files(root_dirs[1])))
 
-ensemble_inclusions <- purrr::map_dfr(file_paths_inclusion,
-                                      .f = function(x) {
-                                        data <- data.table::fread(x)
-                                        data[location == "GB" & included_in_ensemble]
-                                        
-                                        date <- substr(x,1,nchar(x)-4)
-                                        date <- substr(date, nchar(date)-9, nchar(date))
-                                        
-                                        data[, forecast_date := date]
-                                        return(data)
-                                      }) |>
+ensemble_inclusions <- 
+  purrr::map_dfr(file_paths_inclusion,
+                 .f = function(x) {
+                   data <- data.table::fread(x)
+                   data <- data[location == "GB" & included_in_ensemble]
+                   data[, target_type := ifelse(
+                     grepl("case", target_variable), 
+                     "Cases", "Deaths")]
+                   
+                   date <- substr(x,1,nchar(x)-4)
+                   date <- substr(date, nchar(date)-9, nchar(date))
+                   data[, forecast_date := date]
+                   
+                   data <- data[, .SD, .SDcols = c("model", "forecast_date", "target_type")]
+                   return(data)
+                 }) |>
   filter(forecast_date >= time_start, 
          forecast_date <= time_stop)
 

### 2023-02-07 manuscript/manuscript.tex
started first paper draft

% For more detailed article preparation guidelines, please see: http://wellcomeopenresearch.org/for-authors/article-guidelines and http://wellcomeopenresearch.org/for-authors/data-guidelines

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{WellcomeOR_styles}

%% Packages added manually
\usepackage{float}
\usepackage{hyperref}
\usepackage{booktabs}

%% Default: numerical citations
\usepackage[numbers]{natbib}

%% Uncomment this lines for superscript citations instead
% \usepackage[super]{natbib}

%% Uncomment these lines for author-year citations instead
% \usepackage[round]{natbib}
% \let\cite\citep

\begin{document}

\title{Wellcome Open Research Article Template}
\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise. Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
\author[1]{Author Name1}
\author[2]{Author Name2}
\affil[1]{Address of author1}
\affil[2]{Address of author2}

\maketitle
\thispagestyle{fancy}

Please list all authors that played a significant role in the research involved in the article. Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).

Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized. If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.

\begin{abstract}

\textbf{Background}

\textbf{Methods}

\textbf{Results}

\textbf{Conclusions}

\end{abstract}

\section*{Keywords}

Human judgement forecasting, COVID-19, infectious disease,

Please list up to eight keywords to help readers interested in your article find it more easily.

\clearpage

\section*{Introduction}

Infectious disease forecasting has attracted wide-spread attention during the COVID-19 pandemic and has helped inform public health decision making \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}. Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}. Most forecasts submitted to these Forecast Hubs or used in other contexts were generated by computational models, but some efforts have also explored human judgement forecasting as an alternative to mathematical modelling of COVID-19 \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.

Some papers have in the past directly compared the performance of human judgement forecasts and model-based predictions in the context of infectious diseases. \cite{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US. However, a comparable approach performed slightly worse than computtional models at predicting the 2014/15 outbreak of chikungunya in the Americas. \cite{mcandrewChimericForecastingCombining2022} found an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident cases, and worse when predicting incident deaths. \cite{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths. \cite{mcandrewChimericForecastingCombining2022} had between 22 and 61 human forecasters, while the number of participants in \cite{farrowHumanJudgmentApproach2017} ranged between ten and twenty, and were even lower in the cases of \cite{bosseComparingHumanModelbased2022} with only in between four to ten active forecasters.

In some situations, human judgement forecasting has clear advantages relative to computational models. Humans can very quickly provide forecasts, even in situations where data is sparse and lots of parameters are unknown. In addition, humans are able to answer a broader set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model. On the other hand, human judgement forecasting is difficult to scale, and humans may be at a disadvantage at tasks that require complex computations.

Methods that aim to combine human judgement and mathematical modelling are therefore appealing. One straightforward way is to combine separate human judgement and computational model forecasts with the goal of improving predictive performance \citep{mcandrewChimericForecastingCombining2022}. \cite{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required. One approach is to use human forecasts, for example of relevant disease paremeters, as an input to computational modelling. Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustements to model outputs. \cite{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) feed this forecast into a mathematical model in order to obtain forecasts for case and death numbers.

This paper represents a follow-up study to \cite{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021. Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge". This study aims to investigate whether the original findings in \cite{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants. In addition, it explores the approach proposed in \cite{bosseComparingHumanModelbased2022} to ask participants for a forecast of the effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a mathematical model.

\section*{Methods}

\subsection*{Ethics statement}
This study has been approved by the London School of Hygiene \& Tropical Medicine Research Ethics Committee (reference number 22290). Consent from participants was obtained in written form.

\subsection*{European COVID-19 Forecast Hub}

The European Forecast Hub CITATION elicits predictions for various COVID-19 related forecast targets from different research groups every week. Forecasts had to be submitted every Monday 23.59pm GMT. Forecasts are made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon. While forecasts were submitted on Mondays, weeks were defined as epiweeks. ending on a Saturday and starting on Sunday. Forecast horizons were therefore in fact 5, 12, 19 and 26 days. Submissions to the European Forecast Hub follow a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
Every week, forecast submitted to the hub were automatically checked for conformity with the required forecasta and all eligible forecasts combined into different ensembles. Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisiations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles). From the 29th of July on, the default Forecast Hub ensemble became a median-ensemble. The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:ensemble-models}.

Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by European Forecast Hub and sourced from the Johns Hopkins University. Data were subject to reporting artifacts and revisions. The Forecast Hubs marks data as anomalous if in subsequent updates data is changed by more than 5 percent. THIS IS THE CASE FOR ALL DEATH FORECASTS... CAN WE CHECK THIS? 
See Figure MAKE FIGURE WITH DATA AND UPDATES.

\subsection*{Crowd forecasts}

Forecasts of incident cases and deaths linked to COVID-19 in the UK were elicited from individual participants every week through the web application (https://cmmid-lshtm.shinyapps.io/crowd-forecast/) described in \cite{bosseComparingHumanModelbased2022}. The application is based on `R` `shiny` available as an R package called `crowdforecastr` \citep{crowdforecastr}.

The web application offered participants two different ways of making a forecast called 'direct' (or 'classical') and 'Rt' forecast. To make a 'direct' forecast (as described in more detail \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty. In addition to information about past observations, participants could see various metrics and data sourced from Our World in Data \citep{owidcoronavirus}.

SAM TO THE RESCUE
In addition, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$. This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the `R` package `EpiNow2` \citep{epinow2}, APPENDIX INFO?. In order to obtain forecasts for deaths, we convolved predicted infections as implied by the $R_t$ forecast over data-based delay distributions [40, 42–44] to model the time between infection and report date. As $R_t$-estimates up to two weeks prior to the forecast data were deemed uncertain, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date. Participants were therefore asked to estimate / predict four $R_t$ values, four of them in the future.

The crowd forecasts we submitted to the European Forecast Hub were created as an ensemble of individual forecasts every week. In contrast to the ensemble of human forecasts described in \cite{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions. %, following suggestions made in CITATION. 
We submitted three different ensembles: The first one, "epiforecasts-EpiExpert\_direct" (here called "crowd direct") was a quantile-wise median ensemble of all the direct forecasts. "epiforecasts-EpiExpert\_Rt" (here called "crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface. "epiforecasts-EpiExpert" (here called "crowd combined") was a median ensemble of all forecasts, after combining direct and $R_t$ forecast from the same individual by taking a quantile-wise mean. Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as for example forecasts that were zero everywhere).

\subsection{EpiNow2}

SAM TO THE RESCUE, RELEVANT PART OF OLD PAPER PASTED HERE AND COMMENTED OUT
% From old paper: 
% The first of these models, here called “renewal model”, used the renewal equation [41] to predict reported cases and deaths (see details in S1 Text). It estimated the effective reproduction number Rt (the average number of people each person infected at time t is expected to infect in turn) and modelled future infections as a weighted sum of past infection multiplied by Rt. Rt was assumed to stay constant beyond the forecast date, roughly corresponding to continuing the latest exponential trend in infections. On the 9th of November we altered the date when Rt was assumed to be constant from two weeks prior to the date of the forecast to the forecast date, which we found to yield a more stable Rt estimate. Reported case and death notifications were obtained by convolving predicted infections over data-based delay distributions [40, 42–44] to model the time between infection and report date. The renewal model was used to predict cases as well as deaths with forecasts being generated for each target separately. Death forecasts from the renewal model were therefore not informed by past cases. One submission of the renewal model on December 28th 2020 was delayed and therefore not included in the official Forecast hub ensemble.
% 
% Line list data used to inform the prior for the delay from symptom onset to test positive case report or death in the model-based forecasts was sourced from [45] with data available up to the 1st of August. All model fitting was done using Markov-chain Monte Carlo (MCMC) in stan [46] with each location and forecast target being fitted separately.

\subsection*{The UK Crowd Forecasting Challenge}

In order to boost participation compared to our last crowd forecasting study in Germany and Poland CITATION, we announced an official tournament, the "UK Crowd Forecasting Challenge". Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future. Everynone who had submitted a forecast for targets in the UK during the tournament period from the 24th of May 2021 to the 16th of August 2021 was deemed a participant and eligible for a prize. The first prize was 100 GBP, second prize 50 GBP and third prize 25 GBP. For the purpose of the tournament ranking, participants who did not submit a forecast in a given week where assigned the median score of all other participants who submitted a forecast that week. The UK crowd forecasting challenge was announced over Twitter and our networks. 
% (e.g. https://www.voice-global.org/public/opportunities/archived/uk-covid-19-crowd-forecasting-challenge/)
In addition, we created a project website, \url{https://crowdforecastr.org}, made weekly posts on Twitter and sent participants who had registered on the online application weekly emails with a reminder and a summary of their past performance. A public leaderboard was available on our website \url{https://epiforecasts.io}.

Participants were offered two different ways of forecasting. The first option was to make a "direct" forecast of cases and deaths. The second option was to make a forecast of the effective reproduction number $R_t$ which would then be converted to a forecast of cases and deaths using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} via the `R` \citep{R} package `EpiNow2`. Participants could make a forecast for either one or both of these methods and were eligible for prizes twice. Weekly forecasts had to be submitted between Sunday 12pm and Monday 8pm UK time.

\subsection{Analysis}

Our analysis of the data largely follows the one outlined in \cite{bosseComparingHumanModelbased2022}. If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs that found poor predictive performance beyond this horizon [15].

We scored forecasts using the weighted interval score \citep[WIS][]{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic callibration. The WIS is proper scoring rule (with smaller values implying better performance). A forecaster, in expectation, optimises her score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$ and is therefore incentivised to report her true belief. The WIS can be seen as a generalisation of the absolute error to predictive distribution in a quantile-based format and represents an approximation of the continuous ranked probability score (CRPS). The WIS can be decomposed into three separate penalty components: forecast spread (i.e. uncertainty of forecasts), over-prediction and under-prediction. FORMULAS IN APPENDIX. Empirical coverage means the percentage of observations falling inside any given prediction interval (e.g. the cumulative percentage of observed values that fall inside all 50\% prediction intervals). It represents a measure of probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}. Scoring metrics are explained in more detail in S1 Table. Forecasts were evaluated using the scoringutils \citep{bosseEvaluatingForecastsScoringutils2022} package in R.

\cite{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process. We therefore also compute WIS values after transforming all forecasts and observations using the function $f: x \to \log (x + 1)$.

\section*{Results}

\subsection*{Crowd forecast participation}

A total number of 90 participants submitted forecasts (more precisely, forecasts were submitted from 90 different accounts, some of them anonymous). The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51. With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13. The median number of submissions was one, meaning that most forecasters dropped out after their first submission. Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study. The number of direct forecasts was higher than the number of $R_t$ forecasts in all weeks (see Figure \ref{fig:num-forecasters}).

% X of those self-identified as ‘expert’ in either forecasting or epidemiology.

\begin{figure}[H]
\centering
\includegraphics[width=0.99\textwidth]{../output/figures/scores-and-forecasts.png}
\caption{\bf{Test}}
\label{fig:forecasts-scores}
\end{figure}

\subsection*{Case forecasts}

For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the weighted interval score (WIS), with the direct crowd forecasts performing poorest (see Figure \ref{fig:performance} and Table \ref{tab:scores}). When looking at WIS values after applying a log-transformation, all crowd forecasting approaches outperform the Forecast Hub ensemble. Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, KATH} do all forecasts for cases exhibit underdispersion, i.e. forecasts on average are too narrow and not uncertain enough, with empirical coverage below nominal coverage for all models.

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{llcccc}
\toprule
Model & Target & WIS & WIS (log scale) & Coverage 50\% & Coverage 90\%\\
\midrule
EuroCOVIDhub-ensemble & Cases & 38000 & 0.25 & 0.38 & 0.69\\
crowd-direct & Cases & 46000 & 0.23 & 0.46 & 0.62\\
crowd-ensemble & Cases & 40000 & 0.22 & 0.38 & 0.69\\
crowd-rt & Cases & 39000 & 0.24 & 0.31 & 0.69\\
\addlinespace
EuroCOVIDhub-ensemble & Deaths & 57 & 0.16 & 0.62 & 0.85\\
crowd-direct & Deaths & 78 & 0.17 & 0.46 & 0.77\\
crowd-ensemble & Deaths & 61 & 0.16 & 0.38 & 0.77\\
crowd-rt & Deaths & 63 & 0.23 & 0.23 & 0.69\\
\bottomrule
\end{tabular}}
\caption{Table caption}
\label{tab:scores}
\end{table}

\subsection*{Death forecasts}

In terms of WIS the Forecast Hub ensemble also outperformed all human forecasts when forecasting deahts. This is true as well for WIS values obtained after applying a log-transformation. When looking at scores on the natural scale, direct human forecasts performed worst and also show the largest amount of underprediction. When looking at scores of log-transformed forecasts, $R_t$-crowd forecast showed the worst performance. Empirical coverage for deaths seems broadly comparable for all human forecasts, whereas the Forecast Hub ensemble displays noticeably better coverage for deaths than for cases (Figure \ref{fig:performance}E, F).

% \subsection*{Individual forecasts?}

\section*{Discussion}
The discussion should include the implications of the article results in view of prior work in this field.

\section*{Conclusions}
Please state what you think are the main conclusions that can be realistically drawn from the findings in the paper, taking care not to make claims that cannot be supported.

interface difficult
giving participants the median score maybe not so smart

\subsection*{Author contributions}
In order to give appropriate credit to each author of an article, the individual
contributions of each author to the manuscript should be detailed in this section. We
recommend using author initials and then stating briefly how they contributed.

\subsection*{Competing interests}
The authors declare no competing interests

\subsection*{Grant information}
Please state who funded the work discussed in this article, whether it is your employer,
a grant funder etc. Please do not list funding that you have that is not relevant to this
specific piece of research. For each funder, please state the funder’s name, the grant
number where applicable, and the individual to whom the grant was assigned.

\subsection*{Acknowledgements}
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided earlier
(e.g. someone or an organisation that provided writing assistance). Please state how
they contributed; authors should obtain permission to acknowledge from all those
mentioned in the Acknowledgements section.

Acknowledgement to all forecasters

\appendix
\section{Supplementary information}

% \renewcommand{\thefigure}{SI.\arabic{figure}}
% \setcounter{figure}{0}
% \renewcommand{\thetable}{SI.\arabic{table}} \setcounter{table}{0}

\begin{figure}[H]
\centering
\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters-week.png}
\caption{\bf{Test}}
\label{fig:num-forecasters}
\end{figure}

\clearpage

{\small\bibliographystyle{unsrtnat}
\bibliography{sample}}

\bigskip
References can be listed in any standard referencing style that uses a numbering system
(i.e. not Harvard referencing style), and should be consistent between references within
a given article.

Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files. Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography. This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own. Use the \verb|\cite| command  to create in-text citations, like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.

% See this guide for more information on BibTeX:
% http://libguides.mit.edu/content.php?pid=55482&sid=406343

% For more author guidance please see:
% http://wellcomeopenresearch.org/for-authors/article-guidelines

% When all authors are happy with the paper, use the 
% ‘Submit to WELLCOME OPEN RESEARCH' button from the menu above
% to submit directly to the open life science journal Wellcome Open Research.

% Please note that this template results in a draft pre-submission PDF document.
% Articles will be professionally typeset when accepted for publication.

% We hope you find the Wellcome Open Research Overleaf template useful,
% please let us know if you have any feedback using the help menu above.

\end{document}

### 2023-02-07 manuscript/manuscript.tex
add citations

@@
 When looking at WIS values after applying a log-transformation, all crowd forecasting approaches outperform the Forecast Hub ensemble.
-Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, KATH} do all forecasts for cases exhibit underdispersion, i.e.
+Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a} do all forecasts for cases exhibit underdispersion, i.e.
 forecasts on average are too narrow and not uncertain enough, with empirical coverage below nominal coverage for all models.
@@
 {\small\bibliographystyle{unsrtnat}
-\bibliography{sample}}
+\bibliography{software, uk-forecasting-challenge}}
 
 \bigskip
-References can be listed in any standard referencing style that uses a numbering system
-(i.e. not Harvard referencing style), and should be consistent between references within
-a given article.
-
-Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files.
-Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography.
-This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own.
-Use the \verb|\cite| command  to create in-text citations, like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.
 

### 2023-02-08 Analysis.rmd
update manuscript, add data revision plots, add screenshots of the app, add citations

@@
                                             "Cases - log", "Deaths - log")))
-weekly_truth <- fread("data/weekly-truth.csv")
-daily_truth <- fread("data/daily-truth.csv")
+weekly_truth_revised <- fread("data/weekly-truth.csv")
+daily_truth_revised <- fread("data/daily-truth.csv")
+weekly_truth <- fread("data/weekly-truth-original.csv")
+daily_truth <- fread("data/daily-truth-original.csv")
+
 ensemble_inclusions <- fread("data/ensemble-inclusions.csv")
+data_revisions <- fread("data/data-revisions-deaths.csv")
 
 time_start <- as.Date("2021-05-24")
-time_stop <- as.Date("2021-08-16")
+time_stop <- as.Date("2021-08-16") + 4 * 7 - 2
 
@@
 ```{r}
+plot_data <- function(weekly_truth, daily_truth) {
   weekly_truth |>
@@
   labs(y = "Observed value", x = "Date", color = "Data status")
+}
+
+plot_data_revised <- plot_data(weekly_truth_revised, daily_truth_revised)
+plot_data <- plot_data(weekly_truth, daily_truth)
+
+ggsave("output/figures/plot-data-revised.png", plot = plot_data_revised, 
+       width = 7, height = 5)
+
+ggsave("output/figures/plot-data.png", plot = plot_data, 
+       width = 7, height = 3.5)
+
+```
+
+Show data revisions
+
+```{r}
+
+plot_weekly_revisions <- weekly_truth |>
+  rbind(weekly_truth_revised) |>
+  filter(target_type == "Deaths") |>
+  select(target_end_date, true_value, target_type) |>
+  group_by(target_end_date, target_type) |>
+  mutate(diff = c(NA, diff(true_value))) |>
+  filter(!is.na(diff)) |>
+  ggplot(aes(x = target_end_date, y = diff)) + 
+  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") + 
+  geom_line() + 
+  geom_point(size = 0.7) + 
+  facet_wrap(~ target_type, ncol = 1, scale = "free_y") +
+  labs(y = "Difference in data", x = "Date") + 
+  coord_cartesian(ylim = c(-500, 500)) + 
+  theme_scoringutils()
+
+ggsave("output/figures/data-revisions.png", plot = plot_weekly_revisions, 
+       width = 7, height = 2.5)
+
 ```
@@
 ```{r}
-
 available <- function(forecasts, filter, name) {
@@
 
-## Number of forecasts
+## Number of forecasts and ensemble members
 avail <- 
   available(forecasts, filter = models_direct, name = "Direct") |>
-  rbind(available(forecasts, filter = models_rt, name = "Rt"))
+  rbind(available(forecasts, filter = models_rt, name = "Rt")) |>
+  mutate(group = "Human forecast")
 
-plot_num_forecasters <- avail |>
-  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
-             fill = type)) +
-  geom_bar(stat = "identity", color = "white") + 
-  facet_wrap(~ target_type) + 
-  labs(x = "Forecast date", fill = "Forecast method") + 
-  theme_scoringutils()
-
-ggsave("output/figures/num-forecasters-week.png", plot = plot_num_forecasters, 
-       width = 7, height = 3)
-
-## Number of Hub ensemble members
-plot_ensemble_members <- ensemble_inclusions |>
+avail_ensemble <- ensemble_inclusions |>
   mutate(prediction = 1, true_value = 1) |>
@@
                     collapse = c("horizon", "quantile", "target_end_date")) |>
-  ggplot(aes(x = forecast_date, y = `Number forecasts`)) +
-  geom_bar(stat = "identity", fill = "grey60", color = "white") + 
+  mutate(group = "Forecast Hub")
+
+plot_number_forecasters <- avail |>
+  pivot_wider(names_from = type, values_from = `Number forecasts`) |>
+  mutate(Rt = Direct + Rt) |>
+  pivot_longer(cols = c(Direct, Rt), names_to = "type",
+               values_to = "Number forecasts") |>
+  as.data.table() |>
+  rbind(avail_ensemble, fill = TRUE) |>
+  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
+             group = group,
+             fill = group, 
+             alpha = type)) +
+  geom_bar(stat = "identity", 
+           color = "white",
+           linewidth = 0.2,
+           width = 5,
+           position = position_dodge(width = 5)) + 
   facet_wrap(~ target_type) + 
-  labs(x = "Forecast date", fill = "Forecast method") + 
-  theme_scoringutils()
+  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
+  labs(x = "Forecast date", fill = "Forecast type", 
+       alpha = "Forecast method") + 
+  theme_scoringutils() +
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin())
 
-ggsave("output/figures/num-ensemble-members.png", plot = plot_ensemble_members, 
-       width = 7, height = 3)
+ggsave("output/figures/num-forecasters.png", plot = plot_number_forecasters, 
+       width = 7, height = 3.5)
 
@@
 
+```{r}
+forecasts$forecast_date |>
+  unique()
+
+forecasts$target_end_date |>
+  unique()
+```
+
 Interesting questions

### 2023-02-08 manuscript/manuscript.tex
update manuscript, add data revision plots, add screenshots of the app, add citations

@@
 \usepackage{booktabs}
+\usepackage{longtable}
 
@@
 Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
-\author[1]{Author Name1}
-\author[2]{Author Name2}
-\affil[1]{Address of author1}
+
+\author[1, 2, 3]{Anonymous Alpaca}
+\author[2]{Benevolent Brachiosaurus}
+\affil[1]{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom}
+\affil[2]{Centre for the Mathematical Modelling of Infectious Diseases, London, United Kingdom}
+\affil[3]{NIHR Health Protection Research Unit in Modelling \& Health Economics}
 \affil[2]{Address of author2}
@@
 
-Please list all authors that played a significant role in the research involved in the article.
+% Please list all authors that played a significant role in the research involved in the article.
 Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).
+
+\begin{abstract}
 
@@
 If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.
-
-\begin{abstract}
 
@@
 
-Please list up to eight keywords to help readers interested in your article find it more easily.
+% Please list up to eight keywords to help readers interested in your article find it more easily.
 
@@
 
-Infectious disease forecasting has attracted wide-spread attention during the COVID-19 pandemic and has helped inform public health decision making \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}.
-Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
-Most forecasts submitted to these Forecast Hubs or used in other contexts were generated by computational models, but some efforts have also explored human judgement forecasting as an alternative to mathematical modelling of COVID-19 \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
+Infectious disease modelling and forecasting has attracted wide-spread attention during the COVID-19 pandemic and helped inform decision making in public health organisations and governments \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}.
+% Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+Most forecasts were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
 
-Some papers have in the past directly compared the performance of human judgement forecasts and model-based predictions in the context of infectious diseases.
-\cite{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
+Past research found that in the context of infectious disease forecasting, human judgement forecasts could achieve predictive performance broadly comparable to forecasts generated based on mathematical modelling, in particular when forecasting incident cases.
+\citet{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
 However, a comparable approach performed slightly worse than computtional models at predicting the 2014/15 outbreak of chikungunya in the Americas.
-\cite{mcandrewChimericForecastingCombining2022} found an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident cases, and worse when predicting incident deaths.
-\cite{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths.
-\cite{mcandrewChimericForecastingCombining2022} had between 22 and 61 human forecasters, while the number of participants in \cite{farrowHumanJudgmentApproach2017} ranged between ten and twenty, and were even lower in the cases of \cite{bosseComparingHumanModelbased2022} with only in between four to ten active forecasters.
+\citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths.
+Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident cases, and worse when predicting incident deaths.
+\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
 
@@
 Humans can very quickly provide forecasts, even in situations where data is sparse and lots of parameters are unknown.
-In addition, humans are able to answer a broader set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
-On the other hand, human judgement forecasting is difficult to scale, and humans may be at a disadvantage at tasks that require complex computations.
+In addition, humans are able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
+On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, and humans may be at a disadvantage at tasks that require complex computations.
+
+% Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
 
@@
 Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustements to model outputs.
-\cite{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) feed this forecast into a mathematical model in order to obtain forecasts for case and death numbers.
+\citet{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) and to feed this forecast into a mathematical model in order to obtain forecasts for case and death numbers.
 
-This paper represents a follow-up study to \cite{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021.
-Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge".
-This study aims to investigate whether the original findings in \cite{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants.
-In addition, it explores the approach proposed in \cite{bosseComparingHumanModelbased2022} to ask participants for a forecast of the effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a mathematical model.
+This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021.
+Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
+All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, German and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants.
+In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a mathematical model.
+
+% Define the terms model, forecaster, crowd forecast, ensemble.
 
@@
 
-\subsection*{European COVID-19 Forecast Hub}
+\subsection*{Interaction with the European Forecast Hub}
 
-The European Forecast Hub CITATION elicits predictions for various COVID-19 related forecast targets from different research groups every week.
+The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} elicits predictions for various COVID-19 related forecast targets from different research groups every week.
 Forecasts had to be submitted every Monday 23.59pm GMT.
@@
 From the 29th of July on, the default Forecast Hub ensemble became a median-ensemble.
-The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:ensemble-models}.
+The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:num-forecasters}.
 
@@
 
-\subsection*{Crowd forecasts}
+\subsection*{Human judgement forecasts}
 
 Forecasts of incident cases and deaths linked to COVID-19 in the UK were elicited from individual participants every week through the web application (https://cmmid-lshtm.shinyapps.io/crowd-forecast/) described in \cite{bosseComparingHumanModelbased2022}.
-The application is based on `R` `shiny` available as an R package called `crowdforecastr` \citep{crowdforecastr}.
+The application is based on \textsf{R} \cite{R} \texttt{shiny} \cite{shiny} and is available as an \textsf{R} package called \texttt{crowdforecastr} \citep{crowdforecastr}.
 
@@
 In addition to information about past observations, participants could see various metrics and data sourced from Our World in Data \citep{owidcoronavirus}.
+Figure \ref{fig:screenshot-classical} shows a screenshot of the forecast interface for direct forecasts.
 
 SAM TO THE RESCUE
+
 In addition, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
-This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the `R` package `EpiNow2` \citep{epinow2}, APPENDIX INFO?.
-In order to obtain forecasts for deaths, we convolved predicted infections as implied by the $R_t$ forecast over data-based delay distributions [40, 42–44] to model the time between infection and report date.
+This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the \textsf{R} package \texttt{EpiNow2} \citep{epinow2}, APPENDIX INFO?.
+In order to obtain forecasts for deaths, we convolved predicted infections as implied by the $R_t$ forecast over data-based delay distributions \cite{epinow2, sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} to model the time between infection and report date.
 As $R_t$-estimates up to two weeks prior to the forecast data were deemed uncertain, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
 Participants were therefore asked to estimate / predict four $R_t$ values, four of them in the future.
+Upon pressing a button participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
+However, due to the high computational complexity and necessary waiting times, participants could not preview the death forecast implied by their current input for $R_t$.
+Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
 
-The crowd forecasts we submitted to the European Forecast Hub were created as an ensemble of individual forecasts every week.
-In contrast to the ensemble of human forecasts described in \cite{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions.
+Every week, we submitted an ensemble of individual forecasts ("crowd forecast") to the European Forecast Hub.
+In contrast to the ensemble of human forecasts described in \citet{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions.
 %, following suggestions made in CITATION.
-We submitted three different ensembles: The first one, "epiforecasts-EpiExpert\_direct" (here called "crowd direct") was a quantile-wise median ensemble of all the direct forecasts.
-"epiforecasts-EpiExpert\_Rt" (here called "crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface.
-"epiforecasts-EpiExpert" (here called "crowd combined") was a median ensemble of all forecasts, after combining direct and $R_t$ forecast from the same individual by taking a quantile-wise mean.
-Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as for example forecasts that were zero everywhere).
+We submitted three different ensembles: The first one, "epiforecasts-EpiExpert\_direct" (here called "Crowd direct") was a quantile-wise median ensemble of all the direct forecasts.
+"epiforecasts-EpiExpert\_Rt" (here called "Crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface.
+"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts, after combining direct and $R_t$ forecast from the same individual by taking a quantile-wise mean.
+Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
 
-\subsection{EpiNow2}
+\subsection*{EpiNow2?}
 
-SAM TO THE RESCUE, RELEVANT PART OF OLD PAPER PASTED HERE AND COMMENTED OUT
+SAM TO THE RESCUE, RELEVANT PART OF OLD PAPER PASTED HERE AND COMMENTED OUT. DO WE NEED AN EXTRA SECTION, OR SHOULD THIS BE INCLDUED ABOVE?
+
 % From old paper: 
@@
 
-In order to boost participation compared to our last crowd forecasting study in Germany and Poland CITATION, we announced an official tournament, the "UK Crowd Forecasting Challenge".
+In order to boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022}, we announced an official tournament, the "UK Crowd Forecasting Challenge".
 Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
@@
 A public leaderboard was available on our website \url{https://epiforecasts.io}.
-
-Participants were offered two different ways of forecasting.
-The first option was to make a "direct" forecast of cases and deaths.
-The second option was to make a forecast of the effective reproduction number $R_t$ which would then be converted to a forecast of cases and deaths using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} via the `R` \citep{R} package `EpiNow2`.
-Participants could make a forecast for either one or both of these methods and were eligible for prizes twice.
+Participants could choose to make a direct forecast as well as an $R_t$ forecast and were eligible for prizes twice.
 Weekly forecasts had to be submitted between Sunday 12pm and Monday 8pm UK time.
 
-\subsection{Analysis}
+\subsection*{Analysis}
 
-Our analysis of the data largely follows the one outlined in \cite{bosseComparingHumanModelbased2022}.
+Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}.
 If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs that found poor predictive performance beyond this horizon [15].
 
-We scored forecasts using the weighted interval score \citep[WIS][]{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic callibration.
+We scored forecasts using the weighted interval score (WIS, formulas given in the SI in Section \ref{sec:wis}) \cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic callibration.
 The WIS is proper scoring rule (with smaller values implying better performance).
@@
 It represents a measure of probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
-Scoring metrics are explained in more detail in S1 Table.
-Forecasts were evaluated using the scoringutils \citep{bosseEvaluatingForecastsScoringutils2022} package in R.
+Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
 
-\cite{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
-We therefore also compute WIS values after transforming all forecasts and observations using the function $f: x \to \log (x + 1)$.
+\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
+We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
+In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance.png}
+\caption{\bf{Test}}
+\label{fig:performance}
+\end{figure}
+
 \subsection*{Case forecasts}
 
-For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the weighted interval score (WIS), with the direct crowd forecasts performing poorest (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-When looking at WIS values after applying a log-transformation, all crowd forecasting approaches outperform the Forecast Hub ensemble.
-Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a} do all forecasts for cases exhibit underdispersion, i.e.
-forecasts on average are too narrow and not uncertain enough, with empirical coverage below nominal coverage for all models.
+For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the weighted interval score (WIS).
+On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
+Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July \ref{fig:forecasts-scores}.
+Direct forecasts tended to be higher throughout the study period and especially during the peak than both the $R_t$ forecasts and the Forecast Hub ensemble and therefore showed the poorest performance (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+When looking at WIS values on the log scale (that is, when log-transforming forecasts and observations prior to applying the WIS), scores are more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
+On the log scale, the Forecast Hub ensemble performs slightly worse than all of the human forecasting approaches.
+Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all crowd forecasts as well the Forecast Hub ensemble, exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
 
@@
 
-In terms of WIS the Forecast Hub ensemble also outperformed all human forecasts when forecasting deahts.
+In terms of the WIS the Forecast Hub ensemble also outperformed all human forecasts when forecasting deaths.
 This is true as well for WIS values obtained after applying a log-transformation.
-When looking at scores on the natural scale, direct human forecasts performed worst and also show the largest amount of underprediction.
-When looking at scores of log-transformed forecasts, $R_t$-crowd forecast showed the worst performance.
+From June to the end of July, direct forecasts of deaths were noticeable higher than $R_t$ death forecasts, whereas in August direct death forecasts were substantially lower than $R_t$ death forecasts.
+On the natural scale, the largest scores occured at the end of the study period when deaths were highest.
+Direct forecasts, which most strongly underpredicted deaths in August therefore showed the poorest performance overall.
+On the log scale, however, underprediction in June and July when death incidences were lower, received more weight and direct crowd forecast outperformed $R_t$ forecasts.
+\textbf{CHECK WITH OTHER GROUND TRUTH DATA?}.
 Empirical coverage for deaths seems broadly comparable for all human forecasts, whereas the Forecast Hub ensemble displays noticeably better coverage for deaths than for cases (Figure \ref{fig:performance}E, F).
@@
 
+- hard to draw conclusions, performance changes depending on how you look at the data. 
+- headline result not supported? 
+- still some evidence that forecasting deaths is easier for computers
+-
+
+- Rt forecasting worked reasonably well. The interface was a bit clunky. Required more expertise. But a potential way to reduce the workload.
+
+- interesting that combining Rt and direct forecasts for cases improved performance.
+
+Limitations
+- The tournament setting may have distorted results as people were more eager to participate
+- ensemble was better
+
 \section*{Conclusions}
@@
 
+Seems that a model of human forecasters is alright. Combining human and model based output is still interesting. 
 interface difficult
-giving participants the median score maybe not so smart
 
@@
 
+% =================================== Appendix =============================== %
+
 \appendix
-\section{Supplementary information}
-
+\section*{Supplementary information}
 % \renewcommand{\thefigure}{SI.\arabic{figure}}
@@
 
+\subsection*{Weighted interval score}
+label{sec:wis}
+The weighted interval score (smaller values are better) is a proper scoring rule for quantile forecasts.
+It converges to the continuos ranked probability score (which itself is a generalisation of the absolute error to probabilistic forecasts) for an increasing number of intervals.
+The score can be decomposed into a dispersion (uncertainty) component and penalties for over- and underprediction.
+For a single interval, the score is computed as
+  $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
+  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e.
+the lower and upper bound of a single prediction interval.
+For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,
+  $$WIS = \frac{1}{K + 0.5} \cdot \left( w_0 \cdot |y - m| + \sum_{k = 1}^{K} w_k \cdot IS_{\alpha}(F, y) \right), $$
+  where $w_k$ is a weight for every interval. Usually, $w_k = \frac{\alpha_k}{2}$ and $w_0 = 0.5$.
+
 \begin{figure}[H]
 \centering
-\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters-week.png}
-\caption{\bf{Test}}
+\includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-classical.png}
+\caption{\bf{Screenshot of the direct forecasting interface.}}
+\label{fig:screenshot-classical}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-rt-app.png}
+\caption{\bf{Screenshot of the $R_t$ forecasting interface.}}
+\label{fig:screenshot-rt}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
+\caption{\bf{Number of forecasts across the study period.}}
 \label{fig:num-forecasters}

### 2023-02-08 get-paper-data.R

@@ -10,18 +10,98 @@ library(scoringutils)
 library(here)
 library(stringr)
 library(tidyr)
+library("gh")
+library("lubridate")
+library("readr")
+
 
 analysis_date <- Sys.Date()
 time_start <- as.Date("2021-05-24")
 time_stop <- as.Date("2021-08-16")
 
 ## -------------------------------------------------------------------------- ##  
-##                  Download and process Forecast Hub data                    ##
+##                                Download Hub data                           ##
 ## -------------------------------------------------------------------------- ##
 
 # ----------------------------- Renew Forecast Hub data ---------------------- #
 # bash fetch-data.sh
 
+
+# ------------------------------ Obtain data revisions ----------------------- #
+earliest_date <- NULL
+min_data <- as.Date("2021-03-08") ## earliest date to pot in data
+
+sources <-
+  c(Deaths = "JHU")
+
+target_variables <-
+  c(Deaths = "inc death")
+
+owner <- "epiforecasts"
+repo <- "covid19-forecast-hub-europe"
+path <- vapply(names(sources), function(x) {
+  paste("data-truth", sources[[x]],
+        paste0("truth_", sources[[x]], "-Incident ", x, ".csv"),
+        sep = "/")
+}, "")
+names(path) <- names(sources)
+
+data <- list()
+for (source in names(sources)) {
+  query <- "/repos/{owner}/{repo}/commits?path={path}"
+  if (!is.null(earliest_date)) {
+    query <- paste0(query, "&since={date}")
+  }
+  
+  commits <-
+    gh::gh(query,
+           owner = owner,
+           repo = repo,
+           path = path[source],
+           date = earliest_date,
+           .limit = Inf
+    )
+  
+  shas <- vapply(commits, "[[", "", "sha")
+  dates <- vapply(commits, function(x) x[["commit"]][["author"]][["date"]], "")
+  dates <- as_date(ymd_hms(dates))
+  
+  ## keep multiples of 7 since today
+  select_commits <- which(as.integer(max(dates) - dates) %% 7 == 0)
+  
+  data[[source]] <-
+    lapply(
+      select_commits,
+      function(id)
+        readr::read_csv(
+          URLencode(
+            paste("https://raw.githubusercontent.com", owner, repo,
+                  shas[id], path[source], sep = "/")),
+          show_col_types = FALSE) %>%
[…]

### 2023-02-10 get-paper-data.R
update data including data revisions

@@ -27,81 +27,126 @@ time_stop <- as.Date("2021-08-16")
 # bash fetch-data.sh
 
 
-# ------------------------------ Obtain data revisions ----------------------- #
-earliest_date <- NULL
-min_data <- as.Date("2021-03-08") ## earliest date to pot in data
-
-sources <-
-  c(Deaths = "JHU")
-
-target_variables <-
-  c(Deaths = "inc death")
-
-owner <- "epiforecasts"
-repo <- "covid19-forecast-hub-europe"
-path <- vapply(names(sources), function(x) {
-  paste("data-truth", sources[[x]],
-        paste0("truth_", sources[[x]], "-Incident ", x, ".csv"),
-        sep = "/")
-}, "")
-names(path) <- names(sources)
-
-data <- list()
-for (source in names(sources)) {
-  query <- "/repos/{owner}/{repo}/commits?path={path}"
-  if (!is.null(earliest_date)) {
-    query <- paste0(query, "&since={date}")
+# ---------------load truth data using the covidHubutils package ------------- #
+if (file.exists("data/weekly-truth-revised.csv")) {
+  truth_revised <- fread("data/weekly-truth-revised.csv")
+  daily_truth_revised <- fread("data/daily-truth-revised.csv")
+  
+} else {
+  clean_truth <- function(data) {
+    data |>
+      filter(target_variable %in% c("inc case", "inc death")) |>
+      mutate(target_variable = ifelse(target_variable == "inc case", 
+                                      "Cases", "Deaths")) |>
+      rename(target_type = target_variable, 
+             true_value = value) |>
+      select(-model) |>
+      filter(location == "GB") |>
+      filter(target_end_date >= "2021-01-01",
+             target_end_date <= "2022-01-01")
   }
   
-  commits <-
-    gh::gh(query,
-           owner = owner,
-           repo = repo,
-           path = path[source],
-           date = earliest_date,
-           .limit = Inf
+  truth_revised <- covidHubUtils::load_truth(hub = "ECDC") |>
+    clean_truth()
+  daily_truth_revised <- covidHubUtils::load_truth(hub = "ECDC", temporal_resolution = "daily") |>
+    clean_truth()
+  
+  truth_revised <- flag_anomalies(
+    mutate(truth_revised, target_end_date = as.character(target_end_date))
     )
   
-  shas <- vapply(commits, "[[", "", "sha")
[…]

### 2023-02-10 Analysis.rmd
update code and figures

@@
 
-colors <- RColorBrewer::brewer.pal(n = 10, name = "Set1")
+colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")
 
@@
 ```{r}
+
+add_type_and_scale <- function(data) {
+  data |>
+    mutate(type_and_scale = paste(target_type, "-", scale)) |>
+    mutate(type_and_scale = factor(type_and_scale, 
+                                   levels = c("Cases - natural", "Deaths - natural", 
+                                              "Cases - log", "Deaths - log")))
+}
+
 forecasts <- fread("data/forecast-data.csv")
 scores <- fread("data/all-scores-crowd-forecasts.csv") |>
-    mutate(type_and_scale = factor(type_and_scale, 
-                                 levels = c("Cases - natural", "Deaths - natural", 
-                                            "Cases - sqrt", "Deaths - sqrt", 
-                                            "Cases - log", "Deaths - log")))
-weekly_truth_revised <- fread("data/weekly-truth.csv")
-daily_truth_revised <- fread("data/daily-truth.csv")
+  add_type_and_scale()
+
+forecasts_revised <- fread("data/forecast-data-revised.csv")    
+scores_revised <- fread(here("data", "all-scores-crowd-forecasts-revised.csv")) |>
+  add_type_and_scale()
+
+weekly_truth_revised <- fread("data/weekly-truth-revised.csv")
+daily_truth_revised <- fread("data/daily-truth-revised.csv")
 weekly_truth <- fread("data/weekly-truth-original.csv")
@@
 
+Participant forecasts 
 ```{r}
-
 models <- forecasts$model |> unique()
@@
 
-plot_df |>
+plot_participant_forecasts <- plot_df |>
   ggplot(aes(x = target_end_date)) +
@@
   theme_scoringutils()
+
+ggsave("output/figures/participant-forecasts.png", plot = plot_participant_forecasts, 
+       width = 7, height = 5)
 ```
 
+Median forecasts ensemble
 ```{r}
@@
 
-comparison_plot(scores |>
-                  filter(horizon == 2), 
+comparison_plot(scores, 
                 models = c("crowd-direct", "EuroCOVIDhub-ensemble"))
@@
 
+plot_performance <- function(scores) {
   df <- scores |>
-  filter(model %in% models_non_human, 
-         scale %in% c("log", "natural")) |>
+    filter(scale %in% c("log", "natural")) |>
     summarise_scores(by = c("horizon", "model", 
@@
 
-cov <- forecasts |>
-  filter(model %in% models_non_human) |>
-  score() |>
-  add_coverage(by = c("model", "target_type", "horizon")) |>
-  summarise_scores(by = c("model", "target_type", "range", "horizon"))
-
-cov_plot <- function(cov_scores, filter_range) {
-  cov_scores |>
+  cov_plot <- function(scores, filter_range) {
+    scores |>
       filter(range == filter_range) |>
+      summarise_scores(by = c("model", "target_type", "range", "horizon")) |>
       ggplot(aes(x = horizon, y = coverage, color = model)) + 
@@
 
-p_cov50 <- cov_plot(cov, 50)
-p_cov90 <- cov_plot(cov, 90)
+  p_cov50 <- cov_plot(scores, 50)
+  p_cov90 <- cov_plot(scores, 90)
 
@@
 
-p_performance
+  return(p_performance)
+}
+
+p_performance <- plot_performance(
+  filter(scores, model %in% models_non_human))
 
@@
 
+```{r}
+# helper function to score forecasts both on the natural and on the log scale
+score_and_add_log <- function(forecasts) {
+  forecasts |>
+  mutate(scale = "natural") |>
+  rbind(forecasts |>
+          mutate(
+            scale = "log", 
+            true_value = log(true_value + 1), 
+            prediction = log(pmax(prediction, 0) + 1)
+          )) |>
+  score(metrics = c("interval_score", "coverage")) |>
+  add_coverage(by = c("target_type", "horizon", "method")) |>
+    add_type_and_scale() |>
+  as.data.table() 
+}
+
+# helper function that filters the forecast data such that we only have 
+# forecasts left where a single person made a forecast both for direct and rt
+filter_comparison_rt_direct <- function(forecasts) {
+  forecasts_comparison <- forecasts |>
+  filter(model %in% models_human) |>
+  mutate(rt_name = ifelse(
+    grepl("(Rt)", model), 
+    gsub("\ \\(Rt\\)", "", model), 
+    NA
+  )) |>
+  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
+  group_by(forecast_date, target_type) |>
+  filter(model %in% rt_name | rt_name %in% model) |>
+  group_by(forecast_date, target_type, method) |>
+  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile)
+}
+
+forecasts_comparison <- filter_comparison_rt_direct(forecasts)
+scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
+
+forecasts_comparison_revised <- filter_comparison_rt_direct(forecasts_revised)
+scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)
+
+scores_forecasts_comparison |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type", "type_and_scale", "name", "method", "scale")) |>
+  ggplot(aes(y = interval_score, x = name, group = method, color = method)) + 
+  geom_point() + 
+  geom_line() + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  theme_scoringutils()
+
+plot_rt_direct_individual <- function(scores) {
+ scores |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
+  arrange(target_type, name, method) |>
+  group_by(target_type, name, scale) |>
+  mutate(diff = rep(diff(interval_score), 2)) |>
+  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
+  mutate(method = stringr::str_to_title(method)) |>
+  ggplot(aes(y = interval_score, x = method, group = name)) + 
+  geom_point(size = 0.5) + 
+  geom_line(aes(color = improvement), linewidth = 0.3) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
+  scale_y_continuous(labels = label_fn) + 
+  theme_scoringutils()
+}
+
+p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)
+
+ggsave("output/figures/comparison-direct-rt-individual.png", 
+       plot = p_comparison_direct_rt_individual, 
+       width = 7, height = 5)
+
+p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)
+
+ggsave("output/figures/comparison-direct-rt-individual-revised.png", 
+       plot = p_comparison_direct_rt_individual, 
+       width = 7, height = 5)
+
+# performance plot for an ensemble of the subset of human and direct forecasts
+forecasts_comparable_ensemble <- forecasts_comparison |>
+  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+  summarise(prediction = median(prediction)) |>
+  mutate(model = str_to_title(method)) |>
+  as.data.table()
+
+scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)
+
+p_performance_comparison <- plot_performance(
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
+       width = 7, height = 8)
+
+# performance plot for an ensemble of the subset of human and direct forecasts
+# using revised death data
+
+forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
+  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+  summarise(prediction = median(prediction)) |>
+  mutate(model = str_to_title(method)) |>
+  as.data.table()
+
+scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)
+
+p_performance_comparison_revised <- plot_performance(
+  scores_comparable_ensemble_revised
+)
+
+ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison, 
+       width = 7, height = 8)
+
+scores_comparable_ensemble |>
+  filter(horizon == 2) |>
+  summarise_scores(by = c("target_type", "horizon", "model", "scale"))
+
+scores |>
+  filter(scale != "sqrt") |>
+  filter(model %in% models_non_human) |>
+  filter(horizon == 2,
+         scale == "log") |>
+  summarise_scores(by = c("target_type", "horizon", "model", "scale"))
+
+```
+
 Table with performance
 
-```{r, eval=FALSE}
-
-coverage <- cov |>
-  select(model, target_type, horizon, coverage_50, coverage_90) |>
-  unique()
-
+```{r, performance-table}
+make_performance_table <- function(scores) {
   scores |>
-  filter(model %in% models_non_human, 
-         scale %in% c("log", "natural")) |>
+    filter(scale %in% c("log", "natural")) |>
     summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
-  inner_join(coverage) |>
     filter(horizon == 2) |>
@@
           align = c("l", "l", rep("c", 4)),
+          caption = "Performance for two-week-ahead forecasts \\label{tab:scores}",
           linesep = c('', '', '', '\\addlinespace'),
@@
                                     "hold_position")) 
+}
+
+scores |>
+  filter(model %in% models_non_human) |>
+  make_performance_table() |>
+  cat(file = 'manuscript/performance-table.tex')
+
+scores_revised |>
+  filter(model %in% models_non_human) |>
+  make_performance_table() |>
+  cat(file = 'manuscript/performance-table-revised.tex')
+
+scores_comparable_ensemble |>
+  make_performance_table() |>
+  cat(file = 'manuscript/performance-table-comparison-rt.tex')
+
 ```
@@
 ```{r}
+
+plot_scores_and_pred <- function(forecasts, scores) {
 
@@
     forecasts |>
-    filter(model %in% models_non_human, 
-           horizon == 2) |>
+      filter(horizon == 2) |>
       filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
-    dcast(... ~ quantile, value.var = "prediction") |>
+      pivot_wider(values_from = prediction, names_from = quantile) |>
+      # dcast(... ~ quantile, value.var = "prediction") |>
       mutate(target_type = paste(target_type, "-", scale)) |>
@@
     scores |>
-    filter(model %in% models_non_human, 
-           horizon == 2) |>
+      filter(horizon == 2) |>
       summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
@@
 
+  return(p_scores_and_pred)  
+}
+
+p_scores_and_pred <- plot_scores_and_pred(
+  filter(forecasts, model %in% models_non_human),
+  filter(scores, model %in% models_non_human)
+)
+
 ggsave("output/figures/scores-and-forecasts.png", plot = p_scores_and_pred, 
+       width = 9, height = 9)
+
+p_scores_and_pred_comparison <- plot_scores_and_pred(
+  forecasts_comparable_ensemble,
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
+       plot = p_scores_and_pred_comparison, 
        width = 9, height = 9)

### 2023-02-10 manuscript/manuscript.tex
update results in paper

@@
 "epiforecasts-EpiExpert\_Rt" (here called "Crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface.
-"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts, after combining direct and $R_t$ forecast from the same individual by taking a quantile-wise mean.
+"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts together, meaning that a participant could enter the the median ensemble twice if she had submitted both a direct and an $R_t$ forecast.
 Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
@@
 If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs that found poor predictive performance beyond this horizon [15].
+We analysed all forecasts together separated by forecasting forecast target (cases or deaths).
+In addition, we compared the performance of the direct vs.
+$R_t$ forecasting approach using instances where we had both a direct forecasts and an $R_t$ forecast from the same person.
+Those forecasts were combined using a quantile-wise median separately for direct and $R_t$ forecasts such that we obtained a modified direct human ensemble and a modified $R_t$ ensemble that included corresponding forecasts from the same forecaster.
 
@@
 uncertainty of forecasts), over-prediction and under-prediction.
-FORMULAS IN APPENDIX.
 Empirical coverage means the percentage of observations falling inside any given prediction interval (e.g.
@@
 On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
-Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July \ref{fig:forecasts-scores}.
-Direct forecasts tended to be higher throughout the study period and especially during the peak than both the $R_t$ forecasts and the Forecast Hub ensemble and therefore showed the poorest performance (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-When looking at WIS values on the log scale (that is, when log-transforming forecasts and observations prior to applying the WIS), scores are more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
-On the log scale, the Forecast Hub ensemble performs slightly worse than all of the human forecasting approaches.
-Similarly to what was observed previously \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all crowd forecasts as well the Forecast Hub ensemble, exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
+Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July \ref{fig:forecasts-scores}.$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+When looking at WIS values on the log scale,
+%(that is, when log-transforming forecasts and observations prior to applying the WIS), 
+scores are more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
+On the log scale, all human forecast ensembles performed sligthly better than the Forecast Hub ensemble.
+Similarly to what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
+This was the case most severely for the $R_t$ forecasts.
 
-\begin{table}[!h]
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}{llcccc}
-\toprule
-Model & Target & WIS & WIS (log scale) & Coverage 50\% & Coverage 90\%\\
-\midrule
-EuroCOVIDhub-ensemble & Cases & 38000 & 0.25 & 0.38 & 0.69\\
-crowd-direct & Cases & 46000 & 0.23 & 0.46 & 0.62\\
-crowd-ensemble & Cases & 40000 & 0.22 & 0.38 & 0.69\\
-crowd-rt & Cases & 39000 & 0.24 & 0.31 & 0.69\\
-\addlinespace
-EuroCOVIDhub-ensemble & Deaths & 57 & 0.16 & 0.62 & 0.85\\
-crowd-direct & Deaths & 78 & 0.17 & 0.46 & 0.77\\
-crowd-ensemble & Deaths & 61 & 0.16 & 0.38 & 0.77\\
-crowd-rt & Deaths & 63 & 0.23 & 0.23 & 0.69\\
-\bottomrule
-\end{tabular}}
-\caption{Table caption}
-\label{tab:scores}
-\end{table}
+\input performance-table.tex
 
@@
 
-In terms of the WIS the Forecast Hub ensemble also outperformed all human forecasts when forecasting deaths.
-This is true as well for WIS values obtained after applying a log-transformation.
-From June to the end of July, direct forecasts of deaths were noticeable higher than $R_t$ death forecasts, whereas in August direct death forecasts were substantially lower than $R_t$ death forecasts.
-On the natural scale, the largest scores occured at the end of the study period when deaths were highest.
-Direct forecasts, which most strongly underpredicted deaths in August therefore showed the poorest performance overall.
+In terms of the WIS on the natural scale, direct human forecasts performed best, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance}B and Table \ref{tab:scores}).
+From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts.
+On the natural scale, the highest scores were concentrated towards the end of the study period, where death incidences were highest, while on the log scale, scores were more evenly distributed across the study period.
+On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
+Note that the picture changes when looking at the revised data which was corrected upwards compared to the original data(DO WE WANT TO DO THAT?
+IS THAT MEANINGFUL?) (see Figure \ref{fig:data-revisions}).
+For the revised data, direct forecasts performed significantly worse than both the Forecast Hub ensemble and the $R_t$ forecasts on the log scale, while still performing better than the $R_t$ forecasts on the natural scale (see Table \ref{tab:scores-revised}).
+%For revised data: On the natural scale, the largest scores occured at the end of the study period when deaths were highest.
 On the log scale, however, underprediction in June and July when death incidences were lower, received more weight and direct crowd forecast outperformed $R_t$ forecasts.
-\textbf{CHECK WITH OTHER GROUND TRUTH DATA?}.
-Empirical coverage for deaths seems broadly comparable for all human forecasts, whereas the Forecast Hub ensemble displays noticeably better coverage for deaths than for cases (Figure \ref{fig:performance}E, F).
+$R_t$ forecasts show strong underdispersion with empirical coverage of the 50\% prediction intervals being close to zero and smaller than 50\% for the 90\% prediction intervals (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+The Forecast Hub ensemble was overdispersed with empirical coverage exceeding nominal coverage, and for the direct human forecasts, empirical coverage was close to nominal coverage.
+
+% For data revision: deaths seems broadly comparable for all human forecasts, whereas the Forecast Hub ensemble displays noticeably better coverage for deaths than for cases (Figure \ref{fig:performance}E, F).
 
 % \subsection*{Individual forecasts?}
+
+\subsection*{Rt forecasts}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/comparison-direct-rt-individual.png}
+\caption{\bf{Change in WIS for forecasters making either direct or $R_t$ forecasts.
+The absolute level for a given forecaster compared to others is not meaningful as forecasters differ in the amounts of forecasts they have submitted.}}
+\label{fig:comparison-direct-rt-individual}
+\end{figure}
+
+For cases, where participants could observe the case forecast implied their $R_t$ forecast, predictive performance was broadly comparable between direct and $R_t$ forecasts for most forecasters (see Figure \ref{comparison-direct-rt-individual}).
+For deaths, predictive performance was significantly worse for most forecasters, suggesting that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not work well.
+IMPORTANTLY THOUGH, THE PICTURE DOES LOOK DIFFERNT WHEN LOOKING AT THE REVISED DATA.
+When looking at an ensemble including only those data points where a single forecaster has submitted both a direct and an $R_t$ forecast, we found a similar picture.
+Performance for cases was broadly comparable, with direct forecasts slighly better than $R_t$ forecasts on the natural scale and slightly worse on the log scale.
+On the log scale, the ensemble of $R_t$ forecasts performed significantly worse than the ensemble of corresponding direct forecasts both on the natural and on the log scale.
 
@@
 
-- hard to draw conclusions, performance changes depending on how you look at the data. 
+- hard to draw conclusions, performance changes depending on how you look at the data. Different horizons look different. 
 - headline result not supported? 

### 2023-02-13 get-paper-data.R
add expert status to data

@@ -202,7 +202,7 @@ clean_forecasts <- function(forecasts) {
     mutate(horizon = as.numeric(substring(target, 1, 1))) |>
     filter(type == "quantile") |>
     select(location, forecast_date, quantile, prediction, 
-           horizon, model, target_end_date, target, target_type) |>
+           horizon, model, target_end_date, target, target_type, expert) |>
     filter(forecast_date >= "2021-05-24", 
            forecast_date <= "2021-08-16") |>
     filter(model != "EpiNow2") |>
@@ -313,6 +313,7 @@ if (file.exists("data/forecast-data.csv")) {
       "EpiNow2"
     )
   ) |>
+    mutate(expert = NA) |>
     clean_forecasts()
   
   root_dirs <- c(here::here("data", "crowd-direct-forecast", "processed-forecast-data"),

### 2023-02-15 manuscript/manuscript.tex
add discussion and conclusion

@@
 
-\title{Wellcome Open Research Article Template}
-\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
+\title{Human Judgement Forecasting of COVID-19 in the UK}
+%\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
 Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
@@
 Infectious disease modelling and forecasting has attracted wide-spread attention during the COVID-19 pandemic and helped inform decision making in public health organisations and governments \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}.
-% Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+% Beginning in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
 Most forecasts were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
@@
 \citet{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
-However, a comparable approach performed slightly worse than computtional models at predicting the 2014/15 outbreak of chikungunya in the Americas.
+However, a comparable approach performed slightly worse than computational models at predicting the 2014/15 outbreak of chikungunya in the Americas.
 \citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths.
@@
 
-% Beginining in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+% Beginning in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
 
@@
 \cite{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
-One approach is to use human forecasts, for example of relevant disease paremeters, as an input to computational modelling.
-Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustements to model outputs.
+One approach is to use human forecasts, for example of relevant disease parameters, as an input to computational modelling.
+Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustments to model outputs.
 \citet{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) and to feed this forecast into a mathematical model in order to obtain forecasts for case and death numbers.
@@
 Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
-All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, German and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
 This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants.
@@
 Submissions to the European Forecast Hub follow a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
-Every week, forecast submitted to the hub were automatically checked for conformity with the required forecasta and all eligible forecasts combined into different ensembles.
+Every week, forecast submitted to the hub were automatically checked for conformity with the required forecasts and all eligible forecasts combined into different ensembles.
 Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisiations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
 From the 29th of July on, the default Forecast Hub ensemble became a median-ensemble.
-The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:num-forecasters}.
+The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:num-forecasters} in the SI.
 
-Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by European Forecast Hub and sourced from the Johns Hopkins University.
+Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by European Forecast Hub and sourced from the Johns Hopkins University (JHU).
 Data were subject to reporting artifacts and revisions.
-The Forecast Hubs marks data as anomalous if in subsequent updates data is changed by more than 5 percent.
-THIS IS THE CASE FOR ALL DEATH FORECASTS...
-CAN WE CHECK THIS?
-See Figure MAKE FIGURE WITH DATA AND UPDATES.
+All data points are marked as anomalous by the European Forecast Hub if in subsequent updates data is changed by more than 5 percent.
+In August 2022 JHU switched the data source for their death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency.
+The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
+The data and revisions are displayed in Figure \ref{fig:plot-data}.
+If not otherwise stated, all results are derived based on the original data available in 2021.
 
@@
 
+When signing up, we asked participants to self-identify as "experts", asking them to tick a box if they worked in infectious disease modelling or had professional experience in any related field.
+
 SAM TO THE RESCUE
@@
 As $R_t$-estimates up to two weeks prior to the forecast data were deemed uncertain, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
-Participants were therefore asked to estimate / predict four $R_t$ values, four of them in the future.
-Upon pressing a button participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
+Participants were therefore asked to estimate/predict six $R_t$ values, four of them in the future.
+Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
 However, due to the high computational complexity and necessary waiting times, participants could not preview the death forecast implied by their current input for $R_t$.
@@
 %, following suggestions made in CITATION.
-We submitted three different ensembles: The first one, "epiforecasts-EpiExpert\_direct" (here called "Crowd direct") was a quantile-wise median ensemble of all the direct forecasts.
+We submitted three different ensembles to the Hub: The first one, "epiforecasts-EpiExpert\_direct" (here called "Crowd direct") was a quantile-wise median ensemble of all the direct forecasts.
 "epiforecasts-EpiExpert\_Rt" (here called "Crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface.
-"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts together, meaning that a participant could enter the the median ensemble twice if she had submitted both a direct and an $R_t$ forecast.
+"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts together.
+A participant could enter the "Crowd combined" ensemble twice if she had submitted both a direct and an $R_t$ forecast.
 Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
@@
 Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
-Everynone who had submitted a forecast for targets in the UK during the tournament period from the 24th of May 2021 to the 16th of August 2021 was deemed a participant and eligible for a prize.
+Everyone who had submitted a forecast for targets in the UK during the tournament period from the 24th of May 2021 to the 16th of August 2021 was deemed a participant and eligible for a prize.
 The first prize was 100 GBP, second prize 50 GBP and third prize 25 GBP.
-For the purpose of the tournament ranking, participants who did not submit a forecast in a given week where assigned the median score of all other participants who submitted a forecast that week.
+Participant performance was determined using the mean weighted interval score (WIS) on the log scale (see details in the next Section, %\ref{sec:analysis}),
+averaged across forecast dates, horizons and forecast targets.
+For the purpose of the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
 The UK crowd forecasting challenge was announced over Twitter and our networks.
@@
 A public leaderboard was available on our website \url{https://epiforecasts.io}.
-Participants could choose to make a direct forecast as well as an $R_t$ forecast and were eligible for prizes twice.
+Participants could choose to make a direct forecast as well as an $R_t$ forecast and were counted as two separate forecasters and eligible for prizes twice.
 Weekly forecasts had to be submitted between Sunday 12pm and Monday 8pm UK time.
@@
 \subsection*{Analysis}
+\label{sec:analysis}
 
 Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}.
-If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs that found poor predictive performance beyond this horizon [15].
-We analysed all forecasts together separated by forecasting forecast target (cases or deaths).
-In addition, we compared the performance of the direct vs.
+If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022a, bracherShorttermForecastingCOVID192021}.
+We analysed all forecasts stratified by forecast target (cases or deaths), forecast horizon, and forecast approach.
+We compared the performance of the direct vs.
 $R_t$ forecasting approach using instances where we had both a direct forecasts and an $R_t$ forecast from the same person.
-Those forecasts were combined using a quantile-wise median separately for direct and $R_t$ forecasts such that we obtained a modified direct human ensemble and a modified $R_t$ ensemble that included corresponding forecasts from the same forecaster.
 
-We scored forecasts using the weighted interval score (WIS, formulas given in the SI in Section \ref{sec:wis}) \cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic callibration.
-The WIS is proper scoring rule (with smaller values implying better performance).
-A forecaster, in expectation, optimises her score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$ and is therefore incentivised to report her true belief.
-The WIS can be seen as a generalisation of the absolute error to predictive distribution in a quantile-based format and represents an approximation of the continuous ranked probability score (CRPS).
-The WIS can be decomposed into three separate penalty components: forecast spread (i.e.
+For self-reported "experts" and "non-experts", a simple comparison of scores would be confounded by individual differences in participation and the timing of individual forecasts.
+We therefore compared the performance of self-reported "experts" vs.
+"non-experts" by creating and evaluating two modified median ensembles, one including only "experts" and the other only "non-experts".
+
+% Those forecasts were combined using a quantile-wise median separately for direct and $R_t$ forecasts such that we obtained a modified direct human ensemble and a modified $R_t$ ensemble that included only corresponding forecasts from the same forecaster.
+
+We scored forecasts using the weighted interval score (WIS, formulas given in the SI) %in Section \ref{sec:wis}) 
+\cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration.
+The WIS is a proper scoring rule (with smaller values implying better performance).
+A forecaster, in expectation, optimises her score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report her true belief.
+The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
+The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
 uncertainty of forecasts), over-prediction and under-prediction.
-Empirical coverage means the percentage of observations falling inside any given prediction interval (e.g.
+Empirical coverage refers to the percentage of observations falling inside any given prediction interval (e.g.
 the cumulative percentage of observed values that fall inside all 50\% prediction intervals).
@@
 
-\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
+\citet{bosseTransformationForecastsEvaluating2023} suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
 We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
@@
 A total number of 90 participants submitted forecasts (more precisely, forecasts were submitted from 90 different accounts, some of them anonymous).
-The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51.
+Out of 90 participants, 21 self-identified as "experts", i.e.
+stated they had professional experience in infectious disease modelling or a related field.
+
+The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51 (compared to 6, 2, and 10 in \cite{bosseComparingHumanModelbased2022}).
 With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13.
-The median number of submissions was one, meaning that most forecasters dropped out after their first submission.
+The median number of submissions from a single individual was one, meaning that most forecasters dropped out after their first submission.
 Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study.
@@
 
-% X of those self-identified as ‘expert’ in either forecasting or epidemiology.
-
-\begin{figure}[H]
+\begin{figure*}
 \centering
 \includegraphics[width=0.99\textwidth]{../output/figures/scores-and-forecasts.png}
-\caption{\bf{Test}}
+\caption{\bf{Forecasts and corresponding WIS for 2-week ahead forecasts of cases and deaths from COVID-19 in the UK.} A: 50\% prediction intervals (coloured bars) and observed values (black line and points) for cases and deaths on the natural scale.
+B: Corresponding WIS values, decomposed into dispersion, overprediction and underprediction.
+C: 50\% prediction intervals on the log scale, i.e.
+after applying the natural logarithm to all forecasts and observations.
+D: Corresponding WIS on the log scale, i.e.
+the WIS applied to the log-transformed forecasts and observations.}
 \label{fig:forecasts-scores} 
-\end{figure}
+\end{figure*}
 
-\begin{figure}[H]
+\begin{figure*}
 \centering
 \includegraphics[width=0.99\textwidth]{../output/figures/performance.png}
-\caption{\bf{Test}}
+\caption{\bf{Predictive performance across forecast horizons.} A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
+E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
 \label{fig:performance}
-\end{figure}
+\end{figure*}
 
@@
 
-For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the weighted interval score (WIS).
+For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
 On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
-Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July \ref{fig:forecasts-scores}.$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
+$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
 When looking at WIS values on the log scale,
@@
 scores are more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
-On the log scale, all human forecast ensembles performed sligthly better than the Forecast Hub ensemble.
+On the log scale, all human forecast ensembles performed slightly better than the Forecast Hub ensemble.
 Similarly to what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
-This was the case most severely for the $R_t$ forecasts.
+Underdispersion was most severe for the $R_t$ forecasts.
 
@@
 
-In terms of the WIS on the natural scale, direct human forecasts performed best, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance}B and Table \ref{tab:scores}).
-From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts.
+In terms of the WIS on the natural scale, direct human forecasts performed best, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
 On the natural scale, the highest scores were concentrated towards the end of the study period, where death incidences were highest, while on the log scale, scores were more evenly distributed across the study period.
 On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
-Note that the picture changes when looking at the revised data which was corrected upwards compared to the original data(DO WE WANT TO DO THAT?
+%Note that the picture changes when looking at the revised data which was corrected upwards compared to the original data(DO WE WANT TO DO THAT?
 IS THAT MEANINGFUL?) (see Figure \ref{fig:data-revisions}).
 For the revised data, direct forecasts performed significantly worse than both the Forecast Hub ensemble and the $R_t$ forecasts on the log scale, while still performing better than the $R_t$ forecasts on the natural scale (see Table \ref{tab:scores-revised}).
-%For revised data: On the natural scale, the largest scores occured at the end of the study period when deaths were highest.
+%For revised data: On the natural scale, the largest scores occurred at the end of the study period when deaths were highest.
 On the log scale, however, underprediction in June and July when death incidences were lower, received more weight and direct crowd forecast outperformed $R_t$ forecasts.
@@
 
-\begin{figure}[H]
+\begin{figure*}
 \centering
 \includegraphics[width=0.99\textwidth]{../output/figures/comparison-direct-rt-individual.png}
-\caption{\bf{Change in WIS for forecasters making either direct or $R_t$ forecasts.
-The absolute level for a given forecaster compared to others is not meaningful as forecasters differ in the amounts of forecasts they have submitted.}}
+\caption{\bf{Comparison of predictive performance of individual forecasters using either the direct forecasting or $R_t$ interface}.
+Comparisons are based only on those instances where forecasters have submitted a prediction using both interfaces.
+The absolute level for a given forecaster relative to others is not meaningful as forecasters differ in the amounts of forecasts they have submitted and when.}
 \label{fig:comparison-direct-rt-individual}
-\end{figure}
+\end{figure*}
 
-For cases, where participants could observe the case forecast implied their $R_t$ forecast, predictive performance was broadly comparable between direct and $R_t$ forecasts for most forecasters (see Figure \ref{comparison-direct-rt-individual}).
-For deaths, predictive performance was significantly worse for most forecasters, suggesting that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not work well.
-IMPORTANTLY THOUGH, THE PICTURE DOES LOOK DIFFERNT WHEN LOOKING AT THE REVISED DATA.
-When looking at an ensemble including only those data points where a single forecaster has submitted both a direct and an $R_t$ forecast, we found a similar picture.
-Performance for cases was broadly comparable, with direct forecasts slighly better than $R_t$ forecasts on the natural scale and slightly worse on the log scale.
+For cases, where participants could observe the case forecast implied by their $R_t$ forecast, predictive performance was similar between corresponding direct and $R_t$ forecasts for most forecasters who had submitted both (see Figure \ref{fig:comparison-direct-rt-individual}).
+For deaths, predictive performance of their $R_t$ forecasts was significantly worse than their corresponding direct forecasts for most forecasters, suggesting that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately.
+This inadequacy may be related to the change in the case fatality ratio over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK.
+%IMPORTANTLY THOUGH, THE PICTURE DOES LOOK DIFFERNT WHEN LOOKING AT THE REVISED DATA.
+
+% When looking at an ensemble including only those data points where a single forecaster has submitted both a direct and an $R_t$ forecast, we found a similar picture.
+Performance for cases was broadly comparable, with direct forecasts slightly better than $R_t$ forecasts on the natural scale and slightly worse on the log scale.
 On the log scale, the ensemble of $R_t$ forecasts performed significantly worse than the ensemble of corresponding direct forecasts both on the natural and on the log scale.
 
+\subsection*{Experts and Non-Experts}
+
+A comparison of "experts" and "non-experts" (self-reported experience in infectious disease modelling or a related field) showed broadly comparable performance, both on the natural and the log scale (see Figure \ref{fig:performance-experts}).
+Precise results were susceptible to whether or not $R_t$ forecasts were included, with performance of "experts" on deaths being slightly worse than performance of "non-experts" when only looking at direct forecasts.
+
+% As a direct comparison of individual scores between "experts" and "non-experts" is biased by missing forecasts,
+
 \section*{Discussion}
-The discussion should include the implications of the article results in view of prior work in this field.
+% The discussion should include the implications of the article results in view of prior work in this field.
 
-- hard to draw conclusions, performance changes depending on how you look at the data. Different horizons look different. 
-- headline result not supported? 
-- still some evidence that forecasting deaths is easier for computers
--
+% \paragraph{Summary}
 
-- Rt forecasting worked reasonably well. The interface was a bit clunky. Required more expertise. But a potential way to reduce the workload.
+In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
+Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deahts and b) forecasts of the effective reproduction number $R_t$ which were mapped to cases using the renewal equation and convolved over a delay distribution to obtain deaths.
+We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
+For deaths, performance of direct human forecasts was comparable to that of the Forecast Hub ensemble, while $R_t$ forecasts performed significantly worse.
+Model rankings were not always robust, but sometimes changed depending on how forecasts are evaluated.
+For example, model rankings could differ when focusing on a different forecast horizon, or analysing WIS values on the natural scale vs.
+on the log scale (i.e.
+after log-transforming both forecasts and observations).
 
-- interesting that combining Rt and direct forecasts for cases improved performance.
+% \paragraph{Literature context}
 
-Limitations
-- The tournament setting may have distorted results as people were more eager to participate
-- ensemble was better
+In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
+They hypothesised that computational models might have an advantage over human forecsters when predicting deaths, benefitting from the ability to model the delays and exact epidemiological relationships between different leading and lagged indicators.
+\citet{mcandrewChimericForecastingCombining2022} similarly found that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
+We could not replicate these findings, and found that in terms of the WIS on the natural scale, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differ).
+Looking at scores on the log scale suggests overall comparable performance of the Hub ensemble and the our direct human forecasts on both cases and deaths.
+Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to the rise of the Delta variant in the UK starting in the beginning of May.
+The Delta variant had a higher case fatality ratio (CFR) than previous variants CITATION.
+Humans may have found it easier than computational models to adapt to the fact that the relationship between cases and deaths continuously changed during the study period.
+
+% \paragraph{Effort to recruit participants}.
+
+Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to recruit a large number of participants.
+Focused public outreach efforts such as creating a dedicated website, anouncing an official tournament, providing a public leaderboard, sending weekly emails with details on past performance and weekly announcements on Twitter, did noticeably increase participation compared to the previous study in Germany and Poland.
+Nevertheless, particaption was still relatively low and most forecasters only submitted a single forecast.
+\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms such as Metaculus or Good Judgement Open that provide access to a more elegant user interface and a large existing user base may be helpful in recruting a larger number of participants.
+
+% \paragraph{Expertise}
+
+In our study, an ensemble of only self-reported "experts" did not outperform "non-experts", and even performed worse when looking only at direct forecasts of deaths (but comparably when including $R_t$ forecasts).
+Past expert forecasts of COVID-19 \cite{recchiaHowWellDid2021} had found predictions from experts to outperform those of non-experts.
+Our results should be taken with care in light of low sample sizes, and given that expert status was self-reported.
+Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
+
+% \paragraph{Rt forecasting}
+
+This study explored a novel method of forecasting infectious diseases that combines a human forecast of the effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths.
+One appeal of this approach is that the forecaster only needs to provide one quantity that directly represents what she believes about the force of the epidemic in the future.
+Computational modelling would then take care of dealing with details such as reporting delays, generation intervals, or the relationship between different indicators.
+This could help reduce cognitive load, at least for forecasters who have an intuitive understanding of $R_t$.
+Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
+In our study, $R_t$ forecasts of cases were comparable to direct forecasts.
+However, given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality just forecasting cases and merely updated their $R_t$ forecast so they would obtain the desired case forecast.
+$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
+\textbf{SAM TO THE RESCUE} The relationship between infections and deaths used to map from $R_t$ to deaths was automatically estimated from past data, and forecasters had no way of adjusting it.
+The rise of the Delta variant with a subsequent gradual change in the case fatality ratio (CFR) may have affected $R_t$ forecasts, as the changing CFR was not directly accounted for and estimated distributions from past data only slowly updated to new circumstances.
+One potential solution for this would be to allow humans to adjust the CFR and other model parameters manually as part of their forecast.
+
+% \paragraph{Susceptibility of results}
+
+Overall, results of our study should be taken with some caution due to a number of important limitations.
+Firstly, our study was restricted to one location and to a relatively short period of thirteen weeks.
+Secondly, there are many confounding factors that can influence results.
+These include for example the fact that different participants made forecasts at different points in time and that subgroups of interest (e.g.
+"experts", or $R_t$ forecasts) had different numbers of forecasters.
+In addition, there are many researcher degrees of freedom that influence results, for example how indidual forecasts are combined to an ensemble and how forecasts are evaluated.
+Prizes to the human forecasters, for example were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
+Had we chosen to instead measure WIS on the natural scale, rankings and payouts would have been different.
 
 \section*{Conclusions}
-Please state what you think are the main conclusions that can be realistically drawn from the findings in the paper, taking care not to make claims that cannot be supported.
+% Please state what you think are the main conclusions that can be realistically drawn from the findings in the paper, taking care not to make claims that cannot be supported.
 
-Seems that a model of human forecasters is alright. Combining human and model based output is still interesting. 
-interface difficult
+Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles can produce forecasts that are on par with those of an ensemble of computational models.
+Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited.
+Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths.
+However, the approach has appealing properties and better results are likely achievable.
+More research is required to obtain a better understanding of the role of subject matter expertise in infectious disease forecasting.
+Our results underline that it is difficult to evaluate forecast performance devoid of context that helps inform what a good or a bad forecast is.
+Different ways to look at the data let different forecasts appear good or bad.
+Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture.
 
@@
 
+% ================================= References =============================== %
+\clearpage
+
+{\small\bibliographystyle{unsrtnat}
+\bibliography{software, uk-forecasting-challenge}}
+
+\bigskip
+\clearpage
+
 % =================================== Appendix =============================== %
+
+% Make sure that Appendix tables and Figures are separate
+% https://tex.stackexchange.com/questions/520193/is-it-possible-to-except-figures-in-the-appendix-from-being-repositioned-by-endf
+\processdelayedfloats
+\csname efloat@restorefloats\endcsname
 
@@
 \section*{Supplementary information}
-% \renewcommand{\thefigure}{SI.\arabic{figure}}
-% \setcounter{figure}{0}
-% \renewcommand{\thetable}{SI.\arabic{table}} \setcounter{table}{0}
+\renewcommand{\thefigure}{SI.\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{SI.\arabic{table}} \setcounter{table}{0}
 
 \subsection*{Weighted interval score}
-label{sec:wis}
+\label{sec:wis}
+
 The weighted interval score (smaller values are better) is a proper scoring rule for quantile forecasts.
-It converges to the continuos ranked probability score (which itself is a generalisation of the absolute error to probabilistic forecasts) for an increasing number of intervals.
+It converges to the continuous ranked probability score (which itself is a generalisation of the absolute error to probabilistic forecasts) for an increasing number of intervals.
 The score can be decomposed into a dispersion (uncertainty) component and penalties for over- and underprediction.
@@
 
-\begin{figure}[H]
+\input SI/epinow2.tex
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/plot-data.png}
+\caption{\bf{Observed cases and deaths of COVID-19 in the UK}.
+A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021.
+Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.
+Red dots represent days for which the original data and the revised data disagreed by more than five percent.
+B: Revised data available as of February 14 2023.
+In August, Johns Hopkins University that provided the data switched the data stream for their death forecasts to reflect the number of death certificates that mentioned COVID-19 rather than the number of people who died within 28 days of a positive test.
+C: Difference between the original and revised weekly death numbers.}
+\label{fig:plot-data}
+\end{figure*}
+
+\begin{figure*}
 \centering
@@
 \label{fig:screenshot-classical}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[H]
+\begin{figure*}
 \centering
@@
 \label{fig:screenshot-rt}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[H]
+\begin{figure*}
 \centering
@@
 \label{fig:num-forecasters}
-\end{figure}
+\end{figure*}
 
-\clearpage
-
-{\small\bibliographystyle{unsrtnat}
-\bibliography{software, uk-forecasting-challenge}}
-
-\bigskip
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
+\caption{\bf{Predictive performance of self-reported "experts" and "non-experts" across forecast horizons.} Forecasts from "experts" and "non-experts" were combined to two separate median ensembles, including both direct and $R_t$ forecasts.
+A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
+E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
+\label{fig:performance-experts}
+\end{figure*}
 

### 2023-02-15 manuscript/SI/epinow2.tex
update code and figures

% % Options for packages loaded elsewhere
% \PassOptionsToPackage{unicode}{hyperref}
% \PassOptionsToPackage{hyphens}{url}
% %
% \documentclass[
% ]{article}
% \usepackage{amsmath,amssymb}
% \usepackage{lmodern}
% \usepackage{iftex}
% \ifPDFTeX
%   \usepackage[T1]{fontenc}
%   \usepackage[utf8]{inputenc}
%   \usepackage{textcomp} % provide euro and other symbols
% \else % if luatex or xetex
%   \usepackage{unicode-math}
%   \defaultfontfeatures{Scale=MatchLowercase}
%   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
% \fi
% % Use upquote if available, for straight quotes in verbatim environments
% \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% \IfFileExists{microtype.sty}{% use microtype if available
%   \usepackage[]{microtype}
%   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
% }{}
% \makeatletter
% \@ifundefined{KOMAClassName}{% if non-KOMA class
%   \IfFileExists{parskip.sty}{%
%     \usepackage{parskip}
%   }{% else
%     \setlength{\parindent}{0pt}
%     \setlength{\parskip}{6pt plus 2pt minus 1pt}}
% }{% if KOMA class
%   \KOMAoptions{parskip=half}}
% \makeatother
% \usepackage{xcolor}
% \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
% \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
% \hypersetup{
%   pdftitle={Text S1. Further details on the semi-mechanistic forecasting models},
%   hidelinks,
%   pdfcreator={LaTeX via pandoc}}
% \urlstyle{same} % disable monospaced font for URLs
% \usepackage[margin=1in]{geometry}
% \usepackage{longtable,booktabs,array}
% \usepackage{calc} % for calculating minipage widths
% % Correct order of tables after \paragraph or \subparagraph
% \usepackage{etoolbox}
% \makeatletter
% \patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% \makeatother
% % Allow footnotes in longtable head/foot
% \IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
% \makesavenoteenv{longtable}
% \usepackage{graphicx}
% \makeatletter
% \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
% \def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
% \makeatother
% % Scale images if necessary, so that they will not overflow the page
% % margins by default, and it is still possible to overwrite the defaults
% % using explicit options in \includegraphics[width, height, ...]{}
% \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% % Set default figure placement to htbp
% \makeatletter
% \def\fps@figure{htbp}
% \makeatother
% \setlength{\emergencystretch}{3em} % prevent overfull lines
% \providecommand{\tightlist}{%
%   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
% \setcounter{secnumdepth}{5}
% \newlength{\cslhangindent}
% \setlength{\cslhangindent}{1.5em}
% \newlength{\csllabelwidth}
% \setlength{\csllabelwidth}{3em}
% \newlength{\cslentryspacingunit} % times entry-spacing
% \setlength{\cslentryspacingunit}{\parskip}
% \newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
%  {% don't indent paragraphs
%   \setlength{\parindent}{0pt}
%   % turn on hanging indent if param 1 is 1
%   \ifodd #1
%   \let\oldpar\par
%   \def\par{\hangindent=\cslhangindent\oldpar}
%   \fi
%   % set entry spacing
%   \setlength{\parskip}{#2\cslentryspacingunit}
%  }%
%  {}
% \usepackage{calc}
% \newcommand{\CSLBlock}[1]{#1\hfill\break}
% \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
% \newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
% \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \ifLuaTeX
%   \usepackage{selnolig}  % disable illegal ligatures
% \fi
% 
% \title{Text S1. Further details on the semi-mechanistic forecasting models}
% \author{}
% \date{\vspace{-2.5em}}
% 
% \begin{document}
% \maketitle
% 
% {
% \setcounter{tocdepth}{2}
% \tableofcontents
% }
% \hypertarget{text-s1.-further-details-on-the-semi-mechanistic-forecasting-models}{%
% \subsection{Text S1. Further details on the semi-mechanistic forecasting models}\label{text-s1.-further-details-on-the-semi-mechanistic-forecasting-models}}

\hypertarget{renewal-equation-model}{%
\subsubsection{Renewal equation model}\label{renewal-equation-model}}

The model was initialised prior to the first observed data point by assuming constant exponential growth for the mean of assumed delays from infection to case report.

\begin{align}
  I_{t} &= I_0 \exp  \left(r t \right)  \\
  I_0 &\sim \mathcal{LN}(\log I_{obs}, 0.2) \\
  r &\sim \mathcal{LN}(r_{obs}, 0.2) 
\end{align}

Where \(I_{obs}\) and \(r_{obs}\) are estimated from the first week of observed data. For the time window of the observed data infections were then modelled by weighting previous infections by the generation time and scaling by the instantaneous reproduction number. These infections were then convolved to cases by date (\(O_t\)) and cases by date of report (\(D_t\)) using log-normal delay distributions. This model can be defined mathematically as follows,

\begin{align}
  \log R_{t} &= \log R_{t-1} + \mathrm{GP}_t \\
  I_t &= R_t \sum_{\tau = 1}^{15} w(\tau | \mu_{w}, \sigma_{w}) I_{t - \tau} \\
  O_t &= \sum_{\tau = 0}^{15} \xi_{O}(\tau | \mu_{\xi_{O}}, \sigma_{\xi_{O}}) I_{t-\tau} \\
  D_t &= \alpha \sum_{\tau = 0}^{15} \xi_{D}(\tau | \mu_{\xi_{D}}, \sigma_{\xi_{D}}) O_{t-\tau} \\ 
  C_t &\sim \mathrm{NB}\left(\omega_{(t \mod 7)}D_t, \phi\right)
\end{align}

Where,
\begin{align}
     w &\sim \mathcal{G}(\mu_{w}, \sigma_{w}) \\
    \xi_{O} &\sim \mathcal{LN}(\mu_{\xi_{O}}, \sigma_{\xi_{O}}) \\
    \xi_{D} &\sim \mathcal{LN}(\mu_{\xi_{D}}, \sigma_{\xi_{D}}) 
\end{align}

This model used the following priors for cases,

\begin{align}
     R_0 &\sim \mathcal{LN}(0.079, 0.18) \\
    \mu_w &\sim \mathcal{N}(3.6, 0.7) \\
    \sigma_w &\sim \mathcal{N}(3.1, 0.8) \\
    \mu_{\xi_{O}} &\sim \mathcal{N}(1.62, 0.064) \\
    \sigma_{\xi_{O}} &\sim \mathcal{N}(0.418, 0.069) \\
    \mu_{\xi_{D}} &\sim \mathcal{N}(0.614, 0.066) \\
    \sigma_{\xi_{D}} &\sim \mathcal{N}(1.51, 0.048) \\
    \alpha &\sim \mathcal{N}(0.25, 0.05) \\
    \frac{\omega}{7} &\sim \mathrm{Dirichlet}(1, 1, 1, 1, 1, 1, 1) \\
    \phi &\sim \frac{1}{\sqrt{\mathcal{N}(0, 1)}}
\end{align}

and updated the reporting process as follows when forecasting deaths,

\begin{align}
    \mu_{\xi_{D}} &\sim \mathcal{N}(2.29, 0.076) \\
    \sigma_{\xi_{D}} &\sim \mathcal{N}(0.76, 0.055) \\
    \alpha &\sim \mathcal{N}(0.005, 0.0025) 
\end{align}

\(\alpha\), \(\mu\), \(\sigma\), and \(\phi\) were truncated to be greater than 0 and with \(\xi\), and \(w\) normalised to sum to 1.

The prior for the generation time was sourced from (Ganyani et al. 2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior (Lauer et al. 2020) for \(\xi_{O}\). This resulted in a gamma-distributed generation time with mean 3.6 days (standard deviation (SD) 0.7), and SD of 3.1 days (SD 0.8) for all estimates. We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data (Xu et al. 2020; Abbott, Sherratt, et al. 2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution (Abbott, Hellewell, et al. 2020; epiforecasts.io/covid 2020; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}''} n.d.; Stan Development Team 2020).

\(GP_t\) is an approximate Hilbert space Gaussian process as defined in (Riutort-Mayol et al. 2020) using a Matern 3/2 kernel using a boundary factor of 1.5 and 17 basis functions (20\% of the number of days used in fitting). The length scale of the Gaussian process was given a log-normal prior with a mean of 21 days, and a standard deviation of 7 days truncated to be greater than 3 days and less than 60 days. The magnitude of the Gaussian process was assumed to be normally distributed centred at 0 with a standard deviation of 0.1.

From the forecast time horizon (\(T\)) and onwards the last value of the Gaussian process was used (hence \(R_t\) was assumed to be fixed) and latent infections were adjusted to account for the proportion of the population that was susceptible to infection as follows,

\begin{equation}
    I_t = (N - I^c_{t-1}) \left(1 - \exp \left(\frac{-I'_t}{N - I^c_{T}}\right)\right),
\end{equation}

where \(I^c_t = \sum_{s< t} I_s\) are cumulative infections by \(t-1\) and \(I'_t\) are the unadjusted infections defined above. This adjustment is based on that implemented in the \texttt{epidemia} R package (Scott et al. 2020; Bhatt et al., n.d.).

\hypertarget{convolution-model}{%
\paragraph{Convolution model}\label{convolution-model}}

The convolution model shares the same observation model as the renewal model but rather than assuming that an observation is predicted by itself using the renewal equation instead assumes that it is predicted entirely by another observation after some parametric delay. It can be defined mathematically as follows,

\begin{equation} 
    D_{t} \sim \mathrm{NB}\left(\omega_{(t \mod 7)} \alpha \sum_{\tau = 0}^{30} \xi(\tau | \mu, \sigma) C_{t-\tau},  \phi \right)
\end{equation}

with the following priors,

\begin{align}
    \frac{\omega}{7} &\sim \mathrm{Dirichlet}(1, 1, 1, 1, 1, 1, 1) \\
    \alpha &\sim \mathcal{N}(0.01, 0.02) \\
    \xi &\sim \mathcal{LN}(\mu, \sigma) \\
    \mu &\sim \mathcal{N}(2.5, 0.5) \\
\sigma &\sim \mathcal{N}(0.47, 0.2) \\
\phi &\sim \frac{1}{\sqrt{\mathcal{N}(0, 1)}}
\end{align}

with \(\alpha\), \(\mu\), \(\sigma\), and \(\phi\) truncated to be greater than 0 and with \(\xi\) normalised such that \(\sum_{\tau = 0}^{30} \xi(\tau | \mu, \sigma) = 1\).

\hypertarget{model-fitting}{%
\subsubsection{Model fitting}\label{model-fitting}}

Both models were implemented using the \texttt{EpiNow2} R package (version 1.3.3) (Abbott, Hellewell, et al. 2020). Each forecast target was fitted independently for each model using Markov-chain Monte Carlo (MCMC) in stan (Stan Development Team 2020). A minimum of 4 chains were used with a warmup of 250 samples for the renewal equation-based model and 1000 samples for the convolution model. 2000 samples total post warmup were used for the renewal equation model and 4000 samples for the convolution model. Different settings were chosen for each model to optimise compute time contingent on convergence. Convergence was assessed using the R hat diagnostic (Stan Development Team 2020). For the convolution model forecast the case forecast from the renewal equation model was used in place of observed cases beyond the forecast horizon using 1000 posterior samples. 12 weeks of data was used for both models though only 3 weeks of data were included in the likelihood for the convolution model.

\clearpage

\hypertarget{refs}{}
% \begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-epinow2}{}}%
Abbott, Sam, Joel Hellewell, Joe Hickson, James Munday, Katelyn Gostic, Peter Ellis, Katharine Sherratt, et al. 2020. {``EpiNow2: Estimate Real-Time Case Counts and Time-Varying Epidemiological Parameters.''} \emph{-} - (-): --. \url{https://doi.org/10.5281/zenodo.3957489}.

\leavevmode\vadjust pre{\hypertarget{ref-covidregionaldata}{}}%
Abbott, Sam, Katharine Sherratt, Jonnie Bevan, Hamish Gibbs, Joel Hellewell, James Munday, Patrick Barks, Paul Campbell, Flavio Finger, and Sebastian Funk. 2020. {``Covidregionaldata: Subnational Data for the Covid-19 Outbreak.''} \emph{-} - (-): --. \url{https://doi.org/10.5281/zenodo.3957539}.

\leavevmode\vadjust pre{\hypertarget{ref-bhattSemiMechanisticBayesianModeling}{}}%
Bhatt, Samir, Neil Ferguson, Seth Flaxman, Axel Gandy, Swapnil Mishra, and James A Scott. n.d. {``Semi-{Mechanistic Bayesian} Modeling of {COVID-19} with {Renewal Processes},''} 14.

\leavevmode\vadjust pre{\hypertarget{ref-epiforecasts.ioux2fcovidCovid19TemporalVariation2020}{}}%
epiforecasts.io/covid. 2020. {``Covid-19: {Temporal} Variation in Transmission During the {COVID-19} Outbreak.''} {Covid-19}. 2020. \url{https://epiforecasts.io/covid/}.

\leavevmode\vadjust pre{\hypertarget{ref-EvaluatingUseReproduction}{}}%
{``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}.''} n.d. Accessed May 30, 2021. \url{https://www.medrxiv.org/content/10.1101/2020.10.18.20214585v1}.

\leavevmode\vadjust pre{\hypertarget{ref-generationinterval}{}}%
Ganyani, Tapiwa, Cecile Kremer, Dongxuan Chen, Andrea Torneri, Christel Faes, Jacco Wallinga, and Niel Hens. 2020. {``Estimating the Generation Interval for Coronavirus Disease (COVID-19) Based on Symptom Onset Data, March 2020.''} \emph{Eurosurveillance} 25 (17).

\leavevmode\vadjust pre{\hypertarget{ref-incubationperiod}{}}%
Lauer, Stephen A, Kyra H Grantz, Qifang Bi, Forrest K Jones, Qulu Zheng, Hannah R Meredith, Andrew S Azman, Nicholas G Reich, and Justin Lessler. 2020. {``The Incubation Period of Coronavirus Disease 2019 (COVID-19) from Publicly Reported Confirmed Cases: Estimation and Application.''} \emph{Annals of Internal Medicine} 172 (9): 577--82.

\leavevmode\vadjust pre{\hypertarget{ref-approxGP}{}}%
Riutort-Mayol, Gabriel, Paul-Christian Bürkner, Michael R. Andersen, Arno Solin, and Aki Vehtari. 2020. {``Practical Hilbert Space Approximate Bayesian Gaussian Processes for Probabilistic Programming.''} \url{https://arxiv.org/abs/2004.11408}.

\leavevmode\vadjust pre{\hypertarget{ref-epidemia}{}}%
Scott, James A., Axel Gandy, Swapnil Mishra, Juliette Unwin, Seth Flaxman, and Samir Bhatt. 2020. {``Epidemia: Modeling of Epidemics Using Hierarchical Bayesian Models.''} \url{https://imperialcollegelondon.github.io/epidemia/}.

\leavevmode\vadjust pre{\hypertarget{ref-rstan}{}}%
Stan Development Team. 2020. {``RStan: The r Interface to Stan.''} \url{http://mc-stan.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-kraemer2020epidemiological}{}}%
Xu, Bo, Bernardo Gutierrez, Sarah Hill, Samuel Scarpino, Alyssa Loskill, Jessie Wu, Kara Sewalk, et al. 2020. {``Epidemiological Data from the nCoV-2019 Outbreak: Early Descriptions from Publicly Available Data.''} \url{http://virological.org/t/epidemiological-data-from-the-ncov-2019-outbreak-early-descriptions-from-publicly-available-data/337}.

% \end{CSLReferences}
% 
% \end{document}

### 2023-02-16 manuscript/manuscript.tex
Add abstract and link to code

@@
 
-Abstracts should be up to 300 words and provide a succinct summary of the article.
+%Abstracts should be up to 300 words and provide a succinct summary of the article.
 Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
@@
 \textbf{Background}
+Accurate forecasts of infectious diseases can aid public health decision-making.
+In the past, two studies found ensembles of human judgement forecasts of COVID-19 to show predictive performance comparable to ensembles of computational models, at least when predicting case incidences.
+We present a follow-up to a study conducted in Germany and Poland and investigate a novel approach to combine human judgement and epidemiological modelling.
 
 \textbf{Methods}
+From May 24th to August 16th 2021, we elicited weekly one to four-week-ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
+Forecasts were elicited as part of a tournament, the "UK Crowd Forecasting Challenge", and a median ensemble of all forecasts was submitted to the European Forecast Hub.
+Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
+This forecast was automatically mapped to cases and deaths using the renewal equation and a convolution over a delay distribution.
+Forecasts were scored using the weighted interval score on the original forecasts, as well as after applying the natural logarithm to both forecasts and observations.
 
 \textbf{Results}
+The ensemble of human forecasters overall performed comparably to the offical European Forecast Hub ensemble on both cases and deaths, although results were susceptible to changes in details of the evaluation.
+$R_t$ foreasts performed comparably to direct forecasts on cases, but worse on deaths.
 
 \textbf{Conclusions}
+We find more evidence that human judgement can be useful in predicting the large-scale evolution of an infectious disease such as COVID-19.
+The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
 
@@
 Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
+All code and data, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
 

### 2023-02-16 manuscript/manuscript.tex
correct errors in manuscript, add citation

@@
 
-\textbf{Background}
+\paragraph{Background}
+
 Accurate forecasts of infectious diseases can aid public health decision-making.
@@
 
-\textbf{Methods}
+\paragraph{Methods}
+
 From May 24th to August 16th 2021, we elicited weekly one to four-week-ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
@@
 
-\textbf{Results}
+\paragraph{Results}
+
 The ensemble of human forecasters overall performed comparably to the offical European Forecast Hub ensemble on both cases and deaths, although results were susceptible to changes in details of the evaluation.
-$R_t$ foreasts performed comparably to direct forecasts on cases, but worse on deaths.
+$R_t$ foreasts performed comparably to direct forecasts ona cases, but worse on deaths.
+Self-identified experts did not outperform non-experts.
 
-\textbf{Conclusions}
-We find more evidence that human judgement can be useful in predicting the large-scale evolution of an infectious disease such as COVID-19.
+\paragraph{Conclusions}
+
+Human judgement can be useful in predicting the large-scale evolution of an infectious disease such as COVID-19.
 The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
@@
 Forecasts are made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
-While forecasts were submitted on Mondays, weeks were defined as epiweeks.
-ending on a Saturday and starting on Sunday.
+While forecasts were submitted on Mondays, weeks were defined as epiweeks, ending on a Saturday and starting on Sunday.
 Forecast horizons were therefore in fact 5, 12, 19 and 26 days.
 Submissions to the European Forecast Hub follow a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
-Every week, forecast submitted to the hub were automatically checked for conformity with the required forecasts and all eligible forecasts combined into different ensembles.
+Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
 Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisiations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
@@
 
-Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by European Forecast Hub and sourced from the Johns Hopkins University (JHU).
+Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by the European Forecast Hub and sourced from the Johns Hopkins University (JHU).
 Data were subject to reporting artifacts and revisions.
@@
 The data and revisions are displayed in Figure \ref{fig:plot-data}.
-If not otherwise stated, all results are derived based on the original data available in 2021.
+All results presented here were derived based on the original data available in 2021.
 
@@
 
-Forecasts of incident cases and deaths linked to COVID-19 in the UK were elicited from individual participants every week through the web application (https://cmmid-lshtm.shinyapps.io/crowd-forecast/) described in \cite{bosseComparingHumanModelbased2022}.
+Forecasts of incident cases and deaths linked to COVID-19 in the UK were elicited from individual participants every week through a web application (https://cmmid-lshtm.shinyapps.io/crowd-forecast/) described in \cite{bosseComparingHumanModelbased2022}.
 The application is based on \textsf{R} \cite{R} \texttt{shiny} \cite{shiny} and is available as an \textsf{R} package called \texttt{crowdforecastr} \citep{crowdforecastr}.
 
-The web application offered participants two different ways of making a forecast called 'direct' (or 'classical') and 'Rt' forecast.
-To make a 'direct' forecast (as described in more detail \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty.
+When signing up, we asked participants to self-identify as "experts", asking them to tick a box if they worked in infectious disease modelling or had professional experience in any related field.
+
+The web application offered participants two different ways of making a forecast, called 'direct' (or 'classical') and 'Rt' forecast.
+To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty.
 In addition to information about past observations, participants could see various metrics and data sourced from Our World in Data \citep{owidcoronavirus}.
@@
 
-When signing up, we asked participants to self-identify as "experts", asking them to tick a box if they worked in infectious disease modelling or had professional experience in any related field.
+\textbf{SAM TO THE RESCUE}
 
-SAM TO THE RESCUE
-
-In addition, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
-This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the \textsf{R} package \texttt{EpiNow2} \citep{epinow2}, APPENDIX INFO?.
+In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
+This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the \textsf{R} package \texttt{EpiNow2} \citep{epinow2} (see details in the SI).
 In order to obtain forecasts for deaths, we convolved predicted infections as implied by the $R_t$ forecast over data-based delay distributions \cite{epinow2, sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} to model the time between infection and report date.
@@
 We scored forecasts using the weighted interval score (WIS, formulas given in the SI) %in Section \ref{sec:wis}) 
-\cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration.
+\cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
 The WIS is a proper scoring rule (with smaller values implying better performance).
@@
 uncertainty of forecasts), over-prediction and under-prediction.
-Empirical coverage refers to the percentage of observations falling inside any given prediction interval (e.g.
-the cumulative percentage of observed values that fall inside all 50\% prediction intervals).
-It represents a measure of probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
+Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
+the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
 Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
@@
 
-\citet{bosseTransformationForecastsEvaluating2023} suggested to transform forecasts and observations using the natural logarithm prior prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
+\citet{bosseTransformationForecastsEvaluating2023} suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
 We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
@@
 For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+The difference between the Hub ensemble and the human judgement forecasts steadily increased with increasing forecast horizon.
 On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
@@
 On the log scale, all human forecast ensembles performed slightly better than the Forecast Hub ensemble.
+For a four week forecast horizon, however, the Forecast Hub ensemble performed slightly better than human forecasts.
 Similarly to what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
@@
 
-In terms of the WIS on the natural scale, direct human forecasts performed best, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
+In terms of the WIS on the natural scale, direct human forecasts performed best at a two-week forecast horizon, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
+For four-week-ahead death forecasts, the Hub ensemble performed slightly better than other approaches.
 On the natural scale, the highest scores were concentrated towards the end of the study period, where death incidences were highest, while on the log scale, scores were more evenly distributed across the study period.
@@
 In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
-Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deahts and b) forecasts of the effective reproduction number $R_t$ which were mapped to cases using the renewal equation and convolved over a delay distribution to obtain deaths.
+Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were mapped to cases using the renewal equation and convolved over a delay distribution to obtain deaths.
 We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
@@
 \citet{mcandrewChimericForecastingCombining2022} similarly found that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
-We could not replicate these findings, and found that in terms of the WIS on the natural scale, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differ).
+We could not replicate these findings, and found that in terms of the WIS on the natural scale, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed).
 Looking at scores on the log scale suggests overall comparable performance of the Hub ensemble and the our direct human forecasts on both cases and deaths.
 Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to the rise of the Delta variant in the UK starting in the beginning of May.
-The Delta variant had a higher case fatality ratio (CFR) than previous variants CITATION.
+The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
 Humans may have found it easier than computational models to adapt to the fact that the relationship between cases and deaths continuously changed during the study period.

### 2023-02-17 Analysis.rmd
test commit to test overleaf sync

@@
 colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")
+```
 
-```
+# Define Helper Functions, Load Data
 
@@
 models_direct <- models_human[!grepl("(Rt)", models_human)]
+models_expert <- forecasts |>
+  filter(expert) |>
+  pull(model) |> unique() |> sort()
+
+update_expert_status <- function(data) {
+  data |>
+    mutate(expert = ifelse(
+      model %in% models_expert | model %in% paste(models_expert, "(Rt)"), 
+      TRUE, 
+      FALSE
+    ))
+}
 
@@
 
+# Figure SI.1: Data and data revisions
+data revisions: https://github.com/CSSEGISandData/COVID-19/issues/6021
+https://coronavirus.data.gov.uk/details/whats-new 
+https://coronavirus.data.gov.uk/details/whats-new/record/965a7923-dacf-42ed-abb9-3a972dbf1835
+
 ```{r}
@@
   weekly_truth |>
+    mutate(status = str_to_title(status)) |>
   ggplot(aes(x = target_end_date, y = true_value)) +
-  annotate("rect", xmin = time_start, xmax = time_stop, ymin = 0, ymax = Inf,
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
            fill = "darkolivegreen3", alpha = .1) + 
@@
   scale_colour_manual(values = c("red", "black")) + 
-  labs(y = "Observed value", x = "Date", color = "Data status")
+  labs(y = "Original data", x = "Date", color = "Data status")
 }
 
-plot_data_revised <- plot_data(weekly_truth_revised, daily_truth_revised)
-plot_data <- plot_data(weekly_truth, daily_truth)
+p_data_revised <- plot_data(weekly_truth_revised |>
+                                 filter(target_type == "Deaths"), 
+                               daily_truth_revised |>
+                                 filter(target_type == "Deaths")) + 
+  ylab("Revised data")
+p_data <- plot_data(weekly_truth, daily_truth)
 
-ggsave("output/figures/plot-data-revised.png", plot = plot_data_revised, 
-       width = 7, height = 5)
-
-ggsave("output/figures/plot-data.png", plot = plot_data, 
-       width = 7, height = 3.5)
-
-```
-
-Show data revisions
-
-```{r}
-
-plot_weekly_revisions <- weekly_truth |>
+p_weekly_revisions <- weekly_truth |>
   rbind(weekly_truth_revised) |>
@@
   ggplot(aes(x = target_end_date, y = diff)) + 
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
+           fill = "darkolivegreen3", alpha = .1) +
   geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") + 
@@
   facet_wrap(~ target_type, ncol = 1, scale = "free_y") +
-  labs(y = "Difference in data", x = "Date") + 
+  labs(y = "Weekly data revisions", x = "Date") + 
   coord_cartesian(ylim = c(-500, 500)) + 
@@
 
-ggsave("output/figures/data-revisions.png", plot = plot_weekly_revisions, 
-       width = 7, height = 2.5)
+p_all_data <- p_data / 
+  p_data_revised / 
+  p_weekly_revisions +
+  plot_annotation(tag_levels = "A") + 
+  plot_layout(guides = "collect", heights = c(2.2, 1, 1)) & 
+  theme(legend.position = "bottom")
 
+p_all_data
+
+ggsave("output/figures/plot-data.png", plot = p_all_data, 
+       width = 7, height = 8)
 ```
 
-Participant forecasts 
-```{r}
-models <- forecasts$model |> unique()
-
-horizon_0 <- weekly_truth |>
-  filter(target_end_date >= "2021-04-01") |>
-  mutate(prediction = true_value, 
-         forecast_date = target_end_date + 2, 
-         horizon = 0, 
-         quantile = 0.5, 
-         model = list(models)) |>
-  tidyr::unnest(cols = model)
-
-plot_df <- forecasts |>
-  filter(quantile == 0.5) |>
-  rbind(horizon_0) |>
-  filter(target_end_date >= "2021-04-01", 
-         target_end_date <= "2021-10-01") |>
-  filter(!(model %in% c("crowd-rt", "EpiExpert-ensemble"))) |>
-  filter(!(grepl("(Rt)", model))) |>
-  mutate(colouring = ifelse(model == "crowd-ensemble", 
-                            "Crowd ensemble", 
-                            "individual particpant"))
-
-plot_participant_forecasts <- plot_df |>
-  ggplot(aes(x = target_end_date)) +
-  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
-                color = "Individual participant"), 
-            linewidth = .1) + 
-  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
-            aes(y = prediction, group = interaction(forecast_date, model), 
-                color = "Crowd ensemble"),
-            linewidth = .4) + 
-  geom_line(aes(y = true_value),
-            linewidth = .4, color = "black") + 
-  geom_point(aes(y = true_value),
-             size = .5, color = "black") + 
-  scale_color_manual(values = c(colors[1], "grey70")) + 
-  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
-  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
-  theme_scoringutils()
-
-ggsave("output/figures/participant-forecasts.png", plot = plot_participant_forecasts, 
-       width = 7, height = 5)
-```
-
-Median forecasts ensemble
-```{r}
-
-forecasts |>
-  filter(quantile == 0.5) |>
-  rbind(horizon_0) |>
-  filter(target_end_date >= "2021-04-01", 
-         target_end_date <= "2021-10-01") |>
-  filter((model %in% c("crowd-ensemble", "EuroCOVIDhub-ensemble"))) |>
-  ggplot(aes(x = target_end_date)) +
-  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
-                color = model), 
-            linewidth = .4) + 
-  geom_line(aes(y = true_value),
-            linewidth = .4, color = "black") + 
-  geom_point(aes(y = true_value),
-             size = .5, color = "black") + 
-  scale_color_manual(values = colors[1:2]) +
-  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
-  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
-  theme_scoringutils()
-```
-
-```{r}
-
-comparison_plot <- function(scores, models = c("crowd-direct", "EuroCOVIDhub-ensemble")) {
-  mean_wis <- scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models) |>
-  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale")) |>
-  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale"),
-                   fun = signif, digits = 3) |>
-  mutate(wis_mean = paste("Mean:", label_fn(interval_score)))
-
-scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models) |>
-  arrange(scale, target_type, model) |>
-  ggplot(aes(y = interval_score, x = model)) + 
-  geom_violin(aes(fill = model), color = NA,
-         alpha = 0.2) + 
-  geom_boxplot(alpha = 0.5) + 
-  scale_fill_brewer(palette = "Set1", name = "Forecast target") +
-
-  geom_text(data = mean_wis, 
-            aes(label = wis_mean, y = -Inf), 
-            vjust   = -1) + 
-  facet_wrap(~ type_and_scale, scale = "free") + 
-  scale_y_continuous(labels = label_fn, 
-                     expand = expansion(mult = c(0.15, 0.05))) + 
-  labs(y = "Weighted interval score", x = "Forecaster") + 
-  theme_scoringutils() + 
-  theme(legend.position = "None")
-}
-
-comparison_plot(scores, 
-                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))
-
-comparison_plot(scores |>
-                  filter(horizon == 2), 
-                models = c("crowd-direct", "crowd-rt", "crowd-ensemble"))
-
-```
-
-## Weekly participation
+# Figure SI.4: Weekly participation humans and hub ensemble
 
@@
   mutate(prediction = 1, true_value = 1) |>
+  filter(forecast_date <= "2021-08-16") |>
+  unique() |>
   avail_forecasts(by = c("forecast_date", "target_type"), 
@@
       geom_line(show.legend = FALSE) + 
-      geom_point(show.legend = FALSE) + 
+      geom_point(show.legend = FALSE, size = 0.7) + 
       geom_hline(yintercept = filter_range / 100, 
@@
 ggsave("output/figures/performance.png", plot = p_performance, 
-       width = 7, height = 8)
+       width = 7, height = 6)
 
@@
        plot = p_comparison_direct_rt_individual, 
-       width = 7, height = 5)
+       width = 7, height = 4)
 
@@
     mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
+    arrange(target_type, model) |>
     kable(format = "latex", booktabs = TRUE, 
@@
 
+## Look at expert forecasts
+
+# ensemble of experts and non-experts
+
+```{r}
+
+create_expert_ensemble <- function(forecasts) {
+
+  forecasts_expert_ensemble <- forecasts |>
+    filter(model %in% models_human) |>
+    update_expert_status() |>
+    group_by(target_type, horizon, expert, quantile, true_value, forecast_date, target_end_date) |>
+    summarise(prediction = median(prediction), 
+              n = n()) |>
+    mutate(model = ifelse(expert, "Expert", "Non-Expert")) |>
+    as.data.table()
+
+  scores_expert_ensemble <- forecasts_expert_ensemble |>
+    mutate(method = model) |>
+    score_and_add_log()
+
+  return(list(forecasts = forecasts_expert_ensemble, 
+              scores = scores_expert_ensemble))
+}
+
+expert_ensemble <- create_expert_ensemble(filter(forecasts))
+
+# plot with number of experts and non-experts every week
+expert_ensemble$forecasts |>
+  filter(horizon == 2) |>
+  group_by(model, forecast_date, target_type) |>
+  summarise(n = unique(n)) |>
+  ggplot(aes(x = forecast_date, y = n, colour = model)) + 
+  geom_line() + 
+  geom_point() + 
+  theme_scoringutils() + 
+  facet_wrap(~ target_type)
+
+expert_ensemble$scores |>
+   make_performance_table()
+
+p_scores_expert <- plot_performance(expert_ensemble$scores)
+ggsave("output/figures/performance-expert.png", 
+       plot = p_scores_expert, 
+       width = 7, height = 6)
+
+expert_ensemble_direct <- forecasts |>
+  filter(model %in% models_direct) |>
+  create_expert_ensemble()
+
+expert_ensemble_direct$scores |>
+   make_performance_table()
+
+p_scores_expert_direct <- plot_performance(expert_ensemble_direct$scores)
+ggsave("output/figures/performance-expert-direct.png", 
+       plot = p_scores_expert_direct, 
+       width = 7, height = 6)
+
+# plot_scores_and_pred(
+#   forecasts_expert_ensemble,
+#   scores_expert_ensemble
+# )
+
+```
+
+```{r}
+scores_df <- scores  |>
+  update_expert_status() |>
+  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
+  filter(model %in% models_human) |>
+  filter(horizon == 2) |>
+  summarise_scores()
+
+mean_wis <- scores_df |>
+  group_by(scale, expert, target_type, type_and_scale) |>
+  summarise(wis_mean = signif(median(interval_score), 3)) |>
+  mutate(wis_mean = paste("Mean:", label_fn(wis_mean)))
+
+scores_df |>   
+  ggplot(aes(x = expert, y = interval_score)) +
+  geom_violin(alpha = 0.2) + 
+  geom_boxplot(alpha = 0.5) + 
+  # scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
+  geom_text(data = mean_wis, 
+            aes(label = unique(wis_mean), y = -Inf), 
+            vjust   = -1) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  scale_y_continuous(labels = label_fn, 
+                     expand = expansion(mult = c(0.15, 0.05))) + 
+  theme_scoringutils()
+
+scores_df |>
+  summarise_scores(by = c("expert", "target_type", "scale", "horizon")) |>
+  rename(model = expert) |>
+  make_performance_table()
+
+```
+
 ```{r}
@@
 - Hub ensemble switched from mean to median on 2021-07-19
-- did we do do mean or median? 
-- how did we end up with the crowd ensemble
+
+# Parking lot
+Median forecasts for a 1:4 horizon
+```{r}
+models <- forecasts$model |> unique()
+
+horizon_0 <- weekly_truth |>
+  filter(target_end_date >= "2021-04-01") |>
+  mutate(prediction = true_value, 
+         forecast_date = target_end_date + 2, 
+         horizon = 0, 
+         quantile = 0.5, 
+         model = list(models)) |>
+  tidyr::unnest(cols = model)
+
+plot_df <- forecasts |>
+  filter(quantile == 0.5) |>
+  rbind(horizon_0, fill = TRUE) |>
+  filter(target_end_date >= "2021-04-01", 
+         target_end_date <= "2021-10-01") |>
+  filter(!(model %in% c("crowd-rt", "EpiExpert-ensemble"))) |>
+  filter(!(grepl("(Rt)", model))) |>
+  mutate(colouring = ifelse(model == "crowd-ensemble", 
+                            "Crowd ensemble", 
+                            "individual particpant"))
+
+plot_participant_forecasts <- plot_df |>
+  ggplot(aes(x = target_end_date)) +
+  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
+                color = "Individual participant"), 
+            linewidth = .1) + 
+  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
+            aes(y = prediction, group = interaction(forecast_date, model), 
+                color = "Crowd ensemble"),
+            linewidth = .4) + 
+  geom_line(aes(y = true_value),
+            linewidth = .4, color = "black") + 
+  geom_point(aes(y = true_value),
+             size = .5, color = "black") + 
+  scale_color_manual(values = c(colors[1], "grey70")) + 
+  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
+  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
+  theme_scoringutils()
+
+ggsave("output/figures/participant-forecasts.png", plot = plot_participant_forecasts, 
+       width = 7, height = 5)
+```
+
+Median forecasts combined human ensemble and Forecast Hub ensemble 1:4 horizon
+```{r}
+forecasts |>
+  filter(quantile == 0.5) |>
+  rbind(horizon_0, fill = TRUE) |>
+  filter(target_end_date >= "2021-04-01", 
+         target_end_date <= "2021-10-01") |>
+  filter((model %in% c("crowd-ensemble", "EuroCOVIDhub-ensemble"))) |>
+  ggplot(aes(x = target_end_date)) +
+  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
+                color = model), 
+            linewidth = .4) + 
+  geom_line(aes(y = true_value),
+            linewidth = .4, color = "black") + 
+  geom_point(aes(y = true_value),
+             size = .5, color = "black") + 
+  scale_color_manual(values = colors[1:2]) +
+  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
+  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
+  theme_scoringutils()
+```
+
+## Violin plot to compare scores
+
+```{r}
+comparison_plot <- function(scores, models = c("crowd-direct", "EuroCOVIDhub-ensemble")) {
+  mean_wis <- scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models) |>
+  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale")) |>
+  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale"),
+                   fun = signif, digits = 3) |>
+  mutate(wis_mean = paste("Mean:", label_fn(interval_score)))
+
+scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models) |>
+  arrange(scale, target_type, model) |>
+  ggplot(aes(y = interval_score, x = model)) + 
+  geom_violin(aes(fill = model), color = NA,
+         alpha = 0.2) + 
+  geom_boxplot(alpha = 0.5) + 
+  scale_fill_brewer(palette = "Set1", name = "Forecast target") +
+
+  geom_text(data = mean_wis, 
+            aes(label = wis_mean, y = -Inf), 
+            vjust   = -1) + 
+  facet_wrap(~ type_and_scale, scale = "free") + 
+  scale_y_continuous(labels = label_fn, 
+                     expand = expansion(mult = c(0.15, 0.05))) + 
+  labs(y = "Weighted interval score", x = "Forecaster") + 
+  theme_scoringutils() + 
+  theme(legend.position = "None")
+}
+
+comparison_plot(scores, 
+                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))
+
+comparison_plot(scores |>
+                  filter(horizon == 2), 
+                models = c("crowd-direct", "crowd-rt", "crowd-ensemble"))
+
+```

### 2023-02-28 Analysis.rmd
second test commit to test overleaf sync
Add standardised scores to performance table

@@
 colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")
+
 ```
@@
 
+```{r}
+scores_standardised <- scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models_non_human) |>
+  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
+  filter(horizon == 2) |>
+  group_by(target_type, scale) |>
+  mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
+  group_by(model, target_type, scale) |>
+  summarise(mean = mean(interval_score)) |>
+  pivot_wider(names_from = scale, values_from = mean, names_prefix = "stand_") |>
+  arrange(target_type, model)
+```
+
 Table with performance
@@
 ```{r, performance-table}
-make_performance_table <- function(scores) {
+make_performance_table <- function(scores, standardised_scores) {
   scores |>
@@
     select(model, target_type, natural, log, coverage_50, coverage_90) |>
+    inner_join(standardised_scores) |>
     mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
     arrange(target_type, model) |>
+    mutate(natural = paste0(natural, " (", stand_natural, ")"), 
+           log = paste0(log, " (", stand_log, ")")) |>
+    select(-stand_log, -stand_natural) |>
     kable(format = "latex", booktabs = TRUE, 
@@
   filter(model %in% models_non_human) |>
-  make_performance_table() |>
+  make_performance_table(standardised_scores = scores_standardised) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
   cat(file = 'manuscript/performance-table.tex')

### 2023-02-28 manuscript/manuscript.tex
second test commit to test overleaf sync
comments from Sam on manuscript

@@
 \begin{abstract}
-
 %Abstracts should be up to 300 words and provide a succinct summary of the article.
@@
 In the past, two studies found ensembles of human judgement forecasts of COVID-19 to show predictive performance comparable to ensembles of computational models, at least when predicting case incidences.
-We present a follow-up to a study conducted in Germany and Poland and investigate a novel approach to combine human judgement and epidemiological modelling.
+We present a follow-up to a study conducted in Germany and Poland and investigate a novel joint approach to combine human judgement and epidemiological modelling.
 
@@
 Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
-This forecast was automatically mapped to cases and deaths using the renewal equation and a convolution over a delay distribution.
+This forecast was automatically mapped to cases and deaths using the \texttt{EpiNow2} \textsf{R} package and associated model.
 Forecasts were scored using the weighted interval score on the original forecasts, as well as after applying the natural logarithm to both forecasts and observations.
@@
 
-The ensemble of human forecasters overall performed comparably to the offical European Forecast Hub ensemble on both cases and deaths, although results were susceptible to changes in details of the evaluation.
-$R_t$ foreasts performed comparably to direct forecasts ona cases, but worse on deaths.
-Self-identified experts did not outperform non-experts.
+The ensemble of human forecasters overall performed comparably to the official European Forecast Hub ensemble on both cases and deaths, although results were susceptible to changes in details of the evaluation.
+$R_t$ forecasts performed comparably to direct forecasts on cases, but worse on deaths.
+Self-identified experts did not outperform non-experts though the sample size was small.
 
@@
 The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
+Combinations of human and computational forecasts hold potential but present real-world challenges that need to be solved.
 
@@
 % Beginning in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
-Most forecasts were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
+Most forecasts used to inform decision making were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative or in combination \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
 
@@
 \citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
+It is also important to note that in the majority of these studies it is difficult to unpick the reliance of human forecasters on the computational models they were being compared to or to similar models so we should expect performance to at least equal with computational models in most settings.
 
@@
 In addition, humans are able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
-On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, and humans may be at a disadvantage at tasks that require complex computations.
+On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, humans may be at a disadvantage at tasks that require complex computations.
+Also, the use of human judgement forecasts by decision makers may be complicated by the lack of understanding of the basis on which they were made, if forecasters don't also provide a rational for their predictions.
 
@@
 
-Methods that aim to combine human judgement and mathematical modelling are therefore appealing.
-One straightforward way is to combine separate human judgement and computational model forecasts with the goal of improving predictive performance \citep{mcandrewChimericForecastingCombining2022}.
+Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting mathematical modelling as truly independent is perhaps misleading as the vast majority of computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage.
+One straightforward method is to combine separate human judgement and computational model forecasts with the goal of improving predictive performance in the form of an ensemble.
+This has been widely shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
 \cite{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
@@
 Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustments to model outputs.
-\citet{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) and to feed this forecast into a mathematical model in order to obtain forecasts for case and death numbers.
+\citet{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) and to then feed this forecast into a mathematical simulation model in order to obtain forecasts for case and death numbers.
 
@@
 In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a mathematical model.
+We describe this approach as human in the loop computational modelling and consider it a formalisation of often practiced manual intervention in computational forecasts.
 
@@
 Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
-Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisiations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
+Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
 From the 29th of July on, the default Forecast Hub ensemble became a median-ensemble.
-The number of models included in the Forecast Hub ensemble is shown in Figure \ref{fig:num-forecasters} in the SI.
+The median number of models included in the Forecast Hub ensemble was nine for cases and ten for deaths (see Figure \ref{fig:num-forecasters} in the SI).
 
@@
 To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty.
-In addition to information about past observations, participants could see various metrics and data sourced from Our World in Data \citep{owidcoronavirus}.
+In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
 Figure \ref{fig:screenshot-classical} shows a screenshot of the forecast interface for direct forecasts.
 
-\textbf{SAM TO THE RESCUE}
+In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline forecast produced by the \texttt{EpiNow2} \textsf{R} package effective reproduction number model \citep{epinow2} which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
+This forecast was then translated into forecasts of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+\texttt{EpiNow2} implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007}  generative process for latent infections where the effective reproduction number is modelled as a Gaussian process.
+These latent infections are then convolved with delay distributions, such as the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
+This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022} [CITE ECDC] [CITE SPI-M paper] and used to inform policy makers via reproduction number estimates [CITE one of the SPI-M reproduction number summary papers or kaths paper on multiple surveillance Rt esimates].
+Further details are given in the SI.
 
-In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
-This forecast was then translated into forecasts of cases using the renewal equation \citep{fraserEstimatingIndividualHousehold2007} as implemented in the \textsf{R} package \texttt{EpiNow2} \citep{epinow2} (see details in the SI).
-In order to obtain forecasts for deaths, we convolved predicted infections as implied by the $R_t$ forecast over data-based delay distributions \cite{epinow2, sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} to model the time between infection and report date.
-As $R_t$-estimates up to two weeks prior to the forecast data were deemed uncertain, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
-Participants were therefore asked to estimate/predict six $R_t$ values, four of them in the future.
+In order to obtain forecasts for deaths, we similarly convolved observed and predicted cases as implied by the $R_t$ forecast over data-based delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} to model the time between infection and report date  and applied a case fatality ratio estimated based on the relationship between observed and estimated reported cases and deaths again using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+Again further details are given in the SI and in our previous work \citetp{bosseComparingHumanModelbased2022}.
+
+As $R_t$-estimates up to two weeks prior to the forecast data were uncertain due to the delay between infection and case report, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
+Participants were therefore asked to estimate/predict six $R_t$ values, four of them beyond the forecast horizon.
 Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
-However, due to the high computational complexity and necessary waiting times, participants could not preview the death forecast implied by their current input for $R_t$.
+However, due to the computational complexity and resource limitations, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the assumed case fatality ratio or delay between reported cases and reported deaths.
 Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
@@
 
-\subsection*{EpiNow2?}
-
-SAM TO THE RESCUE, RELEVANT PART OF OLD PAPER PASTED HERE AND COMMENTED OUT. DO WE NEED AN EXTRA SECTION, OR SHOULD THIS BE INCLDUED ABOVE?
-
-% From old paper: 
-% The first of these models, here called “renewal model”, used the renewal equation [41] to predict reported cases and deaths (see details in S1 Text).
-It estimated the effective reproduction number Rt (the average number of people each person infected at time t is expected to infect in turn) and modelled future infections as a weighted sum of past infection multiplied by Rt.
-Rt was assumed to stay constant beyond the forecast date, roughly corresponding to continuing the latest exponential trend in infections.
-On the 9th of November we altered the date when Rt was assumed to be constant from two weeks prior to the date of the forecast to the forecast date, which we found to yield a more stable Rt estimate.
-Reported case and death notifications were obtained by convolving predicted infections over data-based delay distributions [40, 42–44] to model the time between infection and report date.
-The renewal model was used to predict cases as well as deaths with forecasts being generated for each target separately.
-Death forecasts from the renewal model were therefore not informed by past cases.
-One submission of the renewal model on December 28th 2020 was delayed and therefore not included in the official Forecast hub ensemble.
-% 
-% Line list data used to inform the prior for the delay from symptom onset to test positive case report or death in the model-based forecasts was sourced from [45] with data available up to the 1st of August.
-All model fitting was done using Markov-chain Monte Carlo (MCMC) in stan [46] with each location and forecast target being fitted separately.
-
 \subsection*{The UK Crowd Forecasting Challenge}
 
-In order to boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022}, we announced an official tournament, the "UK Crowd Forecasting Challenge".
+In order to boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
 Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
@@
 
-Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}.
+Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022} except for our extension to also consider the natural log of forecasts as a proxy for the exponential growth rate \citet{bosseTransformationForecastsEvaluating2023}.
 If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022a, bracherShorttermForecastingCOVID192021}.
@@
 
-The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51 (compared to 6, 2, and 10 in \cite{bosseComparingHumanModelbased2022}).
+The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51.
+This was higher than the number of participants in \cite{bosseComparingHumanModelbased2022} (which had a median number of 6, a minimum of 2, and a maximum 10).
 With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13.
-The median number of submissions from a single individual was one, meaning that most forecasters dropped out after their first submission.
+The median number of submissions from a single individual was one, meaning that similar to \citep{bosseComparingHumanModelbased2022} most forecasters dropped out after their first submission.
 Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study.
 The number of direct forecasts was higher than the number of $R_t$ forecasts in all weeks (see Figure \ref{fig:num-forecasters}).
+
+\begin{figure*}[ht]
+%\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance.png}
+\caption{\bf{Predictive performance across forecast horizons.} A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
+E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
+\label{fig:performance}
+\end{figure*}
 
@@
 \label{fig:forecasts-scores} 
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance.png}
-\caption{\bf{Predictive performance across forecast horizons.} A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
-E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance}
 \end{figure*}
@@
 
-In terms of the WIS on the natural scale, direct human forecasts performed best at a two-week forecast horizon, while $R_t$ forecasts performed significantly worse than all other approaches (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+In terms of the WIS on the natural scale, direct human forecasts performed best at a two-week forecast horizon, while $R_t$ forecasts performed significantly worse than all other approaches (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
 From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
@@
 On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
-%Note that the picture changes when looking at the revised data which was corrected upwards compared to the original data(DO WE WANT TO DO THAT?
-IS THAT MEANINGFUL?) (see Figure \ref{fig:data-revisions}).
-For the revised data, direct forecasts performed significantly worse than both the Forecast Hub ensemble and the $R_t$ forecasts on the log scale, while still performing better than the $R_t$ forecasts on the natural scale (see Table \ref{tab:scores-revised}).
-%For revised data: On the natural scale, the largest scores occurred at the end of the study period when deaths were highest.
-On the log scale, however, underprediction in June and July when death incidences were lower, received more weight and direct crowd forecast outperformed $R_t$ forecasts.
+
 $R_t$ forecasts show strong underdispersion with empirical coverage of the 50\% prediction intervals being close to zero and smaller than 50\% for the 90\% prediction intervals (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
@@
 For cases, where participants could observe the case forecast implied by their $R_t$ forecast, predictive performance was similar between corresponding direct and $R_t$ forecasts for most forecasters who had submitted both (see Figure \ref{fig:comparison-direct-rt-individual}).
-For deaths, predictive performance of their $R_t$ forecasts was significantly worse than their corresponding direct forecasts for most forecasters, suggesting that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately.
-This inadequacy may be related to the change in the case fatality ratio over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK.
-%IMPORTANTLY THOUGH, THE PICTURE DOES LOOK DIFFERNT WHEN LOOKING AT THE REVISED DATA.
-
-% When looking at an ensemble including only those data points where a single forecaster has submitted both a direct and an $R_t$ forecast, we found a similar picture.
-Performance for cases was broadly comparable, with direct forecasts slightly better than $R_t$ forecasts on the natural scale and slightly worse on the log scale.
-On the log scale, the ensemble of $R_t$ forecasts performed significantly worse than the ensemble of corresponding direct forecasts both on the natural and on the log scale.
+For deaths, where forecasters could not see the incidence forecast implied by their $R_t$ forecast or manually adjust the case fatality rate, performance of the $R_t$ forecasts was significantly worse than the corresponding direct forecasts for most forecasters.
+This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings [CITE SOPHIE's paper].
+This may be related to the change in the case fatality ratio over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK which was not accounted for in the simple model used and not available to be altered by forecasters.
 
@@
 In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
-Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were mapped to cases using the renewal equation and convolved over a delay distribution to obtain deaths.
+Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on output from an open source effective reproduction number estimation model and then used to generate simulated reported cases and deaths using the same model and a simple extension which assumed a static case fatality ration and a delay between cases and reports that was estimated from the data.
 We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
@@
 In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
-They hypothesised that computational models might have an advantage over human forecsters when predicting deaths, benefitting from the ability to model the delays and exact epidemiological relationships between different leading and lagged indicators.
+They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and exact epidemiological relationships between different leading and lagged indicators.
 \citet{mcandrewChimericForecastingCombining2022} similarly found that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
-We could not replicate these findings, and found that in terms of the WIS on the natural scale, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed).
-Looking at scores on the log scale suggests overall comparable performance of the Hub ensemble and the our direct human forecasts on both cases and deaths.
+We could not replicate these findings, and found that in terms of the WIS \textit{on the natural scale}, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed).
+Looking at scores \textit{on the log scale} suggests overall comparable performance of the Hub ensemble and the direct human forecasts on both cases and deaths.
 Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to the rise of the Delta variant in the UK starting in the beginning of May.
@@
 Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to recruit a large number of participants.
-Focused public outreach efforts such as creating a dedicated website, anouncing an official tournament, providing a public leaderboard, sending weekly emails with details on past performance and weekly announcements on Twitter, did noticeably increase participation compared to the previous study in Germany and Poland.
-Nevertheless, particaption was still relatively low and most forecasters only submitted a single forecast.
-\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms such as Metaculus or Good Judgement Open that provide access to a more elegant user interface and a large existing user base may be helpful in recruting a larger number of participants.
+Focused public outreach efforts such as creating a dedicated website, announcing an official tournament, providing a public leaderboard, sending weekly emails with details on past performance and weekly announcements on Twitter, did noticeably increase participation compared to the previous study in Germany and Poland.
+Nevertheless, participation was still relatively low and most forecasters only submitted a single forecast.
+\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that provide access to a more elegant user interface and a large existing user base may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
 
@@
 One appeal of this approach is that the forecaster only needs to provide one quantity that directly represents what she believes about the force of the epidemic in the future.
-Computational modelling would then take care of dealing with details such as reporting delays, generation intervals, or the relationship between different indicators.
-This could help reduce cognitive load, at least for forecasters who have an intuitive understanding of $R_t$.
+Computational modelling would then take care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, or the relationship between different indicators.
+This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$.
 Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
@@
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-\textbf{SAM TO THE RESCUE} The relationship between infections and deaths used to map from $R_t$ to deaths was automatically estimated from past data, and forecasters had no way of adjusting it.
+The relationship between cases and deaths used to forecast deaths from cases was estimated from past data, forecasters had no way of adjusting it, and could not visualise the forecasts in the app.
 The rise of the Delta variant with a subsequent gradual change in the case fatality ratio (CFR) may have affected $R_t$ forecasts, as the changing CFR was not directly accounted for and estimated distributions from past data only slowly updated to new circumstances.
@@
 "experts", or $R_t$ forecasts) had different numbers of forecasters.
-In addition, there are many researcher degrees of freedom that influence results, for example how indidual forecasts are combined to an ensemble and how forecasts are evaluated.
+In addition, there are many researcher degrees of freedom that influence results, for example how individual forecasts are combined to an ensemble and how forecasts are evaluated.
 Prizes to the human forecasters, for example were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
@@
 Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles can produce forecasts that are on par with those of an ensemble of computational models.
-Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited.
+Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly with those submitting models to the European Forecast Hub not being properly incentivised to update their forecasting methodology.
 Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths.
@@
 Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture.
+Allowing researchers free reign to design their own evaluations post-hoc runs the risk of allowing for motivated reasoning and more work should go in to preventing this.
 
@@
 \centering
+\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
+\caption{\bf{Number of forecasts across the study period.}}
+\label{fig:num-forecasters}
+\end{figure*}
+
+\begin{figure*}
+\centering
 \includegraphics[width=0.99\textwidth]{../output/figures/plot-data.png}
@@
 \label{fig:screenshot-rt}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
-\caption{\bf{Number of forecasts across the study period.}}
-\label{fig:num-forecasters}
 \end{figure*}

### 2023-03-01 Analysis.rmd
update performance table with standard deviations

--- 2023-02-17 Analysis.rmd
@@
 
+```{r}
+scores_summarised <- scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models_non_human) |>
+  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
+  group_by(target_type, scale, horizon) |>
+  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
+  group_by(model, target_type, scale, horizon) |>
+  summarise(mean = mean(interval_score), 
+            sd = sd(interval_score)) |>
+  pivot_wider(id_cols = c(model, target_type, horizon), 
+              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
+  arrange(target_type, model)
+```
+
 Table with performance
@@
 ```{r, performance-table}
-make_performance_table <- function(scores) {
+make_performance_table <- function(scores, scores_summarised, 
+                                   hor = 2) {
   scores |>
@@
     summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
-    filter(horizon == 2) |>
-    select(model, target_type, scale, interval_score, coverage_50, coverage_90) |>
+    select(model, target_type, scale, interval_score, coverage_50, coverage_90, horizon) |>
     pivot_wider(names_from = scale, values_from = interval_score) |>
-    select(model, target_type, natural, log, coverage_50, coverage_90) |>
+    select(model, target_type, natural, log, coverage_50, coverage_90, horizon) |>
+    inner_join(scores_summarised) |>
     mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
     arrange(target_type, model) |>
+    mutate(natural = label_fn(natural), 
+           log = label_fn(log)) |>
+    mutate(natural = paste0(natural, " (", label_fn(sd_stand_natural), ")"), 
+           log = paste0(log, " (", label_fn(sd_stand_log), ")")) |>
+    filter(horizon == hor) |>
+    select(-mean_stand_log, -sd_stand_log, -mean_stand_natural, -sd_stand_natural, -horizon) |>
     kable(format = "latex", booktabs = TRUE, 
@@
   filter(model %in% models_non_human) |>
-  make_performance_table() |>
+  make_performance_table(scores_summarised = scores_summarised) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
   cat(file = 'manuscript/performance-table.tex')
 
-scores_revised |>
+scores |>
   filter(model %in% models_non_human) |>
-  make_performance_table() |>
-  cat(file = 'manuscript/performance-table-revised.tex')
+  make_performance_table(scores_summarised = scores_summarised, hor = 4) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-horizon-4.tex')
 
-scores_comparable_ensemble |>
-  make_performance_table() |>
-  cat(file = 'manuscript/performance-table-comparison-rt.tex')
+# scores_revised |>
+#   filter(model %in% models_non_human) |>
+#   make_performance_table() |>
+#   cat(file = 'manuscript/performance-table-revised.tex')
+# 
+# 
+# scores_comparable_ensemble |>
+#   make_performance_table() |>
+#   cat(file = 'manuscript/performance-table-comparison-rt.tex')
 

### 2023-03-01 manuscript/manuscript.tex
update manuscript

@@
 In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
+To make results easier to interpret, we also report normalised scores.
+We normalised scores for a given target (cases or deaths) and scale (natural or log) by calculating a Z-score, i.e.
+by subtracting the mean of all scores for the same target and scale, and dividing by the standard deviation.
 
@@
 The number of direct forecasts was higher than the number of $R_t$ forecasts in all weeks (see Figure \ref{fig:num-forecasters}).
+
+\input performance-table.tex
 
@@
 
+Overall scores for all forecasting approaches (Forecast Hub ensemble, direct crowd forecasts, $R_t$ crowd forecasts, combined crowd forecast) are very close to each other.
+
 For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
@@
 Underdispersion was most severe for the $R_t$ forecasts.
-
-\input performance-table.tex
 

### 2023-03-20 manuscript/manuscript.tex
address comments

@@
 
-\author[1, 2, 3]{Anonymous Alpaca}
-\author[2]{Benevolent Brachiosaurus}
+\author[1, 2, 3]{Nikos I. Bosse}
+\author[1, 2]{Sam Abbott}
+\author[1, 4]{Edwin van Leeuwen}
+\author[5]{Anne Cori}
+\author[2]{Johannes Bracher}
+\author[1, 2, 3]{Sebastian Funk}
+
 \affil[1]{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom}
@@
 \affil[3]{NIHR Health Protection Research Unit in Modelling \& Health Economics}
-\affil[2]{Address of author2}
+\affil[4]{Modelling Economics Unit and NIHR Health Protection Research Unit in Modelling \& Health Economics, UK Health Security Agency, London, United Kingdom}
+\affil[5]{MRC Centre for Outbreak Analysis and Modelling, Department of Infectious Disease Epidemiology, School of Public Health, Imperial College London, London, United Kingdom}
+\affil[6]{Chair of Statistical Methods and Econometrics, Karlsruhe Institute of Technology, Karlsruhe, Germany}
+\affil[7]{Computational Statistics Group, Heidelberg Institute for Theoretical Studies, Heidelberg, Germany}
 
@@
 Forecasts were elicited as part of a tournament, the "UK Crowd Forecasting Challenge", and a median ensemble of all forecasts was submitted to the European Forecast Hub.
-Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
-This forecast was automatically mapped to cases and deaths using the \texttt{EpiNow2} \textsf{R} package and associated model.
+Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$ based on an estimated baseline.
+This was then used to forecast cases and deaths using simulation methods from the \texttt{EpiNow2} \textsf{R} package.
 Forecasts were scored using the weighted interval score on the original forecasts, as well as after applying the natural logarithm to both forecasts and observations.
@@
 
-Human judgement can be useful in predicting the large-scale evolution of an infectious disease such as COVID-19.
+Human judgement can be useful in predicting the spread of an infectious disease such as COVID-19.
 The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
@@
 \citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
-It is also important to note that in the majority of these studies it is difficult to unpick the reliance of human forecasters on the computational models they were being compared to or to similar models so we should expect performance to at least equal with computational models in most settings.
+It is important to note that in previous studies (and also this one) human forecasters were free to use any resources, including computational models, in the process of creating a forecast, making it difficult to completely separate human judgement and computational modelling.
 
-In some situations, human judgement forecasting has clear advantages relative to computational models.
-Humans can very quickly provide forecasts, even in situations where data is sparse and lots of parameters are unknown.
-In addition, humans are able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
-On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, humans may be at a disadvantage at tasks that require complex computations.
-Also, the use of human judgement forecasts by decision makers may be complicated by the lack of understanding of the basis on which they were made, if forecasters don't also provide a rational for their predictions.
+In some situations, human judgement forecasting may have advantages relative to computational models.
+Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and lots of parameters are unknown.
+Humans are generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
+On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, and humans may be at a disadvantage at tasks that strongly benefit from the ability to perform complex computations.
+Also, the use of human judgement forecasts by decision makers may be complicated by the lack of clarity of the basis on which they were made.
 
@@
 
-Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting mathematical modelling as truly independent is perhaps misleading as the vast majority of computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage.
-One straightforward method is to combine separate human judgement and computational model forecasts with the goal of improving predictive performance in the form of an ensemble.
+Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting this as a binary choice is misleading as most computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage and many human forecasters make use of simple base rate based approaches which are in reality simple models.
+One explicit method to combine separate human judgement and computational model forecasts with the goal of improving predictive performance is an ensemble.
 This has been widely shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
@@
 One approach is to use human forecasts, for example of relevant disease parameters, as an input to computational modelling.
-Another approach is to use mathematical modelling as an input to human judgement, for example by giving experts the option to make post-hoc adjustments to model outputs.
-\citet{bosseComparingHumanModelbased2022} proposed to ask human forecasters to predict the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) and to then feed this forecast into a mathematical simulation model in order to obtain forecasts for case and death numbers.
+Another approach is to use mathematical modelling in explicit combination with human judgement, for example by giving experts the option to make post-hoc adjustments to model outputs.
+\citet{bosseComparingHumanModelbased2022} proposed asking human forecasters to forecast the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) based on modelled estimates and to then use this forecast in a mathematical simulation model in order to obtain forecasts for observed case and death numbers.
 
@@
 This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants.
-In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a mathematical model.
+In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the estimated effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a simulation model.
 We describe this approach as human in the loop computational modelling and consider it a formalisation of often practiced manual intervention in computational forecasts.
@@
 \section*{Methods}
-
-\subsection*{Ethics statement}
-This study has been approved by the London School of Hygiene \& Tropical Medicine Research Ethics Committee (reference number 22290). Consent from participants was obtained in written form.
 
@@
 
-In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline forecast produced by the \texttt{EpiNow2} \textsf{R} package effective reproduction number model \citep{epinow2} which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
-This forecast was then translated into forecasts of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-\texttt{EpiNow2} implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007}  generative process for latent infections where the effective reproduction number is modelled as a Gaussian process.
+In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline estimate produced by the \texttt{EpiNow2} \textsf{R} package effective reproduction number model \citep{epinow2} which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
+This forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+\texttt{EpiNow2} implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007}  generative process for latent infections where the effective reproduction number is modelled as Gaussian process on the residuals.
 These latent infections are then convolved with delay distributions, such as the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
@@
 
-In order to obtain forecasts for deaths, we similarly convolved observed and predicted cases as implied by the $R_t$ forecast over data-based delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} to model the time between infection and report date  and applied a case fatality ratio estimated based on the relationship between observed and estimated reported cases and deaths again using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-Again further details are given in the SI and in our previous work \citetp{bosseComparingHumanModelbased2022}.
+In order to obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+Further details are given in the SI and in our previous work \citep{bosseComparingHumanModelbased2022}.
 
-As $R_t$-estimates up to two weeks prior to the forecast data were uncertain due to the delay between infection and case report, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
+As $R_t$-estimates up to at least two weeks prior to the forecast data were uncertain due to depending on partially complete observations of underlying infections, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
 Participants were therefore asked to estimate/predict six $R_t$ values, four of them beyond the forecast horizon.
 Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
-However, due to the computational complexity and resource limitations, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the assumed case fatality ratio or delay between reported cases and reported deaths.
+However, due to computational complexity and lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
 Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
@@
 Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
-All code and data, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
+All code and data used for this analysis, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
+All code used to submit the forecasts to the European Forecast Hub is available at \url{https://github.com/epiforecasts/europe-covid-forecast}.
 
@@
 by subtracting the mean of all scores for the same target and scale, and dividing by the standard deviation.
+
+\subsection*{Ethics statement}
+This study has been approved by the London School of Hygiene \& Tropical Medicine Research Ethics Committee (reference number 22290). Consent from participants was obtained in written form.
 
@@
 Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study.
-The number of direct forecasts was higher than the number of $R_t$ forecasts in all weeks (see Figure \ref{fig:num-forecasters}).
+The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}).
 
@@
 
-Overall scores for all forecasting approaches (Forecast Hub ensemble, direct crowd forecasts, $R_t$ crowd forecasts, combined crowd forecast) are very close to each other.
+Mean scores for the four forecasting approaches were close to each other, with a mean WIS for two-week-ahead-forecasts of incident cases on the natural scale (log scale) of 38k (0.25) for the Hub-ensemble, 39k (0.24) for the direct crowd forecasts, 40k (0.22) for the combined crowd forecast and 46k (0.23) for the $R_t$ forecasts (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
+Differences in mean scores between the four were small compared to the standard deviations of the scores within each forecasting approach on the natural (56k, 67k, 69k, 75k) and on the log scale (0.22, 0.27, 0.25, 0.24).
+% Could add a sentence that scores were not significant using a paired wilcoxon mann whitney test. 
+On the natural scale, the Forecast Hub ensemble was slightly ahead of the crowd forecasts, with the difference gradually increasing with increasing forecast horizon (see Figure \ref{fig:performance}).
+On the log scale, the Forecast Hub ensemble was slightly worse than all crowd forecasting approaches at a two week forecast horizon, but again slightly ahead for forecasts four weeks into the future.
 
-For two-week-ahead-forecasts of incident cases, the Forecast Hub ensemble outperformed all crowd forecasting ensembles as measured by the WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-The difference between the Hub ensemble and the human judgement forecasts steadily increased with increasing forecast horizon.
 On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
-Average scores are therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
-$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+Average scores were therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
+$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale.
 When looking at WIS values on the log scale,
 %(that is, when log-transforming forecasts and observations prior to applying the WIS), 
-scores are more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
-On the log scale, all human forecast ensembles performed slightly better than the Forecast Hub ensemble.
-For a four week forecast horizon, however, the Forecast Hub ensemble performed slightly better than human forecasts.
+scores were more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
 Similarly to what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
@@
 
-In terms of the WIS on the natural scale, direct human forecasts performed best at a two-week forecast horizon, while $R_t$ forecasts performed significantly worse than all other approaches (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
+Overall, performance for two-week-ahead-forecasts of incident deaths was comparable for the Hub-ensemble, the direct crowd forecasts and the combined forecasts, while performance for the $R_t$ forecast was significantly worse.
+On the natural scale (log scale), mean WIS values were 38 (0.22) for the Hub ensemble, 34 (0.27) for the direct forecasts, 40 (0.22) for the combined crowd forecast, and 80 (0.25) for the $R_t$ forecasts (Table \ref{tab:scores}).
+With the exception of the $R_t$ crowd forecasts, differences between forecasting approaches were again small compared to 
+standard deviations of
+the scores within each forecasting approach on the natural scale (27, 31, 42, 73) and on the log scale (0.036, 0.081, 0.073, 0.13). 
+%On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
+
+On the natural scale, direct forecasts had a slightly better WIS than the Hub ensemble and the other two human forecast ensembles for a two-week-ahead horizon, while the Hub ensemble was again slightly ahead for a four week-ahead horizon (see Figure \ref{fig:performance}).
+The combined crowd forecast (containing both the direct and the $R_t$ forecasts) overall performed slightly worse than the direct crowd forecast on the natural scale, but much better than the $R_t$ forecasts.
+On the log scale, the combined crowd forecasts even performed best among the four approaches.
+
 From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
-For four-week-ahead death forecasts, the Hub ensemble performed slightly better than other approaches.
 On the natural scale, the highest scores were concentrated towards the end of the study period, where death incidences were highest, while on the log scale, scores were more evenly distributed across the study period.
-On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
-
 $R_t$ forecasts show strong underdispersion with empirical coverage of the 50\% prediction intervals being close to zero and smaller than 50\% for the 90\% prediction intervals (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
 The Forecast Hub ensemble was overdispersed with empirical coverage exceeding nominal coverage, and for the direct human forecasts, empirical coverage was close to nominal coverage.
-
-% For data revision: deaths seems broadly comparable for all human forecasts, whereas the Forecast Hub ensemble displays noticeably better coverage for deaths than for cases (Figure \ref{fig:performance}E, F).
-
-% \subsection*{Individual forecasts?}
 
@@
 For deaths, where forecasters could not see the incidence forecast implied by their $R_t$ forecast or manually adjust the case fatality rate, performance of the $R_t$ forecasts was significantly worse than the corresponding direct forecasts for most forecasters.
-This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings [CITE SOPHIE's paper].
-This may be related to the change in the case fatality ratio over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK which was not accounted for in the simple model used and not available to be altered by forecasters.
+This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings \citep{meakinComparativeAssessmentMethods2022}.
+This may be related to changes in the age decomposition of cases and the change in the case fatality ratio (CFR) over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK.
+This change was not accounted for in the simple model used and the CFR could not be adjusted by forecasters.
 
@@
 In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
-Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on output from an open source effective reproduction number estimation model and then used to generate simulated reported cases and deaths using the same model and a simple extension which assumed a static case fatality ration and a delay between cases and reports that was estimated from the data.
-We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
-For deaths, performance of direct human forecasts was comparable to that of the Forecast Hub ensemble, while $R_t$ forecasts performed significantly worse.
-Model rankings were not always robust, but sometimes changed depending on how forecasts are evaluated.
-For example, model rankings could differ when focusing on a different forecast horizon, or analysing WIS values on the natural scale vs.
-on the log scale (i.e.
-after log-transforming both forecasts and observations).
+Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model to generate simulated reported cases and deaths and an extension which estimated a fixed case fatality ratio and a log normal delay between cases and reports.
+For case numbers, performance of the human judgement forecasts was broadly comparable to the ensemble of all models submitted to the European Forecast Hub, with results changing slightly depending on the forecast horizon and whether or not forecasts were transformed before scoring.
+The Hub ensemble tended to perform slightly better on longer forecast horizons and when evaluated on the natural scale, while human judgment forecasts tended to perform better for shorter horizons and when evaluated on the log scale.
+%We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
+For deaths, performance of direct human forecasts was broadly comparable to that of the Forecast Hub ensemble (with results again slightly changing depending on the forecast horizons and whether or not forecasts were transformed before scoring), while $R_t$ forecasts performed significantly worse.
+The ensemble that combined $R_t$ and direct forecasts performed only slightly worse than the direct forecasts on the natural scale and was the best performing model on the log scale, in spite of the poor performance of the $R_t$ forecasts.
 
@@
 
-In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
-They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and exact epidemiological relationships between different leading and lagged indicators.
-\citet{mcandrewChimericForecastingCombining2022} similarly found that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
-We could not replicate these findings, and found that in terms of the WIS \textit{on the natural scale}, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed).
-Looking at scores \textit{on the log scale} suggests overall comparable performance of the Hub ensemble and the direct human forecasts on both cases and deaths.
+In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
+They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators.
+\citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
+Results in this study did not show the same pattern.
+We found that in terms of the WIS \textit{on the natural scale}, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed, as did results for the $R_t$ forecasts).
+Looking at scores \textit{on the log scale}, our study suggests overall comparable performance of the Hub ensemble and the direct human forecasts on both cases and deaths.
 Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to the rise of the Delta variant in the UK starting in the beginning of May.
 The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
-Humans may have found it easier than computational models to adapt to the fact that the relationship between cases and deaths continuously changed during the study period.
+One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters.
+Given the variability in scores across forecast horizons and evaluation methods, result should be interpreted with care.
 
@@
 
-This study explored a novel method of forecasting infectious diseases that combines a human forecast of the effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths.
-One appeal of this approach is that the forecaster only needs to provide one quantity that directly represents what she believes about the force of the epidemic in the future.
+This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths.
+One appeal of this approach is that the forecaster can directly forecast the generative process.
 Computational modelling would then take care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, or the relationship between different indicators.
@@
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-The relationship between cases and deaths used to forecast deaths from cases was estimated from past data, forecasters had no way of adjusting it, and could not visualise the forecasts in the app.
-The rise of the Delta variant with a subsequent gradual change in the case fatality ratio (CFR) may have affected $R_t$ forecasts, as the changing CFR was not directly accounted for and estimated distributions from past data only slowly updated to new circumstances.
+The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
+The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Cup.
+Forecasters had no way of observing the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
+Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
+Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon.
+Another potential solution would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
+%The relationship between cases and deaths used to forecast 
+%In addition .
+The gradual change in the CFR was not accounted for in the computational model underying the $R_t$ forecasts, and estimated distributions from past data only slowly updated to new circumstances.
 One potential solution for this would be to allow humans to adjust the CFR and other model parameters manually as part of their forecast.
@@
 
-Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles can produce forecasts that are on par with those of an ensemble of computational models.
-Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly with those submitting models to the European Forecast Hub not being properly incentivised to update their forecasting methodology.
-Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths.
-However, the approach has appealing properties and better results are likely achievable.
+Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models can show similar performance.
+One possible interpretation is that a mixed crowd of human forecaster can produce a viable alternative or addition to an ensemble of mathematical models created by experts.
+Another interpretation is that an ensemble of fairly automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
+All previous studies comparing human judgement forecasts and computational models only ran over short periods of time and the majority of them struggled with recruitment and upkeep.
+Meanwhile, COVID-19 Forecast Hubs have attracted continuous submissions for almost three years and were able to consistently provide high-quality forecasts.
+
+Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
+Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology.
+Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths in this study.
+However, we only presented a prototype of a forecasting approach, which, while having appealing properties, proved challenging to implement.
+Subsequent iterations and improvements could likely achieve better results.
 More research is required to obtain a better understanding of the role of subject matter expertise in infectious disease forecasting.
@@
 Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture.
-Allowing researchers free reign to design their own evaluations post-hoc runs the risk of allowing for motivated reasoning and more work should go in to preventing this.
+Furthermore, evaluating forecasts post-hoc leaves the researchers with many degrees of freedom to make decisions that strongly affect which models look good and there is a high risk of allowing for motivated reasoning.
+More emphasis should be put on measures that prevent this, e.g.
+by establishing common standards for evaluations, pre-registering studies, and making it a norm to display a variety of metrics.
 
@@
 \subsection*{Competing interests}
-The authors declare no competing interests
+The authors declare no competing interests.
 
@@
 \subsection*{Acknowledgements}
-This section should acknowledge anyone who contributed to the research or the
-article but who does not qualify as an author based on the criteria provided earlier
-(e.g. someone or an organisation that provided writing assistance). Please state how
-they contributed; authors should obtain permission to acknowledge from all those
-mentioned in the Acknowledgements section.
-
-Acknowledgement to all forecasters
+We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russel Bradshaw, Sebastian Funk, and Akira Endo.
+All winners donated their prizes.
 

### 2023-03-20 Analysis.rmd
update figure code

@@
 
+avail |> 
+  group_by(target_type, type) |>
+  summarise(
+    mean = mean(`Number forecasts`), 
+    median = median(`Number forecasts`), 
+    min = min(`Number forecasts`), 
+    max = max(`Number forecasts`)
+  )
+
 ## number of unique participants
@@
       filter(target_type == type_filter) |>
-      ggplot(aes(x = reorder(Model, (interval_score)), y = value, fill = Model, 
+      ggplot(aes(x = Model, 
+                 # x = reorder(Model, (interval_score)), 
+                 y = value, fill = Model, 
                  group = Model, alpha = `WIS components`)) + 
@@
       theme_scoringutils() + 
+      scale_y_continuous(label = label_fn) +
       # theme(axis.text.x = element_blank()) + 
@@
        width = 7, height = 6)
+
+scores$model |> unique()
 
@@
 
+# Sinificance and bootstrap analysis
+
+```{r}
+scores |>
+  filter(model %in% models_non_human) |>
+  filter(horizon == 2) |>
+  summarise_scores() |>
+  select(model, forecast_date, type_and_scale, interval_score) |>
+  pivot_wider(names_from = model, values_from = interval_score) |>
+  group_by(type_and_scale) |>
+  summarise(p1 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-direct`, paired = TRUE)$p.value, 
+            p2 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-ensemble`, paired = TRUE)$p.value, 
+            p3 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)
+
+wilcox.test(1:20, 2:21) |> str()
+
+```
+
 Table with performance
@@
 
+```
+
+```{r}
+epinow2_data |>
+  score()
 ```

### 2023-03-20 get-paper-data.R

@@ -192,8 +192,8 @@ flag_anomalies <- function(data, analysis_date = Sys.Date()) {
 }
 
 # -------------------- helper function to clean forecasts -------------------- #
-clean_forecasts <- function(forecasts) {
-  forecasts |>
+clean_forecasts <- function(forecasts, filter_epinow2 = TRUE) {
+  out <- forecasts |>
     mutate(target_type = ifelse(grepl("death", target), "Deaths", "Cases")) |>
     mutate(forecast_date := calc_submission_due_date(forecast_date)) |>
     mutate(target_end_date = as.character(target_end_date)) |>
@@ -205,8 +205,14 @@ clean_forecasts <- function(forecasts) {
            horizon, model, target_end_date, target, target_type, expert) |>
     filter(forecast_date >= "2021-05-24", 
            forecast_date <= "2021-08-16") |>
-    filter(model != "EpiNow2") |>
     select(-target)
+  
+  if (filter_epinow2) {
+    out <- out |> 
+      filter(model != "EpiNow2")
+  }
+  
+  return(out)
 }
 
 
@@ -298,6 +304,14 @@ load_forecasts <- function(directories,
 if (file.exists("data/forecast-data.csv")) {
   data <- fread("data/forecast-data.csv")
 } else {
+  
+  epinow2_data <- 
+    load_forecasts(directories = here("data", "submissions", "rt-forecasts"), 
+                   model_names = "EpiNow2") |>
+    mutate(expert = NA) |>
+    clean_forecasts(filter_epinow2 = FALSE)
+  
+  fwrite(epinow2_data, file = "data/epinow2-forecasts.csv")
   
   forecasts <- load_forecasts(
     directories = c(

### 2023-03-22 Analysis.rmd
update tables, add citations

@@
 ```{r, performance-table}
-make_performance_table <- function(scores, scores_summarised, 
-                                   hor = 2) {
+make_performance_table <- function(
+    scores, scores_summarised, 
+    hor = 2, 
+    cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
+) {
   scores |>
@@
     inner_join(scores_summarised) |>
-    mutate(across(c(where(is.numeric)), signif, digits = 2)) |>
+    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
     arrange(target_type, model) |>
@@
           align = c("l", "l", rep("c", 4)),
-          caption = "Performance for two-week-ahead forecasts \\label{tab:scores}",
+          caption = cap,
           linesep = c('', '', '', '\\addlinespace'),
@@
   filter(model %in% models_non_human) |>
-  make_performance_table(scores_summarised = scores_summarised) %>%
+  make_performance_table(
+    scores_summarised = scores_summarised, 
+    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
+    ) %>%
   gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
@@
   filter(model %in% models_non_human) |>
-  make_performance_table(scores_summarised = scores_summarised, hor = 4) %>%
+  make_performance_table(
+    scores_summarised = scores_summarised, 
+    hor = 4, 
+    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
+  ) %>%
   gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>

### 2023-03-22 manuscript/manuscript.tex

@@
 \author[1, 2]{Sam Abbott}
+\author[6, 7]{Johannes Bracher}
 \author[1, 4]{Edwin van Leeuwen}
 \author[5]{Anne Cori}
-\author[2]{Johannes Bracher}
 \author[1, 2, 3]{Sebastian Funk}
@@
 \citet{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
-However, a comparable approach performed slightly worse than computational models at predicting the 2014/15 outbreak of chikungunya in the Americas.
+However, a comparable approach performed worse than computational models at predicting the 2014/15 outbreak of chikungunya in the Americas.
 \citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths.
@@
 In some situations, human judgement forecasting may have advantages relative to computational models.
-Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and lots of parameters are unknown.
+Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are unknown.
 Humans are generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
@@
 This has been widely shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
-\cite{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
+\citet{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022, swallowChallengesEstimationUncertainty2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
 One approach is to use human forecasts, for example of relevant disease parameters, as an input to computational modelling.
@@
 
-The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} elicits predictions for various COVID-19 related forecast targets from different research groups every week.
-Forecasts had to be submitted every Monday 23.59pm GMT.
-Forecasts are made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
-While forecasts were submitted on Mondays, weeks were defined as epiweeks, ending on a Saturday and starting on Sunday.
+The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} started in March 2021 to elicit predictions for various COVID-19 related forecast targets from different research groups every week.
+The forecasts evaluated in this study were submitted every Monday before 23.59pm GMT between May 24 2021 and August 16 2021.
+Forecasts were made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
+While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday and starting on Sunday.
 Forecast horizons were therefore in fact 5, 12, 19 and 26 days.
-Submissions to the European Forecast Hub follow a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
+Submissions to the European Forecast Hub followed a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
 Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
 Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
-From the 29th of July on, the default Forecast Hub ensemble became a median-ensemble.
-The median number of models included in the Forecast Hub ensemble was nine for cases and ten for deaths (see Figure \ref{fig:num-forecasters} in the SI).
+From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
+The median number of models included in the Forecast Hub ensemble during the study period was nine for cases and ten for deaths (see Figure \ref{fig:num-forecasters} in the SI).
 
@@
 Data were subject to reporting artifacts and revisions.
-All data points are marked as anomalous by the European Forecast Hub if in subsequent updates data is changed by more than 5 percent.
+All data points were marked as anomalous retrospectively by the European Forecast Hub if in subsequent updates data was changed by more than 5 percent.
 In August 2022 JHU switched the data source for their death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency.
@@
 The data and revisions are displayed in Figure \ref{fig:plot-data}.
-All results presented here were derived based on the original data available in 2021.
+All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).
 
@@
 The web application offered participants two different ways of making a forecast, called 'direct' (or 'classical') and 'Rt' forecast.
-To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty.
+To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty at each forecast horizon.
+
+Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in weekly log observed forecasts (i.e.
+as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
 In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
@@
 In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline estimate produced by the \texttt{EpiNow2} \textsf{R} package effective reproduction number model \citep{epinow2} which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
-This forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-\texttt{EpiNow2} implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007}  generative process for latent infections where the effective reproduction number is modelled as Gaussian process on the residuals.
-These latent infections are then convolved with delay distributions, such as the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
+The estimate produced by \texttt{EpiNow2} was shown as the default forecast and could be adjusted by the user.
+The resulting $R_t$ forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}, which implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007} generative process for latent infections.
+We chose a Gaussian Process prior with mean 0 for the first differences of the effective reproduction number in time, implying that in the absence of informative data the reproduction number would remain constant on average, with uncertainty increasing with the temporal distance to informative data points.
+Latent infections were convolved with delay distributions representing the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
 This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022} [CITE ECDC] [CITE SPI-M paper] and used to inform policy makers via reproduction number estimates [CITE one of the SPI-M reproduction number summary papers or kaths paper on multiple surveillance Rt esimates].
@@
 
-As $R_t$-estimates up to at least two weeks prior to the forecast data were uncertain due to depending on partially complete observations of underlying infections, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
+As $R_t$-estimates up to at least two weeks prior to the forecast data were uncertain due to their dependence on partially complete observations of underlying infections given the delays from infection to report, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
 Participants were therefore asked to estimate/predict six $R_t$ values, four of them beyond the forecast horizon.
@@
 "epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts together.
-A participant could enter the "Crowd combined" ensemble twice if she had submitted both a direct and an $R_t$ forecast.
+A participant could enter the "Crowd combined" ensemble twice if they had submitted both a direct and an $R_t$ forecast.
 Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
@@
 
-Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022} except for our extension to also consider the natural log of forecasts as a proxy for the exponential growth rate \citet{bosseTransformationForecastsEvaluating2023}.
+Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022} except for our extension to also consider the natural log of forecasts as a proxy for the exponential growth rate \cite{bosseTransformationForecastsEvaluating2023}.
 If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022a, bracherShorttermForecastingCOVID192021}.
@@
 \cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
-The WIS is a proper scoring rule (with smaller values implying better performance).
-A forecaster, in expectation, optimises her score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report her true belief.
+The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
+A forecaster, in expectation, optimises their score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report their true belief.
 The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
 The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
-uncertainty of forecasts), over-prediction and under-prediction.
+uncertainty of forecasts), overprediction and underprediction.
 Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
 the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
-Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
-All code and data used for this analysis, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
-All code used to submit the forecasts to the European Forecast Hub is available at \url{https://github.com/epiforecasts/europe-covid-forecast}.
 
@@
 by subtracting the mean of all scores for the same target and scale, and dividing by the standard deviation.
+
+Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
+All code and data used for this analysis, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
+All code used to submit the forecasts to the European Forecast Hub is available at \url{https://github.com/epiforecasts/europe-covid-forecast}.
 
@@
 % Could add a sentence that scores were not significant using a paired wilcoxon mann whitney test. 
-On the natural scale, the Forecast Hub ensemble was slightly ahead of the crowd forecasts, with the difference gradually increasing with increasing forecast horizon (see Figure \ref{fig:performance}).
+On the natural scale, the Forecast Hub ensemble was ahead of the crowd forecasts (WIS two weeks ahead: 38k for the Hub ensemble and 40k for the combined human ensemble), with the difference gradually increasing with increasing forecast horizon (see Figure \ref{fig:performance}).
 On the log scale, the Forecast Hub ensemble was slightly worse than all crowd forecasting approaches at a two week forecast horizon, but again slightly ahead for forecasts four weeks into the future.
@@
 Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model to generate simulated reported cases and deaths and an extension which estimated a fixed case fatality ratio and a log normal delay between cases and reports.
-For case numbers, performance of the human judgement forecasts was broadly comparable to the ensemble of all models submitted to the European Forecast Hub, with results changing slightly depending on the forecast horizon and whether or not forecasts were transformed before scoring.
-The Hub ensemble tended to perform slightly better on longer forecast horizons and when evaluated on the natural scale, while human judgment forecasts tended to perform better for shorter horizons and when evaluated on the log scale.
+For case numbers, performance of the human judgement forecasts was broadly comparable to the ensemble of all models submitted to the European Forecast Hub, with results changing depending on the forecast horizon and whether or not forecasts were transformed before scoring.
+The Hub ensemble tended to perform better on longer forecast horizons and when evaluated on the natural scale, while human judgment forecasts tended to perform better for shorter horizons and when evaluated on the log scale.
 %We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
-For deaths, performance of direct human forecasts was broadly comparable to that of the Forecast Hub ensemble (with results again slightly changing depending on the forecast horizons and whether or not forecasts were transformed before scoring), while $R_t$ forecasts performed significantly worse.
+For deaths, performance of direct human forecasts was broadly comparable to that of the Forecast Hub ensemble (with results again changing depending on the forecast horizons and whether or not forecasts were transformed before scoring), while $R_t$ forecasts performed significantly worse.
 The ensemble that combined $R_t$ and direct forecasts performed only slightly worse than the direct forecasts on the natural scale and was the best performing model on the log scale, in spite of the poor performance of the $R_t$ forecasts.
@@
 
-Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to recruit a large number of participants.
+Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to retain a large number of participants.
 Focused public outreach efforts such as creating a dedicated website, announcing an official tournament, providing a public leaderboard, sending weekly emails with details on past performance and weekly announcements on Twitter, did noticeably increase participation compared to the previous study in Germany and Poland.
-Nevertheless, participation was still relatively low and most forecasters only submitted a single forecast.
-\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that provide access to a more elegant user interface and a large existing user base may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
+Nevertheless, retaining participants beyond the initial recruitment proved challenging, and most forecasters only submitted a single forecast.
+\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that provide access to a large existing user base may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
 
@@
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
-The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Cup.
+The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last X weeks of data, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
+The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021}, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
 Forecasters had no way of observing the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
@@
 Another potential solution would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
+Future work could expose forecasters to different such options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
 %The relationship between cases and deaths used to forecast 
@@
 Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
-Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology.
+Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology in order to maximise utility, predictive performance, or both.
 Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths in this study.
@@
 \subsection*{Competing interests}
-The authors declare no competing interests.
+Nikos I. Bosse is a part-time employee of Metaculus, an online prediction platform.
 

### 2023-03-22 Analysis.rmd
Add tables for experts and non-experts

@@
 
+expert_scores_summarised <- expert_ensemble$scores |>
+  filter(scale %in% c("log", "natural")) |>
+  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
+  group_by(target_type, scale, horizon) |>
+  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
+  group_by(model, target_type, scale, horizon) |>
+  summarise(mean = mean(interval_score), 
+            sd = sd(interval_score)) |>
+  pivot_wider(id_cols = c(model, target_type, horizon), 
+              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
+  arrange(target_type, model)
+
 expert_ensemble$scores |>
-   make_performance_table()
+   make_performance_table(
+     scores_summarised = expert_scores_summarised,
+    cap = "Performance for two-week-ahead forecasts of experts and non-experts. Standard deviations are given in brackets. \\label{tab:scores-experts}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-experts.tex')
+
+expert_ensemble$scores |>
+   make_performance_table(
+     scores_summarised = expert_scores_summarised,
+    cap = "Performance for four-week-ahead forecasts of experts and non-experts. Standard deviations are given in brackets. \\label{tab:scores-experts-4}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-experts-4.tex')
 

### 2023-03-22 manuscript/manuscript.tex
Updates from Overleaf

@@
 
-On the natural scale, direct forecasts had a slightly better WIS than the Hub ensemble and the other two human forecast ensembles for a two-week-ahead horizon, while the Hub ensemble was again slightly ahead for a four week-ahead horizon (see Figure \ref{fig:performance}).
-The combined crowd forecast (containing both the direct and the $R_t$ forecasts) overall performed slightly worse than the direct crowd forecast on the natural scale, but much better than the $R_t$ forecasts.
+On the natural scale, direct forecasts had a better WIS than the Hub ensemble and the other two human forecast ensembles for a two-week-ahead horizon (direct: 34, Hub ensemble: 38, combined crowd: 40, $R_t$: 80), while the Hub ensemble was again slightly ahead for a four week-ahead horizon (direct: 88, Hub ensemble: 85, combined crowd: 98, $R_t$: 150, see Figure \ref{fig:performance} and Table \ref{tab:scores-4}).
+The combined crowd forecast (containing both the direct and the $R_t$ forecasts) overall performed worse than the direct crowd forecast on the natural scale, but much better than the $R_t$ forecasts.
 On the log scale, the combined crowd forecasts even performed best among the four approaches.
@@
 
+\input performance-table-horizon-4.tex
+
 \begin{figure*}

### 2023-03-27 Analysis.rmd
restructure code file

@@
 ---
+
+# Setup, define Helper Functions, Load Data
+================================================================================
 
@@
 
-```
-
-# Define Helper Functions, Load Data
-
-```{r}
+# Helper functions
+# ==============================================================================
 
@@
 
-# Figure SI.1: Data and data revisions
-data revisions: https://github.com/CSSEGISandData/COVID-19/issues/6021
-https://coronavirus.data.gov.uk/details/whats-new 
-https://coronavirus.data.gov.uk/details/whats-new/record/965a7923-dacf-42ed-abb9-3a972dbf1835
+# Table 1: Performance
+================================================================================
 
-```{r}
-plot_data <- function(weekly_truth, daily_truth) {
-  weekly_truth |>
-    mutate(status = str_to_title(status)) |>
-  ggplot(aes(x = target_end_date, y = true_value)) +
-  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
-           fill = "darkolivegreen3", alpha = .1) + 
-  geom_line() + 
-  geom_point(size = 0.8, aes(color = status)) + 
-  geom_bar(data = daily_truth |>
-             mutate(true_value = true_value * 7), 
-           stat = "identity", 
-           fill = "skyblue3", alpha = 0.4) + 
-  facet_wrap(~ target_type, 
-             nrow = 2, scale = "free_y") + 
-  scale_y_continuous(labels = label_fn) +
-  theme_scoringutils() + 
-  scale_colour_manual(values = c("red", "black")) + 
-  labs(y = "Original data", x = "Date", color = "Data status")
+```{r, performance-table}
+scores_summarised <- scores |>
+  filter(scale %in% c("log", "natural")) |>
+  filter(model %in% models_non_human) |>
+  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
+  group_by(target_type, scale, horizon) |>
+  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
+  group_by(model, target_type, scale, horizon) |>
+  summarise(mean = mean(interval_score), 
+            sd = sd(interval_score)) |>
+  pivot_wider(id_cols = c(model, target_type, horizon), 
+              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
+  arrange(target_type, model)
+
+make_performance_table <- function(
+    scores, scores_summarised, 
+    hor = 2, 
+    cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
+) {
+  scores |>
+    filter(scale %in% c("log", "natural")) |>
+    summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
+    select(model, target_type, scale, interval_score, coverage_50, coverage_90, horizon) |>
+    pivot_wider(names_from = scale, values_from = interval_score) |>
+    select(model, target_type, natural, log, coverage_50, coverage_90, horizon) |>
+    inner_join(scores_summarised) |>
+    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
+    arrange(target_type, model) |>
+    mutate(natural = label_fn(natural), 
+           log = label_fn(log)) |>
+    mutate(natural = paste0(natural, " (", label_fn(sd_stand_natural), ")"), 
+           log = paste0(log, " (", label_fn(sd_stand_log), ")")) |>
+    filter(horizon == hor) |>
+    select(-mean_stand_log, -sd_stand_log, -mean_stand_natural, -sd_stand_natural, -horizon) |>
+    kable(format = "latex", booktabs = TRUE, 
+          align = c("l", "l", rep("c", 4)),
+          caption = cap,
+          linesep = c('', '', '', '\\addlinespace'),
+          col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
+                        "Coverage 50%", "Coverage 90%")) |> 
+    kable_styling(latex_options = c("scale_down", 
+                                    "hold_position")) 
 }
 
-p_data_revised <- plot_data(weekly_truth_revised |>
-                                 filter(target_type == "Deaths"), 
-                               daily_truth_revised |>
-                                 filter(target_type == "Deaths")) + 
-  ylab("Revised data")
-p_data <- plot_data(weekly_truth, daily_truth)
+scores |>
+  filter(model %in% models_non_human) |>
+  make_performance_table(
+    scores_summarised = scores_summarised, 
+    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table.tex')
 
-p_weekly_revisions <- weekly_truth |>
-  rbind(weekly_truth_revised) |>
-  filter(target_type == "Deaths") |>
-  select(target_end_date, true_value, target_type) |>
-  group_by(target_end_date, target_type) |>
-  mutate(diff = c(NA, diff(true_value))) |>
-  filter(!is.na(diff)) |>
-  ggplot(aes(x = target_end_date, y = diff)) + 
-  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
-           fill = "darkolivegreen3", alpha = .1) +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") + 
-  geom_line() + 
-  geom_point(size = 0.7) + 
-  facet_wrap(~ target_type, ncol = 1, scale = "free_y") +
-  labs(y = "Weekly data revisions", x = "Date") + 
-  coord_cartesian(ylim = c(-500, 500)) + 
-  theme_scoringutils()
+scores |>
+  filter(model %in% models_non_human) |>
+  make_performance_table(
+    scores_summarised = scores_summarised, 
+    hor = 4, 
+    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
+  ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-horizon-4.tex')
 
-p_all_data <- p_data / 
-  p_data_revised / 
-  p_weekly_revisions +
-  plot_annotation(tag_levels = "A") + 
-  plot_layout(guides = "collect", heights = c(2.2, 1, 1)) & 
-  theme(legend.position = "bottom")
+# scores_revised |>
+#   filter(model %in% models_non_human) |>
+#   make_performance_table() |>
+#   cat(file = 'manuscript/performance-table-revised.tex')
+# 
+# 
+# scores_comparable_ensemble |>
+#   make_performance_table() |>
+#   cat(file = 'manuscript/performance-table-comparison-rt.tex')
 
-p_all_data
-
-ggsave("output/figures/plot-data.png", plot = p_all_data, 
-       width = 7, height = 8)
 ```
 
-# Figure SI.4: Weekly participation humans and hub ensemble
-
-```{r}
-available <- function(forecasts, filter, name) {
-  forecasts |>
-    filter(model %in% filter) |>
-    avail_forecasts(by = c("forecast_date", "target_type"), 
-                    collapse = c("horizon", "quantile", "target_end_date")) |>
-    mutate(type = name)
-}
-
-## Number of forecasts and ensemble members
-avail <- 
-  available(forecasts, filter = models_direct, name = "Direct") |>
-  rbind(available(forecasts, filter = models_rt, name = "Rt")) |>
-  mutate(group = "Human forecast")
-
-avail_ensemble <- ensemble_inclusions |>
-  mutate(prediction = 1, true_value = 1) |>
-  filter(forecast_date <= "2021-08-16") |>
-  unique() |>
-  avail_forecasts(by = c("forecast_date", "target_type"), 
-                    collapse = c("horizon", "quantile", "target_end_date")) |>
-  mutate(group = "Forecast Hub")
-
-plot_number_forecasters <- avail |>
-  pivot_wider(names_from = type, values_from = `Number forecasts`) |>
-  mutate(Rt = Direct + Rt) |>
-  pivot_longer(cols = c(Direct, Rt), names_to = "type",
-               values_to = "Number forecasts") |>
-  as.data.table() |>
-  rbind(avail_ensemble, fill = TRUE) |>
-  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
-             group = group,
-             fill = group, 
-             alpha = type)) +
-  geom_bar(stat = "identity", 
-           color = "white",
-           linewidth = 0.2,
-           width = 5,
-           position = position_dodge(width = 5)) + 
-  facet_wrap(~ target_type) + 
-  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
-  labs(x = "Forecast date", fill = "Forecast type", 
-       alpha = "Forecast method") + 
-  theme_scoringutils() +
-  theme(legend.position = "bottom", 
-        legend.box="vertical", legend.margin = margin())
-
-ggsave("output/figures/num-forecasters.png", plot = plot_number_forecasters, 
-       width = 7, height = 3.5)
-
-avail |> 
-  group_by(target_type, type) |>
-  summarise(
-    mean = mean(`Number forecasts`), 
-    median = median(`Number forecasts`), 
-    min = min(`Number forecasts`), 
-    max = max(`Number forecasts`)
-  )
-
-## number of unique participants
-unique_participants <- gsub("\ \\(Rt\\)", "", models_human) |> unique()
-length(unique_participants)
-
-## Number of unique participants in every week
-forecasts |>
-  filter(model %in% models_human) |>
-  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
-  select(model, forecast_date) |>
-  group_by(forecast_date) |>
-  summarise(n = length(unique(model))) |>
-  summary()
-
-## number of times a forecaster submitted
-forecasts |>
-  filter(model %in% models_human) |>
-  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
-  select(forecast_date, model) |>
-  group_by(model) |>
-  summarise(n = length(unique(forecast_date))) |>
-  summary()
-```
-
-## Performance
+# Figure 1: Predictive performance across horizons (summary)
+================================================================================
 
@@
        width = 7, height = 6)
-
-scores$model |> unique()
-
 ```
 
-```{r}
-# helper function to score forecasts both on the natural and on the log scale
-score_and_add_log <- function(forecasts) {
-  forecasts |>
-  mutate(scale = "natural") |>
-  rbind(forecasts |>
-          mutate(
-            scale = "log", 
-            true_value = log(true_value + 1), 
-            prediction = log(pmax(prediction, 0) + 1)
-          )) |>
-  score(metrics = c("interval_score", "coverage")) |>
-  add_coverage(by = c("target_type", "horizon", "method")) |>
-    add_type_and_scale() |>
-  as.data.table() 
-}
-
-# helper function that filters the forecast data such that we only have 
-# forecasts left where a single person made a forecast both for direct and rt
-filter_comparison_rt_direct <- function(forecasts) {
-  forecasts_comparison <- forecasts |>
-  filter(model %in% models_human) |>
-  mutate(rt_name = ifelse(
-    grepl("(Rt)", model), 
-    gsub("\ \\(Rt\\)", "", model), 
-    NA
-  )) |>
-  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
-  group_by(forecast_date, target_type) |>
-  filter(model %in% rt_name | rt_name %in% model) |>
-  group_by(forecast_date, target_type, method) |>
-  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile)
-}
-
-forecasts_comparison <- filter_comparison_rt_direct(forecasts)
-scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
-
-forecasts_comparison_revised <- filter_comparison_rt_direct(forecasts_revised)
-scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)
-
-scores_forecasts_comparison |> 
-  filter(horizon == 2) |>
-  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
-  summarise_scores(by = c("target_type", "type_and_scale", "name", "method", "scale")) |>
-  ggplot(aes(y = interval_score, x = name, group = method, color = method)) + 
-  geom_point() + 
-  geom_line() + 
-  facet_wrap(~ type_and_scale, scale = "free_y") + 
-  theme_scoringutils()
-
-plot_rt_direct_individual <- function(scores) {
- scores |> 
-  filter(horizon == 2) |>
-  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
-  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
-  arrange(target_type, name, method) |>
-  group_by(target_type, name, scale) |>
-  mutate(diff = rep(diff(interval_score), 2)) |>
-  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
-  mutate(method = stringr::str_to_title(method)) |>
-  ggplot(aes(y = interval_score, x = method, group = name)) + 
-  geom_point(size = 0.5) + 
-  geom_line(aes(color = improvement), linewidth = 0.3) + 
-  facet_wrap(~ type_and_scale, scale = "free_y") + 
-  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
-  scale_y_continuous(labels = label_fn) + 
-  theme_scoringutils()
-}
-
-p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)
-
-ggsave("output/figures/comparison-direct-rt-individual.png", 
-       plot = p_comparison_direct_rt_individual, 
-       width = 7, height = 4)
-
-p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)
-
-ggsave("output/figures/comparison-direct-rt-individual-revised.png", 
-       plot = p_comparison_direct_rt_individual, 
-       width = 7, height = 5)
-
-# performance plot for an ensemble of the subset of human and direct forecasts
-forecasts_comparable_ensemble <- forecasts_comparison |>
-  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
-  summarise(prediction = median(prediction)) |>
-  mutate(model = str_to_title(method)) |>
-  as.data.table()
-
-scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)
-
-p_performance_comparison <- plot_performance(
-  scores_comparable_ensemble
-)
-
-ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
-       width = 7, height = 8)
-
-# performance plot for an ensemble of the subset of human and direct forecasts
-# using revised death data
-
-forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
-  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
-  summarise(prediction = median(prediction)) |>
-  mutate(model = str_to_title(method)) |>
-  as.data.table()
-
-scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)
-
-p_performance_comparison_revised <- plot_performance(
-  scores_comparable_ensemble_revised
-)
-
-ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison, 
-       width = 7, height = 8)
-
-scores_comparable_ensemble |>
-  filter(horizon == 2) |>
-  summarise_scores(by = c("target_type", "horizon", "model", "scale"))
-
-scores |>
-  filter(scale != "sqrt") |>
-  filter(model %in% models_non_human) |>
-  filter(horizon == 2,
-         scale == "log") |>
-  summarise_scores(by = c("target_type", "horizon", "model", "scale"))
-
-```
-
-```{r}
-scores_summarised <- scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models_non_human) |>
-  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
-  group_by(target_type, scale, horizon) |>
-  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
-  group_by(model, target_type, scale, horizon) |>
-  summarise(mean = mean(interval_score), 
-            sd = sd(interval_score)) |>
-  pivot_wider(id_cols = c(model, target_type, horizon), 
-              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
-  arrange(target_type, model)
-```
-
-# Sinificance and bootstrap analysis
-
-```{r}
-scores |>
-  filter(model %in% models_non_human) |>
-  filter(horizon == 2) |>
-  summarise_scores() |>
-  select(model, forecast_date, type_and_scale, interval_score) |>
-  pivot_wider(names_from = model, values_from = interval_score) |>
-  group_by(type_and_scale) |>
-  summarise(p1 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-direct`, paired = TRUE)$p.value, 
-            p2 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-ensemble`, paired = TRUE)$p.value, 
-            p3 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)
-
-wilcox.test(1:20, 2:21) |> str()
-
-```
-
-Table with performance
-
-```{r, performance-table}
-make_performance_table <- function(
-    scores, scores_summarised, 
-    hor = 2, 
-    cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
-) {
-  scores |>
-    filter(scale %in% c("log", "natural")) |>
-    summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
-    select(model, target_type, scale, interval_score, coverage_50, coverage_90, horizon) |>
-    pivot_wider(names_from = scale, values_from = interval_score) |>
-    select(model, target_type, natural, log, coverage_50, coverage_90, horizon) |>
-    inner_join(scores_summarised) |>
-    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
-    arrange(target_type, model) |>
-    mutate(natural = label_fn(natural), 
-           log = label_fn(log)) |>
-    mutate(natural = paste0(natural, " (", label_fn(sd_stand_natural), ")"), 
-           log = paste0(log, " (", label_fn(sd_stand_log), ")")) |>
-    filter(horizon == hor) |>
-    select(-mean_stand_log, -sd_stand_log, -mean_stand_natural, -sd_stand_natural, -horizon) |>
-    kable(format = "latex", booktabs = TRUE, 
-          align = c("l", "l", rep("c", 4)),
-          caption = cap,
-          linesep = c('', '', '', '\\addlinespace'),
-          col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
-                        "Coverage 50%", "Coverage 90%")) |> 
-    kable_styling(latex_options = c("scale_down", 
-                                    "hold_position")) 
-}
-
-scores |>
-  filter(model %in% models_non_human) |>
-  make_performance_table(
-    scores_summarised = scores_summarised, 
-    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
-    ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table.tex')
-
-scores |>
-  filter(model %in% models_non_human) |>
-  make_performance_table(
-    scores_summarised = scores_summarised, 
-    hor = 4, 
-    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
-  ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table-horizon-4.tex')
-
-# scores_revised |>
-#   filter(model %in% models_non_human) |>
-#   make_performance_table() |>
-#   cat(file = 'manuscript/performance-table-revised.tex')
-# 
-# 
-# scores_comparable_ensemble |>
-#   make_performance_table() |>
-#   cat(file = 'manuscript/performance-table-comparison-rt.tex')
-
-```
-
-```{r}
-epinow2_data |>
-  score()
-```
+# Figure 2: Forecasts and corresponding WIS
+================================================================================
 
@@
        width = 9, height = 9)
+```
 
-p_scores_and_pred_comparison <- plot_scores_and_pred(
-  forecasts_comparable_ensemble,
-  scores_comparable_ensemble
+# Figure 3: Comparison performance direct vs. Rt
+================================================================================
+
+```{r}
+# helper function to score forecasts both on the natural and on the log scale
+score_and_add_log <- function(forecasts) {
+  forecasts |>
+  mutate(scale = "natural") |>
+  rbind(forecasts |>
+          mutate(
+            scale = "log", 
+            true_value = log(true_value + 1), 
+            prediction = log(pmax(prediction, 0) + 1)
+          )) |>
+  score(metrics = c("interval_score", "coverage")) |>
+  add_coverage(by = c("target_type", "horizon", "method")) |>
+    add_type_and_scale() |>
+  as.data.table() 
+}
+
+# helper function that filters the forecast data such that we only have 
+# forecasts left where a single person made a forecast both for direct and rt
+filter_comparison_rt_direct <- function(forecasts) {
+  forecasts_comparison <- forecasts |>
+  filter(model %in% models_human) |>
+  mutate(rt_name = ifelse(
+    grepl("(Rt)", model), 
+    gsub("\ \\(Rt\\)", "", model), 
+    NA
+  )) |>
+  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
+  group_by(forecast_date, target_type) |>
+  filter(model %in% rt_name | rt_name %in% model) |>
+  group_by(forecast_date, target_type, method) |>
+  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile)
+}
+
+forecasts_comparison <- filter_comparison_rt_direct(forecasts)
+scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
+
+plot_rt_direct_individual <- function(scores) {
+ scores |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
+  arrange(target_type, name, method) |>
+  group_by(target_type, name, scale) |>
+  mutate(diff = rep(diff(interval_score), 2)) |>
+  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
+  mutate(method = stringr::str_to_title(method)) |>
+  ggplot(aes(y = interval_score, x = method, group = name)) + 
+  geom_point(size = 0.5) + 
+  geom_line(aes(color = improvement), linewidth = 0.3) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
+  scale_y_continuous(labels = label_fn) + 
+  theme_scoringutils()
+}
+
+p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)
+
+ggsave("output/figures/comparison-direct-rt-individual.png", 
+       plot = p_comparison_direct_rt_individual, 
+       width = 7, height = 4)
+```
+
+================================================================================
+#                                  Appendix                                    #
+================================================================================
+
+# Figure SI.1: Number of forecasters across the study period
+================================================================================
+- Figure SI.1
+- numbers of forecasters used in the text
+
+```{r}
+available <- function(forecasts, filter, name) {
+  forecasts |>
+    filter(model %in% filter) |>
+    avail_forecasts(by = c("forecast_date", "target_type"), 
+                    collapse = c("horizon", "quantile", "target_end_date")) |>
+    mutate(type = name)
+}
+
+## Number of forecasts and ensemble members
+avail <- 
+  available(forecasts, filter = models_direct, name = "Direct") |>
+  rbind(available(forecasts, filter = models_rt, name = "Rt")) |>
+  mutate(group = "Human forecast")
+
+avail_ensemble <- ensemble_inclusions |>
+  mutate(prediction = 1, true_value = 1) |>
+  filter(forecast_date <= "2021-08-16") |>
+  unique() |>
+  avail_forecasts(by = c("forecast_date", "target_type"), 
+                    collapse = c("horizon", "quantile", "target_end_date")) |>
+  mutate(group = "Forecast Hub")
+
+plot_number_forecasters <- avail |>
+  pivot_wider(names_from = type, values_from = `Number forecasts`) |>
+  mutate(Rt = Direct + Rt) |>
+  pivot_longer(cols = c(Direct, Rt), names_to = "type",
+               values_to = "Number forecasts") |>
+  as.data.table() |>
+  rbind(avail_ensemble, fill = TRUE) |>
+  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
+             group = group,
+             fill = group, 
+             alpha = type)) +
+  geom_bar(stat = "identity", 
+           color = "white",
+           linewidth = 0.2,
+           width = 5,
+           position = position_dodge(width = 5)) + 
+  facet_wrap(~ target_type) + 
+  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
+  labs(x = "Forecast date", fill = "Forecast type", 
+       alpha = "Forecast method") + 
+  theme_scoringutils() +
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin())
+
+ggsave("output/figures/num-forecasters.png", plot = plot_number_forecasters, 
+       width = 7, height = 3.5)
+
+avail |> 
+  group_by(target_type, type) |>
+  summarise(
+    mean = mean(`Number forecasts`), 
+    median = median(`Number forecasts`), 
+    min = min(`Number forecasts`), 
+    max = max(`Number forecasts`)
   )
 
-ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
-       plot = p_scores_and_pred_comparison, 
-       width = 9, height = 9)
+## number of unique participants
+unique_participants <- gsub("\ \\(Rt\\)", "", models_human) |> unique()
+length(unique_participants)
 
+## Number of unique participants in every week
+forecasts |>
+  filter(model %in% models_human) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  select(model, forecast_date) |>
+  group_by(forecast_date) |>
+  summarise(n = length(unique(model))) |>
+  summary()
+
+## number of times a forecaster submitted
+forecasts |>
+  filter(model %in% models_human) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  select(forecast_date, model) |>
+  group_by(model) |>
+  summarise(n = length(unique(forecast_date))) |>
+  summary()
 ```
 
-## Look at expert forecasts
+# Figure SI.2: Data and data revisions
+================================================================================
 
-# ensemble of experts and non-experts
+Sources for data revisions: 
+- https://github.com/CSSEGISandData/COVID-19/issues/6021
+- https://coronavirus.data.gov.uk/details/whats-new 
+- https://coronavirus.data.gov.uk/details/whats-new/record/965a7923-dacf-42ed-abb9-3a972dbf1835
 
 ```{r}
+plot_data <- function(weekly_truth, daily_truth) {
+  weekly_truth |>
+    mutate(status = str_to_title(status)) |>
+  ggplot(aes(x = target_end_date, y = true_value)) +
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
+           fill = "darkolivegreen3", alpha = .1) + 
+  geom_line() + 
+  geom_point(size = 0.8, aes(color = status)) + 
+  geom_bar(data = daily_truth |>
+             mutate(true_value = true_value * 7), 
+           stat = "identity", 
+           fill = "skyblue3", alpha = 0.4) + 
+  facet_wrap(~ target_type, 
+             nrow = 2, scale = "free_y") + 
+  scale_y_continuous(labels = label_fn) +
+  theme_scoringutils() + 
+  scale_colour_manual(values = c("red", "black")) + 
+  labs(y = "Original data", x = "Date", color = "Data status")
+}
 
+p_data_revised <- plot_data(weekly_truth_revised |>
+                                 filter(target_type == "Deaths"), 
+                               daily_truth_revised |>
+                                 filter(target_type == "Deaths")) + 
+  ylab("Revised data")
+p_data <- plot_data(weekly_truth, daily_truth)
+
+p_weekly_revisions <- weekly_truth |>
+  rbind(weekly_truth_revised) |>
+  filter(target_type == "Deaths") |>
+  select(target_end_date, true_value, target_type) |>
+  group_by(target_end_date, target_type) |>
+  mutate(diff = c(NA, diff(true_value))) |>
+  filter(!is.na(diff)) |>
+  ggplot(aes(x = target_end_date, y = diff)) + 
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
+           fill = "darkolivegreen3", alpha = .1) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") + 
+  geom_line() + 
+  geom_point(size = 0.7) + 
+  facet_wrap(~ target_type, ncol = 1, scale = "free_y") +
+  labs(y = "Weekly data revisions", x = "Date") + 
+  coord_cartesian(ylim = c(-500, 500)) + 
+  theme_scoringutils()
+
+p_all_data <- p_data / 
+  p_data_revised / 
+  p_weekly_revisions +
+  plot_annotation(tag_levels = "A") + 
+  plot_layout(guides = "collect", heights = c(2.2, 1, 1)) & 
+  theme(legend.position = "bottom")
+
+p_all_data
+
+  ggsave("output/figures/plot-data.png", plot = p_all_data, 
+       width = 7, height = 8)
+```
+
+# Figure SI.3 + SI.4: Screenshots from the forecasting app
+================================================================================
+
+# Figure SI.5: Performance of "experts" and "non-experts"
+================================================================================
+Separate ensemble of experts and non-experts
+
+```{r}
 create_expert_ensemble <- function(forecasts) {
@@
      scores_summarised = expert_scores_summarised,
+     hor = 4,
      cap = "Performance for four-week-ahead forecasts of experts and non-experts. Standard deviations are given in brackets. \\label{tab:scores-experts-4}"
@@
 
+```
+
+# Unused analyses
+================================================================================
+
+# Analysis on a comparable ensemble of Rt and direct forecasts
+```{r}
+# performance plot for an ensemble of the subset of human and direct forecasts
+forecasts_comparable_ensemble <- forecasts_comparison |>
+  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+  summarise(prediction = median(prediction)) |>
+  mutate(model = str_to_title(method)) |>
+  as.data.table()
+
+scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)
+
+p_performance_comparison <- plot_performance(
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
+       width = 7, height = 8)
+
+p_scores_and_pred_comparison <- plot_scores_and_pred(
+  forecasts_comparable_ensemble,
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
+       plot = p_scores_and_pred_comparison, 
+       width = 9, height = 9)
+
+```
+
+# Analyses on revised data
+```{r}
+# forecasts_comparison_revised <- filter_comparison_rt_direct(forecasts_revised)
+# scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)
+
+# p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)
+# 
+# ggsave("output/figures/comparison-direct-rt-individual-revised.png", 
+#        plot = p_comparison_direct_rt_individual, 
+#        width = 7, height = 5)
+
+# performance plot for an ensemble of the subset of human and direct forecasts
+# using revised death data
+
+# forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
+#   group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+#   summarise(prediction = median(prediction)) |>
+#   mutate(model = str_to_title(method)) |>
+#   as.data.table()
+# 
+# scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)
+# 
+# p_performance_comparison_revised <- plot_performance(
+#   scores_comparable_ensemble_revised
+# )
+# 
+# ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison, 
+#        width = 7, height = 8)
+```
+
+# Significance and bootstrap analysis
+
+```{r}
+scores |>
+  filter(model %in% models_non_human) |>
+  filter(horizon == 2) |>
+  summarise_scores() |>
+  select(model, forecast_date, type_and_scale, interval_score) |>
+  pivot_wider(names_from = model, values_from = interval_score) |>
+  group_by(type_and_scale) |>
+  summarise(p1 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-direct`, paired = TRUE)$p.value, 
+            p2 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-ensemble`, paired = TRUE)$p.value, 
+            p3 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)
+```
+
+```{r}
+epinow2_data |>
+  score()
+```
+
+```{r}
+# Only look at experts with direct forecasts
 expert_ensemble_direct <- forecasts |>
@@
 expert_ensemble_direct$scores |>
+  summarise_scores(c("model", "target_type", "horizon", "scale"))
    make_performance_table()
@@
        width = 7, height = 6)
-
-# plot_scores_and_pred(
-#   forecasts_expert_ensemble,
-#   scores_expert_ensemble
-# )
-
 ```
-
-```{r}
-scores_df <- scores  |>
-  update_expert_status() |>
-  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
-  filter(model %in% models_human) |>
-  filter(horizon == 2) |>
-  summarise_scores()
-
-mean_wis <- scores_df |>
-  group_by(scale, expert, target_type, type_and_scale) |>
-  summarise(wis_mean = signif(median(interval_score), 3)) |>
-  mutate(wis_mean = paste("Mean:", label_fn(wis_mean)))
-
-scores_df |>   
-  ggplot(aes(x = expert, y = interval_score)) +
-  geom_violin(alpha = 0.2) + 
-  geom_boxplot(alpha = 0.5) + 
-  # scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
-  geom_text(data = mean_wis, 
-            aes(label = unique(wis_mean), y = -Inf), 
-            vjust   = -1) + 
-  facet_wrap(~ type_and_scale, scale = "free_y") + 
-  scale_y_continuous(labels = label_fn, 
-                     expand = expansion(mult = c(0.15, 0.05))) + 
-  theme_scoringutils()
-
-scores_df |>
-  summarise_scores(by = c("expert", "target_type", "scale", "horizon")) |>
-  rename(model = expert) |>
-  make_performance_table()
-
-```
-
-```{r}
-forecasts$forecast_date |>
-  unique()
-
-forecasts$target_end_date |>
-  unique()
-```
-
-Interesting questions
-- [x] performance for cases and deaths
-- [x] on the natural scale and log scale
-
-- performance experts vs. non-experts
-- performance Rt vs. non-Rt forecast
-- need a table with number of participants
-- add calibration plot (maybe just copy the plot from the Germany / Poland paper)
-- need to find the numbers of member models
-
-Methods
-- Hub ensemble switched from mean to median on 2021-07-19
 

### 2023-03-27 Analysis.rmd
Clean code, update main table

@@
 
+```
+
+UPDATED VERSION
+```{r, performance-table2}
+make_performance_table <- function(
+    scores, 
+    hor = 2, 
+    model_filter = models_non_human,
+    cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
+) {
+  df <- scores |>
+    filter(scale %in% c("log", "natural"), 
+           model %in% model_filter, 
+           horizon %in% hor)
+
+  sd <- df |>
+    summarise_scores(by = c("model", "target_type", "scale", "horizon"), 
+                     fun = stats::sd) |>
+    select(model, target_type, scale, horizon, wis_sd = interval_score)
+
+  df |>
+    summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
+    select(model, target_type, scale, interval_score, coverage_50, coverage_90, horizon) |>
+    inner_join(sd) |>
+    group_by(target_type, scale) |>
+    mutate(relative = interval_score / interval_score[model == "EuroCOVIDhub-ensemble"]) |>
+    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
+    mutate(across(c(where(is.numeric)), \(x) label_fn(x))) |> 
+    # mutate(interval_score = paste0(interval_score, " (", relative, "), sd: ", wis_sd)) |>
+    # select(-relative, -wis_sd) |>
+    pivot_wider(names_from = scale, values_from = c(interval_score, wis_sd, relative)) |>
+    arrange(target_type, model) |>
+    select(model, target_type, interval_score_natural, relative_natural, wis_sd_natural, interval_score_log, relative_log, wis_sd_log, coverage_50, coverage_90) |>
+    kable(format = "latex", booktabs = TRUE, 
+          align = c("l", "l", rep("c", 4)),
+          caption = cap,
+          linesep = c('', '', '', '\\addlinespace'),
+          col.names = c("Model", "Target", "abs.", "rel.", "sd", "abs.", "rel.", "sd", 
+                        "Coverage 50%", "Coverage 90%")) |> 
+    kable_styling(latex_options = c("scale_down", 
+                                    "hold_position")) |>
+    add_header_above(c(" " = 2, "WIS" = 3, "WIS - log scale" = 3, " " = 2)) 
+}
+
+scores |>
+  filter(model %in% models_non_human) |>
+  make_performance_table(
+    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table.tex')
+
+scores |>
+  filter(model %in% models_non_human) |>
+  make_performance_table(
+    hor = 4, 
+    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
+  ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-horizon-4.tex')
 ```
@@
 
-# Unused analyses
-================================================================================
-
-# Analysis on a comparable ensemble of Rt and direct forecasts
-```{r}
-# performance plot for an ensemble of the subset of human and direct forecasts
-forecasts_comparable_ensemble <- forecasts_comparison |>
-  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
-  summarise(prediction = median(prediction)) |>
-  mutate(model = str_to_title(method)) |>
-  as.data.table()
-
-scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)
-
-p_performance_comparison <- plot_performance(
-  scores_comparable_ensemble
-)
-
-ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
-       width = 7, height = 8)
-
-p_scores_and_pred_comparison <- plot_scores_and_pred(
-  forecasts_comparable_ensemble,
-  scores_comparable_ensemble
-)
-
-ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
-       plot = p_scores_and_pred_comparison, 
-       width = 9, height = 9)
-
-```
-
-# Analyses on revised data
-```{r}
-# forecasts_comparison_revised <- filter_comparison_rt_direct(forecasts_revised)
-# scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)
-
-# p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)
-# 
-# ggsave("output/figures/comparison-direct-rt-individual-revised.png", 
-#        plot = p_comparison_direct_rt_individual, 
-#        width = 7, height = 5)
-
-# performance plot for an ensemble of the subset of human and direct forecasts
-# using revised death data
-
-# forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
-#   group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
-#   summarise(prediction = median(prediction)) |>
-#   mutate(model = str_to_title(method)) |>
-#   as.data.table()
-# 
-# scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)
-# 
-# p_performance_comparison_revised <- plot_performance(
-#   scores_comparable_ensemble_revised
-# )
-# 
-# ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison, 
-#        width = 7, height = 8)
-```
-
-# Significance and bootstrap analysis
-
+# Unused - Significance analysis
+# ============================================================================ #
 ```{r}
@@
             p3 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)
-```
-
-```{r}
-epinow2_data |>
-  score()
-```
-
-```{r}
-# Only look at experts with direct forecasts
-expert_ensemble_direct <- forecasts |>
-  filter(model %in% models_direct) |>
-  create_expert_ensemble()
-
-expert_ensemble_direct$scores |>
-  summarise_scores(c("model", "target_type", "horizon", "scale"))
-   make_performance_table()
-
-p_scores_expert_direct <- plot_performance(expert_ensemble_direct$scores)
-ggsave("output/figures/performance-expert-direct.png", 
-       plot = p_scores_expert_direct, 
-       width = 7, height = 6)
-```
-
-# Parking lot
-Median forecasts for a 1:4 horizon
-```{r}
-models <- forecasts$model |> unique()
-
-horizon_0 <- weekly_truth |>
-  filter(target_end_date >= "2021-04-01") |>
-  mutate(prediction = true_value, 
-         forecast_date = target_end_date + 2, 
-         horizon = 0, 
-         quantile = 0.5, 
-         model = list(models)) |>
-  tidyr::unnest(cols = model)
-
-plot_df <- forecasts |>
-  filter(quantile == 0.5) |>
-  rbind(horizon_0, fill = TRUE) |>
-  filter(target_end_date >= "2021-04-01", 
-         target_end_date <= "2021-10-01") |>
-  filter(!(model %in% c("crowd-rt", "EpiExpert-ensemble"))) |>
-  filter(!(grepl("(Rt)", model))) |>
-  mutate(colouring = ifelse(model == "crowd-ensemble", 
-                            "Crowd ensemble", 
-                            "individual particpant"))
-
-plot_participant_forecasts <- plot_df |>
-  ggplot(aes(x = target_end_date)) +
-  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
-                color = "Individual participant"), 
-            linewidth = .1) + 
-  geom_line(data = filter(plot_df, model == "crowd-ensemble"), 
-            aes(y = prediction, group = interaction(forecast_date, model), 
-                color = "Crowd ensemble"),
-            linewidth = .4) + 
-  geom_line(aes(y = true_value),
-            linewidth = .4, color = "black") + 
-  geom_point(aes(y = true_value),
-             size = .5, color = "black") + 
-  scale_color_manual(values = c(colors[1], "grey70")) + 
-  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
-  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
-  theme_scoringutils()
-
-ggsave("output/figures/participant-forecasts.png", plot = plot_participant_forecasts, 
-       width = 7, height = 5)
-```
-
-Median forecasts combined human ensemble and Forecast Hub ensemble 1:4 horizon
-```{r}
-forecasts |>
-  filter(quantile == 0.5) |>
-  rbind(horizon_0, fill = TRUE) |>
-  filter(target_end_date >= "2021-04-01", 
-         target_end_date <= "2021-10-01") |>
-  filter((model %in% c("crowd-ensemble", "EuroCOVIDhub-ensemble"))) |>
-  ggplot(aes(x = target_end_date)) +
-  geom_line(aes(y = prediction, group = interaction(forecast_date, model), 
-                color = model), 
-            linewidth = .4) + 
-  geom_line(aes(y = true_value),
-            linewidth = .4, color = "black") + 
-  geom_point(aes(y = true_value),
-             size = .5, color = "black") + 
-  scale_color_manual(values = colors[1:2]) +
-  facet_wrap(~ target_type, scale = "free_y", ncol = 1) + 
-  labs(x = "Date", y = "Median forecast and observed values", color = "Forecaster") + 
-  theme_scoringutils()
-```
-
-## Violin plot to compare scores
-
-```{r}
-comparison_plot <- function(scores, models = c("crowd-direct", "EuroCOVIDhub-ensemble")) {
-  mean_wis <- scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models) |>
-  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale")) |>
-  summarise_scores(by = c("model", "scale", "target_type", "type_and_scale"),
-                   fun = signif, digits = 3) |>
-  mutate(wis_mean = paste("Mean:", label_fn(interval_score)))
-
-scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models) |>
-  arrange(scale, target_type, model) |>
-  ggplot(aes(y = interval_score, x = model)) + 
-  geom_violin(aes(fill = model), color = NA,
-         alpha = 0.2) + 
-  geom_boxplot(alpha = 0.5) + 
-  scale_fill_brewer(palette = "Set1", name = "Forecast target") +
-
-  geom_text(data = mean_wis, 
-            aes(label = wis_mean, y = -Inf), 
-            vjust   = -1) + 
-  facet_wrap(~ type_and_scale, scale = "free") + 
-  scale_y_continuous(labels = label_fn, 
-                     expand = expansion(mult = c(0.15, 0.05))) + 
-  labs(y = "Weighted interval score", x = "Forecaster") + 
-  theme_scoringutils() + 
-  theme(legend.position = "None")
-}
-
-comparison_plot(scores, 
-                models = c("crowd-direct", "EuroCOVIDhub-ensemble"))
-
-comparison_plot(scores |>
-                  filter(horizon == 2), 
-                models = c("crowd-direct", "crowd-rt", "crowd-ensemble"))
-
-```

### 2023-03-27 old-unused-analyses.R

@@ -1,0 +1,207 @@
+
+# Analsyses on revised data
+# ============================================================================ #
+
+scores_forecasts_comparison_revised <- score_and_add_log(forecasts_comparison_revised)
+
+p_comparison_direct_rt_individual_reivsed <- plot_rt_direct_individual(scores_forecasts_comparison_revised)
+
+ggsave("output/figures/comparison-direct-rt-individual-revised.png",
+       plot = p_comparison_direct_rt_individual,
+       width = 7, height = 5)
+
+# performance plot for an ensemble of the subset of human and direct forecasts
+# using revised death data
+
+forecasts_comparable_ensemble_revised <- forecasts_comparison_revised |>
+  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+  summarise(prediction = median(prediction)) |>
+  mutate(model = str_to_title(method)) |>
+  as.data.table()
+
+scores_comparable_ensemble_revised <- score_and_add_log(forecasts_comparable_ensemble_revised)
+
+p_performance_comparison_revised <- plot_performance(
+  scores_comparable_ensemble_revised
+)
+
+ggsave("output/figures/performance-rt-comparison-revised.png", plot = p_performance_comparison,
+       width = 7, height = 8)
+
+
+
+
+
+  
+# Analysis on a comparable ensemble of Rt and direct forecasts
+# ============================================================================ #
+
+# performance plot for an ensemble of the subset of human and direct forecasts
+forecasts_comparable_ensemble <- forecasts_comparison |>
+  group_by(target_type, horizon, method, quantile, true_value, forecast_date, target_end_date) |>
+  summarise(prediction = median(prediction)) |>
+  mutate(model = str_to_title(method)) |>
+  as.data.table()
+
+scores_comparable_ensemble <- score_and_add_log(forecasts_comparable_ensemble)
+
+p_performance_comparison <- plot_performance(
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/performance-rt-comparison.png", plot = p_performance_comparison, 
+       width = 7, height = 8)
+
+p_scores_and_pred_comparison <- plot_scores_and_pred(
+  forecasts_comparable_ensemble,
+  scores_comparable_ensemble
+)
+
+ggsave("output/figures/scores-and-forecasts-rt-comparison.png", 
[…]

### 2023-03-27 manuscript/manuscript.tex

@@
 Latent infections were convolved with delay distributions representing the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
-This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022} [CITE ECDC] [CITE SPI-M paper] and used to inform policy makers via reproduction number estimates [CITE one of the SPI-M reproduction number summary papers or kaths paper on multiple surveillance Rt esimates].
+This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022} [CITE ECDC] and used to inform policy makers via reproduction number estimates [CITE one of the SPI-M reproduction number summary papers or kaths paper on multiple surveillance Rt esimates].
 Further details are given in the SI.
@@
 Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
-However, due to computational complexity and lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
+However, due to lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
 Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
@@
 In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
-To make results easier to interpret, we also report normalised scores.
+%To make results easier to interpret, we also report normalised scores.
 We normalised scores for a given target (cases or deaths) and scale (natural or log) by calculating a Z-score, i.e.
@@
 
-A comparison of "experts" and "non-experts" (self-reported experience in infectious disease modelling or a related field) showed broadly comparable performance, both on the natural and the log scale (see Figure \ref{fig:performance-experts}).
-Precise results were susceptible to whether or not $R_t$ forecasts were included, with performance of "experts" on deaths being slightly worse than performance of "non-experts" when only looking at direct forecasts.
+A comparison of "experts" and "non-experts" (self-reported experience in infectious disease modelling or a related field) showed similar performance, both on the natural (WIS two weeks ahead: 43k for experts vs.
+43k for non-experts for cases, 41 vs.
+46 for deaths) and the log scale (0.24 vs.
+0.26 for cases, 0.16 vs.
+0.13 for deaths, see Figure \ref{fig:performance-experts}) and Table \ref{tab:scores-expert}.
+Precise results changed depending on whether or not direct forecasts, $R_t$ forecasts or both were scored.
 
@@
 In our study, $R_t$ forecasts of cases were comparable to direct forecasts.
-However, given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality just forecasting cases and merely updated their $R_t$ forecast so they would obtain the desired case forecast.
+However, given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality directly forecasting cases.
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
@@
 One possible interpretation is that a mixed crowd of human forecaster can produce a viable alternative or addition to an ensemble of mathematical models created by experts.
-Another interpretation is that an ensemble of fairly automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
+Another interpretation is that an ensemble of automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
 All previous studies comparing human judgement forecasts and computational models only ran over short periods of time and the majority of them struggled with recruitment and upkeep.
-Meanwhile, COVID-19 Forecast Hubs have attracted continuous submissions for almost three years and were able to consistently provide high-quality forecasts.
+Meanwhile, COVID-19 Forecast Hubs have attracted continuous submissions for almost three years and were able to consistently provide forecasts of comparable quality.
 
-Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both ways is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
+Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
 Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology in order to maximise utility, predictive performance, or both.
@@
 
+\input{manuscript/performance-table-experts}
+
+\input{manuscript/performance-table-experts-4}
+
 \begin{figure*}

### 2023-03-27 Analysis.rmd
change order of models in plots and table

@@
 models_human <- models_all[!(models_all %in% models_non_human)]
+models_non_human <- factor(models_non_human, 
+                           levels = c("EuroCOVIDhub-ensemble", 
+                                      "crowd-ensemble", 
+                                      "crowd-direct", 
+                                      "crowd-rt"))
 models_rt <- models_human[grepl("(Rt)", models_human)]
@@
     pivot_wider(names_from = scale, values_from = c(interval_score, wis_sd, relative)) |>
+    mutate(model = factor(model, labels = levels(models_non_human))) |>
     arrange(target_type, model) |>
-    select(model, target_type, interval_score_natural, relative_natural, wis_sd_natural, interval_score_log, relative_log, wis_sd_log, coverage_50, coverage_90) |>
+    select(model, target_type, 
+           interval_score_natural, relative_natural, wis_sd_natural, 
+           interval_score_log, relative_log, wis_sd_log, 
+           coverage_50, coverage_90) |>
     kable(format = "latex", booktabs = TRUE, 
@@
                                     "hold_position")) |>
-    add_header_above(c(" " = 2, "WIS" = 3, "WIS - log scale" = 3, " " = 2)) 
+    add_header_above(c(" " = 2, "WIS - natural" = 3, "WIS - log scale" = 3, " " = 2)) 
 }
@@
 
-p_performance <- plot_performance(
-  filter(scores, model %in% models_non_human))
+p_performance <- scores |>
+  filter(model %in% models_non_human) |>
+  mutate(model = factor(model, labels = levels(models_non_human))) |>
+  plot_performance()
 
@@
 p_scores_and_pred <- plot_scores_and_pred(
-  filter(forecasts, model %in% models_non_human),
-  filter(scores, model %in% models_non_human)
+  filter(forecasts, model %in% models_non_human) |>
+    mutate(model = factor(model, labels = levels(models_non_human))),
+  filter(scores, model %in% models_non_human) |>
+    mutate(model = factor(model, labels = levels(models_non_human)))
 )

### 2023-03-27 manuscript/manuscript.tex

@@
 Participants were therefore asked to estimate/predict six $R_t$ values, four of them beyond the forecast horizon.
+In order to obtain sample trajectories needed as input for the simulation model, we drew 1000 samples from the six provided distributions.
+These samples were ordered and corresponding samples treated as one sample trajectory.
+Samples for daily values were obtained by linearly interpolating between weekly samples.
+
 Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
@@
 The first prize was 100 GBP, second prize 50 GBP and third prize 25 GBP.
-Participant performance was determined using the mean weighted interval score (WIS) on the log scale (see details in the next Section, %\ref{sec:analysis}),
+Participant performance was determined using the mean weighted interval score (WIS) on the log scale (see details in the next Section), %\ref{sec:analysis}),
 averaged across forecast dates, horizons and forecast targets.
@@
 
-Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022} except for our extension to also consider the natural log of forecasts as a proxy for the exponential growth rate \cite{bosseTransformationForecastsEvaluating2023}.
+Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}. We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
+\cite{bracherEvaluatingEpidemicForecasts2021}.
+The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
+A forecaster, in expectation, optimises their score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report their true belief.
+The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
+The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
+uncertainty of forecasts), overprediction and underprediction.
+
+\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
+We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
+In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
+To make scores easier to interpret, we report relative WIS scores, where the average score for a given model was divided by the average score for the European Forecast Hub ensemble.
+
+In addition to the WIS, we used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
+Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
+the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
+
 If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022a, bracherShorttermForecastingCOVID192021}.
@@
 % Those forecasts were combined using a quantile-wise median separately for direct and $R_t$ forecasts such that we obtained a modified direct human ensemble and a modified $R_t$ ensemble that included only corresponding forecasts from the same forecaster.
-
-We scored forecasts using the weighted interval score (WIS, formulas given in the SI) %in Section \ref{sec:wis}) 
-\cite{bracherEvaluatingEpidemicForecasts2021} and used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
-The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
-A forecaster, in expectation, optimises their score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report their true belief.
-The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
-The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
-uncertainty of forecasts), overprediction and underprediction.
-Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
-the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
-
-\citet{bosseTransformationForecastsEvaluating2023} suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
-We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
-In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
-%To make results easier to interpret, we also report normalised scores.
-We normalised scores for a given target (cases or deaths) and scale (natural or log) by calculating a Z-score, i.e.
-by subtracting the mean of all scores for the same target and scale, and dividing by the standard deviation.
 

### 2023-03-28 Analysis.rmd
update tables

@@
     pivot_wider(names_from = scale, values_from = c(interval_score, wis_sd, relative)) |>
-    mutate(model = factor(model, labels = levels(models_non_human))) |>
+    mutate(model = factor(model, levels = levels(models_non_human))) |>
     arrange(target_type, model) |>

### 2023-03-28 Analysis.rmd
update table formatting

@@
     mutate(relative = interval_score / interval_score[model == "EuroCOVIDhub-ensemble"]) |>
-    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
+    mutate(across(c(where(is.numeric)), \(x) round(signif(x, digits = 3), digits = 2))) |>
     mutate(across(c(where(is.numeric)), \(x) label_fn(x))) |> 
@@
   make_performance_table(
-    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
+    cap = "Performance for two-week-ahead forecasts. Vallues have been cut to three significant digits and rounded.  \\label{tab:scores}"
     ) %>%
@@
     hor = 4, 
-    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
+    cap = "Performance for four-week-ahead forecasts. Vallues have been cut to three significant digits and rounded \\label{tab:scores-4}"
   ) %>%

### 2023-03-28 Analysis.rmd
Added plots with model ranks

@@
   filter(model %in% models_non_human) |>
-  mutate(model = factor(model, labels = levels(models_non_human))) |>
+  mutate(model = factor(model, levels = levels(models_non_human))) |>
   plot_performance()
@@
   filter(forecasts, model %in% models_non_human) |>
-    mutate(model = factor(model, labels = levels(models_non_human))),
+    mutate(model = factor(model, levels = levels(models_non_human))),
   filter(scores, model %in% models_non_human) |>
-    mutate(model = factor(model, labels = levels(models_non_human)))
+    mutate(model = factor(model, levels = levels(models_non_human)))
 )
@@
 
+# Ranks
+
+```{r}
+
+plot_ranks <- function(scores, hor = 2) {
+  scores |>
+    filter(model %in% models_non_human) |>
+    mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+    summarise_scores() |>
+    select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
+    group_by(target_type, scale, horizon, forecast_date) |>
+    mutate(rank = rank(interval_score)) |>
+    filter(horizon == hor) |>
+    ggplot(aes(y = model, group = rank, fill = rank)) +
+    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
+    facet_wrap(~ type_and_scale) + 
+    scale_fill_continuous(trans='reverse') + 
+    theme_scoringutils() + 
+    labs(y = "", x = "Frequency", fill = "Rank") + 
+    theme(legend.position = "right")
+}
+
+p_ranks <- plot_ranks(scores)
+ggsave("output/figures/performance-ranks.png", 
+       plot = p_ranks, 
+       width = 7, height = 3.4)
+
+p_ranks_4 <- plot_ranks(scores, hor = 4)
+ggsave("output/figures/performance-ranks-4.png", 
+       plot = p_ranks_4, 
+       width = 7, height = 3.4)
+
+```
+
 # Unused - Significance analysis

### 2023-03-28 manuscript/manuscript.tex
Updates from Overleaf

@@
 
+\input performance-table-horizon-4.tex
+
 \begin{figure*}[ht]
@@
 
-Mean scores for the four forecasting approaches were close to each other, with a mean WIS for two-week-ahead-forecasts of incident cases on the natural scale (log scale) of 38k (0.25) for the Hub-ensemble, 39k (0.24) for the direct crowd forecasts, 40k (0.22) for the combined crowd forecast and 46k (0.23) for the $R_t$ forecasts (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
+Mean scores for the four forecasting approaches were close to each other, with a mean WIS relative to the Hub ensemble (=1) of
+
+for two-week-ahead-forecasts of incident cases on the natural scale (log scale) of 38k (0.25) for the Hub-ensemble, 39k (0.24) for the direct crowd forecasts, 40k (0.22) for the combined crowd forecast and 46k (0.23) for the $R_t$ forecasts (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
 Differences in mean scores between the four were small compared to the standard deviations of the scores within each forecasting approach on the natural (56k, 67k, 69k, 75k) and on the log scale (0.22, 0.27, 0.25, 0.24).
@@
 
-\input performance-table-horizon-4.tex
+%table 4 weeks
 

### 2023-03-28 Analysis.rmd
update tables for expert performance

--- 2023-03-28 Analysis.rmd
@@
   make_performance_table(
-    cap = "Performance for two-week-ahead forecasts. Vallues have been cut to three significant digits and rounded.  \\label{tab:scores}"
+    cap = "Performance for two-week-ahead forecasts. Values have been cut to three significant digits and rounded.  \\label{tab:scores}"
     ) %>%
@@
     hor = 4, 
-    cap = "Performance for four-week-ahead forecasts. Vallues have been cut to three significant digits and rounded \\label{tab:scores-4}"
+    cap = "Performance for four-week-ahead forecasts. Values have been cut to three significant digits and rounded \\label{tab:scores-4}"
   ) %>%
@@
 expert_ensemble$scores |>
+  rbind(filter(scores, model == "EuroCOVIDhub-ensemble"), fill = TRUE) |>
    make_performance_table(
-     scores_summarised = expert_scores_summarised,
-    cap = "Performance for two-week-ahead forecasts of experts and non-experts. Standard deviations are given in brackets. \\label{tab:scores-experts}"
+     model_filter = c("Expert", "Non-Expert", "EuroCOVIDhub-ensemble"),
+    cap = "Performance for two-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts}"
     ) %>%
@@
 expert_ensemble$scores |>
+  rbind(filter(scores, model == "EuroCOVIDhub-ensemble"), fill = TRUE) |>
    make_performance_table(
-     scores_summarised = expert_scores_summarised,
-     hor = 4,
-     cap = "Performance for four-week-ahead forecasts of experts and non-experts. Standard deviations are given in brackets. \\label{tab:scores-experts-4}"
+     model_filter = c("Expert", "Non-Expert", "EuroCOVIDhub-ensemble"),
+     cap = "Performance for four-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts-4}"
     ) %>%
@@
        width = 7, height = 6)
+
+plot_ranks(expert_ensemble$scores)
+
+```
+
+# Ranks
+
+```{r}
+
+plot_ranks <- function(scores, hor = 2) {
+  scores |>
+    summarise_scores() |>
+    select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
+    group_by(target_type, scale, horizon, forecast_date) |>
+    mutate(rank = rank(interval_score)) |>
+    filter(horizon == hor) |>
+    ggplot(aes(y = model, group = rank, fill = rank)) +
+    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
+    facet_wrap(~ type_and_scale) + 
+    scale_fill_continuous(trans='reverse') + 
+    theme_scoringutils() + 
+    labs(y = "", x = "Frequency", fill = "Rank") + 
+    theme(legend.position = "right")
+}
+
+p_ranks <- scores |>
+  filter(model %in% models_non_human) |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks()
+ggsave("output/figures/performance-ranks.png", 
+       plot = p_ranks, 
+       width = 7, height = 3.4)
+
+p_ranks_4 <- scores |>
+  filter(model %in% models_non_human) |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks(hor = 4)
+ggsave("output/figures/performance-ranks-4.png", 
+       plot = p_ranks_4, 
+       width = 7, height = 3.4)
 

### 2023-03-28 manuscript/manuscript.tex

--- 2023-03-27 manuscript/manuscript.tex
@@
 
-Mean scores for the four forecasting approaches were close to each other, with a mean WIS for two-week-ahead-forecasts of incident cases on the natural scale (log scale) of 38k (0.25) for the Hub-ensemble, 39k (0.24) for the direct crowd forecasts, 40k (0.22) for the combined crowd forecast and 46k (0.23) for the $R_t$ forecasts (see Table \ref{tab:scores} and Figure \ref{fig:performance}).
-Differences in mean scores between the four were small compared to the standard deviations of the scores within each forecasting approach on the natural (56k, 67k, 69k, 75k) and on the log scale (0.22, 0.27, 0.25, 0.24).
-% Could add a sentence that scores were not significant using a paired wilcoxon mann whitney test. 
-On the natural scale, the Forecast Hub ensemble was ahead of the crowd forecasts (WIS two weeks ahead: 38k for the Hub ensemble and 40k for the combined human ensemble), with the difference gradually increasing with increasing forecast horizon (see Figure \ref{fig:performance}).
-On the log scale, the Forecast Hub ensemble was slightly worse than all crowd forecasting approaches at a two week forecast horizon, but again slightly ahead for forecasts four weeks into the future.
+All forecasting approaches exhibited underdispersion when predicting cases, meaning that forecasts on average were too narrow and not uncertain enough.
+Empirical coverage was below nominal coverage for all forecasts and forecast horizons.
+For 50\% prediction intervals, empirical coverage was worst for the direct crowd forecasts (0.31), best for the $R_t$ forecasts (0.46) and in between for the Hub ensemble and the crowd ensemble (both 0.38, see Table \ref{tab:scores}).
+For 90\% prediction intervals, coverage was worst for the $R_t$ forecasts (0.62) and slightly better for the other approaches (all 0.69).
+Coverage for all forecasts deteriorated further with increasing forecast horizon (see Figure \ref{fig:performance}).
+%This underdisperion for case forecasts is in line with what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion,
 
-On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increases or decreases with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
+In terms of WIS on the log scale, all human forecasting approaches outperformed the Forecast Hub ensemble for two week ahead forecasts of cases (see Figure \ref{fig:performance}).
+WIS values relative to the Hub ensemble (=1) were 0.91 for the combined crowd ensemble, 0.96 for the direct crowd forecasts and 0.93 for the $R_t$ forecasts (see Table \ref{tab:scores}).
+In terms of WIS on the natural scale, however, the Hub ensemble was ahead of all human forecasting approaches.
+Relative WIS values on the natural scale for two week ahead forecasts were 1.05 for the combined crowd ensemble, 1.03 for the direct crowd forecasts and 1.2 for the $R_t$ forecasts.
+
+Performance of the Hub ensemble relative to the human forecasting approaches improved with increasing forecast horizon (see Figure \ref{fig:performance}).
+For a four-week-ahead forecast horizon, the Hub ensemble outperformed all other approaches both on the log scale (rel.
+WIS values the human forecasts of 1.02, 1.05, 1.06) and on the natural scale (rel.
+WIS values of 1.21, 1.25, 1.3) (see Table \ref{tab:scores-4}).
+
+In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ show a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}).
+Both the Hub ensemble and the $R_t$ forecast are most often in the first place, but also most often in the last place.
+The direct forecasts places relatively equally in places 1-4.
+The crowd ensemble never places fourth, but also has the lowest number of first places.
+Model ranks only change marginally when switching between the log and the natural scale.
+
+When comparing WIS values on the log scale with those on the natural scale, scores were more equally distributed across the study period on the log scale and more weight was given to forecasts in June and July which underpredicted the extent to which case number would rise (see Figure \ref{fig:forecasts-scores}).
+On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increased or decreased with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
 Average scores were therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
-$R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, and therefore showed the poorest performance in terms of WIS on the natural scale.
-When looking at WIS values on the log scale,
-%(that is, when log-transforming forecasts and observations prior to applying the WIS), 
-scores were more equally distributed across the study period and more weight is given to forecasts in June and July which underpredicted the extent to which case number would rise.
-Similarly to what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion, meaning that forecasts on average were too narrow and not uncertain enough, with empirical coverage below nominal coverage.
-Underdispersion was most severe for the $R_t$ forecasts.
+Switching between scores on the log and on the natural scale had the strongest effect on the $R_t$ forecasts, which had a relative WIS value of 0.96 on the log scale and 1.2 on the natural scale.
+The $R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak and were most severely penalised for overprediction (see Figure \ref{fig:performance}A).
+On the log scale, underprediction played a larger role, favouring the $R_t$ forecasts (see Figure \ref{fig:performance}C).
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks.png}
+\caption{\bf{Performance ranks}....}
+\label{fig:performance-ranks}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks-4.png}
+\caption{\bf{Performance ranks}....}
+\label{fig:performance-ranks-4}
+\end{figure*}
 
@@
 
-Overall, performance for two-week-ahead-forecasts of incident deaths was comparable for the Hub-ensemble, the direct crowd forecasts and the combined forecasts, while performance for the $R_t$ forecast was significantly worse.
-On the natural scale (log scale), mean WIS values were 38 (0.22) for the Hub ensemble, 34 (0.27) for the direct forecasts, 40 (0.22) for the combined crowd forecast, and 80 (0.25) for the $R_t$ forecasts (Table \ref{tab:scores}).
-With the exception of the $R_t$ crowd forecasts, differences between forecasting approaches were again small compared to 
-standard deviations of
-the scores within each forecasting approach on the natural scale (27, 31, 42, 73) and on the log scale (0.036, 0.081, 0.073, 0.13). 
-%On the log scale, direct human forecasts, the combined crowd ensemble and the Forecast Hub ensemble showed comparable performance, while $R_t$ forecasts performed again significantly worse.
+All forecasting approaches except the $R_t$ forecasts showed higher empirical coverage for deaths than for cases (see Figure \ref{fig:performance}).
+For 50\% prediction intervals, the Hub ensemble exceeded the nominal coverage noticeably (0.77) (see Table \ref{tab:scores}).
+$R_t$ forecasts failed to get close to nominal coverage (0.15), while the combined crowd ensemble and the direct forecasts had empirical coverage close to nominal coverage (both 0.54).
+For 90\% prediction intervals, the Hub ensemble again exceeded nominal coverage and covered all observations (1) while the $R_t$ forecasts again failed to get close to nominal coverage (0.46).
+The crowd ensemble exhibited a some underdispersion (0.77) while the direct forcecasts almost reached nominal coverage for two week ahead forecasts of deaths (0.85).
 
-On the natural scale, direct forecasts had a better WIS than the Hub ensemble and the other two human forecast ensembles for a two-week-ahead horizon (direct: 34, Hub ensemble: 38, combined crowd: 40, $R_t$: 80), while the Hub ensemble was again slightly ahead for a four week-ahead horizon (direct: 88, Hub ensemble: 85, combined crowd: 98, $R_t$: 150, see Figure \ref{fig:performance} and Table \ref{tab:scores-4}).
-The combined crowd forecast (containing both the direct and the $R_t$ forecasts) overall performed worse than the direct crowd forecast on the natural scale, but much better than the $R_t$ forecasts.
-On the log scale, the combined crowd forecasts even performed best among the four approaches.
+In terms of WIS on the log scale for two week ahead predictions of deaths, the combined crowd ensemble (0.07) and the direct crowd forecasts (0.99) were slightly ahead of the Hub ensemble, while the $R_t$ forecasts performed noticeably worse (1.98) (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
+Interestingly, combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them.
+In terms of WIS on the natural scale, only the direct forecasts (0.89) performed better for two week ahead death predictions than the Hub ensemble, while the combined crowd ensemble performed a bit worse (1.06) and the $R_t$ forecasts again noticeably worse (2.1).
 
-From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
-On the natural scale, the highest scores were concentrated towards the end of the study period, where death incidences were highest, while on the log scale, scores were more evenly distributed across the study period.
-$R_t$ forecasts show strong underdispersion with empirical coverage of the 50\% prediction intervals being close to zero and smaller than 50\% for the 90\% prediction intervals (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-The Forecast Hub ensemble was overdispersed with empirical coverage exceeding nominal coverage, and for the direct human forecasts, empirical coverage was close to nominal coverage.
+% Maybe say something about increasing forecast horizons. But the pattern is unclear.
+
+In terms of relative model ranks for two week ahead death forecasts, the $R_t$ forecasts took the fourth place most often, while the direct forecasts placed first most often (see Figure \ref{fig:performance-ranks)}.
+Again, the crowd ensemble never placed fourth.
+
+When comparing scores on the log and on the natural scale, scores on the log scale are again more evenly distributed across the study period.
+On the natural scale, high scores were concentrated around the end of the study period, were death incidences were highest (see Figure \ref{fig:forecasts-scores}).
 
@@
 For cases, where participants could observe the case forecast implied by their $R_t$ forecast, predictive performance was similar between corresponding direct and $R_t$ forecasts for most forecasters who had submitted both (see Figure \ref{fig:comparison-direct-rt-individual}).
-For deaths, where forecasters could not see the incidence forecast implied by their $R_t$ forecast or manually adjust the case fatality rate, performance of the $R_t$ forecasts was significantly worse than the corresponding direct forecasts for most forecasters.
-This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings \citep{meakinComparativeAssessmentMethods2022}.
+For deaths, where forecasters could not see the incidence forecast implied by their $R_t$ forecast or manually adjust the case fatality rate, performance of the $R_t$ forecasts was significantly worse.
+From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
+In particular, $R_t$ forecasts for deaths were worse than the corresponding direct death forecasts for most forecasters (see Figure \ref{fig:comparison-direct-rt-individual}).
+Figure \ref{fig:comparison-direct-rt-individual} suggests that changing from the direct forecasting method to $R_t$ forecasting for cases may have improved scores for better forecasters and decreased scores for worse forecasters, although sample sizes are too small to draw definitive conclusions.
+
+%This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings \citep{meakinComparativeAssessmentMethods2022}.
 This may be related to changes in the age decomposition of cases and the change in the case fatality ratio (CFR) over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK.
@@
 The ensemble that combined $R_t$ and direct forecasts performed only slightly worse than the direct forecasts on the natural scale and was the best performing model on the log scale, in spite of the poor performance of the $R_t$ forecasts.
+
+%This underdisperion for case forecasts is in line with what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion,
 

### 2023-03-28 Analysis.rmd
update expert tables

@@
     model_filter = models_non_human,
+    baseline = "EuroCOVIDhub-ensemble",
     cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
@@
            horizon %in% hor)
+
+  n_linesep <- length(unique(df$model))
 
@@
     group_by(target_type, scale) |>
-    mutate(relative = interval_score / interval_score[model == "EuroCOVIDhub-ensemble"]) |>
+    mutate(relative = interval_score / interval_score[model == baseline]) |>
     mutate(across(c(where(is.numeric)), \(x) round(signif(x, digits = 3), digits = 2))) |>
@@
     pivot_wider(names_from = scale, values_from = c(interval_score, wis_sd, relative)) |>
-    mutate(model = factor(model, levels = levels(models_non_human))) |>
     arrange(target_type, model) |>
@@
           caption = cap,
-          linesep = c('', '', '', '\\addlinespace'),
+          linesep = c(rep('', (n_linesep - 1)), '\\addlinespace'),
           col.names = c("Model", "Target", "abs.", "rel.", "sd", "abs.", "rel.", "sd", 
@@
   filter(model %in% models_non_human) |>
+  mutate(model = factor(model, levels = levels(models_non_human))) |>
   make_performance_table(
@@
 expert_ensemble$scores |>
-  rbind(filter(scores, model == "EuroCOVIDhub-ensemble"), fill = TRUE) |>
+  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
+  mutate(model = factor(model, levels = c("crowd-ensemble", "Expert", "Non-Expert"))) |>
    make_performance_table(
-     model_filter = c("Expert", "Non-Expert", "EuroCOVIDhub-ensemble"),
+     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
+     baseline = "crowd-ensemble",
     cap = "Performance for two-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts}"

### 2023-03-28 Analysis.rmd
update expert table

@@
 
-expert_scores_summarised <- expert_ensemble$scores |>
-  filter(scale %in% c("log", "natural")) |>
-  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
-  group_by(target_type, scale, horizon) |>
-  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
-  group_by(model, target_type, scale, horizon) |>
-  summarise(mean = mean(interval_score), 
-            sd = sd(interval_score)) |>
-  pivot_wider(id_cols = c(model, target_type, horizon), 
-              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
-  arrange(target_type, model)
-
 expert_ensemble$scores |>
@@
 expert_ensemble$scores |>
-  rbind(filter(scores, model == "EuroCOVIDhub-ensemble"), fill = TRUE) |>
+  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
    make_performance_table(
-     model_filter = c("Expert", "Non-Expert", "EuroCOVIDhub-ensemble"),
+     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
+     baseline = "crowd-ensemble",
+     hor = 4,
      cap = "Performance for four-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts-4}"

### 2023-03-28 manuscript/manuscript.tex
Updates from Overleaf

@@
 
+An median ensemble of two week ahead forecasts made only by "experts" or "non-experts" performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Table \ref{tab:scores-experts}).
+
+For cases two week ahead, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
+WIS values \textit{relative to the combined crowd ensemble} were 1.08 for "experts" and 1.14 for "non-experts" on the log scale and 1.03 for "experts" and 1.14 for "non-experts" on the natural scale.
+
 A comparison of "experts" and "non-experts" (self-reported experience in infectious disease modelling or a related field) showed similar performance, both on the natural (WIS two weeks ahead: 43k for experts vs.

### 2023-03-29 Analysis.rmd
Reorganise code that checks for the number of participants, add plot about expert participation

@@
 library(forcats)
+library(scales)
 
 colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")
+colors <- hex <- hue_pal()(4)
 
@@
   )
+}
+
+# helper function to score forecasts both on the natural and on the log scale
+score_and_add_log <- function(forecasts) {
+  forecasts |>
+  mutate(scale = "natural") |>
+  rbind(forecasts |>
+          mutate(
+            scale = "log", 
+            true_value = log(true_value + 1), 
+            prediction = log(pmax(prediction, 0) + 1)
+          )) |>
+  score(metrics = c("interval_score", "coverage")) |>
+  add_coverage(by = c("target_type", "horizon", "method")) |>
+    add_type_and_scale() |>
+  as.data.table() 
 }
@@
 ```{r}
-# helper function to score forecasts both on the natural and on the log scale
-score_and_add_log <- function(forecasts) {
-  forecasts |>
-  mutate(scale = "natural") |>
-  rbind(forecasts |>
-          mutate(
-            scale = "log", 
-            true_value = log(true_value + 1), 
-            prediction = log(pmax(prediction, 0) + 1)
-          )) |>
-  score(metrics = c("interval_score", "coverage")) |>
-  add_coverage(by = c("target_type", "horizon", "method")) |>
-    add_type_and_scale() |>
-  as.data.table() 
-}
-
 # helper function that filters the forecast data such that we only have 
@@
 
-## Number of forecasts and ensemble members
+## Number of forecasts 
+# this gives the total with people being double counted if they submit both
+# Rt and direct forecasts
 avail <- 
@@
        alpha = "Forecast method") + 
+  scale_fill_manual(values = colors[1:2]) +
   theme_scoringutils() +
@@
 
-avail |> 
-  group_by(target_type, type) |>
-  summarise(
-    mean = mean(`Number forecasts`), 
-    median = median(`Number forecasts`), 
-    min = min(`Number forecasts`), 
-    max = max(`Number forecasts`)
-  )
-
-## number of unique participants
+## overall number of unique participants
 unique_participants <- gsub("\ \\(Rt\\)", "", models_human) |> unique()
@@
 
-## Number of unique participants in every week
+## overall number of experts
 forecasts |>
   filter(model %in% models_human) |>
+  update_expert_status() |>
   mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
-  select(model, forecast_date) |>
-  group_by(forecast_date) |>
-  summarise(n = length(unique(model))) |>
-  summary()
+  select(model, expert) |>
+  unique() |>
+  group_by(expert) |>
+  summarise(n = length(unique(model)))
 
@@
   summary()
+
+# helper function
+sum_number <- function(data) {
+  data |> 
+    mutate(n = n()) |>
+    summarise(
+      mean = mean(`n`), 
+      median = median(`n`), 
+      min = min(`n`), 
+      max = max(`n`)
+    )
+}
+
+nums <- forecasts |>
+  update_expert_status() |>
+  filter(model %in% models_human) |>
+  mutate(method = ifelse(grepl("(Rt)", model), "Rt", "direct")) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  select(model, forecast_date, target_type, expert, method) |>
+  unique()
+
+# number of forecasts, (double counting participants who made a forecast for both Rt and direct)
+nums |>
+  group_by(target_type, forecast_date, method) |>
+  summarise(n = n()) |>
+  group_by(target_type, method) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  )
+
+# unqiue forecasts for cases and deaths without double counting
+nums |>
+  select(-method) |>
+  unique() |>
+  group_by(forecast_date, target_type) |>
+  summarise(n = n()) |>
+  group_by(target_type) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  )
+
+# unique participants every week (without double counting Rt and direct)
+nums |>
+  select(-method, -target_type) |>
+  unique() |>
+  group_by(forecast_date) |>
+  summarise(n = n()) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  )
+
+# number of forecasts from experts and non-experts every week (allowing double counting)
+nums |>
+  group_by(forecast_date, target_type, expert) |>
+  summarise(n = n()) |>
+  group_by(target_type, expert) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  )
+
+p_num_experts <- nums |>
+  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
+  group_by(target_type, forecast_date, expert, method) |>
+  summarise(n = n()) |>
+  pivot_wider(names_from = method, values_from = n) |>
+  mutate(Rt = Rt + direct) |>
+  pivot_longer(cols = c(Rt, direct), names_to = "method", values_to = "n") |>
+  ggplot(aes(x = forecast_date, y = n, 
+             group = expert,
+             fill = expert, 
+             alpha = method)) +
+  geom_bar(stat = "identity", 
+           color = "white",
+           linewidth = 0.2,
+           width = 5,
+           position = position_dodge(width = 5)) + 
+  facet_wrap(~ target_type) + 
+  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
+  labs(x = "Forecast date", fill = "Forecast type", 
+       alpha = "Forecast method") + 
+  theme_scoringutils() +
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin())
+
+ggsave("output/figures/num-forecasters-expert.png", plot = p_num_experts, 
+       width = 7, height = 3.5)
+
+# forecasts |> 
+#   update_expert_status() |>
+#   mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
+#   filter(forecast_date == "2021-06-07", 
+#          model %in% models_human, horizon == 1, 
+#          quantile == 0.5) |> 
+#   mutate(method = ifelse(grepl("(Rt)", model), "Rt", "direct")) |>
+#   mutate(model = gsub("\ \\(Rt\\)", "", model)) |> 
+#   group_by(target_type, method, expert) |> 
+#   summarise(n = length(unique(model)))
+
 ```

### 2023-03-29 manuscript/manuscript.tex

--- 2023-03-28 manuscript/manuscript.tex
@@
 The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}).
+The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths).
 
@@
 
-A comparison of "experts" and "non-experts" (self-reported experience in infectious disease modelling or a related field) showed similar performance, both on the natural (WIS two weeks ahead: 43k for experts vs.
-43k for non-experts for cases, 41 vs.
-46 for deaths) and the log scale (0.24 vs.
-0.26 for cases, 0.16 vs.
-0.13 for deaths, see Figure \ref{fig:performance-experts}) and Table \ref{tab:scores-expert}.
-Precise results changed depending on whether or not direct forecasts, $R_t$ forecasts or both were scored.
+%Mention calbration
 
-% As a direct comparison of individual scores between "experts" and "non-experts" is biased by missing forecasts,
+% Mention number of experts and non-experts
+
+An median ensemble of two week ahead forecasts made only by "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts}).
+
+For cases two week ahead, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
+WIS values \textit{relative to the combined crowd ensemble} were 1.08 for "experts" and 1.14 for "non-experts" on the log scale and 1.06 for "experts" and 1.07 for "non-experts" on the natural scale (see Table \ref{tab:scores-experts}).
+
+For two week ahead forecasts of deaths, "experts" performed worse than "non-experts" in terms of WIS on the log scale (WIS relative to the combined crowd ensemble: 1.29 vs.
+1.06), but better on the natural scale (1.03 vs.
+1.14).
+
+For four weeks ahead forecasts of cases, the combined ensemble outperformed both "experts" and "non-experts" on the log scale as well as on the natural scale.
+"Experts" performed better than "non-experts" both on the log scale (1.08 for "experts" vs.
+1.21 for "non-experts") and on the natural scale (1.04 vs.
+1.07).
+For four week ahead forecasts of deaths, "Experts" performed better than "Non-experts" on the log scale (1.17 vs.
+1.18) as well as on the natural scale (0.95 vs.
+1.15).
 
@@
 Nevertheless, retaining participants beyond the initial recruitment proved challenging, and most forecasters only submitted a single forecast.
-\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that provide access to a large existing user base may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
+\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that access to a large existing user base and greater resources may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
 

### 2023-03-30 Analysis.rmd
merge figures with number of forecasters with same figure for experts

@@
 ================================================================================
+- also includes Table SI.1
 
@@
 
-plot_number_forecasters <- avail |>
-  pivot_wider(names_from = type, values_from = `Number forecasts`) |>
-  mutate(Rt = Direct + Rt) |>
-  pivot_longer(cols = c(Direct, Rt), names_to = "type",
-               values_to = "Number forecasts") |>
-  as.data.table() |>
-  rbind(avail_ensemble, fill = TRUE) |>
-  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
-             group = group,
-             fill = group, 
-             alpha = type)) +
-  geom_bar(stat = "identity", 
-           color = "white",
-           linewidth = 0.2,
-           width = 5,
-           position = position_dodge(width = 5)) + 
-  facet_wrap(~ target_type) + 
-  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
-  labs(x = "Forecast date", fill = "Forecast type", 
-       alpha = "Forecast method") + 
-  scale_fill_manual(values = colors[1:2]) +
-  theme_scoringutils() +
-  theme(legend.position = "bottom", 
-        legend.box="vertical", legend.margin = margin())
-
-ggsave("output/figures/num-forecasters.png", plot = plot_number_forecasters, 
-       width = 7, height = 3.5)
-
 ## overall number of unique participants
@@
 
-p_num_experts <- nums |>
-  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
-  group_by(target_type, forecast_date, expert, method) |>
-  summarise(n = n()) |>
-  pivot_wider(names_from = method, values_from = n) |>
-  mutate(Rt = Rt + direct) |>
-  pivot_longer(cols = c(Rt, direct), names_to = "method", values_to = "n") |>
-  ggplot(aes(x = forecast_date, y = n, 
-             group = expert,
-             fill = expert, 
-             alpha = method)) +
+plot_number_forecasters <- avail |>
+  pivot_wider(names_from = type, values_from = `Number forecasts`) |>
+  mutate(Rt = Direct + Rt) |>
+  pivot_longer(cols = c(Direct, Rt), names_to = "type",
+               values_to = "Number forecasts") |>
+  as.data.table() |>
+  rbind(avail_ensemble, fill = TRUE) |>
+  ggplot(aes(x = forecast_date, y = `Number forecasts`, 
+             group = group,
+             fill = group, 
+             alpha = type)) +
   geom_bar(stat = "identity", 
@@
        alpha = "Forecast method") + 
+  scale_fill_manual(values = c(
+    "Expert" = "#984EA3", 
+    "Non-Expert" = "#377EB8", 
+    "Forecast Hub" = "#F8766D", 
+    "Human forecast" = "#7CAE00"), 
+    breaks = c("Expert", "Non-Expert", "Forecast Hub", "Human forecast")
+    ) +
   theme_scoringutils() +
@@
 
-ggsave("output/figures/num-forecasters-expert.png", plot = p_num_experts, 
-       width = 7, height = 3.5)
+p_num_experts <- nums |>
+  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
+  group_by(target_type, forecast_date, expert, method) |>
+  mutate(method = ifelse(method == "direct", "Direct", "Rt")) |>
+  summarise(n = n()) |>
+  pivot_wider(names_from = method, values_from = n) |>
+  mutate(Rt = Rt + Direct) |>
+  pivot_longer(cols = c(Rt, Direct), names_to = "method", values_to = "n") |>
+  ggplot(aes(x = forecast_date, y = n, 
+             group = expert,
+             fill = expert, 
+             alpha = method)) +
+  geom_bar(stat = "identity", 
+           color = "white",
+           linewidth = 0.2,
+           width = 5,
+           position = position_dodge(width = 5)
+           ) + 
+  facet_wrap(~ target_type) + 
+  scale_alpha_manual(values = c(1, 0.5), na.translate = FALSE) + 
+  scale_fill_manual(values = c(
+    "Expert" = "#984EA3", 
+    "Non-Expert" = "#377EB8", 
+    "Forecast Hub" = "#F8766D", 
+    "Human forecast" = "#7CAE00"), 
+    breaks = c("Expert", "Non-Expert", "Forecast Hub", "Human forecast")
+    ) +
+  labs(x = "Forecast date", fill = "Forecast type", 
+       alpha = "Forecast method") + 
+  theme_scoringutils() +
+  theme(legend.position = "bottom", 
+        legend.box="vertical", legend.margin = margin()) + 
+  labs(y = "Number forecasts")
+
+p_nums <- plot_number_forecasters /
+  p_num_experts +
+  plot_annotation(tag_levels = "A") &
+  theme(legend.position = "right")
+
+ggsave("output/figures/num-forecasters.png", plot = p_nums, 
+       width = 7, height = 5)
 
@@
 
-plot_ranks <- function(scores, hor = 2) {
-  scores |>
+ranks <- scores |>
+  filter(model %in% models_non_human) |>
   summarise_scores() |>
@@
   group_by(target_type, scale, horizon, forecast_date) |>
-    mutate(rank = rank(interval_score)) |>
+  mutate(rank = rank(interval_score))
+
+plot_ranks <- function(ranks, hor = 2) {
+  ranks |>
     filter(horizon == hor) |>
@@
 
-p_ranks <- scores |>
-  filter(model %in% models_non_human) |>
+p_ranks <- ranks |>
   mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
@@
 
-p_ranks_4 <- scores |>
-  filter(model %in% models_non_human) |>
+p_ranks_4 <- ranks |>
   mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
@@
        width = 7, height = 3.4)
+
+ranks |>
+  filter(horizon == 2) |>
+  mutate(model = factor(model, levels = (levels(models_non_human)))) |>
+  group_by(model, target_type, scale, rank) |>
+  summarise(n = n()) |>
+  mutate(rel = n() / 13) |>
+  pivot_wider(names_from = rank, values_from = c(n, rel)) |>
+  {\(.) {replace(.,is.na(.),0)}}() |>
+  mutate(n_1 = paste0(n_1, " (", round(rel_1, 2), ")"), 
+         n_2 = paste0(n_2, " (", round(rel_2, 2), ")"), 
+         n_3 = paste0(n_3, " (", round(rel_3, 2), ")"), 
+         n_4 = paste0(n_4, " (", round(rel_4, 2), ")")) |>
+  select(-c(rel_1, rel_2, rel_3, rel_4))
 

### 2023-03-30 manuscript/manuscript.tex
Updates from Overleaf

@@
 One explicit method to combine separate human judgement and computational model forecasts with the goal of improving predictive performance is an ensemble.
-This has been widely shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
+This has been shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
 \citet{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022, swallowChallengesEstimationUncertainty2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
@@
 The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} started in March 2021 to elicit predictions for various COVID-19 related forecast targets from different research groups every week.
-The forecasts evaluated in this study were submitted every Monday before 23.59pm GMT between May 24 2021 and August 16 2021.
+The forecasts evaluated in this study were submitted every Monday before 11.59pm GMT between May 24 2021 and August 16 2021.
 Forecasts were made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
@@
 From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
-The median number of models included in the Forecast Hub ensemble during the study period was nine for cases and ten for deaths (see Figure \ref{fig:num-forecasters} in the SI).
+The median number of models included in the Forecast Hub ensemble during the study period was 9 for cases and 10 for deaths (see Figure \ref{fig:num-forecasters} in the SI).
 
@@
 
-The web application offered participants two different ways of making a forecast, called 'direct' (or 'classical') and 'Rt' forecast.
+The web application offered participants two different ways of making a forecast, called 'direct' (or 'classical') and '$R_t$ forecast'.
 To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty at each forecast horizon.
@@
 
-In addition to the 'direct' forecasts, we implemented a second forecasting method ('Rt forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline estimate produced by the \texttt{EpiNow2} \textsf{R} package effective reproduction number model \citep{epinow2} which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
+In addition to the 'direct' forecasts, we implemented a second forecasting method ('$R_t$ forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline estimate produced by the \texttt{EpiNow2} \citep{epinow2} \textsf{R} \cite{R} package effective reproduction number model which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
 The estimate produced by \texttt{EpiNow2} was shown as the default forecast and could be adjusted by the user.
-The resulting $R_t$ forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}, which implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007} generative process for latent infections.
+The resulting $R_t$ forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package, which implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007} generative process for latent infections.
 We chose a Gaussian Process prior with mean 0 for the first differences of the effective reproduction number in time, implying that in the absence of informative data the reproduction number would remain constant on average, with uncertainty increasing with the temporal distance to informative data points.
 Latent infections were convolved with delay distributions representing the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
-This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022} [CITE ECDC] and used to inform policy makers via reproduction number estimates [CITE one of the SPI-M reproduction number summary papers or kaths paper on multiple surveillance Rt esimates].
+This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
 Further details are given in the SI.
@@
 
-Every week, we submitted an ensemble of individual forecasts ("crowd forecast") to the European Forecast Hub.
+Every week, we submitted an ensemble of individual forecasts to the European Forecast Hub.
 In contrast to the ensemble of human forecasts described in \citet{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions.
 %, following suggestions made in CITATION.
-We submitted three different ensembles to the Hub: The first one, "epiforecasts-EpiExpert\_direct" (here called "Crowd direct") was a quantile-wise median ensemble of all the direct forecasts.
-"epiforecasts-EpiExpert\_Rt" (here called "Crowd Rt" was a median ensemble of all forecasts made through the $R_t$ interface.
-"epiforecasts-EpiExpert" (here called "Crowd combined") was a median ensemble of all forecasts together.
-A participant could enter the "Crowd combined" ensemble twice if they had submitted both a direct and an $R_t$ forecast.
+We submitted three different ensembles to the Hub: The first one, "epiforecasts-EpiExpert\_direct" (here called "direct crowd forecast" or "crowd-direct") was a quantile-wise median ensemble of all the direct forecasts.
+"epiforecasts-EpiExpert\_Rt" (here called "$R_t$ forecast" or "crowd-rt") was a median ensemble of all forecasts made through the $R_t$ interface.
+"epiforecasts-EpiExpert" (here called "combined crowd ensemble" or "crowd-ensemble") was a median ensemble of all forecasts together.
+A participant could enter the combined crowd ensemble twice if they had submitted both a direct and an $R_t$ forecast.
 Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
@@
 To make scores easier to interpret, we report relative WIS scores, where the average score for a given model was divided by the average score for the European Forecast Hub ensemble.
+In addition, we computed ranks based on WIS values.
 
-In addition to the WIS, we used the empirical coverage of all central 50\% and 90\% prediction intervals to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}.
+In order to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}, we used the empirical coverage of all central 50\% and 90\% prediction intervals.
 Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
@@
 Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study.
-The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}).
-The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths).
+The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A).
+The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths) (see Figure \ref{fig:num-forecasters}B).
 
 \input performance-table.tex
-
-\input performance-table-horizon-4.tex
 
@@
 All forecasting approaches exhibited underdispersion when predicting cases, meaning that forecasts on average were too narrow and not uncertain enough.
-Empirical coverage was below nominal coverage for all forecasts and forecast horizons.
+Empirical coverage for case forecasts was below nominal coverage for all forecasting approaches for forecasts more than one week into the future (see Figure \ref{fig:performance}E,F).
 For 50\% prediction intervals, empirical coverage was worst for the direct crowd forecasts (0.31), best for the $R_t$ forecasts (0.46) and in between for the Hub ensemble and the crowd ensemble (both 0.38, see Table \ref{tab:scores}).
 For 90\% prediction intervals, coverage was worst for the $R_t$ forecasts (0.62) and slightly better for the other approaches (all 0.69).
-Coverage for all forecasts deteriorated further with increasing forecast horizon (see Figure \ref{fig:performance}).
-%This underdisperion for case forecasts is in line with what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion,
+Coverage for all forecasts deteriorated further with increasing forecast horizon (see Figure \ref{fig:performance}E,F).
 
@@
 
-In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ show a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}).
-Both the Hub ensemble and the $R_t$ forecast are most often in the first place, but also most often in the last place.
-The direct forecasts places relatively equally in places 1-4.
-The crowd ensemble never places fourth, but also has the lowest number of first places.
-Model ranks only change marginally when switching between the log and the natural scale.
+In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}).
+Both the Hub ensemble and the $R_t$ forecast were more often in first place than other approaches (4 times each, both on the log and on the natural scale).
+However, they were also most often in the last place (Hub ensemble: 6 on the log scale and 5 on the natural scale, $R_t$: 5 on the log scale and 6 on the natural scale).
+The direct forecasts placed relatively equally in places 1-4.
+The crowd ensemble never placed fourth, but also had the lowest number of first places (2, both on the log and the natural scale).
+Model ranks only changed marginally when switching between the log and the natural scale.
 
@@
 Average scores were therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
+For all forecasting approaches, overprediction was the largest contributor to overall scores (see Figure \ref{fig:performance}A).
+On the log scale, underprediction played a larger role (see Figure \ref{fig:performance}C).
 Switching between scores on the log and on the natural scale had the strongest effect on the $R_t$ forecasts, which had a relative WIS value of 0.96 on the log scale and 1.2 on the natural scale.
-The $R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak and were most severely penalised for overprediction (see Figure \ref{fig:performance}A).
-On the log scale, underprediction played a larger role, favouring the $R_t$ forecasts (see Figure \ref{fig:performance}C).
+The $R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, leading to high scores on the natural scale, but not on the log scale.
 
@@
 For 90\% prediction intervals, the Hub ensemble again exceeded nominal coverage and covered all observations (1) while the $R_t$ forecasts again failed to get close to nominal coverage (0.46).
-The crowd ensemble exhibited a some underdispersion (0.77) while the direct forcecasts almost reached nominal coverage for two week ahead forecasts of deaths (0.85).
+The crowd ensemble exhibited some underdispersion (0.77) while the direct forcecasts almost reached nominal coverage for two week ahead forecasts of deaths (0.85).
 
@@
 
-In terms of relative model ranks for two week ahead death forecasts, the $R_t$ forecasts took the fourth place most often, while the direct forecasts placed first most often (see Figure \ref{fig:performance-ranks)}.
+In terms of relative model ranks for two week ahead death forecasts, the $R_t$ forecasts took the fourth place most often (9 on the log scale and 10 on the natural scale), while the direct forecasts placed first most often (5 on the log scale and 6 on the natural scale, see Figure \ref{fig:performance-ranks}).
 Again, the crowd ensemble never placed fourth.
 
-When comparing scores on the log and on the natural scale, scores on the log scale are again more evenly distributed across the study period.
+When comparing scores on the log and on the natural scale, scores on the log scale were again more evenly distributed across the study period.
 On the natural scale, high scores were concentrated around the end of the study period, were death incidences were highest (see Figure \ref{fig:forecasts-scores}).
@@
 \end{figure*}
-
 For cases, where participants could observe the case forecast implied by their $R_t$ forecast, predictive performance was similar between corresponding direct and $R_t$ forecasts for most forecasters who had submitted both (see Figure \ref{fig:comparison-direct-rt-individual}).
@@
 In particular, $R_t$ forecasts for deaths were worse than the corresponding direct death forecasts for most forecasters (see Figure \ref{fig:comparison-direct-rt-individual}).
-Figure \ref{fig:comparison-direct-rt-individual} suggests that changing from the direct forecasting method to $R_t$ forecasting for cases may have improved scores for better forecasters and decreased scores for worse forecasters, although sample sizes are too small to draw definitive conclusions.
+Changing from the direct forecasting method to $R_t$ forecasting for cases tended to improve scores for better forecasters and decrease scores for worse forecasters, although sample sizes are very small.
+
+Combining direct crowd forecasts and $R_t$ forecasts improved performance on the log scale compared to both direct and $R_t$ forecasts alone across all horizons and target types.
+This was not the case on the natural scale, where direct forecasts performed better than the $R_t$ and the direct forecasts for both cases and deaths across most horizons.
+Only for case forecasts four weeks ahead on the natural scale was the combined ensemble better than the direct forecasts.
+However, even on the natural scale, performance of the combined ensemble was better than the average of the WIS of direct and $R_t$ forecasts.
 
@@
 
-An median ensemble of two week ahead forecasts made only by "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts}).
+A median ensemble of two week ahead forecasts made only by "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts}).
 
+When comparing "experts" and "non-experts", "experts" were better calibrated for forecasts two weeks ahead (see Figure \ref{fig:performance-experts}.
 For cases two week ahead, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
@@
 For four weeks ahead forecasts of cases, the combined ensemble outperformed both "experts" and "non-experts" on the log scale as well as on the natural scale.
-"Experts" performed better than "non-experts" both on the log scale (1.08 for "experts" vs.
+"Experts" performed better than "non-experts" both on the log scale (WIS value relative to the combined crowd ensemble of 1.08 for "experts" vs.
 1.21 for "non-experts") and on the natural scale (1.04 vs.
@@
 % \paragraph{Summary}
-
 In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
 Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model to generate simulated reported cases and deaths and an extension which estimated a fixed case fatality ratio and a log normal delay between cases and reports.
-For case numbers, performance of the human judgement forecasts was broadly comparable to the ensemble of all models submitted to the European Forecast Hub, with results changing depending on the forecast horizon and whether or not forecasts were transformed before scoring.
-The Hub ensemble tended to perform better on longer forecast horizons and when evaluated on the natural scale, while human judgment forecasts tended to perform better for shorter horizons and when evaluated on the log scale.
-%We found that human judgement forecasts of case numbers, regardless of how they were obtained, showed performance broadly comparable to the ensemble of all forecasts submitted to the European Forecast Hub.
-For deaths, performance of direct human forecasts was broadly comparable to that of the Forecast Hub ensemble (with results again changing depending on the forecast horizons and whether or not forecasts were transformed before scoring), while $R_t$ forecasts performed significantly worse.
-The ensemble that combined $R_t$ and direct forecasts performed only slightly worse than the direct forecasts on the natural scale and was the best performing model on the log scale, in spite of the poor performance of the $R_t$ forecasts.
-
-%This underdisperion for case forecasts is in line with what was observed previously for case forecasts \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, all forecasting approaches exhibited underdispersion,
-
-% \paragraph{Literature context}
-
-In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
-They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators.
-\citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
-Results in this study did not show the same pattern.
-We found that in terms of the WIS \textit{on the natural scale}, direct human forecasts were slightly worse than the Forecast Hub ensemble for cases, and slightly better for deaths (although results for greater forecast horizons differed, as did results for the $R_t$ forecasts).
-Looking at scores \textit{on the log scale}, our study suggests overall comparable performance of the Hub ensemble and the direct human forecasts on both cases and deaths.
-Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to the rise of the Delta variant in the UK starting in the beginning of May.
-The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
-One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters.
-Given the variability in scores across forecast horizons and evaluation methods, result should be interpreted with care.
 
 % \paragraph{Effort to recruit participants}. 
-
 Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to retain a large number of participants.
@@
 
-% \paragraph{Expertise}
+%Calibration
+In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, we found almost all forecasts for cases to be underdispersed (i.e.
+too narrow and overconfident).
+Empirical coverage for death forecasts was higher for all forecasting approaches except the $R_t$ forecasts.
 
-In our study, an ensemble of only self-reported "experts" did not outperform "non-experts", and even performed worse when looking only at direct forecasts of deaths (but comparably when including $R_t$ forecasts).
-Past expert forecasts of COVID-19 \cite{recchiaHowWellDid2021} had found predictions from experts to outperform those of non-experts.
-Our results should be taken with care in light of low sample sizes, and given that expert status was self-reported.
-Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
+%WIS scores
+For forecasts of cases two weeks ahead, performance of the human judgement forecasts was better than the European Forecast Hub ensemble in terms of WIS on the log scale, and worse in terms of WIS on the natural scale.
+For forecasts of deaths two weeks ahead, direct human forecasts and the combined crowd ensemble performed better than the Hub ensemble on the log scale.
+On the natural scale, the combined crowd ensemble performed worse than the Hub ensemble, while the direct crowd forecasts still performed better.
+$R_t$ forecasts for deaths performed noticeably worse than all other approaches both on the log and on the natural scale.
+
+In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
+They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators.
+\citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
+Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it.
+In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, on worse on the natural scale.
+Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths.
+Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to changes in the age distribution of cases and the rise of the Delta variant in the UK starting in the beginning of May.
+The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
+One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters.
 
@@
 Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
-In our study, $R_t$ forecasts of cases were comparable to direct forecasts.
-However, given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality directly forecasting cases.
+In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more.
+Sample sizes, however, were very low.
+Given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality directly forecasting cases. 
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last X weeks of data, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
+The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
 The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021}, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
@@
 Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon.
-Another potential solution would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
+Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters.
+Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons.
+One potential way to disentangle the effect of the convolution model from the forecasts would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
 Future work could expose forecasters to different such options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
-%The relationship between cases and deaths used to forecast 
-%In addition .
-The gradual change in the CFR was not accounted for in the computational model underying the $R_t$ forecasts, and estimated distributions from past data only slowly updated to new circumstances.
-One potential solution for this would be to allow humans to adjust the CFR and other model parameters manually as part of their forecast.
+
+%Combining Rt and direct
+Combining forecasts from "experts" and "non-experts" led to better performance for forecasts two weeks ahead for cases as well as deaths, and both on the log scale and on the natural scale.
+Combining direct forecasts and $R_t$ forecasts led to better performance on the log scale, but not on the natural scale.
+However, WIS values of the combined forecast were always lower than the average of the separate WIS values.
+This suggests that combining different forecasts can be beneficial in many instances, although there may be differences in terms of WIS on the log and the natural scale.
+In particular, WIS values on the natural scale may be more susceptible to models that would tend to overshoot and miss the peak, while WIS on the log scale may be more affected by models that underpredict and miss upswings \citep{bosseTransformationForecastsEvaluating2023}.
+
+% \paragraph{Expertise}
+
+Past expert forecasts of COVID-19 \cite{recchiaHowWellDid2021} had found predictions from experts to outperform those of non-experts.
+In our study, an ensemble of self-reported "experts" outperformed an ensemble of "non-experts" when forecasting cases two weeks ahead, both on the log scale and on the natural scale.
+When forecasting deaths two weeks ahead, "experts" performed worse than "non-experts" on the log scale, but better on the natural scale.
+Forecasts for "experts" tended to be better calibrated than non-experts.
+However results should be taken with care in light of relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths), and given that expert status was self-reported.
+Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
+
+%results should be interpreted with caution. Overall, it seems that ensembles comprised of either human forecasts or computational ensembles are able to produce forecasts of similar quality.
 
@@
 
-Overall, results of our study should be taken with some caution due to a number of important limitations.
+Overall, results of our study should be taken with caution due to a number of important limitations.
 Firstly, our study was restricted to one location and to a relatively short period of thirteen weeks.
-Secondly, there are many confounding factors that can influence results.
-These include for example the fact that different participants made forecasts at different points in time and that subgroups of interest (e.g.
+Secondly, there were many confounding factors that likely influence results.
+These include for example the fact that different participants made forecasts at different points in time (with the median forecaster only submitting a single forecast) and that subgroups of interest (e.g.
 "experts", or $R_t$ forecasts) had different numbers of forecasters.
-In addition, there are many researcher degrees of freedom that influence results, for example how individual forecasts are combined to an ensemble and how forecasts are evaluated.
+In most instances, differences in scores between forecast approaches were small compared to the variance of scores within a single approach.
+In addition, there were many researcher degrees of freedom that could influence findings, for example how individual forecasts were combined to an ensemble.
+Results were strongly influenced by the choices made during the evaluation and for example changed depending on which horizon was chosen for comparisons and whether or not forecasts were transformed before scoring.
 Prizes to the human forecasters, for example were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
@@
 
-Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models can show similar performance.
+Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
 One possible interpretation is that a mixed crowd of human forecaster can produce a viable alternative or addition to an ensemble of mathematical models created by experts.
@@
 \subsection*{Acknowledgements}
-We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russel Bradshaw, Sebastian Funk, and Akira Endo.
+We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russel Bradshaw, Sebastian Funk (an author of this study), and Akira Endo.
 All winners donated their prizes.
@@
 
-\input{manuscript/performance-table-experts}
-
-\input{manuscript/performance-table-experts-4}
-
 \begin{figure*}
@@
 \includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
-\caption{\bf{Number of forecasts across the study period.}}
+\caption{\bf{Number of forecasts across the study period.} A: number of forecasts included in the Hub ensemble and the combined crowd ensemble.
+B: number of forecasts by "experts" and "non-experts".
+Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
 \label{fig:num-forecasters}
@@
 \end{figure*}
+
+\input performance-table-horizon-4.tex
+
+\input{manuscript/performance-table-experts}
+
+\input{manuscript/performance-table-experts-4}
 

### 2023-04-10 Analysis.rmd
reorganise code, add some analysis about the numbers of experts and non-experts using Rt, add citations

@@
 - also includes Table SI.1
-
-```{r, performance-table}
-scores_summarised <- scores |>
-  filter(scale %in% c("log", "natural")) |>
-  filter(model %in% models_non_human) |>
-  summarise_scores(by = c("model", "target_type", "scale", "horizon", "forecast_date")) |>
-  group_by(target_type, scale, horizon) |>
-  # mutate(interval_score = (interval_score - mean(interval_score)) / sd(interval_score)) |>
-  group_by(model, target_type, scale, horizon) |>
-  summarise(mean = mean(interval_score), 
-            sd = sd(interval_score)) |>
-  pivot_wider(id_cols = c(model, target_type, horizon), 
-              names_from = scale, values_from = c(mean, sd), names_prefix = "stand_") |>
-  arrange(target_type, model)
-
-make_performance_table <- function(
-    scores, scores_summarised, 
-    hor = 2, 
-    cap = "Performance for two-week-ahead forecasts \\label{tab:scores}"
-) {
-  scores |>
-    filter(scale %in% c("log", "natural")) |>
-    summarise_scores(by = c("model", "target_type", "scale", "horizon")) |>
-    select(model, target_type, scale, interval_score, coverage_50, coverage_90, horizon) |>
-    pivot_wider(names_from = scale, values_from = interval_score) |>
-    select(model, target_type, natural, log, coverage_50, coverage_90, horizon) |>
-    inner_join(scores_summarised) |>
-    mutate(across(c(where(is.numeric)), \(x) signif(x, digits = 2))) |>
-    arrange(target_type, model) |>
-    mutate(natural = label_fn(natural), 
-           log = label_fn(log)) |>
-    mutate(natural = paste0(natural, " (", label_fn(sd_stand_natural), ")"), 
-           log = paste0(log, " (", label_fn(sd_stand_log), ")")) |>
-    filter(horizon == hor) |>
-    select(-mean_stand_log, -sd_stand_log, -mean_stand_natural, -sd_stand_natural, -horizon) |>
-    kable(format = "latex", booktabs = TRUE, 
-          align = c("l", "l", rep("c", 4)),
-          caption = cap,
-          linesep = c('', '', '', '\\addlinespace'),
-          col.names = c("Model", "Target", "WIS", "WIS (log scale)", 
-                        "Coverage 50%", "Coverage 90%")) |> 
-    kable_styling(latex_options = c("scale_down", 
-                                    "hold_position")) 
-}
-
-scores |>
-  filter(model %in% models_non_human) |>
-  make_performance_table(
-    scores_summarised = scores_summarised, 
-    cap = "Performance for two-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores}"
-    ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table.tex')
-
-scores |>
-  filter(model %in% models_non_human) |>
-  make_performance_table(
-    scores_summarised = scores_summarised, 
-    hor = 4, 
-    cap = "Performance for four-week-ahead forecasts. Standard deviations are given in brackets. \\label{tab:scores-4}"
-  ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table-horizon-4.tex')
-
-# scores_revised |>
-#   filter(model %in% models_non_human) |>
-#   make_performance_table() |>
-#   cat(file = 'manuscript/performance-table-revised.tex')
-# 
-# 
-# scores_comparable_ensemble |>
-#   make_performance_table() |>
-#   cat(file = 'manuscript/performance-table-comparison-rt.tex')
-
-```
-
 UPDATED VERSION
@@
 
-# Figure 2: Forecasts and corresponding WIS
+# Figure 2: Ranks for different models
+================================================================================
+
+```{r}
+
+ranks <- scores |>
+  filter(model %in% models_non_human) |>
+  summarise_scores() |>
+  select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
+  group_by(target_type, scale, horizon, forecast_date) |>
+  mutate(rank = rank(interval_score))
+
+plot_ranks <- function(ranks, hor = 2) {
+  ranks |>
+    filter(horizon == hor) |>
+    ggplot(aes(y = model, group = rank, fill = rank)) +
+    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
+    facet_wrap(~ type_and_scale) + 
+    scale_fill_continuous(trans='reverse') + 
+    theme_scoringutils() + 
+    labs(y = "", x = "Frequency", fill = "Rank") + 
+    theme(legend.position = "right")
+}
+
+p_ranks <- ranks |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks()
+ggsave("output/figures/performance-ranks.png", 
+       plot = p_ranks, 
+       width = 7, height = 3.4)
+
+p_ranks_4 <- ranks |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks(hor = 4)
+ggsave("output/figures/performance-ranks-4.png", 
+       plot = p_ranks_4, 
+       width = 7, height = 3.4)
+
+ranks |>
+  filter(horizon == 2) |>
+  mutate(model = factor(model, levels = (levels(models_non_human)))) |>
+  group_by(model, target_type, scale, rank) |>
+  summarise(n = n()) |>
+  mutate(rel = n() / 13) |>
+  pivot_wider(names_from = rank, values_from = c(n, rel)) |>
+  {\(.) {replace(.,is.na(.),0)}}() |>
+  mutate(n_1 = paste0(n_1, " (", round(rel_1, 2), ")"), 
+         n_2 = paste0(n_2, " (", round(rel_2, 2), ")"), 
+         n_3 = paste0(n_3, " (", round(rel_3, 2), ")"), 
+         n_4 = paste0(n_4, " (", round(rel_4, 2), ")")) |>
+  select(-c(rel_1, rel_2, rel_3, rel_4))
+
+```
+
+# Figure 3: Forecasts and corresponding WIS
 ================================================================================
@@
 
-# Figure 3: Comparison performance direct vs. Rt
+# Figure 4: Comparison performance direct vs. Rt
 ================================================================================
@@
   group_by(forecast_date, target_type, method) |>
-  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile)
+  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile, expert)
 }
 
-forecasts_comparison <- filter_comparison_rt_direct(forecasts)
+forecasts_comparison <- forecasts |> 
+  update_expert_status() |>
+  filter_comparison_rt_direct()
+
 scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
@@
        width = 7, height = 4)
+
+# check whether there is a difference between experts and non-experts when using Rt / direct
+scores_forecasts_comparison |>
+  summarise_scores(by = c("type_and_scale", "method", "expert")) |>
+  filter(method == "rt") |>
+  arrange(type_and_scale, expert) |>
+  summarise_scores(fun = signif, digits = 2)
+
+scores_forecasts_comparison |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale", "expert")) |>
+  arrange(target_type, name, method) |>
+  group_by(target_type, name, scale) |>
+  mutate(diff = rep(diff(interval_score), 2)) |>
+  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
+  mutate(method = stringr::str_to_title(method)) |>
+  ggplot(aes(y = interval_score, x = method, group = name)) + 
+  geom_point(size = 0.5) + 
+  geom_line(aes(color = expert), linewidth = 0.3) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  labs(y = "WIS", x = "Forecasting method", color = "Expert status") + 
+  scale_y_continuous(labels = label_fn) + 
+  theme_scoringutils()
 ```
@@
 
-# Ranks
-
-```{r}
-
-ranks <- scores |>
-  filter(model %in% models_non_human) |>
-  summarise_scores() |>
-  select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
-  group_by(target_type, scale, horizon, forecast_date) |>
-  mutate(rank = rank(interval_score))
-
-plot_ranks <- function(ranks, hor = 2) {
-  ranks |>
-    filter(horizon == hor) |>
-    ggplot(aes(y = model, group = rank, fill = rank)) +
-    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
-    facet_wrap(~ type_and_scale) + 
-    scale_fill_continuous(trans='reverse') + 
-    theme_scoringutils() + 
-    labs(y = "", x = "Frequency", fill = "Rank") + 
-    theme(legend.position = "right")
-}
-
-p_ranks <- ranks |>
-  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
-  plot_ranks()
-ggsave("output/figures/performance-ranks.png", 
-       plot = p_ranks, 
-       width = 7, height = 3.4)
-
-p_ranks_4 <- ranks |>
-  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
-  plot_ranks(hor = 4)
-ggsave("output/figures/performance-ranks-4.png", 
-       plot = p_ranks_4, 
-       width = 7, height = 3.4)
-
-ranks |>
-  filter(horizon == 2) |>
-  mutate(model = factor(model, levels = (levels(models_non_human)))) |>
-  group_by(model, target_type, scale, rank) |>
-  summarise(n = n()) |>
-  mutate(rel = n() / 13) |>
-  pivot_wider(names_from = rank, values_from = c(n, rel)) |>
-  {\(.) {replace(.,is.na(.),0)}}() |>
-  mutate(n_1 = paste0(n_1, " (", round(rel_1, 2), ")"), 
-         n_2 = paste0(n_2, " (", round(rel_2, 2), ")"), 
-         n_3 = paste0(n_3, " (", round(rel_3, 2), ")"), 
-         n_4 = paste0(n_4, " (", round(rel_4, 2), ")")) |>
-  select(-c(rel_1, rel_2, rel_3, rel_4))
-
-```
-
 # Unused - Significance analysis
@@
   group_by(type_and_scale) |>
-  summarise(p1 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-direct`, paired = TRUE)$p.value, 
-            p2 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-ensemble`, paired = TRUE)$p.value, 
-            p3 = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)
+  summarise("direct-vs-hub" = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-direct`, paired = TRUE)$p.value, 
+            "combined-vs-hub" = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-ensemble`, paired = TRUE)$p.value, 
+            "rt-vs-hub" = wilcox.test(`EuroCOVIDhub-ensemble`,`crowd-rt`, paired = TRUE)$p.value)

### 2023-04-10 Analysis.rmd
experts and non-experts using Rt

@@
 
+# number of experts and non-experts who used Rt forecasts
+nums |>
+  group_by(forecast_date, target_type, expert, method) |>
+  summarise(n = n()) |>
+  group_by(target_type, expert, method) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  )
+exp_method
+
+# proportion of experts and non-experts using Rt
+nums |>
+  group_by(forecast_date, target_type, expert, method) |>
+  summarise(n = n()) |>
+  #calculate proportion
+  group_by(forecast_date, target_type, expert) |>
+  mutate(n = n / sum(n)) |>
+  group_by(expert, method) |>
+  summarise(
+    mean = mean(`n`), 
+    median = median(`n`), 
+    min = min(`n`), 
+    max = max(`n`)
+  ) |>
+  filter(method == "Rt")
+
 plot_number_forecasters <- avail |>

### 2023-04-10 manuscript/manuscript.tex

@@
 
-Accurate forecasts of infectious diseases can aid public health decision-making.
 In the past, two studies found ensembles of human judgement forecasts of COVID-19 to show predictive performance comparable to ensembles of computational models, at least when predicting case incidences.
@@
 Forecasts were elicited as part of a tournament, the "UK Crowd Forecasting Challenge", and a median ensemble of all forecasts was submitted to the European Forecast Hub.
-Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$ based on an estimated baseline.
+Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
 This was then used to forecast cases and deaths using simulation methods from the \texttt{EpiNow2} \textsf{R} package.
@@
 
-The ensemble of human forecasters overall performed comparably to the official European Forecast Hub ensemble on both cases and deaths, although results were susceptible to changes in details of the evaluation.
+The ensemble of human forecasters overall performed comparably to the official European Forecast Hub ensemble on both cases and deaths, although results were sensitive to changes in details of the evaluation.
 $R_t$ forecasts performed comparably to direct forecasts on cases, but worse on deaths.
-Self-identified experts did not outperform non-experts though the sample size was small.
+Self-identified "experts" outperformed "non-experts" in terms of WIS on the log scale, but not on the natural scale.
 
@@
 
-Human judgement can be useful in predicting the spread of an infectious disease such as COVID-19.
+Human judgement forecasts and computational models can produce forecasts of similar quality for infectious disease such as COVID-19.
 The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
@@
 
-Past research found that in the context of infectious disease forecasting, human judgement forecasts could achieve predictive performance broadly comparable to forecasts generated based on mathematical modelling, in particular when forecasting incident cases.
+Past research found that in the context of infectious disease forecasting, human judgement forecasts could achieve predictive performance broadly comparable to forecasts generated based on mathematical modelling, in particular when forecasting incident cases, rather than lagged indicators indicators like deaths.
 \citet{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
 However, a comparable approach performed worse than computational models at predicting the 2014/15 outbreak of chikungunya in the Americas.
-\citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany, but performing worse when predicting incident deaths.
-Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident cases, and worse when predicting incident deaths.
+\citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany and Poland, but performing worse when predicting incident deaths.
+Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident COVID-19 cases, and worse when predicting incident deaths.
 \citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
@@
 In some situations, human judgement forecasting may have advantages relative to computational models.
-Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are unknown.
+Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are unknown or structural breaks have occurred.
 Humans are generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
@@
 
-Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting this as a binary choice is misleading as most computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage and many human forecasters make use of simple base rate based approaches which are in reality simple models.
+Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting this as a binary choice is misleading.
+Most computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage.
+Also, human forecasters often make use of approaches such as calculating a base rate of incidences, or extrapolating current trends, which are in reality equivalent to simple models.
 One explicit method to combine separate human judgement and computational model forecasts with the goal of improving predictive performance is an ensemble.
@@
 
-The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} started in March 2021 to elicit predictions for various COVID-19 related forecast targets from different research groups every week.
+The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} was launched in March 2021 in order to elicit weekly predictions for various COVID-19 related forecast targets from different research groups.
 The forecasts evaluated in this study were submitted every Monday before 11.59pm GMT between May 24 2021 and August 16 2021.
@@
 From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
-The median number of models included in the Forecast Hub ensemble during the study period was 9 for cases and 10 for deaths (see Figure \ref{fig:num-forecasters} in the SI).
+The median number of models included in the Forecast Hub ensemble for the UK during the study period was 9 for cases and 10 for deaths (see Figure \ref{fig:num-forecasters} in the SI).
 
@@
 All data points were marked as anomalous retrospectively by the European Forecast Hub if in subsequent updates data was changed by more than 5 percent.
-In August 2022 JHU switched the data source for their death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency.
+In August 2022 JHU switched the data source for their UK death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency.
 The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
@@
 
-When signing up, we asked participants to self-identify as "experts", asking them to tick a box if they worked in infectious disease modelling or had professional experience in any related field.
+When signing up, participants could self-identify as "experts" if they worked in infectious disease modelling or had professional experience in any related field.
 
@@
 
-Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in weekly log observed forecasts (i.e.
+Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it.
+The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e.
 as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
@@
 
-In addition to the 'direct' forecasts, we implemented a second forecasting method ('$R_t$ forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$ based on a baseline estimate produced by the \texttt{EpiNow2} \citep{epinow2} \textsf{R} \cite{R} package effective reproduction number model which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
+In addition to the 'direct' forecasts, we implemented a second forecasting method ('$R_t$ forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
+This forecast was made based on a baseline estimate produced by the \texttt{EpiNow2} \citep{epinow2} \textsf{R} \cite{R} package effective reproduction number model which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
 The estimate produced by \texttt{EpiNow2} was shown as the default forecast and could be adjusted by the user.
@@
 Every week, we submitted an ensemble of individual forecasts to the European Forecast Hub.
-In contrast to the ensemble of human forecasts described in \citet{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions.
-%, following suggestions made in CITATION.
+In contrast to the ensemble of human forecasts described in \citet{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions, drawing upon insights gained from the COVID-19 Forecast Hubs \citep{rayComparingTrainedUntrained2022}.
 We submitted three different ensembles to the Hub: The first one, "epiforecasts-EpiExpert\_direct" (here called "direct crowd forecast" or "crowd-direct") was a quantile-wise median ensemble of all the direct forecasts.
@@
 Relative WIS values on the natural scale for two week ahead forecasts were 1.05 for the combined crowd ensemble, 1.03 for the direct crowd forecasts and 1.2 for the $R_t$ forecasts.
+This discrepancy between performance on the log scale can be attributed to the fact that case forecasts of the Hub ensemble tended to be lower than forecasts of the human judgement forecasting approaches (see Figure \ref{fig:forecasts-scores}).
+On the natural scale, this led to smaller overprediction penalties on the natural scale, putting it ahead of human forecasts (see Figure \ref{fig:performance}A,C).
+On the log scale, however, it led to large penalties for underprediction.
 
@@
 
-In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}).
+In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}), despite forecasts being about the same or more dispersed (see Figure \ref{fig:performance}).
 Both the Hub ensemble and the $R_t$ forecast were more often in first place than other approaches (4 times each, both on the log and on the natural scale).
@@
 \includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks.png}
-\caption{\bf{Performance ranks}....}
+\caption{\bf{Ranks for all forecasting approaches for two week ahead forecasts}. Colours indicate how often (out of 13 forecasts) a given approach got 1st, 2nd, 3rd, or 4th rank.}
 \label{fig:performance-ranks}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks-4.png}
-\caption{\bf{Performance ranks}....}
-\label{fig:performance-ranks-4}
 \end{figure*}
@@
 All forecasting approaches except the $R_t$ forecasts showed higher empirical coverage for deaths than for cases (see Figure \ref{fig:performance}).
+Forecasts from the Hub ensemble generally tended to be wider than the human forecasts (see Figures \ref{fig:forecasts-scores} and \ref{fig:performance}B,D).
 For 50\% prediction intervals, the Hub ensemble exceeded the nominal coverage noticeably (0.77) (see Table \ref{tab:scores}).
@@
 
-In terms of WIS on the log scale for two week ahead predictions of deaths, the combined crowd ensemble (0.07) and the direct crowd forecasts (0.99) were slightly ahead of the Hub ensemble, while the $R_t$ forecasts performed noticeably worse (1.98) (see Figure \ref{fig:performance} and Table \ref{tab:scores}).
-Interestingly, combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them.
+In terms of WIS on the log scale for two week ahead predictions of deaths, the combined crowd ensemble (0.97) and the direct crowd forecasts (0.99) were marginally ahead of the Hub ensemble, while the $R_t$ forecasts performed noticeably worse (1.98) (see Figure \ref{fig:performance}D and Table \ref{tab:scores}).
+For the Hub ensemble, the dispersion component played by far the largest role, while this was less the case for the human forecasts, which got higher penalties from both over- and underprediction.
+Interestingly, combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them alone on the log scale.
 In terms of WIS on the natural scale, only the direct forecasts (0.89) performed better for two week ahead death predictions than the Hub ensemble, while the combined crowd ensemble performed a bit worse (1.06) and the $R_t$ forecasts again noticeably worse (2.1).
@@
 When comparing scores on the log and on the natural scale, scores on the log scale were again more evenly distributed across the study period.
-On the natural scale, high scores were concentrated around the end of the study period, were death incidences were highest (see Figure \ref{fig:forecasts-scores}).
+On the natural scale, high scores were concentrated around the end of the study period, when death incidences were highest (see Figure \ref{fig:forecasts-scores}).
 
-\subsection*{Rt forecasts}
+\subsection*{$R_t$ forecasts}
 
@@
 
-%Mention calbration
-
 % Mention number of experts and non-experts
 
-A median ensemble of two week ahead forecasts made only by "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts}).
+A median ensemble of two week ahead forecasts restricted to only those made by either "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts} and Figure \ref{fig:num-forecasters}B for a visualisation of participation).
+The median number of "non-experts" was 11 for cases and 10 for deaths, which was higher than the median number of "experts", which was 8 for cases and deaths.
 
-When comparing "experts" and "non-experts", "experts" were better calibrated for forecasts two weeks ahead (see Figure \ref{fig:performance-experts}.
+When comparing "experts" and "non-experts", the ensemble of "experts" was better calibrated for forecasts two weeks ahead (see Figure \ref{fig:performance-experts}.
 For cases two week ahead, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
 WIS values \textit{relative to the combined crowd ensemble} were 1.08 for "experts" and 1.14 for "non-experts" on the log scale and 1.06 for "experts" and 1.07 for "non-experts" on the natural scale (see Table \ref{tab:scores-experts}).
-
 For two week ahead forecasts of deaths, "experts" performed worse than "non-experts" in terms of WIS on the log scale (WIS relative to the combined crowd ensemble: 1.29 vs.
@@
 1.14).
+Both the "expert"- and the "non-expert"-ensemble had similar proportions of $R_t$ forecasts (mean of 32\% for "experts" and 32.2\% for "non-experts" across cases and deaths together).
 
@@
 Nevertheless, retaining participants beyond the initial recruitment proved challenging, and most forecasters only submitted a single forecast.
-\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms, such as Metaculus or Good Judgement Open that access to a large existing user base and greater resources may be helpful in recruiting a larger number of participants though these lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
+\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms that have access to a large existing user base and greater resources may be helpful in recruiting a larger number of participants, though these platforms lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
 
@@
 For forecasts of cases two weeks ahead, performance of the human judgement forecasts was better than the European Forecast Hub ensemble in terms of WIS on the log scale, and worse in terms of WIS on the natural scale.
+This was linked to a tendency of the Hub ensemble to make lower case predictions, which led to lower overprediction penalties on the natural scale, but noticeably higher underprediction penalties on the log scale.
 For forecasts of deaths two weeks ahead, direct human forecasts and the combined crowd ensemble performed better than the Hub ensemble on the log scale.
@@
 
-In their original study conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
+In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
 They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators.
@@
 Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it.
-In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, on worse on the natural scale.
+In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale.
 Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths.
-Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to changes in the age distribution of cases and the rise of the Delta variant in the UK starting in the beginning of May.
+Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to changes in the age distribution of cases during the study period and the rise of the Delta variant in the UK starting in the beginning of May.
 The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
 One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters.
+A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study had used a mean ensemble.
 
@@
 The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021}, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
-Forecasters had no way of observing the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
+Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
 Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
@@
 Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters.
-Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons.
+Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely affected forecasts.
 One potential way to disentangle the effect of the convolution model from the forecasts would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
@@
 Combining direct forecasts and $R_t$ forecasts led to better performance on the log scale, but not on the natural scale.
-However, WIS values of the combined forecast were always lower than the average of the separate WIS values.
 This suggests that combining different forecasts can be beneficial in many instances, although there may be differences in terms of WIS on the log and the natural scale.
@@
 
-Overall, the results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
+The results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
 One possible interpretation is that a mixed crowd of human forecaster can produce a viable alternative or addition to an ensemble of mathematical models created by experts.
@@
 
-Our findings don't suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
+Our findings do not suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
 Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology in order to maximise utility, predictive performance, or both.
@@
 \centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks-4.png}
+\caption{\bf{Ranks for all forecasting approaches for four week ahead forecasts}. Colours indicate how often (out of 13 forecasts) a given approach got 1st, 2nd, 3rd, or 4th rank.}
+\label{fig:performance-ranks-4}
+\end{figure*}
+
+\clearpage
+
+\begin{figure*}
+\centering
 \includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
@@
 \end{figure*}
+
+\input{manuscript/performance-table-experts}
+
+\input{manuscript/performance-table-experts-4}
 

### 2023-04-10 manuscript/SI/epinow2.tex
correct errors in manuscript, add citation
Updates from Overleaf

@@
 % \hypertarget{text-s1.-further-details-on-the-semi-mechanistic-forecasting-models}{%
-% \subsection{Text S1. Further details on the semi-mechanistic forecasting models}\label{text-s1.-further-details-on-the-semi-mechanistic-forecasting-models}}
+% \subsection*{Text S1. Further details on the semi-mechanistic forecasting models}\label{text-s1.-further-details-on-the-semi-mechanistic-forecasting-models}}
 
 \hypertarget{renewal-equation-model}{%
-\subsubsection{Renewal equation model}\label{renewal-equation-model}}
+\subsubsection*{Renewal equation model}\label{renewal-equation-model}}
 
@@
 
-The prior for the generation time was sourced from (Ganyani et al.
-2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior (Lauer et al.
+The prior for the generation time was sourced from \cite{ganyaniEstimatingGenerationInterval2020} XXX(Ganyani et al.
+2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior \citep{lauerIncubationPeriodCoronavirus2020} XXX(Lauer et al.
 2020) for \(\xi_{O}\).
 This resulted in a gamma-distributed generation time with mean 3.6 days (standard deviation (SD) 0.7), and SD of 3.1 days (SD 0.8) for all estimates.
-We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data (Xu et al.
+We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data \citep{kraemer2020epidemiological, covidregionaldata} XXX(Xu et al.
 2020; Abbott, Sherratt, et al.
-2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution (Abbott, Hellewell, et al.
+2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution \citep{epinow2, epiforecasts.io/covidCovid19TemporalVariation2020, sherrattExploringSurveillanceData2021, rstan} XXX(Abbott, Hellewell, et al.
 2020; epiforecasts.io/covid 2020; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}''} n.d.; Stan Development Team 2020).
 
-\(GP_t\) is an approximate Hilbert space Gaussian process as defined in (Riutort-Mayol et al.
+\(GP_t\) is an approximate Hilbert space Gaussian process as defined in \cite{riutort-mayolPracticalHilbertSpace2022} XXX(Riutort-Mayol et al.
 2020) using a Matern 3/2 kernel using a boundary factor of 1.5 and 17 basis functions (20\% of the number of days used in fitting).
@@
 where \(I^c_t = \sum_{s< t} I_s\) are cumulative infections by \(t-1\) and \(I'_t\) are the unadjusted infections defined above.
-This adjustment is based on that implemented in the \texttt{epidemia} R package (Scott et al.
+This adjustment is based on that implemented in the \texttt{epidemia} R package \citep{epidemia, bhattSemiMechanisticBayesianModeling2023} XXX(Scott et al.
 2020; Bhatt et al., n.d.).
@@
 \hypertarget{model-fitting}{%
-\subsubsection{Model fitting}\label{model-fitting}}
+\subsubsection*{Model fitting}\label{model-fitting}}
 
-Both models were implemented using the \texttt{EpiNow2} R package (version 1.3.3) (Abbott, Hellewell, et al.
-2020).
-Each forecast target was fitted independently for each model using Markov-chain Monte Carlo (MCMC) in stan (Stan Development Team 2020).
+Both models were implemented using the \texttt{EpiNow2} R package (version 1.3.3) \citep{epinow2}.
+Each forecast target was fitted independently for each model using Markov-chain Monte Carlo (MCMC) in stan \citep{rstan}.
 A minimum of 4 chains were used with a warmup of 250 samples for the renewal equation-based model and 1000 samples for the convolution model.
@@
 Different settings were chosen for each model to optimise compute time contingent on convergence.
-Convergence was assessed using the R hat diagnostic (Stan Development Team 2020).
+Convergence was assessed using the R hat diagnostic \citep{rstan}.
 For the convolution model forecast the case forecast from the renewal equation model was used in place of observed cases beyond the forecast horizon using 1000 posterior samples.

### 2023-04-12 Analysis.rmd
update rank plot

@@
 ```{r}
-
 ranks <- scores |>

### 2023-04-12 manuscript/manuscript.tex
Updates from Overleaf

@@
 
-From May 24th to August 16th 2021, we elicited weekly one to four-week-ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
-Forecasts were elicited as part of a tournament, the "UK Crowd Forecasting Challenge", and a median ensemble of all forecasts was submitted to the European Forecast Hub.
+From May 24th to August 16th 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
+A median ensemble of all forecasts was submitted to the European Forecast Hub.
 Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
@@
 
-% Beginning in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
-
 Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting this as a binary choice is misleading.
@@
 
-% Define the terms model, forecaster, crowd forecast, ensemble.
-
 \section*{Methods}
@@
 Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
-Until the 12th of July 2021 the default Hub ensemble shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
+Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
 From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
@@
 \citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
-We therefore also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
+We, therefore, also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
 In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
-To make scores easier to interpret, we report relative WIS scores, where the average score for a given model was divided by the average score for the European Forecast Hub ensemble.
+To make scores easier to interpret, we report relative WIS scores, where the average score for a given model was divided by the average score for the European Forecast Hub ensemble ("EuroCOVIDhub-ensemble").
 In addition, we computed ranks based on WIS values.

### 2023-04-12 manuscript/SI/epinow2.tex

@@
 
-The prior for the generation time was sourced from \cite{ganyaniEstimatingGenerationInterval2020} XXX(Ganyani et al.
-2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior \citep{lauerIncubationPeriodCoronavirus2020} XXX(Lauer et al.
-2020) for \(\xi_{O}\).
+The prior for the generation time was sourced from \cite{ganyaniEstimatingGenerationInterval2020} but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior \citep{lauerIncubationPeriodCoronavirus2020} for \(\xi_{O}\).
 This resulted in a gamma-distributed generation time with mean 3.6 days (standard deviation (SD) 0.7), and SD of 3.1 days (SD 0.8) for all estimates.
-We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data \citep{kraemer2020epidemiological, covidregionaldata} XXX(Xu et al.
-2020; Abbott, Sherratt, et al.
-2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution \citep{epinow2, epiforecasts.io/covidCovid19TemporalVariation2020, sherrattExploringSurveillanceData2021, rstan} XXX(Abbott, Hellewell, et al.
-2020; epiforecasts.io/covid 2020; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}''} n.d.; Stan Development Team 2020).
+We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data \citep{kraemer2020epidemiological, covidregionaldata} and combining the posteriors for the mean and standard deviation of the log-normal distribution \citep{epinow2, epiforecasts.io/covidCovid19TemporalVariation2020, sherrattExploringSurveillanceData2021, rstan}.
 
-\(GP_t\) is an approximate Hilbert space Gaussian process as defined in \cite{riutort-mayolPracticalHilbertSpace2022} XXX(Riutort-Mayol et al.
-2020) using a Matern 3/2 kernel using a boundary factor of 1.5 and 17 basis functions (20\% of the number of days used in fitting).
+\(GP_t\) is an approximate Hilbert space Gaussian process as defined in \cite{riutort-mayolPracticalHilbertSpace2022} using a Matern 3/2 kernel using a boundary factor of 1.5 and 17 basis functions (20\% of the number of days used in fitting).
 The length scale of the Gaussian process was given a log-normal prior with a mean of 21 days, and a standard deviation of 7 days truncated to be greater than 3 days and less than 60 days.
@@
 where \(I^c_t = \sum_{s< t} I_s\) are cumulative infections by \(t-1\) and \(I'_t\) are the unadjusted infections defined above.
-This adjustment is based on that implemented in the \texttt{epidemia} R package \citep{epidemia, bhattSemiMechanisticBayesianModeling2023} XXX(Scott et al.
-2020; Bhatt et al., n.d.).
+This adjustment is based on that implemented in the \texttt{epidemia} R package \citep{epidemia, bhattSemiMechanisticBayesianModeling2023}.
 
@@
 
-\hypertarget{refs}{}
+%\hypertarget{refs}{}
 % \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-epinow2}{}}%
-Abbott, Sam, Joel Hellewell, Joe Hickson, James Munday, Katelyn Gostic, Peter Ellis, Katharine Sherratt, et al.
+%\leavevmode\vadjust pre{\hypertarget{ref-epinow2}{}}%
+%Abbott, Sam, Joel Hellewell, Joe Hickson, James Munday, Katelyn Gostic, Peter Ellis, Katharine Sherratt, et al.
 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-covidregionaldata}{}}%
-Abbott, Sam, Katharine Sherratt, Jonnie Bevan, Hamish Gibbs, Joel Hellewell, James Munday, Patrick Barks, Paul Campbell, Flavio Finger, and Sebastian Funk.
+%\leavevmode\vadjust pre{\hypertarget{ref-covidregionaldata}{}}%
+%Abbott, Sam, Katharine Sherratt, Jonnie Bevan, Hamish Gibbs, Joel Hellewell, James Munday, Patrick Barks, Paul Campbell, Flavio Finger, and Sebastian Funk.
 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bhattSemiMechanisticBayesianModeling}{}}%
-Bhatt, Samir, Neil Ferguson, Seth Flaxman, Axel Gandy, Swapnil Mishra, and James A Scott. n.d. {``Semi-{Mechanistic Bayesian} Modeling of {COVID-19} with {Renewal Processes},''} 14.
+%\leavevmode\vadjust pre{\hypertarget{ref-bhattSemiMechanisticBayesianModeling}{}}%
+%Bhatt, Samir, Neil Ferguson, Seth Flaxman, Axel Gandy, Swapnil Mishra, and James A Scott. n.d. {``Semi-{Mechanistic Bayesian} Modeling of {COVID-19} with {Renewal Processes},''} 14.
 
-\leavevmode\vadjust pre{\hypertarget{ref-epiforecasts.ioux2fcovidCovid19TemporalVariation2020}{}}%
-epiforecasts.io/covid. 2020. {``Covid-19: {Temporal} Variation in Transmission During the {COVID-19} Outbreak.''} {Covid-19}. 2020. \url{https://epiforecasts.io/covid/}.
+%\leavevmode\vadjust pre{\hypertarget{ref-epiforecasts.ioux2fcovidCovid19TemporalVariation2020}{}}%
+%epiforecasts.io/covid. 2020. {``Covid-19: {Temporal} Variation in Transmission During the {COVID-19} Outbreak.''} {Covid-19}. 2020. \url{https://epiforecasts.io/covid/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-EvaluatingUseReproduction}{}}%
-{``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}.''} n.d.
+%\leavevmode\vadjust pre{\hypertarget{ref-EvaluatingUseReproduction}{}}%
+%{``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid-19} Outbreak in {England} \textbar{} {medRxiv}.''} n.d.
 Accessed May 30, 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-generationinterval}{}}%
-Ganyani, Tapiwa, Cecile Kremer, Dongxuan Chen, Andrea Torneri, Christel Faes, Jacco Wallinga, and Niel Hens.
+%\leavevmode\vadjust pre{\hypertarget{ref-generationinterval}{}}%
+%Ganyani, Tapiwa, Cecile Kremer, Dongxuan Chen, Andrea Torneri, Christel Faes, Jacco Wallinga, and Niel Hens.
 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-incubationperiod}{}}%
-Lauer, Stephen A, Kyra H Grantz, Qifang Bi, Forrest K Jones, Qulu Zheng, Hannah R Meredith, Andrew S Azman, Nicholas G Reich, and Justin Lessler.
+%\leavevmode\vadjust pre{\hypertarget{ref-incubationperiod}{}}%
+%Lauer, Stephen A, Kyra H Grantz, Qifang Bi, Forrest K Jones, Qulu Zheng, Hannah R Meredith, Andrew S Azman, Nicholas G Reich, and Justin Lessler.
 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-approxGP}{}}%
-Riutort-Mayol, Gabriel, Paul-Christian Bürkner, Michael R.
+%\leavevmode\vadjust pre{\hypertarget{ref-approxGP}{}}%
+%Riutort-Mayol, Gabriel, Paul-Christian Bürkner, Michael R.
 Andersen, Arno Solin, and Aki Vehtari.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-epidemia}{}}%
-Scott, James A., Axel Gandy, Swapnil Mishra, Juliette Unwin, Seth Flaxman, and Samir Bhatt.
+%\leavevmode\vadjust pre{\hypertarget{ref-epidemia}{}}%
+%Scott, James A., Axel Gandy, Swapnil Mishra, Juliette Unwin, Seth Flaxman, and Samir Bhatt.
 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rstan}{}}%
-Stan Development Team. 2020. {``RStan: The r Interface to Stan.''} \url{http://mc-stan.org/}.
+%\leavevmode\vadjust pre{\hypertarget{ref-rstan}{}}%
+%Stan Development Team. 2020. {``RStan: The r Interface to Stan.''} \url{http://mc-stan.org/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kraemer2020epidemiological}{}}%
-Xu, Bo, Bernardo Gutierrez, Sarah Hill, Samuel Scarpino, Alyssa Loskill, Jessie Wu, Kara Sewalk, et al.
+%\leavevmode\vadjust pre{\hypertarget{ref-kraemer2020epidemiological}{}}%
+%Xu, Bo, Bernardo Gutierrez, Sarah Hill, Samuel Scarpino, Alyssa Loskill, Jessie Wu, Kara Sewalk, et al.
 2020.

### 2023-04-15 manuscript/manuscript.tex
Updates from Overleaf

@@
 Infectious disease modelling and forecasting has attracted wide-spread attention during the COVID-19 pandemic and helped inform decision making in public health organisations and governments \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}.
-% Beginning in March 2020, forecasts for different COVID-19 targets have been systematically collated by Forecast Hubs in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
 Most forecasts used to inform decision making were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative or in combination \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
@@
 In some situations, human judgement forecasting may have advantages relative to computational models.
-Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are unknown or structural breaks have occurred.
-Humans are generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) or can take factors into account that are hard to encode in a computational model.
+Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are hard to quantify.
+Humans are also generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) and can take factors into account that are hard to encode in a computational model.
 On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, and humans may be at a disadvantage at tasks that strongly benefit from the ability to perform complex computations.
@@
 Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
-All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2021}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
-This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country and with an increased number of participants.
+All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants.
 In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the estimated effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a simulation model.
@@
 The median number of submissions from a single individual was one, meaning that similar to \citep{bosseComparingHumanModelbased2022} most forecasters dropped out after their first submission.
-Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study.
+Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study (S.
+Abbott).
 The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A).
@@
 WIS values relative to the Hub ensemble (=1) were 0.91 for the combined crowd ensemble, 0.96 for the direct crowd forecasts and 0.93 for the $R_t$ forecasts (see Table \ref{tab:scores}).
-In terms of WIS on the natural scale, however, the Hub ensemble was ahead of all human forecasting approaches.
+In contrast, in terms of WIS on the natural scale the Hub ensemble outperformed all human forecasting approaches.
 Relative WIS values on the natural scale for two week ahead forecasts were 1.05 for the combined crowd ensemble, 1.03 for the direct crowd forecasts and 1.2 for the $R_t$ forecasts.
-This discrepancy between performance on the log scale can be attributed to the fact that case forecasts of the Hub ensemble tended to be lower than forecasts of the human judgement forecasting approaches (see Figure \ref{fig:forecasts-scores}).
-On the natural scale, this led to smaller overprediction penalties on the natural scale, putting it ahead of human forecasts (see Figure \ref{fig:performance}A,C).
+The discrepancy between performance on the log and natural scale  can be attributed to case forecasts from the Hub ensemble tending to be lower than forecasts from human judgement approaches (see Figure \ref{fig:forecasts-scores}).
+On the natural scale, this resulted in smaller overprediction penalties, putting it ahead of human forecasts (see Figure \ref{fig:performance}A,C).
 On the log scale, however, it led to large penalties for underprediction.
@@
 
-In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}), despite forecasts being about the same or more dispersed (see Figure \ref{fig:performance}).
+In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ forecast showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}), despite forecasts being about the same or more dispersed (see Figure \ref{fig:performance}).
 Both the Hub ensemble and the $R_t$ forecast were more often in first place than other approaches (4 times each, both on the log and on the natural scale).
@@
 The crowd ensemble never placed fourth, but also had the lowest number of first places (2, both on the log and the natural scale).
-Model ranks only changed marginally when switching between the log and the natural scale.
+Aggregated model ranks only changed marginally when switching between the log and the natural scale (see Figure \ref{fig:performance-ranks}.
 
@@
 For the Hub ensemble, the dispersion component played by far the largest role, while this was less the case for the human forecasts, which got higher penalties from both over- and underprediction.
-Interestingly, combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them alone on the log scale.
-In terms of WIS on the natural scale, only the direct forecasts (0.89) performed better for two week ahead death predictions than the Hub ensemble, while the combined crowd ensemble performed a bit worse (1.06) and the $R_t$ forecasts again noticeably worse (2.1).
-
-% Maybe say something about increasing forecast horizons. But the pattern is unclear.
+Combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them alone on the log scale despite the poor overall performance of the $R_t$ forecasts.
+In terms of WIS on the natural scale, only the direct forecasts (0.89) performed better for two week ahead death predictions than the Hub ensemble, while the combined crowd ensemble performed slightly worse (1.06) and the $R_t$ forecasts again noticeably worse (2.1).
 
@@
 In particular, $R_t$ forecasts for deaths were worse than the corresponding direct death forecasts for most forecasters (see Figure \ref{fig:comparison-direct-rt-individual}).
-Changing from the direct forecasting method to $R_t$ forecasting for cases tended to improve scores for better forecasters and decrease scores for worse forecasters, although sample sizes are very small.
+Changing from the direct forecasting method to $R_t$ forecasting for cases tended to improve scores for better forecasters and decrease scores for worse forecasters, although sample sizes and the size of the observed effect are both small.
 
@@
 
-When comparing "experts" and "non-experts", the ensemble of "experts" was better calibrated for forecasts two weeks ahead (see Figure \ref{fig:performance-experts}.
-For cases two week ahead, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
+When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}.
+For cases, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
 WIS values \textit{relative to the combined crowd ensemble} were 1.08 for "experts" and 1.14 for "non-experts" on the log scale and 1.06 for "experts" and 1.07 for "non-experts" on the natural scale (see Table \ref{tab:scores-experts}).
-For two week ahead forecasts of deaths, "experts" performed worse than "non-experts" in terms of WIS on the log scale (WIS relative to the combined crowd ensemble: 1.29 vs.
+For deaths, "experts" performed worse than "non-experts" in terms of WIS on the log scale (WIS relative to the combined crowd ensemble: 1.29 vs.
 1.06), but better on the natural scale (1.03 vs.
@@
 In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
-Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model to generate simulated reported cases and deaths and an extension which estimated a fixed case fatality ratio and a log normal delay between cases and reports.
+Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model, along with a second model relating cases and deaths from the same source, to simulate reported cases and deaths.
 
@@
 In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, we found almost all forecasts for cases to be underdispersed (i.e.
-too narrow and overconfident).
-Empirical coverage for death forecasts was higher for all forecasting approaches except the $R_t$ forecasts.
+too narrow/overconfident).
+Empirical coverage for death forecasts was higher than the corresponding coverage for cases for all forecasting approaches except the $R_t$ forecasts.
 
@@
 Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths.
-Relatively good performance of direct human judgement forecasts for death forecasts compared to previous studies may be related to changes in the age distribution of cases during the study period and the rise of the Delta variant in the UK starting in the beginning of May.
-The Delta variant had a higher case fatality ratio (CFR) than previous variants \cite{linDiseaseSeverityClinical2021}.
-One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters.
-A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study had used a mean ensemble.
+During the study period, the case fatality ratio (CFR) changed quite quickly.
+This was related to the rise of the Delta variant in the UK in the beginning of May with its higher case fatality rate \citep{bastIncreasedRiskHospitalisation2021}.
+In addition, the age distribution of cases changed over the course of the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
+One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update .
+A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.
 
@@
 This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths.
-One appeal of this approach is that the forecaster can directly forecast the generative process.
-Computational modelling would then take care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, or the relationship between different indicators.
+One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour.
+Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators.
 This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$.
+Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions.
 Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
@@
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and not evolving at all over the forecast period.
-The CFR, however, changed quite quickly during the study period, both due to the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021}, as well as changes in the age distribution of cases in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
+The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon.
+However, as mentioned before, the CFR evolved during the study period.
 Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
 Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
-Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon.
+Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period.
 Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters.
-Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely affected forecasts.
-One potential way to disentangle the effect of the convolution model from the forecasts would have been to use the human forecasts for cases as an input to the computational model, which could then have estimated deaths as a convolution of cases over a delay distribution.
-Future work could expose forecasters to different such options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
+Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts.
+One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths.
+Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
 
@@
 Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
-
-%results should be interpreted with caution. Overall, it seems that ensembles comprised of either human forecasts or computational ensembles are able to produce forecasts of similar quality.
+However, as we used ad-hoc recruitment in our networks many of these self-identified experts are likely to be infectious disease modellers.
 
 % \paragraph{Susceptibility of results}
-
 Overall, results of our study should be taken with caution due to a number of important limitations.
@@
 Secondly, there were many confounding factors that likely influence results.
-These include for example the fact that different participants made forecasts at different points in time (with the median forecaster only submitting a single forecast) and that subgroups of interest (e.g.
+These include the fact that different participants made forecasts at different points in time (with the median forecaster only submitting a single forecast) and that subgroups of interest (e.g.
 "experts", or $R_t$ forecasts) had different numbers of forecasters.
 In most instances, differences in scores between forecast approaches were small compared to the variance of scores within a single approach.
-In addition, there were many researcher degrees of freedom that could influence findings, for example how individual forecasts were combined to an ensemble.
-Results were strongly influenced by the choices made during the evaluation and for example changed depending on which horizon was chosen for comparisons and whether or not forecasts were transformed before scoring.
-Prizes to the human forecasters, for example were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
-Had we chosen to instead measure WIS on the natural scale, rankings and payouts would have been different.
+In addition, there were many researcher degrees of freedom that could influence findings, for example how individual forecasts were combined to create an ensemble.
+Results were influenced by choices made during the evaluation with, for example, some conclusions  depending on forecast horizon and the transformation used prior to scoring.
+Highlighting this, prizes to the human forecasters were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
+Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and payouts would have been different.
 
@@
 
-The results of our study seem broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
-One possible interpretation is that a mixed crowd of human forecaster can produce a viable alternative or addition to an ensemble of mathematical models created by experts.
-Another interpretation is that an ensemble of automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
-All previous studies comparing human judgement forecasts and computational models only ran over short periods of time and the majority of them struggled with recruitment and upkeep.
+The results of our study are broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
+One interpretation of these findings is that a mixed crowd of human forecaster can produce a viable alternative or complement to an ensemble of mathematical models created by experts.
+An altnerative interpretation is that an ensemble of automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
+This study, and all previous studies, comparing human judgement forecasts and computational models only ran over short periods of time and the majority of them struggled with recruitment and upkeep.
 Meanwhile, COVID-19 Forecast Hubs have attracted continuous submissions for almost three years and were able to consistently provide forecasts of comparable quality.
 
-Our findings do not suggest that humans are necessarily at a general disadvantage compared to computational models at predicting death numbers, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
+Our findings do not suggest that humans are necessarily at a general disadvantage compared to computational models at predicting reported deaths, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
 Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology in order to maximise utility, predictive performance, or both.
@@
 Our results underline that it is difficult to evaluate forecast performance devoid of context that helps inform what a good or a bad forecast is.
-Different ways to look at the data let different forecasts appear good or bad.
-Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture.
-Furthermore, evaluating forecasts post-hoc leaves the researchers with many degrees of freedom to make decisions that strongly affect which models look good and there is a high risk of allowing for motivated reasoning.
+Different ways to look at the data let different forecasts appear better or worse.
+Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture as we have attempted to present in this work.
+Furthermore, evaluating forecasts post-hoc leaves the researchers with many degrees of freedom to make decisions that affect which models look good and there is a risk of allowing for motivated reasoning.
 More emphasis should be put on measures that prevent this, e.g.
-by establishing common standards for evaluations, pre-registering studies, and making it a norm to display a variety of metrics.
+by establishing common standards for evaluations, pre-registering studies, and making it a norm to display a variety of standard metrics.
 

### 2023-05-04 manuscript/manuscript.tex
Some manuscript corrections

@@
 
-\title{Human Judgement Forecasting of COVID-19 in the UK}
+\title{Human judgement forecasting of COVID-19 in the UK}
 %\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
@@
 
-From May 24th to August 16th 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
+From May 24th to August 16th, 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
 A median ensemble of all forecasts was submitted to the European Forecast Hub.
@@
 Human judgement forecasts and computational models can produce forecasts of similar quality for infectious disease such as COVID-19.
-The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
+The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a ``good" forecast is dependent on the forecast consumer.
 Combinations of human and computational forecasts hold potential but present real-world challenges that need to be solved.
@@
 Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident COVID-19 cases, and worse when predicting incident deaths.
-\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
+\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit many participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
 It is important to note that in previous studies (and also this one) human forecasters were free to use any resources, including computational models, in the process of creating a forecast, making it difficult to completely separate human judgement and computational modelling.
@@
 Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
-All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022a}.
+All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}.
 This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants.
@@
 
-The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022a} was launched in March 2021 in order to elicit weekly predictions for various COVID-19 related forecast targets from different research groups.
+The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022} was launched in March 2021 in order to elicit weekly predictions for various COVID-19 related forecast targets from different research groups.
 The forecasts evaluated in this study were submitted every Monday before 11.59pm GMT between May 24 2021 and August 16 2021.
 Forecasts were made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
-While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday and starting on Sunday.
+While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday, and starting on Sunday.
 Forecast horizons were therefore in fact 5, 12, 19 and 26 days.
@@
 Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
-Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
+Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
 From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
@@
 Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it.
-The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e.
-as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
+The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e., as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
 In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
@@
 Latent infections were convolved with delay distributions representing the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
-This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
+This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
 Further details are given in the SI.
 
-In order to obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+To obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
 Further details are given in the SI and in our previous work \citep{bosseComparingHumanModelbased2022}.
@@
 
-In order to boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
+To boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
 Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
@@
 averaged across forecast dates, horizons and forecast targets.
-For the purpose of the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
+For the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
 The UK crowd forecasting challenge was announced over Twitter and our networks.
@@
 
-Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}. We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
+We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
 \cite{bracherEvaluatingEpidemicForecasts2021}.
@@
 The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
-The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
-uncertainty of forecasts), overprediction and underprediction.
+The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e., uncertainty of forecasts), overprediction and underprediction.
 
-\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
+\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS to better reflect the exponential nature of the underlying disease process.
 We, therefore, also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
@@
 In order to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}, we used the empirical coverage of all central 50\% and 90\% prediction intervals.
-Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
-the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
+Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g., the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
 
-If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022a, bracherShorttermForecastingCOVID192021}.
+If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021}.
 We analysed all forecasts stratified by forecast target (cases or deaths), forecast horizon, and forecast approach.
@@
 A total number of 90 participants submitted forecasts (more precisely, forecasts were submitted from 90 different accounts, some of them anonymous).
-Out of 90 participants, 21 self-identified as "experts", i.e.
-stated they had professional experience in infectious disease modelling or a related field.
+Out of 90 participants, 21 self-identified as "experts", i.e., stated they had professional experience in infectious disease modelling or a related field.
 
@@
 B: Corresponding WIS values, decomposed into dispersion, overprediction and underprediction.
-C: 50\% prediction intervals on the log scale, i.e.
-after applying the natural logarithm to all forecasts and observations.
-D: Corresponding WIS on the log scale, i.e.
-the WIS applied to the log-transformed forecasts and observations.}
+C: 50\% prediction intervals on the log scale, i.e., after applying the natural logarithm to all forecasts and observations.
+D: Corresponding WIS on the log scale, i.e., the WIS applied to the log-transformed forecasts and observations.}
 \label{fig:forecasts-scores} 
@@
 
-\subsection*{Experts and Non-Experts}
+\subsection*{Experts and non-experts}
 
@@
 %Calibration
-In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022a}, we found almost all forecasts for cases to be underdispersed (i.e.
-too narrow/overconfident).
+In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022}, we found almost all forecasts for cases to be underdispersed (i.e., too narrow/overconfident).
 Empirical coverage for death forecasts was higher than the corresponding coverage for cases for all forecasting approaches except the $R_t$ forecasts.
@@
 Sample sizes, however, were very low.
-Given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality directly forecasting cases. 
+Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
@@
 Forecasts for "experts" tended to be better calibrated than non-experts.
-However results should be taken with care in light of relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths), and given that expert status was self-reported.
+However results should be taken with care considering relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths) and given that expert status was self-reported.
 Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
@@
 % \paragraph{Susceptibility of results}
-Overall, results of our study should be taken with caution due to a number of important limitations.
+Overall, results of our study should be taken with caution due to several important limitations.
 Firstly, our study was restricted to one location and to a relatively short period of thirteen weeks.
@@
 Highlighting this, prizes to the human forecasters were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
-Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and payouts would have been different.
+Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and pay-outs would have been different.
 
@@
 \subsection*{Author contributions}
-In order to give appropriate credit to each author of an article, the individual
-contributions of each author to the manuscript should be detailed in this section. We
-recommend using author initials and then stating briefly how they contributed.
+NIB contributed to the conceptualization, data curation, formal analysis, investigation, methodology, software development, validation, visualization, and original draft preparation of the manuscript, as well as its review and editing.
+SA contributed to the conceptualization, data curation, investigation, methodology, software development, supervision, and review and editing of the manuscript.
+JB contributed to the supervision, review, and editing of the manuscript.
+EvL contributed to the conceptualization, supervision, and review and editing of the manuscript.
+AC contributed to the conceptualization, supervision, and review and editing of the manuscript.
+SF contributed to the conceptualization, funding acquisition, project administration, supervision, and review and editing of the manuscript.
+All authors have read and approved the final version of the manuscript.
 
 \subsection*{Competing interests}
-Nikos I. Bosse is a part-time employee of Metaculus, an online prediction platform.
+NIB is a part-time employee of Metaculus, an online prediction platform.
 
 \subsection*{Grant information}
-Please state who funded the work discussed in this article, whether it is your employer,
-a grant funder etc. Please do not list funding that you have that is not relevant to this
-specific piece of research. For each funder, please state the funder’s name, the grant
-number where applicable, and the individual to whom the grant was assigned.
+NIB received funding from the Health Protection Research Unit (grant code NIHR200908).
+SA's work was funded by the Wellcome Trust (grant: 210758/Z/18/Z).
+AC's acknowledges funding by the NIHR, the Sergei Brin foundation, USAID, and support by the Academy of Medical Sciences Springboard scheme, funded by the AMS, Wellcome Trust, BEIS, the British Heart Foundation and Diabetes UK (REF:SBF005\\1044).
+EvL acknowledges funding by the National Institute for Health Research (NIHR) Health Protection Research Unit (HPRU) in Modelling and Health Economics (grant number NIHR200908) and the European Union's Horizon 2020 research and innovation programme - project EpiPose (101003688).
+The work of JB was supported by the Helmholtz Foundation (https://www.helmholtz.de/) via the SIMCARD Information and Data Science Pilot Project.
+SF's work was supported by the Wellcome Trust (grant: 210758/Z/18/Z).
+This work was supported by the National Institute for Health and Care Research (NIHR) Health Protection Research Unit (HPRU) in Modelling and Health Economics, which is a partnership between the UK Health Security Agency (UKHSA), Imperial College London, and the London School of Hygiene \& Tropical Medicine (grant code NIHR200908).
+The views expressed are those of the authors and not necessarily those of the UK Department of Health and Social Care (DHSC), NIHR, or UKHSA.
 
 \subsection*{Acknowledgements}
-We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russel Bradshaw, Sebastian Funk (an author of this study), and Akira Endo.
+We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russell Bradshaw, Sebastian Funk (an author of this study), and Akira Endo.
 All winners donated their prizes.
+
+\subsection*{Data availability}
+All data and ode are available publicy under a MIT license under \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
 
@@
 % https://tex.stackexchange.com/questions/520193/is-it-possible-to-except-figures-in-the-appendix-from-being-repositioned-by-endf
-\processdelayedfloats
+%\processdelayedfloats
 \csname efloat@restorefloats\endcsname
@@
   $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
-  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e.
-the lower and upper bound of a single prediction interval.
+  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
 For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,

### 2023-05-04 manuscript/manuscript.tex
update data availablity information and citation for SI

@@
 \subsection*{Data availability}
-All data and ode are available publicy under a MIT license under \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
+All data and code are available publicy under a MIT license under \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge} and \url{https://doi.org/10.5281/zenodo.7897257}.
+The data has been published separately here: \url{https://doi.org/10.5281/zenodo.7897289}.
+Supplementary information \cite{bosseSupplementaryInformationHuman2023} to this manuscript is available at \url{https://doi.org/10.5281/zenodo.7897513}.
 

### 2023-05-04 manuscript/manuscript.tex
Updates from Overleaf

--- 2023-05-04 manuscript/manuscript.tex
@@
 \usepackage[numbers]{natbib}
+\bibliographystyle{custom}
 
@@
 The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
-The data and revisions are displayed in Figure \ref{fig:plot-data}.
+The data and revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
 All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).
@@
 In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
-Figure \ref{fig:screenshot-classical} shows a screenshot of the forecast interface for direct forecasts.
+Figure 3 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023} shows a screenshot of the forecast interface for direct forecasts.
 
@@
 This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
-Further details are given in the SI.
+Further details are given in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
 
 To obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-Further details are given in the SI and in our previous work \citep{bosseComparingHumanModelbased2022}.
+Further details are given in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
 
@@
 However, due to lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
-Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
+Figure 4 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023} shows a screenshot of the forecast interface for $R_t$ forecasts.
 
@@
 
-We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
-\cite{bracherEvaluatingEpidemicForecasts2021}.
+We scored forecasts using the weighted interval score
+\cite{bracherEvaluatingEpidemicForecasts2021}. For a single interval, the interval score is computed as 
+  $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
+  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
+For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,
+  $$WIS = \frac{1}{K + 0.5} \cdot \left( w_0 \cdot |y - m| + \sum_{k = 1}^{K} w_k \cdot IS_{\alpha}(F, y) \right), $$
+  where $w_k$ is a weight for every interval. Usually, $w_k = \frac{\alpha_k}{2}$ and $w_0 = 0.5$.
+
 The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
+\caption{\bf{Number of forecasts across the study period.} A: number of forecasts included in the Hub ensemble and the combined crowd ensemble.
+B: number of forecasts by "experts" and "non-experts".
+Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
+\label{fig:num-forecasters}
+\end{figure*}
+
 \input performance-table.tex
@@
 WIS values the human forecasts of 1.02, 1.05, 1.06) and on the natural scale (rel.
-WIS values of 1.21, 1.25, 1.3) (see Table \ref{tab:scores-4}).
+WIS values of 1.21, 1.25, 1.3) (compare Table 1 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}).
 
@@
 
-When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}.
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
+\caption{\bf{Predictive performance of self-reported "experts" and "non-experts" across forecast horizons.} Forecasts from "experts" and "non-experts" were combined to two separate median ensembles, including both direct and $R_t$ forecasts.
+A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
+E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
+\label{fig:performance-experts}
+\end{figure*}
+
+\input{manuscript/performance-table-experts}
+
+When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}).
 For cases, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
@@
 Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
-\label{fig:num-forecasters}
+\label{fig:num-forecasters-SI}
 \end{figure*}
@@
 E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance-experts}
+\label{fig:performance-experts-SI}
 \end{figure*}
 
-\input{manuscript/performance-table-experts}
+%\input{manuscript/performance-table-experts}
 

### 2023-05-04 manuscript/supplementary-information.tex
Updates from Overleaf

--- 2023-05-04 manuscript/manuscript.tex
@@
 
-\title{Human judgement forecasting of COVID-19 in the UK}
+\title{Supplementary Information - Human judgement forecasting of COVID-19 in the UK}
 %\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
@@
 
-\author[1, 2, 3]{Nikos I. Bosse}
-\author[1, 2]{Sam Abbott}
-\author[6, 7]{Johannes Bracher}
-\author[1, 4]{Edwin van Leeuwen}
-\author[5]{Anne Cori}
-\author[1, 2, 3]{Sebastian Funk}
-
-\affil[1]{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom}
-\affil[2]{Centre for the Mathematical Modelling of Infectious Diseases, London, United Kingdom}
-\affil[3]{NIHR Health Protection Research Unit in Modelling \& Health Economics}
-\affil[4]{Modelling Economics Unit and NIHR Health Protection Research Unit in Modelling \& Health Economics, UK Health Security Agency, London, United Kingdom}
-\affil[5]{MRC Centre for Outbreak Analysis and Modelling, Department of Infectious Disease Epidemiology, School of Public Health, Imperial College London, London, United Kingdom}
-\affil[6]{Chair of Statistical Methods and Econometrics, Karlsruhe Institute of Technology, Karlsruhe, Germany}
-\affil[7]{Computational Statistics Group, Heidelberg Institute for Theoretical Studies, Heidelberg, Germany}
-
-\maketitle
 \thispagestyle{fancy}
@@
 
-\begin{abstract}
-%Abstracts should be up to 300 words and provide a succinct summary of the article.
-Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
-Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.
-
-\paragraph{Background}
-
-In the past, two studies found ensembles of human judgement forecasts of COVID-19 to show predictive performance comparable to ensembles of computational models, at least when predicting case incidences.
-We present a follow-up to a study conducted in Germany and Poland and investigate a novel joint approach to combine human judgement and epidemiological modelling.
-
-\paragraph{Methods}
-
-From May 24th to August 16th, 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
-A median ensemble of all forecasts was submitted to the European Forecast Hub.
-Participants could use two distinct interfaces: in one, forecasters submitted a predictive distribution directly, in the other forecasters instead submitted a forecast of the effective reproduction number $R_t$.
-This was then used to forecast cases and deaths using simulation methods from the \texttt{EpiNow2} \textsf{R} package.
-Forecasts were scored using the weighted interval score on the original forecasts, as well as after applying the natural logarithm to both forecasts and observations.
-
-\paragraph{Results}
-
-The ensemble of human forecasters overall performed comparably to the official European Forecast Hub ensemble on both cases and deaths, although results were sensitive to changes in details of the evaluation.
-$R_t$ forecasts performed comparably to direct forecasts on cases, but worse on deaths.
-Self-identified "experts" outperformed "non-experts" in terms of WIS on the log scale, but not on the natural scale.
-
-\paragraph{Conclusions}
-
-Human judgement forecasts and computational models can produce forecasts of similar quality for infectious disease such as COVID-19.
-The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a ``good" forecast is dependent on the forecast consumer.
-Combinations of human and computational forecasts hold potential but present real-world challenges that need to be solved.
-
-\end{abstract}
-
-\section*{Keywords}
-
-Human judgement forecasting, COVID-19, infectious disease,
-
-% Please list up to eight keywords to help readers interested in your article find it more easily.
-
-\clearpage
-
-\section*{Introduction}
-
-Infectious disease modelling and forecasting has attracted wide-spread attention during the COVID-19 pandemic and helped inform decision making in public health organisations and governments \cite{cramerEvaluationIndividualEnsemble2021, venkatramananUtilityHumanJudgment2022}.
-Most forecasts used to inform decision making were based on computational models of COVID-19, but some authors also explored human judgement forecasting as an alternative or in combination \cite{recchiaHowWellDid2021, mcandrewExpertJudgmentModel2022, bosseComparingHumanModelbased2022, mcandrewChimericForecastingCombining2022}.
-
-Past research found that in the context of infectious disease forecasting, human judgement forecasts could achieve predictive performance broadly comparable to forecasts generated based on mathematical modelling, in particular when forecasting incident cases, rather than lagged indicators indicators like deaths.
-\citet{farrowHumanJudgmentApproach2017} found that an aggregate of human predictions outperformed computational models when predicting the 2014/15 and 2015/16 flu season in the US.
-However, a comparable approach performed worse than computational models at predicting the 2014/15 outbreak of chikungunya in the Americas.
-\citet{bosseComparingHumanModelbased2022} found an ensemble of human forecasters to outperform an ensemble of computational models when predicting cases of COVID-19 in Germany and Poland, but performing worse when predicting incident deaths.
-Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident COVID-19 cases, and worse when predicting incident deaths.
-\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit many participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
-It is important to note that in previous studies (and also this one) human forecasters were free to use any resources, including computational models, in the process of creating a forecast, making it difficult to completely separate human judgement and computational modelling.
-
-In some situations, human judgement forecasting may have advantages relative to computational models.
-Human judgment may be particularly useful to provide timely forecasts in situations where data is sparse and many parameters are hard to quantify.
-Humans are also generally able to answer a broad set of question (such as for example the likelihood that a given actor will take some specified action) and can take factors into account that are hard to encode in a computational model.
-On the other hand, human judgement forecasting is difficult to scale due to the time and effort required, and humans may be at a disadvantage at tasks that strongly benefit from the ability to perform complex computations.
-Also, the use of human judgement forecasts by decision makers may be complicated by the lack of clarity of the basis on which they were made.
-
-Methods that aim to combine human judgement and mathematical modelling are therefore appealing, though we note that presenting this as a binary choice is misleading.
-Most computational models in use in epidemiology have at least some element of human judgement supporting their structure or usage.
-Also, human forecasters often make use of approaches such as calculating a base rate of incidences, or extrapolating current trends, which are in reality equivalent to simple models.
-One explicit method to combine separate human judgement and computational model forecasts with the goal of improving predictive performance is an ensemble.
-This has been shown to improve performance across model types \citep{mcandrewChimericForecastingCombining2022}.
-\citet{farrowHumanJudgmentApproach2017, bosseComparingHumanModelbased2022, swallowChallengesEstimationUncertainty2022} and others suggested additional possibilities in the context of infectious diseases that may also help reduce the amount of human effort required.
-One approach is to use human forecasts, for example of relevant disease parameters, as an input to computational modelling.
-Another approach is to use mathematical modelling in explicit combination with human judgement, for example by giving experts the option to make post-hoc adjustments to model outputs.
-\citet{bosseComparingHumanModelbased2022} proposed asking human forecasters to forecast the effective reproduction number $R_t$ (the average number of people an infected person would infect in turn) based on modelled estimates and to then use this forecast in a mathematical simulation model in order to obtain forecasts for observed case and death numbers.
-
-This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021.
-Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
-All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}.
-This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants.
-In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the estimated effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a simulation model.
-We describe this approach as human in the loop computational modelling and consider it a formalisation of often practiced manual intervention in computational forecasts.
-
-\section*{Methods}
-
-\subsection*{Interaction with the European Forecast Hub}
-
-The European COVID-19 Forecast Hub \cite{sherrattPredictivePerformanceMultimodel2022} was launched in March 2021 in order to elicit weekly predictions for various COVID-19 related forecast targets from different research groups.
-The forecasts evaluated in this study were submitted every Monday before 11.59pm GMT between May 24 2021 and August 16 2021.
-Forecasts were made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
-While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday, and starting on Sunday.
-Forecast horizons were therefore in fact 5, 12, 19 and 26 days.
-Submissions to the European Forecast Hub followed a quantile-based format with 23 quantiles of each output measure at levels 0.01, 0.025, 0.05, 0.10, 0.15,…, 0.95, 0.975, 0.99.
-Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
-Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
-From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
-The median number of models included in the Forecast Hub ensemble for the UK during the study period was 9 for cases and 10 for deaths (see Figure \ref{fig:num-forecasters} in the SI).
-
-Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by the European Forecast Hub and sourced from the Johns Hopkins University (JHU).
-Data were subject to reporting artifacts and revisions.
-All data points were marked as anomalous retrospectively by the European Forecast Hub if in subsequent updates data was changed by more than 5 percent.
-In August 2022 JHU switched the data source for their UK death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency.
-The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
-The data and revisions are displayed in Figure \ref{fig:plot-data}.
-All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).
-
-\subsection*{Human judgement forecasts}
-
-Forecasts of incident cases and deaths linked to COVID-19 in the UK were elicited from individual participants every week through a web application (https://cmmid-lshtm.shinyapps.io/crowd-forecast/) described in \cite{bosseComparingHumanModelbased2022}.
-The application is based on \textsf{R} \cite{R} \texttt{shiny} \cite{shiny} and is available as an \textsf{R} package called \texttt{crowdforecastr} \citep{crowdforecastr}.
-
-When signing up, participants could self-identify as "experts" if they worked in infectious disease modelling or had professional experience in any related field.
-
-The web application offered participants two different ways of making a forecast, called 'direct' (or 'classical') and '$R_t$ forecast'.
-To make a 'direct' forecast (as described in more detail in \cite{bosseComparingHumanModelbased2022}), participants selected a predictive distribution (by default a log-normal distribution) and adjusted the median and width of the distribution to change the central estimate and uncertainty at each forecast horizon.
-
-Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it.
-The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e., as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
-In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
-Figure \ref{fig:screenshot-classical} shows a screenshot of the forecast interface for direct forecasts.
-
-In addition to the 'direct' forecasts, we implemented a second forecasting method ('$R_t$ forecasts'), where we asked participants to make a forecast of the effective reproduction number $R_t$.
-This forecast was made based on a baseline estimate produced by the \texttt{EpiNow2} \citep{epinow2} \textsf{R} \cite{R} package effective reproduction number model which we also used in \cite{bosseComparingHumanModelbased2022} as a standalone computational model.
-The estimate produced by \texttt{EpiNow2} was shown as the default forecast and could be adjusted by the user.
-The resulting $R_t$ forecast was then translated into a forecast of cases using the simulation model from the \texttt{EpiNow2} \textsf{R} package, which implements a renewal equation based \citep{fraserEstimatingIndividualHousehold2007} generative process for latent infections.
-We chose a Gaussian Process prior with mean 0 for the first differences of the effective reproduction number in time, implying that in the absence of informative data the reproduction number would remain constant on average, with uncertainty increasing with the temporal distance to informative data points.
-Latent infections were convolved with delay distributions representing the incubation period and reporting delay, and assumed to follow a negative binomial observation model with a day of the week effect to produce an estimate of reported cases.
-This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
-Further details are given in the SI.
-
-To obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-Further details are given in the SI and in our previous work \citep{bosseComparingHumanModelbased2022}.
-
-As $R_t$-estimates up to at least two weeks prior to the forecast data were uncertain due to their dependence on partially complete observations of underlying infections given the delays from infection to report, we also asked participants to submit an estimate of $R_t$ for the two weeks prior to the current forecast date.
-Participants were therefore asked to estimate/predict six $R_t$ values, four of them beyond the forecast horizon.
-In order to obtain sample trajectories needed as input for the simulation model, we drew 1000 samples from the six provided distributions.
-These samples were ordered and corresponding samples treated as one sample trajectory.
-Samples for daily values were obtained by linearly interpolating between weekly samples.
-
-Upon pressing a button, participants could see a preview of the evolution of cases implied by their current $R_t$ forecast.
-However, due to lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
-Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
-
-Every week, we submitted an ensemble of individual forecasts to the European Forecast Hub.
-In contrast to the ensemble of human forecasts described in \citet{bosseComparingHumanModelbased2022}, we used the quantile-wise median, rather than the quantile-wise mean to combine predictions, drawing upon insights gained from the COVID-19 Forecast Hubs \citep{rayComparingTrainedUntrained2022}.
-We submitted three different ensembles to the Hub: The first one, "epiforecasts-EpiExpert\_direct" (here called "direct crowd forecast" or "crowd-direct") was a quantile-wise median ensemble of all the direct forecasts.
-"epiforecasts-EpiExpert\_Rt" (here called "$R_t$ forecast" or "crowd-rt") was a median ensemble of all forecasts made through the $R_t$ interface.
-"epiforecasts-EpiExpert" (here called "combined crowd ensemble" or "crowd-ensemble") was a median ensemble of all forecasts together.
-A participant could enter the combined crowd ensemble twice if they had submitted both a direct and an $R_t$ forecast.
-Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
-
-\subsection*{The UK Crowd Forecasting Challenge}
-
-To boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
-Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
-Everyone who had submitted a forecast for targets in the UK during the tournament period from the 24th of May 2021 to the 16th of August 2021 was deemed a participant and eligible for a prize.
-The first prize was 100 GBP, second prize 50 GBP and third prize 25 GBP.
-Participant performance was determined using the mean weighted interval score (WIS) on the log scale (see details in the next Section), %\ref{sec:analysis}),
-averaged across forecast dates, horizons and forecast targets.
-For the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
-The UK crowd forecasting challenge was announced over Twitter and our networks.
-% (e.g. https://www.voice-global.org/public/opportunities/archived/uk-covid-19-crowd-forecasting-challenge/)
-In addition, we created a project website, \url{https://crowdforecastr.org}, made weekly posts on Twitter and sent participants who had registered on the online application weekly emails with a reminder and a summary of their past performance.
-A public leaderboard was available on our website \url{https://epiforecasts.io}.
-Participants could choose to make a direct forecast as well as an $R_t$ forecast and were counted as two separate forecasters and eligible for prizes twice.
-Weekly forecasts had to be submitted between Sunday 12pm and Monday 8pm UK time.
-
-\subsection*{Analysis}
-\label{sec:analysis}
-
-We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
-\cite{bracherEvaluatingEpidemicForecasts2021}.
-The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
-A forecaster, in expectation, optimises their score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report their true belief.
-The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
-The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e., uncertainty of forecasts), overprediction and underprediction.
-
-\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS to better reflect the exponential nature of the underlying disease process.
-We, therefore, also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
-In the following, we refer to WIS scores obtained without a transformation as "scores on the natural scale", and WIS values obtained after log-transforming forecasts and observations as "scores on the log scale".
-To make scores easier to interpret, we report relative WIS scores, where the average score for a given model was divided by the average score for the European Forecast Hub ensemble ("EuroCOVIDhub-ensemble").
-In addition, we computed ranks based on WIS values.
-
-In order to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}, we used the empirical coverage of all central 50\% and 90\% prediction intervals.
-Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g., the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
-
-If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021}.
-We analysed all forecasts stratified by forecast target (cases or deaths), forecast horizon, and forecast approach.
-We compared the performance of the direct vs.
-$R_t$ forecasting approach using instances where we had both a direct forecasts and an $R_t$ forecast from the same person.
-
-For self-reported "experts" and "non-experts", a simple comparison of scores would be confounded by individual differences in participation and the timing of individual forecasts.
-We therefore compared the performance of self-reported "experts" vs.
-"non-experts" by creating and evaluating two modified median ensembles, one including only "experts" and the other only "non-experts".
-
-% Those forecasts were combined using a quantile-wise median separately for direct and $R_t$ forecasts such that we obtained a modified direct human ensemble and a modified $R_t$ ensemble that included only corresponding forecasts from the same forecaster.
-
-Forecasts were evaluated using the \texttt{scoringutils} \citep{bosseEvaluatingForecastsScoringutils2022} package in \textsf{R}.
-All code and data used for this analysis, including individual-level forecasting data is available at \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
-All code used to submit the forecasts to the European Forecast Hub is available at \url{https://github.com/epiforecasts/europe-covid-forecast}.
-
-\subsection*{Ethics statement}
-This study has been approved by the London School of Hygiene \& Tropical Medicine Research Ethics Committee (reference number 22290). Consent from participants was obtained in written form.
-
-\section*{Results}
-
-\subsection*{Crowd forecast participation}
-
-A total number of 90 participants submitted forecasts (more precisely, forecasts were submitted from 90 different accounts, some of them anonymous).
-Out of 90 participants, 21 self-identified as "experts", i.e., stated they had professional experience in infectious disease modelling or a related field.
-
-The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51.
-This was higher than the number of participants in \cite{bosseComparingHumanModelbased2022} (which had a median number of 6, a minimum of 2, and a maximum 10).
-With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13.
-The median number of submissions from a single individual was one, meaning that similar to \citep{bosseComparingHumanModelbased2022} most forecasters dropped out after their first submission.
-Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study (S.
-Abbott).
-The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A).
-The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths) (see Figure \ref{fig:num-forecasters}B).
-
-\input performance-table.tex
-
-\begin{figure*}[ht]
-%\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance.png}
-\caption{\bf{Predictive performance across forecast horizons.} A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
-E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/scores-and-forecasts.png}
-\caption{\bf{Forecasts and corresponding WIS for 2-week ahead forecasts of cases and deaths from COVID-19 in the UK.} A: 50\% prediction intervals (coloured bars) and observed values (black line and points) for cases and deaths on the natural scale.
-B: Corresponding WIS values, decomposed into dispersion, overprediction and underprediction.
-C: 50\% prediction intervals on the log scale, i.e., after applying the natural logarithm to all forecasts and observations.
-D: Corresponding WIS on the log scale, i.e., the WIS applied to the log-transformed forecasts and observations.}
-\label{fig:forecasts-scores} 
-\end{figure*}
-
-\subsection*{Case forecasts}
-
-All forecasting approaches exhibited underdispersion when predicting cases, meaning that forecasts on average were too narrow and not uncertain enough.
-Empirical coverage for case forecasts was below nominal coverage for all forecasting approaches for forecasts more than one week into the future (see Figure \ref{fig:performance}E,F).
-For 50\% prediction intervals, empirical coverage was worst for the direct crowd forecasts (0.31), best for the $R_t$ forecasts (0.46) and in between for the Hub ensemble and the crowd ensemble (both 0.38, see Table \ref{tab:scores}).
-For 90\% prediction intervals, coverage was worst for the $R_t$ forecasts (0.62) and slightly better for the other approaches (all 0.69).
-Coverage for all forecasts deteriorated further with increasing forecast horizon (see Figure \ref{fig:performance}E,F).
-
-In terms of WIS on the log scale, all human forecasting approaches outperformed the Forecast Hub ensemble for two week ahead forecasts of cases (see Figure \ref{fig:performance}).
-WIS values relative to the Hub ensemble (=1) were 0.91 for the combined crowd ensemble, 0.96 for the direct crowd forecasts and 0.93 for the $R_t$ forecasts (see Table \ref{tab:scores}).
-In contrast, in terms of WIS on the natural scale the Hub ensemble outperformed all human forecasting approaches.
-Relative WIS values on the natural scale for two week ahead forecasts were 1.05 for the combined crowd ensemble, 1.03 for the direct crowd forecasts and 1.2 for the $R_t$ forecasts.
-The discrepancy between performance on the log and natural scale  can be attributed to case forecasts from the Hub ensemble tending to be lower than forecasts from human judgement approaches (see Figure \ref{fig:forecasts-scores}).
-On the natural scale, this resulted in smaller overprediction penalties, putting it ahead of human forecasts (see Figure \ref{fig:performance}A,C).
-On the log scale, however, it led to large penalties for underprediction.
-
-Performance of the Hub ensemble relative to the human forecasting approaches improved with increasing forecast horizon (see Figure \ref{fig:performance}).
-For a four-week-ahead forecast horizon, the Hub ensemble outperformed all other approaches both on the log scale (rel.
-WIS values the human forecasts of 1.02, 1.05, 1.06) and on the natural scale (rel.
-WIS values of 1.21, 1.25, 1.3) (see Table \ref{tab:scores-4}).
-
-In terms of relative model ranks for two week ahead forecasts, the Hub ensemble and the $R_t$ forecast showed a higher variance than the combined crowd ensemble and the direct forecasts (See Figure \ref{fig:performance-ranks}), despite forecasts being about the same or more dispersed (see Figure \ref{fig:performance}).
-Both the Hub ensemble and the $R_t$ forecast were more often in first place than other approaches (4 times each, both on the log and on the natural scale).
-However, they were also most often in the last place (Hub ensemble: 6 on the log scale and 5 on the natural scale, $R_t$: 5 on the log scale and 6 on the natural scale).
-The direct forecasts placed relatively equally in places 1-4.
-The crowd ensemble never placed fourth, but also had the lowest number of first places (2, both on the log and the natural scale).
-Aggregated model ranks only changed marginally when switching between the log and the natural scale (see Figure \ref{fig:performance-ranks}.
-
-When comparing WIS values on the log scale with those on the natural scale, scores were more equally distributed across the study period on the log scale and more weight was given to forecasts in June and July which underpredicted the extent to which case number would rise (see Figure \ref{fig:forecasts-scores}).
-On the natural scale, the WIS as a measure of the absolute distance between forecast and observation increased or decreased with the magnitude of the forecast target \cite{bosseTransformationForecastsEvaluating2023, bracherEvaluatingEpidemicForecasts2021}.
-Average scores were therefore dominated by performance around the peak when cases were highest, in particular by forecasts made on the 19th of July for the 31st of July (see Figure \ref{fig:forecasts-scores}).
-For all forecasting approaches, overprediction was the largest contributor to overall scores (see Figure \ref{fig:performance}A).
-On the log scale, underprediction played a larger role (see Figure \ref{fig:performance}C).
-Switching between scores on the log and on the natural scale had the strongest effect on the $R_t$ forecasts, which had a relative WIS value of 0.96 on the log scale and 1.2 on the natural scale.
-The $R_t$ forecasts tended to be higher than both the direct forecasts and the Forecast Hub ensemble, especially around the peak, leading to high scores on the natural scale, but not on the log scale.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks.png}
-\caption{\bf{Ranks for all forecasting approaches for two week ahead forecasts}. Colours indicate how often (out of 13 forecasts) a given approach got 1st, 2nd, 3rd, or 4th rank.}
-\label{fig:performance-ranks}
-\end{figure*}
-
-\subsection*{Death forecasts}
-
-All forecasting approaches except the $R_t$ forecasts showed higher empirical coverage for deaths than for cases (see Figure \ref{fig:performance}).
-Forecasts from the Hub ensemble generally tended to be wider than the human forecasts (see Figures \ref{fig:forecasts-scores} and \ref{fig:performance}B,D).
-For 50\% prediction intervals, the Hub ensemble exceeded the nominal coverage noticeably (0.77) (see Table \ref{tab:scores}).
-$R_t$ forecasts failed to get close to nominal coverage (0.15), while the combined crowd ensemble and the direct forecasts had empirical coverage close to nominal coverage (both 0.54).
-For 90\% prediction intervals, the Hub ensemble again exceeded nominal coverage and covered all observations (1) while the $R_t$ forecasts again failed to get close to nominal coverage (0.46).
-The crowd ensemble exhibited some underdispersion (0.77) while the direct forcecasts almost reached nominal coverage for two week ahead forecasts of deaths (0.85).
-
-In terms of WIS on the log scale for two week ahead predictions of deaths, the combined crowd ensemble (0.97) and the direct crowd forecasts (0.99) were marginally ahead of the Hub ensemble, while the $R_t$ forecasts performed noticeably worse (1.98) (see Figure \ref{fig:performance}D and Table \ref{tab:scores}).
-For the Hub ensemble, the dispersion component played by far the largest role, while this was less the case for the human forecasts, which got higher penalties from both over- and underprediction.
-Combining the $R_t$ forecasts and the direct forecasts led to an ensemble that performed better than either of them alone on the log scale despite the poor overall performance of the $R_t$ forecasts.
-In terms of WIS on the natural scale, only the direct forecasts (0.89) performed better for two week ahead death predictions than the Hub ensemble, while the combined crowd ensemble performed slightly worse (1.06) and the $R_t$ forecasts again noticeably worse (2.1).
-
-In terms of relative model ranks for two week ahead death forecasts, the $R_t$ forecasts took the fourth place most often (9 on the log scale and 10 on the natural scale), while the direct forecasts placed first most often (5 on the log scale and 6 on the natural scale, see Figure \ref{fig:performance-ranks}).
-Again, the crowd ensemble never placed fourth.
-
-When comparing scores on the log and on the natural scale, scores on the log scale were again more evenly distributed across the study period.
-On the natural scale, high scores were concentrated around the end of the study period, when death incidences were highest (see Figure \ref{fig:forecasts-scores}).
-
-\subsection*{$R_t$ forecasts}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/comparison-direct-rt-individual.png}
-\caption{\bf{Comparison of predictive performance of individual forecasters using either the direct forecasting or $R_t$ interface}.
-Comparisons are based only on those instances where forecasters have submitted a prediction using both interfaces.
-The absolute level for a given forecaster relative to others is not meaningful as forecasters differ in the amounts of forecasts they have submitted and when.}
-\label{fig:comparison-direct-rt-individual}
-\end{figure*}
-For cases, where participants could observe the case forecast implied by their $R_t$ forecast, predictive performance was similar between corresponding direct and $R_t$ forecasts for most forecasters who had submitted both (see Figure \ref{fig:comparison-direct-rt-individual}).
-For deaths, where forecasters could not see the incidence forecast implied by their $R_t$ forecast or manually adjust the case fatality rate, performance of the $R_t$ forecasts was significantly worse.
-From June to the end of July, $R_t$ forecasts overpredicted deaths and were noticeable higher than other forecasts, whereas in August, $R_t$ forecasts underpredicted deaths and were substantially lower than other forecasts (see Figure \ref{fig:forecasts-scores}).
-In particular, $R_t$ forecasts for deaths were worse than the corresponding direct death forecasts for most forecasters (see Figure \ref{fig:comparison-direct-rt-individual}).
-Changing from the direct forecasting method to $R_t$ forecasting for cases tended to improve scores for better forecasters and decrease scores for worse forecasters, although sample sizes and the size of the observed effect are both small.
-
-Combining direct crowd forecasts and $R_t$ forecasts improved performance on the log scale compared to both direct and $R_t$ forecasts alone across all horizons and target types.
-This was not the case on the natural scale, where direct forecasts performed better than the $R_t$ and the direct forecasts for both cases and deaths across most horizons.
-Only for case forecasts four weeks ahead on the natural scale was the combined ensemble better than the direct forecasts.
-However, even on the natural scale, performance of the combined ensemble was better than the average of the WIS of direct and $R_t$ forecasts.
-
-%This suggests that the convolution from latent infections implied by the $R_t$ forecasts to deaths did not capture the relationship between cases and deaths accurately in this instance despite performing well in other settings \citep{meakinComparativeAssessmentMethods2022}.
-This may be related to changes in the age decomposition of cases and the change in the case fatality ratio (CFR) over the course of the study period due to the rise of the Delta variant starting in May 2021 in the UK.
-This change was not accounted for in the simple model used and the CFR could not be adjusted by forecasters.
-
-\subsection*{Experts and non-experts}
-
-% Mention number of experts and non-experts
-
-A median ensemble of two week ahead forecasts restricted to only those made by either "experts" or "non-experts" (determined based on self-reported experience in infectious disease modelling or a related field) performed worse than the combined crowd example, both for cases and deaths and both on the log scale and on the natural scale (see Figure \ref{fig:performance-experts} and Table \ref{tab:scores-experts} and Figure \ref{fig:num-forecasters}B for a visualisation of participation).
-The median number of "non-experts" was 11 for cases and 10 for deaths, which was higher than the median number of "experts", which was 8 for cases and deaths.
-
-When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}.
-For cases, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
-WIS values \textit{relative to the combined crowd ensemble} were 1.08 for "experts" and 1.14 for "non-experts" on the log scale and 1.06 for "experts" and 1.07 for "non-experts" on the natural scale (see Table \ref{tab:scores-experts}).
-For deaths, "experts" performed worse than "non-experts" in terms of WIS on the log scale (WIS relative to the combined crowd ensemble: 1.29 vs.
-1.06), but better on the natural scale (1.03 vs.
-1.14).
-Both the "expert"- and the "non-expert"-ensemble had similar proportions of $R_t$ forecasts (mean of 32\% for "experts" and 32.2\% for "non-experts" across cases and deaths together).
-
-For four weeks ahead forecasts of cases, the combined ensemble outperformed both "experts" and "non-experts" on the log scale as well as on the natural scale.
-"Experts" performed better than "non-experts" both on the log scale (WIS value relative to the combined crowd ensemble of 1.08 for "experts" vs.
-1.21 for "non-experts") and on the natural scale (1.04 vs.
-1.07).
-For four week ahead forecasts of deaths, "Experts" performed better than "Non-experts" on the log scale (1.17 vs.
-1.18) as well as on the natural scale (0.95 vs.
-1.15).
-
-\section*{Discussion}
-% The discussion should include the implications of the article results in view of prior work in this field.
-
-% \paragraph{Summary}
-In this paper, we presented a follow-up study to \cite{bosseComparingHumanModelbased2022}, analysing human judgement forecasts of cases of and deaths from COVID-19 in the United Kingdom submitted to the European COVID-19 Forecast Hub between the 24th of May and the 16th of August 2021.
-Human judgement forecasts were generated using two different forecasting approaches, a) direct forecasts of cases and deaths and b) forecasts of the effective reproduction number $R_t$, which were based on estimates from an open source effective reproduction number estimation model and also relied on this model, along with a second model relating cases and deaths from the same source, to simulate reported cases and deaths.
-
-% \paragraph{Effort to recruit participants}. 
-Just like \citet{bosseComparingHumanModelbased2022} and \citet{farrowHumanJudgmentApproach2017}, this study struggled to retain a large number of participants.
-Focused public outreach efforts such as creating a dedicated website, announcing an official tournament, providing a public leaderboard, sending weekly emails with details on past performance and weekly announcements on Twitter, did noticeably increase participation compared to the previous study in Germany and Poland.
-Nevertheless, retaining participants beyond the initial recruitment proved challenging, and most forecasters only submitted a single forecast.
-\citet{mcandrewChimericForecastingCombining2022} had a higher number of participants, suggesting that making use of existing forecasting platforms that have access to a large existing user base and greater resources may be helpful in recruiting a larger number of participants, though these platforms lack the flexibility and software tooling to run a novel study of this kind in real-time as things stand.
-
-%Calibration
-In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022}, we found almost all forecasts for cases to be underdispersed (i.e., too narrow/overconfident).
-Empirical coverage for death forecasts was higher than the corresponding coverage for cases for all forecasting approaches except the $R_t$ forecasts.
-
-%WIS scores
-For forecasts of cases two weeks ahead, performance of the human judgement forecasts was better than the European Forecast Hub ensemble in terms of WIS on the log scale, and worse in terms of WIS on the natural scale.
-This was linked to a tendency of the Hub ensemble to make lower case predictions, which led to lower overprediction penalties on the natural scale, but noticeably higher underprediction penalties on the log scale.
-For forecasts of deaths two weeks ahead, direct human forecasts and the combined crowd ensemble performed better than the Hub ensemble on the log scale.
-On the natural scale, the combined crowd ensemble performed worse than the Hub ensemble, while the direct crowd forecasts still performed better.
-$R_t$ forecasts for deaths performed noticeably worse than all other approaches both on the log and on the natural scale.
-
-In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths.
-They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators.
-\citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19.
-Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it.
-In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale.
-Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths.
-During the study period, the case fatality ratio (CFR) changed quite quickly.
-This was related to the rise of the Delta variant in the UK in the beginning of May with its higher case fatality rate \citep{bastIncreasedRiskHospitalisation2021}.
-In addition, the age distribution of cases changed over the course of the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
-One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update .
-A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.
-
-% \paragraph{Rt forecasting}
-
-This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths.
-One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour.
-Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators.
-This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$.
-Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions.
-Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
-In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more.
-Sample sizes, however, were very low.
-Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
-$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
-The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon.
-However, as mentioned before, the CFR evolved during the study period.
-Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
-Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
-Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period.
-Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters.
-Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts.
-One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths.
-Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
-
-%Combining Rt and direct
-Combining forecasts from "experts" and "non-experts" led to better performance for forecasts two weeks ahead for cases as well as deaths, and both on the log scale and on the natural scale.
-Combining direct forecasts and $R_t$ forecasts led to better performance on the log scale, but not on the natural scale.
-This suggests that combining different forecasts can be beneficial in many instances, although there may be differences in terms of WIS on the log and the natural scale.
-In particular, WIS values on the natural scale may be more susceptible to models that would tend to overshoot and miss the peak, while WIS on the log scale may be more affected by models that underpredict and miss upswings \citep{bosseTransformationForecastsEvaluating2023}.
-
-% \paragraph{Expertise}
-
-Past expert forecasts of COVID-19 \cite{recchiaHowWellDid2021} had found predictions from experts to outperform those of non-experts.
-In our study, an ensemble of self-reported "experts" outperformed an ensemble of "non-experts" when forecasting cases two weeks ahead, both on the log scale and on the natural scale.
-When forecasting deaths two weeks ahead, "experts" performed worse than "non-experts" on the log scale, but better on the natural scale.
-Forecasts for "experts" tended to be better calibrated than non-experts.
-However results should be taken with care considering relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths) and given that expert status was self-reported.
-Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
-However, as we used ad-hoc recruitment in our networks many of these self-identified experts are likely to be infectious disease modellers.
-
-% \paragraph{Susceptibility of results}
-Overall, results of our study should be taken with caution due to several important limitations.
-Firstly, our study was restricted to one location and to a relatively short period of thirteen weeks.
-Secondly, there were many confounding factors that likely influence results.
-These include the fact that different participants made forecasts at different points in time (with the median forecaster only submitting a single forecast) and that subgroups of interest (e.g.
-"experts", or $R_t$ forecasts) had different numbers of forecasters.
-In most instances, differences in scores between forecast approaches were small compared to the variance of scores within a single approach.
-In addition, there were many researcher degrees of freedom that could influence findings, for example how individual forecasts were combined to create an ensemble.
-Results were influenced by choices made during the evaluation with, for example, some conclusions  depending on forecast horizon and the transformation used prior to scoring.
-Highlighting this, prizes to the human forecasters were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
-Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and pay-outs would have been different.
-
-\section*{Conclusions}
-% Please state what you think are the main conclusions that can be realistically drawn from the findings in the paper, taking care not to make claims that cannot be supported.
-
-The results of our study are broadly consistent with previous studies on human judgement forecasting of COVID-19 and suggest that human crowd ensembles and an ensemble of computational models are able to produce forecasts of similar quality.
-One interpretation of these findings is that a mixed crowd of human forecaster can produce a viable alternative or complement to an ensemble of mathematical models created by experts.
-An altnerative interpretation is that an ensemble of automated models can produce forecasts over the course of several years that are on par with that of an engaged crowd of human forecasters.
-This study, and all previous studies, comparing human judgement forecasts and computational models only ran over short periods of time and the majority of them struggled with recruitment and upkeep.
-Meanwhile, COVID-19 Forecast Hubs have attracted continuous submissions for almost three years and were able to consistently provide forecasts of comparable quality.
-
-Our findings do not suggest that humans are necessarily at a general disadvantage compared to computational models at predicting reported deaths, but evidence in both directions is limited and this is made particularly complex as our study took place during a period of time when CFR estimates were changing rapidly.
-Despite evaluations being public, it remains a challenge to properly incentivise contributors to Forecast Hubs to regularly update their forecasting methodology in order to maximise utility, predictive performance, or both.
-Combining human judgement and epidemiological modelling by mapping $R_t$ forecasts to case and death numbers has not yielded competitive forecasts for deaths in this study.
-However, we only presented a prototype of a forecasting approach, which, while having appealing properties, proved challenging to implement.
-Subsequent iterations and improvements could likely achieve better results.
-More research is required to obtain a better understanding of the role of subject matter expertise in infectious disease forecasting.
-Our results underline that it is difficult to evaluate forecast performance devoid of context that helps inform what a good or a bad forecast is.
-Different ways to look at the data let different forecasts appear better or worse.
-Forecast evaluation therefore either needs to be clearly informed by the needs of forecast consumers to determine what a good forecast is, or it needs a broad array of perspectives to provide a wholistic picture as we have attempted to present in this work.
-Furthermore, evaluating forecasts post-hoc leaves the researchers with many degrees of freedom to make decisions that affect which models look good and there is a risk of allowing for motivated reasoning.
-More emphasis should be put on measures that prevent this, e.g.
-by establishing common standards for evaluations, pre-registering studies, and making it a norm to display a variety of standard metrics.
-
-\subsection*{Author contributions}
-NIB contributed to the conceptualization, data curation, formal analysis, investigation, methodology, software development, validation, visualization, and original draft preparation of the manuscript, as well as its review and editing.
-SA contributed to the conceptualization, data curation, investigation, methodology, software development, supervision, and review and editing of the manuscript.
-JB contributed to the supervision, review, and editing of the manuscript.
-EvL contributed to the conceptualization, supervision, and review and editing of the manuscript.
-AC contributed to the conceptualization, supervision, and review and editing of the manuscript.
-SF contributed to the conceptualization, funding acquisition, project administration, supervision, and review and editing of the manuscript.
-All authors have read and approved the final version of the manuscript.
-
-\subsection*{Competing interests}
-NIB is a part-time employee of Metaculus, an online prediction platform.
-
-\subsection*{Grant information}
-NIB received funding from the Health Protection Research Unit (grant code NIHR200908).
-SA's work was funded by the Wellcome Trust (grant: 210758/Z/18/Z).
-AC's acknowledges funding by the NIHR, the Sergei Brin foundation, USAID, and support by the Academy of Medical Sciences Springboard scheme, funded by the AMS, Wellcome Trust, BEIS, the British Heart Foundation and Diabetes UK (REF:SBF005\\1044).
-EvL acknowledges funding by the National Institute for Health Research (NIHR) Health Protection Research Unit (HPRU) in Modelling and Health Economics (grant number NIHR200908) and the European Union's Horizon 2020 research and innovation programme - project EpiPose (101003688).
-The work of JB was supported by the Helmholtz Foundation (https://www.helmholtz.de/) via the SIMCARD Information and Data Science Pilot Project.
-SF's work was supported by the Wellcome Trust (grant: 210758/Z/18/Z).
-This work was supported by the National Institute for Health and Care Research (NIHR) Health Protection Research Unit (HPRU) in Modelling and Health Economics, which is a partnership between the UK Health Security Agency (UKHSA), Imperial College London, and the London School of Hygiene \& Tropical Medicine (grant code NIHR200908).
-The views expressed are those of the authors and not necessarily those of the UK Department of Health and Social Care (DHSC), NIHR, or UKHSA.
-
-\subsection*{Acknowledgements}
-We thank all forecasters for their participation and want to congratulate the three winners of the forecasting challenge: Russell Bradshaw, Sebastian Funk (an author of this study), and Akira Endo.
-All winners donated their prizes.
-
-\subsection*{Data availability}
-All data and ode are available publicy under a MIT license under \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge}.
-
-% ================================= References =============================== %
-\clearpage
-
-{\small\bibliographystyle{unsrtnat}
-\bibliography{software, uk-forecasting-challenge}}
-
-\bigskip
-\clearpage
-
-% =================================== Appendix =============================== %
-
-% Make sure that Appendix tables and Figures are separate
-% https://tex.stackexchange.com/questions/520193/is-it-possible-to-except-figures-in-the-appendix-from-being-repositioned-by-endf
-%\processdelayedfloats
-\csname efloat@restorefloats\endcsname
-
-\appendix
 \section*{Supplementary information}
@@
 
+\clearpage
+{\small\bibliographystyle{unsrtnat}
+\bibliography{software, uk-forecasting-challenge}}
+
 \end{document}

### 2023-05-04 manuscript/manuscript.tex
small updates to manuscript

@@
 The data has been published separately here: \url{https://doi.org/10.5281/zenodo.7897289}.
-Supplementary information \cite{bosseSupplementaryInformationHuman2023} to this manuscript is available at \url{https://doi.org/10.5281/zenodo.7897513}.
+Supplementary Information \cite{bosseSupplementaryInformationHuman2023} to this manuscript is available at \url{https://doi.org/10.5281/zenodo.7897513}.
 
-% ================================= References =============================== %
 \clearpage
-
 {\small\bibliographystyle{unsrtnat}
@@
 
-\bigskip
-\clearpage
+% ================================= References =============================== %
 
-% =================================== Appendix =============================== %
+% \bigskip
+% \clearpage
 
-% Make sure that Appendix tables and Figures are separate
-% https://tex.stackexchange.com/questions/520193/is-it-possible-to-except-figures-in-the-appendix-from-being-repositioned-by-endf
-%\processdelayedfloats
-\csname efloat@restorefloats\endcsname
+% % =================================== Appendix =============================== %
 
-\appendix
-\section*{Supplementary information}
-\renewcommand{\thefigure}{SI.\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{SI.\arabic{table}} \setcounter{table}{0}
+% % Make sure that Appendix tables and Figures are separate
+% % https://tex.stackexchange.com/questions/520193/is-it-possible-to-except-figures-in-the-appendix-from-being-repositioned-by-endf
+% %\processdelayedfloats
+% \csname efloat@restorefloats\endcsname
 
-\subsection*{Weighted interval score}
-\label{sec:wis}
+% \appendix
+% \section*{Supplementary information}
+% \renewcommand{\thefigure}{SI.\arabic{figure}}
+% \setcounter{figure}{0}
+% \renewcommand{\thetable}{SI.\arabic{table}} \setcounter{table}{0}
 
-The weighted interval score (smaller values are better) is a proper scoring rule for quantile forecasts.
+% \subsection*{Weighted interval score}
+% \label{sec:wis}
+
+% The weighted interval score (smaller values are better) is a proper scoring rule for quantile forecasts.
 It converges to the continuous ranked probability score (which itself is a generalisation of the absolute error to probabilistic forecasts) for an increasing number of intervals.
@@
 For a single interval, the score is computed as
-  $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
-  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
+%   $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
+%   where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
 For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,
-  $$WIS = \frac{1}{K + 0.5} \cdot \left( w_0 \cdot |y - m| + \sum_{k = 1}^{K} w_k \cdot IS_{\alpha}(F, y) \right), $$
-  where $w_k$ is a weight for every interval. Usually, $w_k = \frac{\alpha_k}{2}$ and $w_0 = 0.5$.
+%   $$WIS = \frac{1}{K + 0.5} \cdot \left( w_0 \cdot |y - m| + \sum_{k = 1}^{K} w_k \cdot IS_{\alpha}(F, y) \right), $$
+%   where $w_k$ is a weight for every interval. Usually, $w_k = \frac{\alpha_k}{2}$ and $w_0 = 0.5$.
 
-\input SI/epinow2.tex
+% \input SI/epinow2.tex
 
-%table 4 weeks
+% %table 4 weeks
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
-\caption{\bf{Number of forecasts across the study period.} A: number of forecasts included in the Hub ensemble and the combined crowd ensemble.
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
+% \caption{\bf{Number of forecasts across the study period.} A: number of forecasts included in the Hub ensemble and the combined crowd ensemble.
 B: number of forecasts by "experts" and "non-experts".
 Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
-\label{fig:num-forecasters-SI}
-\end{figure*}
+% \label{fig:num-forecasters-SI}
+% \end{figure*}
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/plot-data.png}
-\caption{\bf{Observed cases and deaths of COVID-19 in the UK}.
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/plot-data.png}
+% \caption{\bf{Observed cases and deaths of COVID-19 in the UK}.
 A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021.
@@
 C: Difference between the original and revised weekly death numbers.}
-\label{fig:plot-data}
-\end{figure*}
+% \label{fig:plot-data}
+% \end{figure*}
 
-\input performance-table-horizon-4.tex
+% \input performance-table-horizon-4.tex
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-classical.png}
-\caption{\bf{Screenshot of the direct forecasting interface.}}
-\label{fig:screenshot-classical}
-\end{figure*}
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-classical.png}
+% \caption{\bf{Screenshot of the direct forecasting interface.}}
+% \label{fig:screenshot-classical}
+% \end{figure*}
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-rt-app.png}
-\caption{\bf{Screenshot of the $R_t$ forecasting interface.}}
-\label{fig:screenshot-rt}
-\end{figure*}
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/screenshot-crowd-rt-app.png}
+% \caption{\bf{Screenshot of the $R_t$ forecasting interface.}}
+% \label{fig:screenshot-rt}
+% \end{figure*}
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks-4.png}
-\caption{\bf{Ranks for all forecasting approaches for four week ahead forecasts}. Colours indicate how often (out of 13 forecasts) a given approach got 1st, 2nd, 3rd, or 4th rank.}
-\label{fig:performance-ranks-4}
-\end{figure*}
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/performance-ranks-4.png}
+% \caption{\bf{Ranks for all forecasting approaches for four week ahead forecasts}. Colours indicate how often (out of 13 forecasts) a given approach got 1st, 2nd, 3rd, or 4th rank.}
+% \label{fig:performance-ranks-4}
+% \end{figure*}
 
-\clearpage
+% \clearpage
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
-\caption{\bf{Predictive performance of self-reported "experts" and "non-experts" across forecast horizons.} Forecasts from "experts" and "non-experts" were combined to two separate median ensembles, including both direct and $R_t$ forecasts.
+% \begin{figure*}
+% \centering
+% \includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
+% \caption{\bf{Predictive performance of self-reported "experts" and "non-experts" across forecast horizons.} Forecasts from "experts" and "non-experts" were combined to two separate median ensembles, including both direct and $R_t$ forecasts.
 A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
 E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance-experts-SI}
-\end{figure*}
+% \label{fig:performance-experts-SI}
+% \end{figure*}
 
-%\input{manuscript/performance-table-experts}
+% %\input{manuscript/performance-table-experts}
 
-\input{manuscript/performance-table-experts-4}
+% \input{manuscript/performance-table-experts-4}
 
-% See this guide for more information on BibTeX:
-% http://libguides.mit.edu/content.php?pid=55482&sid=406343
+% % See this guide for more information on BibTeX:
+% % http://libguides.mit.edu/content.php?pid=55482&sid=406343
 
-% For more author guidance please see:
-% http://wellcomeopenresearch.org/for-authors/article-guidelines
+% % For more author guidance please see:
+% % http://wellcomeopenresearch.org/for-authors/article-guidelines
 
-% When all authors are happy with the paper, use the 
-% ‘Submit to WELLCOME OPEN RESEARCH' button from the menu above
-% to submit directly to the open life science journal Wellcome Open Research.
+% % When all authors are happy with the paper, use the 
+% % ‘Submit to WELLCOME OPEN RESEARCH' button from the menu above
+% % to submit directly to the open life science journal Wellcome Open Research.
 
-% Please note that this template results in a draft pre-submission PDF document.
-% Articles will be professionally typeset when accepted for publication.
+% % Please note that this template results in a draft pre-submission PDF document.
+% % Articles will be professionally typeset when accepted for publication.
 
-% We hope you find the Wellcome Open Research Overleaf template useful,
-% please let us know if you have any feedback using the help menu above.
+% % We hope you find the Wellcome Open Research Overleaf template useful,
+% % please let us know if you have any feedback using the help menu above.
 

### 2023-05-08 manuscript/manuscript.tex
small corrections to WIS

@@
 
-The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
+The WIS is a proper scoring rule yielding non-negative values, with smaller values implying better performance.
 A forecaster, in expectation, optimises their score by providing a predictive distribution $F$ that is equal to the data-generating distribution $G$, and is therefore incentivised to report their true belief.
-The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
-The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e., uncertainty of forecasts), overprediction and underprediction.
+The WIS can be understood as an approximation of the continuous ranked probability score (CRPS, \citealt{gneitingProbabilisticForecastsCalibration2007}) for forecasts in a quantile-based format.
+The CRPS, in turn, represents a generalisation of the absolute error to predictive distributions.
+The WIS can be decomposed into three separate penalty components (corresponding to the three terms in the definition of the interval score): forecast dispersion (i.e., uncertainty of forecasts), overprediction and underprediction.
 

### 2023-06-20 Analysis.rmd
sort code, add numbered figures

@@
 
-# Figure 1: Predictive performance across horizons (summary)
+# Figure 1: Number of forecasters across the study period
 ================================================================================
-
-```{r}
-
-plot_performance <- function(scores) {
-  df <- scores |>
-    filter(scale %in% c("log", "natural")) |>
-    summarise_scores(by = c("horizon", "model", 
-                            "target_type", "scale", "type_and_scale")) |>
-    rename("Dispersion" = dispersion, 
-           "Overprediction" = overprediction, 
-           "Underprediction" = underprediction, 
-           Model = model) |>
-    pivot_longer(names_to = "WIS components",
-                 cols = c(`Dispersion`, `Overprediction`, `Underprediction`)) |>
-    mutate(`WIS components` = fct_relevel(`WIS components`, 
-                                          c("Dispersion", "Underprediction", 
-                                            "Overprediction")))
-
-  component_plot <- function(df, scale_filter, type_filter) {
-    df |>
-      filter(scale == scale_filter) |>
-      filter(target_type == type_filter) |>
-      ggplot(aes(x = Model, 
-                 # x = reorder(Model, (interval_score)), 
-                 y = value, fill = Model, 
-                 group = Model, alpha = `WIS components`)) + 
-      scale_alpha_manual(values = c(1, 0.1, 0.6)) +
-      # scale_fill_manual(values = vals) + 
-      geom_bar(stat = "identity", position = "stack",
-               color = "black",
-               size = 0.1,
-               width = 0.8) +
-      facet_wrap(~ horizon, ncol = 4, 
-                 strip.position = "bottom") + 
-      theme_scoringutils() + 
-      scale_y_continuous(label = label_fn) +
-      # theme(axis.text.x = element_blank()) + 
-      theme(axis.text.x = element_blank(),
-            axis.line.x = element_blank(), 
-            axis.ticks.x = element_blank(),
-            plot.title = element_text(hjust = 0.5,
-                                      size = 8)) + 
-      labs(x = NULL, y = "WIS", title = paste(type_filter, "-", scale_filter)) 
-  }
-
-  p_nat_case <- component_plot(df, "natural", "Cases")
-  p_nat_death <- component_plot(df, "natural", "Deaths")
-  p_log_case <- component_plot(df, "log", "Cases")
-  p_log_death <- component_plot(df, "log", "Deaths")
-
-  cov_plot <- function(scores, filter_range) {
-    scores |>
-      filter(range == filter_range) |>
-      summarise_scores(by = c("model", "target_type", "range", "horizon")) |>
-      ggplot(aes(x = horizon, y = coverage, color = model)) + 
-      geom_line(show.legend = FALSE) + 
-      geom_point(show.legend = FALSE, size = 0.7) + 
-      geom_hline(yintercept = filter_range / 100, 
-                 linetype = "dashed", colour = "grey80") + 
-      facet_wrap(~ target_type, ncol = 2) + 
-      scale_y_continuous(labels = \(x) {paste0(100 * x, "%")}) + 
-      theme_scoringutils() + 
-      labs(x = "Horizon", y = paste("Coverage -", filter_range, "PI"))
-  }
-
-  p_cov50 <- cov_plot(scores, 50)
-  p_cov90 <- cov_plot(scores, 90)
-
-  p_performance <- (p_nat_case + p_nat_death) / 
-    (p_log_case + p_log_death) + 
-    (p_cov50 + p_cov90) + 
-    plot_layout(guides = "collect") &
-    plot_annotation(tag_levels = 'A') & 
-    theme(legend.position = "bottom", 
-          legend.box="vertical", legend.margin = margin())
-
-  return(p_performance)
-}
-
-p_performance <- scores |>
-  filter(model %in% models_non_human) |>
-  mutate(model = factor(model, levels = levels(models_non_human))) |>
-  plot_performance()
-
-ggsave("output/figures/performance.png", plot = p_performance, 
-       width = 7, height = 6)
-```
-
-# Figure 2: Ranks for different models
-================================================================================
-
-```{r}
-ranks <- scores |>
-  filter(model %in% models_non_human) |>
-  summarise_scores() |>
-  select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
-  group_by(target_type, scale, horizon, forecast_date) |>
-  mutate(rank = rank(interval_score))
-
-plot_ranks <- function(ranks, hor = 2) {
-  ranks |>
-    filter(horizon == hor) |>
-    ggplot(aes(y = model, group = rank, fill = rank)) +
-    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
-    facet_wrap(~ type_and_scale) + 
-    scale_fill_continuous(trans='reverse') + 
-    theme_scoringutils() + 
-    labs(y = "", x = "Frequency", fill = "Rank") + 
-    theme(legend.position = "right")
-}
-
-p_ranks <- ranks |>
-  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
-  plot_ranks()
-ggsave("output/figures/performance-ranks.png", 
-       plot = p_ranks, 
-       width = 7, height = 3.4)
-
-p_ranks_4 <- ranks |>
-  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
-  plot_ranks(hor = 4)
-ggsave("output/figures/performance-ranks-4.png", 
-       plot = p_ranks_4, 
-       width = 7, height = 3.4)
-
-ranks |>
-  filter(horizon == 2) |>
-  mutate(model = factor(model, levels = (levels(models_non_human)))) |>
-  group_by(model, target_type, scale, rank) |>
-  summarise(n = n()) |>
-  mutate(rel = n() / 13) |>
-  pivot_wider(names_from = rank, values_from = c(n, rel)) |>
-  {\(.) {replace(.,is.na(.),0)}}() |>
-  mutate(n_1 = paste0(n_1, " (", round(rel_1, 2), ")"), 
-         n_2 = paste0(n_2, " (", round(rel_2, 2), ")"), 
-         n_3 = paste0(n_3, " (", round(rel_3, 2), ")"), 
-         n_4 = paste0(n_4, " (", round(rel_4, 2), ")")) |>
-  select(-c(rel_1, rel_2, rel_3, rel_4))
-
-```
-
-# Figure 3: Forecasts and corresponding WIS
-================================================================================
-
-```{r}
-
-plot_scores_and_pred <- function(forecasts, scores) {
-
-  plot_pred_obs <- function(forecasts, scale = "natural") {
-
-    truth <- weekly_truth |>
-      mutate(target_type = paste(target_type, "-", scale))
-
-    if (scale == "log") {
-      forecasts <- forecasts |>
-        mutate(prediction = log(prediction + 1))
-
-      truth <- truth |>
-        mutate(true_value = log(true_value + 1))
-
-    }
-
-    forecasts |>
-      filter(horizon == 2) |>
-      filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
-      pivot_wider(values_from = prediction, names_from = quantile) |>
-      # dcast(... ~ quantile, value.var = "prediction") |>
-      mutate(target_type = paste(target_type, "-", scale)) |>
-      ggplot(aes(x = target_end_date)) + 
-      geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
-                     size = 1.1,
-                     alpha = 1,
-                     position = position_dodge(width = 4.8), 
-                     show.legend = FALSE) + 
-      geom_line(data = truth |>
-                  filter(target_end_date >= "2021-05-01", 
-                         target_end_date <= "2021-09-01"), 
-                aes(y = true_value)) + 
-      geom_point(data = truth |>
-                   filter(target_end_date >= "2021-05-01", 
-                          target_end_date <= "2021-09-01"), 
-                 aes(y = true_value), size = 0.7) + 
-      theme_scoringutils() +
-      facet_wrap(~ target_type, scales = "free", ncol = 2) + 
-      scale_y_continuous(labels = label_fn) + 
-      theme(legend.position = "bottom") + 
-      labs(y = "Forecasts", x = "Date", colour = "Model")
-  }
-
-  p_pred_nat <- plot_pred_obs(forecasts)
-  p_pred_log <- plot_pred_obs(forecasts, scale = "log")
-
-  plot_wis_components <- function(scores) {
-    scores |>
-      filter(horizon == 2) |>
-      summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
-      select(-interval_score) |>
-      # add dispersion value to produce accurate "stacked" barplot
-      mutate(underprediction = underprediction + dispersion, 
-             overprediction = overprediction + dispersion) |>
-      data.table::melt(measure.vars = c(
-        "overprediction",
-        "underprediction",
-        "dispersion"
-      ),
-      variable.name = "wis_component_name",
-      value.name = "component_value"
-      ) |>
-      mutate(wis_component_name = stringr::str_to_title(wis_component_name)) |>
-      mutate(wis_component_name = factor(
-        wis_component_name, 
-        levels = c("Dispersion", "Overprediction", "Underprediction"))) |>
-      mutate(target_end_date = as.Date(target_end_date)) |>
-      ggplot(aes(x = target_end_date, group = model)) +
-      geom_col(
-        position = "dodge",
-        aes(y = component_value, 
-            fill = model, 
-            group = model, alpha = wis_component_name), 
-        color = "black", 
-        size = 0.1,
-        width = 5.5
-      ) +
-      facet_wrap(~ type_and_scale, scales = "free_y") +
-      theme_scoringutils() +
-      scale_alpha_manual(values = c(1, 0.6, 0.1)) +
-      scale_y_continuous(labels = label_fn) + 
-      expand_limits(x = as.Date(c("2021-05-01", "2021-09-01"))) +
-      labs(y = "WIS", x = "Target end date", 
-           fill = "Forecaster", alpha = "WIS component") +
-      theme(legend.position = "bottom", 
-            legend.box="vertical", legend.margin = margin()) 
-  }
-
-  p_scores_nat <- scores |>
-    filter(scale == "natural") |>
-    plot_wis_components()
-
-  p_scores_log <- scores |>
-    filter(scale == "log") |>
-    plot_wis_components()
-
-  p_scores_and_pred <- p_pred_nat / p_scores_nat / p_pred_log / p_scores_log + 
-    plot_layout(guides = "collect", heights = c(1, 0.8, 1, 0.8)) &
-    plot_annotation(tag_levels = 'A') & 
-    theme(legend.position = "bottom", 
-          legend.box="vertical", legend.margin = margin())
-
-  return(p_scores_and_pred)  
-}
-
-p_scores_and_pred <- plot_scores_and_pred(
-  filter(forecasts, model %in% models_non_human) |>
-    mutate(model = factor(model, levels = levels(models_non_human))),
-  filter(scores, model %in% models_non_human) |>
-    mutate(model = factor(model, levels = levels(models_non_human)))
-)
-
-ggsave("output/figures/scores-and-forecasts.png", plot = p_scores_and_pred, 
-       width = 9, height = 9)
-```
-
-# Figure 4: Comparison performance direct vs. Rt
-================================================================================
-
-```{r}
-# helper function that filters the forecast data such that we only have 
-# forecasts left where a single person made a forecast both for direct and rt
-filter_comparison_rt_direct <- function(forecasts) {
-  forecasts_comparison <- forecasts |>
-  filter(model %in% models_human) |>
-  mutate(rt_name = ifelse(
-    grepl("(Rt)", model), 
-    gsub("\ \\(Rt\\)", "", model), 
-    NA
-  )) |>
-  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
-  group_by(forecast_date, target_type) |>
-  filter(model %in% rt_name | rt_name %in% model) |>
-  group_by(forecast_date, target_type, method) |>
-  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile, expert)
-}
-
-forecasts_comparison <- forecasts |> 
-  update_expert_status() |>
-  filter_comparison_rt_direct()
-
-scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
-
-plot_rt_direct_individual <- function(scores) {
- scores |> 
-  filter(horizon == 2) |>
-  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
-  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
-  arrange(target_type, name, method) |>
-  group_by(target_type, name, scale) |>
-  mutate(diff = rep(diff(interval_score), 2)) |>
-  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
-  mutate(method = stringr::str_to_title(method)) |>
-  ggplot(aes(y = interval_score, x = method, group = name)) + 
-  geom_point(size = 0.5) + 
-  geom_line(aes(color = improvement), linewidth = 0.3) + 
-  facet_wrap(~ type_and_scale, scale = "free_y") + 
-  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
-  scale_y_continuous(labels = label_fn) + 
-  theme_scoringutils()
-}
-
-p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)
-
-ggsave("output/figures/comparison-direct-rt-individual.png", 
-       plot = p_comparison_direct_rt_individual, 
-       width = 7, height = 4)
-
-# check whether there is a difference between experts and non-experts when using Rt / direct
-scores_forecasts_comparison |>
-  summarise_scores(by = c("type_and_scale", "method", "expert")) |>
-  filter(method == "rt") |>
-  arrange(type_and_scale, expert) |>
-  summarise_scores(fun = signif, digits = 2)
-
-scores_forecasts_comparison |> 
-  filter(horizon == 2) |>
-  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
-  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale", "expert")) |>
-  arrange(target_type, name, method) |>
-  group_by(target_type, name, scale) |>
-  mutate(diff = rep(diff(interval_score), 2)) |>
-  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
-  mutate(method = stringr::str_to_title(method)) |>
-  ggplot(aes(y = interval_score, x = method, group = name)) + 
-  geom_point(size = 0.5) + 
-  geom_line(aes(color = expert), linewidth = 0.3) + 
-  facet_wrap(~ type_and_scale, scale = "free_y") + 
-  labs(y = "WIS", x = "Forecasting method", color = "Expert status") + 
-  scale_y_continuous(labels = label_fn) + 
-  theme_scoringutils()
-```
-
-================================================================================
-#                                  Appendix                                    #
-================================================================================
-
-# Figure SI.1: Number of forecasters across the study period
-================================================================================
-- Figure SI.1
 - numbers of forecasters used in the text
@@
   )
-exp_method
 
@@
        width = 7, height = 5)
+ggsave("output/figures/Figure-1.png", plot = p_nums, 
+       width = 7, height = 5)
 
-# forecasts |> 
-#   update_expert_status() |>
-#   mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
-#   filter(forecast_date == "2021-06-07", 
-#          model %in% models_human, horizon == 1, 
-#          quantile == 0.5) |> 
-#   mutate(method = ifelse(grepl("(Rt)", model), "Rt", "direct")) |>
-#   mutate(model = gsub("\ \\(Rt\\)", "", model)) |> 
-#   group_by(target_type, method, expert) |> 
-#   summarise(n = length(unique(model)))
+forecasts |>
+  update_expert_status() |>
+  mutate(expert = ifelse(expert, "Expert", "Non-Expert")) |>
+  filter(forecast_date == "2021-06-07",
+         model %in% models_human, horizon == 1,
+         quantile == 0.5) |>
+  mutate(method = ifelse(grepl("(Rt)", model), "Rt", "direct")) |>
+  mutate(model = gsub("\ \\(Rt\\)", "", model)) |>
+  group_by(target_type, method, expert) |>
+  summarise(n = length(unique(model)))
 
 ```
+
+# Figure 2: Predictive performance across horizons (summary)
+================================================================================
+
+```{r}
+plot_performance <- function(scores) {
+  df <- scores |>
+    filter(scale %in% c("log", "natural")) |>
+    summarise_scores(by = c("horizon", "model", 
+                            "target_type", "scale", "type_and_scale")) |>
+    rename("Dispersion" = dispersion, 
+           "Overprediction" = overprediction, 
+           "Underprediction" = underprediction, 
+           Model = model) |>
+    pivot_longer(names_to = "WIS components",
+                 cols = c(`Dispersion`, `Overprediction`, `Underprediction`)) |>
+    mutate(`WIS components` = fct_relevel(`WIS components`, 
+                                          c("Dispersion", "Underprediction", 
+                                            "Overprediction")))
+
+  component_plot <- function(df, scale_filter, type_filter) {
+    df |>
+      filter(scale == scale_filter) |>
+      filter(target_type == type_filter) |>
+      ggplot(aes(x = Model, 
+                 # x = reorder(Model, (interval_score)), 
+                 y = value, fill = Model, 
+                 group = Model, alpha = `WIS components`)) + 
+      scale_alpha_manual(values = c(1, 0.1, 0.6)) +
+      # scale_fill_manual(values = vals) + 
+      geom_bar(stat = "identity", position = "stack",
+               color = "black",
+               size = 0.1,
+               width = 0.8) +
+      facet_wrap(~ horizon, ncol = 4, 
+                 strip.position = "bottom") + 
+      theme_scoringutils() + 
+      scale_y_continuous(label = label_fn) +
+      # theme(axis.text.x = element_blank()) + 
+      theme(axis.text.x = element_blank(),
+            axis.line.x = element_blank(), 
+            axis.ticks.x = element_blank(),
+            plot.title = element_text(hjust = 0.5,
+                                      size = 8)) + 
+      labs(x = NULL, y = "WIS", title = paste(type_filter, "-", scale_filter)) 
+  }
+
+  p_nat_case <- component_plot(df, "natural", "Cases")
+  p_nat_death <- component_plot(df, "natural", "Deaths")
+  p_log_case <- component_plot(df, "log", "Cases")
+  p_log_death <- component_plot(df, "log", "Deaths")
+
+  cov_plot <- function(scores, filter_range) {
+    scores |>
+      filter(range == filter_range) |>
+      summarise_scores(by = c("model", "target_type", "range", "horizon")) |>
+      ggplot(aes(x = horizon, y = coverage, color = model)) + 
+      geom_line(show.legend = FALSE) + 
+      geom_point(show.legend = FALSE, size = 0.7) + 
+      geom_hline(yintercept = filter_range / 100, 
+                 linetype = "dashed", colour = "grey80") + 
+      facet_wrap(~ target_type, ncol = 2) + 
+      scale_y_continuous(labels = \(x) {paste0(100 * x, "%")}) + 
+      theme_scoringutils() + 
+      labs(x = "Horizon", y = paste("Coverage -", filter_range, "PI"))
+  }
+
+  p_cov50 <- cov_plot(scores, 50)
+  p_cov90 <- cov_plot(scores, 90)
+
+  p_performance <- (p_nat_case + p_nat_death) / 
+    (p_log_case + p_log_death) + 
+    (p_cov50 + p_cov90) + 
+    plot_layout(guides = "collect") &
+    plot_annotation(tag_levels = 'A') & 
+    theme(legend.position = "bottom", 
+          legend.box="vertical", legend.margin = margin())
+
+  return(p_performance)
+}
+
+p_performance <- scores |>
+  filter(model %in% models_non_human) |>
+  mutate(model = factor(model, levels = levels(models_non_human))) |>
+  plot_performance()
+
+ggsave("output/figures/performance.png", plot = p_performance, 
+       width = 7, height = 6)
+ggsave("output/figures/Figure-2.png", plot = p_performance, 
+       width = 7, height = 6)
+
+```
+
+# Figure 3: Forecasts and corresponding WIS
+================================================================================
+
+```{r}
+
+plot_scores_and_pred <- function(forecasts, scores) {
+
+  plot_pred_obs <- function(forecasts, scale = "natural") {
+
+    truth <- weekly_truth |>
+      mutate(target_type = paste(target_type, "-", scale))
+
+    if (scale == "log") {
+      forecasts <- forecasts |>
+        mutate(prediction = log(prediction + 1))
+
+      truth <- truth |>
+        mutate(true_value = log(true_value + 1))
+
+    }
+
+    forecasts |>
+      filter(horizon == 2) |>
+      filter(quantile %in% c(0.025, 0.25, 0.5, 0.75, 0.975)) %>%
+      pivot_wider(values_from = prediction, names_from = quantile) |>
+      # dcast(... ~ quantile, value.var = "prediction") |>
+      mutate(target_type = paste(target_type, "-", scale)) |>
+      ggplot(aes(x = target_end_date)) + 
+      geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
+                     size = 1.1,
+                     alpha = 1,
+                     position = position_dodge(width = 4.8), 
+                     show.legend = FALSE) + 
+      geom_line(data = truth |>
+                  filter(target_end_date >= "2021-05-01", 
+                         target_end_date <= "2021-09-01"), 
+                aes(y = true_value)) + 
+      geom_point(data = truth |>
+                   filter(target_end_date >= "2021-05-01", 
+                          target_end_date <= "2021-09-01"), 
+                 aes(y = true_value), size = 0.7) + 
+      theme_scoringutils() +
+      facet_wrap(~ target_type, scales = "free", ncol = 2) + 
+      scale_y_continuous(labels = label_fn) + 
+      theme(legend.position = "bottom") + 
+      labs(y = "Forecasts", x = "Date", colour = "Model")
+  }
+
+  p_pred_nat <- plot_pred_obs(forecasts)
+  p_pred_log <- plot_pred_obs(forecasts, scale = "log")
+
+  plot_wis_components <- function(scores) {
+    scores |>
+      filter(horizon == 2) |>
+      summarise_scores(by = c("target_end_date", "model", "type_and_scale")) |>
+      select(-interval_score) |>
+      # add dispersion value to produce accurate "stacked" barplot
+      mutate(underprediction = underprediction + dispersion, 
+             overprediction = overprediction + dispersion) |>
+      data.table::melt(measure.vars = c(
+        "overprediction",
+        "underprediction",
+        "dispersion"
+      ),
+      variable.name = "wis_component_name",
+      value.name = "component_value"
+      ) |>
+      mutate(wis_component_name = stringr::str_to_title(wis_component_name)) |>
+      mutate(wis_component_name = factor(
+        wis_component_name, 
+        levels = c("Dispersion", "Overprediction", "Underprediction"))) |>
+      mutate(target_end_date = as.Date(target_end_date)) |>
+      ggplot(aes(x = target_end_date, group = model)) +
+      geom_col(
+        position = "dodge",
+        aes(y = component_value, 
+            fill = model, 
+            group = model, alpha = wis_component_name), 
+        color = "black", 
+        size = 0.1,
+        width = 5.5
+      ) +
+      facet_wrap(~ type_and_scale, scales = "free_y") +
+      theme_scoringutils() +
+      scale_alpha_manual(values = c(1, 0.6, 0.1)) +
+      scale_y_continuous(labels = label_fn) + 
+      expand_limits(x = as.Date(c("2021-05-01", "2021-09-01"))) +
+      labs(y = "WIS", x = "Target end date", 
+           fill = "Forecaster", alpha = "WIS component") +
+      theme(legend.position = "bottom", 
+            legend.box="vertical", legend.margin = margin()) 
+  }
+
+  p_scores_nat <- scores |>
+    filter(scale == "natural") |>
+    plot_wis_components()
+
+  p_scores_log <- scores |>
+    filter(scale == "log") |>
+    plot_wis_components()
+
+  p_scores_and_pred <- p_pred_nat / p_scores_nat / p_pred_log / p_scores_log + 
+    plot_layout(guides = "collect", heights = c(1, 0.8, 1, 0.8)) &
+    plot_annotation(tag_levels = 'A') & 
+    theme(legend.position = "bottom", 
+          legend.box="vertical", legend.margin = margin())
+
+  return(p_scores_and_pred)  
+}
+
+p_scores_and_pred <- plot_scores_and_pred(
+  filter(forecasts, model %in% models_non_human) |>
+    mutate(model = factor(model, levels = levels(models_non_human))),
+  filter(scores, model %in% models_non_human) |>
+    mutate(model = factor(model, levels = levels(models_non_human)))
+)
+
+ggsave("output/figures/scores-and-forecasts.png", plot = p_scores_and_pred, 
+       width = 9, height = 9)
+ggsave("output/figures/Figure-3.png", plot = p_scores_and_pred, 
+       width = 9, height = 9)
+```
+
+# Figure 4: Ranks for different models
+================================================================================
+
+```{r}
+ranks <- scores |>
+  filter(model %in% models_non_human) |>
+  summarise_scores() |>
+  select(model, target_type, horizon, forecast_date, interval_score, scale, type_and_scale) |>
+  group_by(target_type, scale, horizon, forecast_date) |>
+  mutate(rank = rank(interval_score))
+
+plot_ranks <- function(ranks, hor = 2) {
+  ranks |>
+    filter(horizon == hor) |>
+    ggplot(aes(y = model, group = rank, fill = rank)) +
+    geom_bar(position = position_stack(), colour = "white", size = 0.03) + 
+    facet_wrap(~ type_and_scale) + 
+    scale_fill_continuous(trans='reverse') + 
+    theme_scoringutils() + 
+    labs(y = "", x = "Frequency", fill = "Rank") + 
+    theme(legend.position = "right")
+}
+
+p_ranks <- ranks |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks()
+ggsave("output/figures/performance-ranks.png", 
+       plot = p_ranks, 
+       width = 7, height = 3.4)
+ggsave("output/figures/Figure-4.png", 
+       plot = p_ranks, 
+       width = 7, height = 3.4)
+
+p_ranks_4 <- ranks |>
+  mutate(model = factor(model, levels = rev(levels(models_non_human)))) |>
+  plot_ranks(hor = 4)
+ggsave("output/figures/performance-ranks-4.png", 
+       plot = p_ranks_4, 
+       width = 7, height = 3.4)
+
+ranks |>
+  filter(horizon == 2) |>
+  mutate(model = factor(model, levels = (levels(models_non_human)))) |>
+  group_by(model, target_type, scale, rank) |>
+  summarise(n = n()) |>
+  mutate(rel = n() / 13) |>
+  pivot_wider(names_from = rank, values_from = c(n, rel)) |>
+  {\(.) {replace(.,is.na(.),0)}}() |>
+  mutate(n_1 = paste0(n_1, " (", round(rel_1, 2), ")"), 
+         n_2 = paste0(n_2, " (", round(rel_2, 2), ")"), 
+         n_3 = paste0(n_3, " (", round(rel_3, 2), ")"), 
+         n_4 = paste0(n_4, " (", round(rel_4, 2), ")")) |>
+  select(-c(rel_1, rel_2, rel_3, rel_4))
+
+```
+
+# Figure 5: Comparison performance direct vs. Rt
+================================================================================
+
+```{r}
+# helper function that filters the forecast data such that we only have 
+# forecasts left where a single person made a forecast both for direct and rt
+filter_comparison_rt_direct <- function(forecasts) {
+  forecasts_comparison <- forecasts |>
+  filter(model %in% models_human) |>
+  mutate(rt_name = ifelse(
+    grepl("(Rt)", model), 
+    gsub("\ \\(Rt\\)", "", model), 
+    NA
+  )) |>
+  mutate(method = ifelse(grepl("(Rt)", model), "rt", "direct")) |>
+  group_by(forecast_date, target_type) |>
+  filter(model %in% rt_name | rt_name %in% model) |>
+  group_by(forecast_date, target_type, method) |>
+  select(target_end_date, target_type, forecast_date, horizon, model, method, true_value, prediction, quantile, expert)
+}
+
+forecasts_comparison <- forecasts |> 
+  update_expert_status() |>
+  filter_comparison_rt_direct()
+
+scores_forecasts_comparison <- score_and_add_log(forecasts_comparison)
+
+plot_rt_direct_individual <- function(scores) {
+ scores |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale")) |>
+  arrange(target_type, name, method) |>
+  group_by(target_type, name, scale) |>
+  mutate(diff = rep(diff(interval_score), 2)) |>
+  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
+  mutate(method = stringr::str_to_title(method)) |>
+  ggplot(aes(y = interval_score, x = method, group = name)) + 
+  geom_point(size = 0.5) + 
+  geom_line(aes(color = improvement), linewidth = 0.3) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  labs(y = "WIS", x = "Forecasting method", color = "Comparison") + 
+  scale_y_continuous(labels = label_fn) + 
+  theme_scoringutils()
+}
+
+p_comparison_direct_rt_individual <- plot_rt_direct_individual(scores_forecasts_comparison)
+
+ggsave("output/figures/comparison-direct-rt-individual.png", 
+       plot = p_comparison_direct_rt_individual, 
+       width = 7, height = 4)
+ggsave("output/figures/Figure-5.png", 
+       plot = p_comparison_direct_rt_individual, 
+       width = 7, height = 4)
+
+# check whether there is a difference between experts and non-experts when using Rt / direct
+scores_forecasts_comparison |>
+  summarise_scores(by = c("type_and_scale", "method", "expert")) |>
+  filter(method == "rt") |>
+  arrange(type_and_scale, expert) |>
+  summarise_scores(fun = signif, digits = 2)
+
+scores_forecasts_comparison |> 
+  filter(horizon == 2) |>
+  mutate(name = gsub("\ \\(Rt\\)", "", model)) |>
+  summarise_scores(by = c("target_type","type_and_scale", "name", "method", "scale", "expert")) |>
+  arrange(target_type, name, method) |>
+  group_by(target_type, name, scale) |>
+  mutate(diff = rep(diff(interval_score), 2)) |>
+  mutate(improvement = ifelse(diff > 0, "Direct better", "Rt better")) |>
+  mutate(method = stringr::str_to_title(method)) |>
+  ggplot(aes(y = interval_score, x = method, group = name)) + 
+  geom_point(size = 0.5) + 
+  geom_line(aes(color = expert), linewidth = 0.3) + 
+  facet_wrap(~ type_and_scale, scale = "free_y") + 
+  labs(y = "WIS", x = "Forecasting method", color = "Expert status") + 
+  scale_y_continuous(labels = label_fn) + 
+  theme_scoringutils()
+```
+
+# Figure 6: Performance of "experts" and "non-experts"
+================================================================================
+Separate ensemble of experts and non-experts
+
+```{r}
+create_expert_ensemble <- function(forecasts) {
+
+  forecasts_expert_ensemble <- forecasts |>
+    filter(model %in% models_human) |>
+    update_expert_status() |>
+    group_by(target_type, horizon, expert, quantile, true_value, forecast_date, target_end_date) |>
+    summarise(prediction = median(prediction), 
+              n = n()) |>
+    mutate(model = ifelse(expert, "Expert", "Non-Expert")) |>
+    as.data.table()
+
+  scores_expert_ensemble <- forecasts_expert_ensemble |>
+    mutate(method = model) |>
+    score_and_add_log()
+
+  return(list(forecasts = forecasts_expert_ensemble, 
+              scores = scores_expert_ensemble))
+}
+
+expert_ensemble <- create_expert_ensemble(filter(forecasts))
+
+# plot with number of experts and non-experts every week
+expert_ensemble$forecasts |>
+  filter(horizon == 2) |>
+  group_by(model, forecast_date, target_type) |>
+  summarise(n = unique(n)) |>
+  ggplot(aes(x = forecast_date, y = n, colour = model)) + 
+  geom_line() + 
+  geom_point() + 
+  theme_scoringutils() + 
+  facet_wrap(~ target_type)
+
+expert_ensemble$scores |>
+  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
+  mutate(model = factor(model, levels = c("crowd-ensemble", "Expert", "Non-Expert"))) |>
+   make_performance_table(
+     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
+     baseline = "crowd-ensemble",
+    cap = "Performance for two-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-experts.tex')
+
+expert_ensemble$scores |>
+  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
+   make_performance_table(
+     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
+     baseline = "crowd-ensemble",
+     hor = 4,
+     cap = "Performance for four-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts-4}"
+    ) %>%
+  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
+  cat(file = 'manuscript/performance-table-experts-4.tex')
+
+p_scores_expert <- plot_performance(expert_ensemble$scores)
+ggsave("output/figures/performance-expert.png", 
+       plot = p_scores_expert, 
+       width = 7, height = 6)
+ggsave("output/figures/Figure-6.png", 
+       plot = p_scores_expert, 
+       width = 7, height = 6)
+
+plot_ranks(expert_ensemble$scores)
+
+```
+
+================================================================================
+#                                  Appendix                                    #
+================================================================================
 
@@
 
-# Figure SI.5: Performance of "experts" and "non-experts"
-================================================================================
-Separate ensemble of experts and non-experts
-
-```{r}
-create_expert_ensemble <- function(forecasts) {
-
-  forecasts_expert_ensemble <- forecasts |>
-    filter(model %in% models_human) |>
-    update_expert_status() |>
-    group_by(target_type, horizon, expert, quantile, true_value, forecast_date, target_end_date) |>
-    summarise(prediction = median(prediction), 
-              n = n()) |>
-    mutate(model = ifelse(expert, "Expert", "Non-Expert")) |>
-    as.data.table()
-
-  scores_expert_ensemble <- forecasts_expert_ensemble |>
-    mutate(method = model) |>
-    score_and_add_log()
-
-  return(list(forecasts = forecasts_expert_ensemble, 
-              scores = scores_expert_ensemble))
-}
-
-expert_ensemble <- create_expert_ensemble(filter(forecasts))
-
-# plot with number of experts and non-experts every week
-expert_ensemble$forecasts |>
-  filter(horizon == 2) |>
-  group_by(model, forecast_date, target_type) |>
-  summarise(n = unique(n)) |>
-  ggplot(aes(x = forecast_date, y = n, colour = model)) + 
-  geom_line() + 
-  geom_point() + 
-  theme_scoringutils() + 
-  facet_wrap(~ target_type)
-
-expert_ensemble$scores |>
-  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
-  mutate(model = factor(model, levels = c("crowd-ensemble", "Expert", "Non-Expert"))) |>
-   make_performance_table(
-     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
-     baseline = "crowd-ensemble",
-    cap = "Performance for two-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts}"
-    ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table-experts.tex')
-
-expert_ensemble$scores |>
-  rbind(filter(scores, model == "crowd-ensemble"), fill = TRUE) |>
-   make_performance_table(
-     model_filter = c("Expert", "Non-Expert", "crowd-ensemble"),
-     baseline = "crowd-ensemble",
-     hor = 4,
-     cap = "Performance for four-week-ahead forecasts of experts and non-experts. Values have been cut to three significant digits and rounded. \\label{tab:scores-experts-4}"
-    ) %>%
-  gsub("\\{table\\}", "\\{table\\*\\}", x = .) |>
-  cat(file = 'manuscript/performance-table-experts-4.tex')
-
-p_scores_expert <- plot_performance(expert_ensemble$scores)
-ggsave("output/figures/performance-expert.png", 
-       plot = p_scores_expert, 
-       width = 7, height = 6)
-
-plot_ranks(expert_ensemble$scores)
-
-```
-
 # Unused - Significance analysis

### 2023-12-11 Analysis.rmd
Respond to review comments

@@
 library(magrittr)
+# use scoringutils version 1.2.2: remotes::install_github("epiforecasts/scoringutils@v1.2.2")
 library(scoringutils) 
@@
 colors <- RColorBrewer::brewer.pal(n = 9, name = "Set1")
-colors <- hex <- hue_pal()(4)
+colors <- hex <- scales::hue_pal()(4)
 
@@
 
+authors <- c("seb", "seabbs", "nib", "anonymous_Human")
+models_authors <- c(authors, paste(authors, "(Rt)"))
+
 update_expert_status <- function(data) {
@@
 }
+
+```
+
+# Figure 1: overview of the data
+
+```{r}
+plot_data_simple <- function(weekly_truth, daily_truth) {
+  weekly_truth |>
+    mutate(status = str_to_title(status)) |>
+  ggplot(aes(x = target_end_date, y = true_value)) +
+  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
+           fill = "darkolivegreen3", alpha = .1) + 
+  geom_line() + 
+  geom_point(size = 0.8) + 
+  geom_bar(data = daily_truth |>
+             mutate(true_value = true_value * 7), 
+           stat = "identity", 
+           fill = "skyblue3", alpha = 0.4) + 
+  facet_wrap(~ target_type, 
+             nrow = 2, scale = "free_y") + 
+  scale_y_continuous(labels = label_fn) +
+  theme_scoringutils() +
+  labs(y = "Observed values", x = "Date", color = "Data status")
+}
+
+p_data <- plot_data_simple(weekly_truth, daily_truth)
+ggsave("output/figures/overview-data.png", plot = p_data, 
+       width = 7, height = 5)
+
+weekly_truth |>
+  filter(target_end_date >= time_start & target_end_date <= time_stop) |>
+  filter(target_type == "Cases") |>
+  pull(target_end_date)
+
+```
+
+Author participation
+
+```{r}
+
+all <- forecasts |> 
+  filter(!is.na(prediction), 
+         model %in% models_human,
+         horizon == 2) |>
+  avail_forecasts(by = c("model", "target_type"))
+
+authors <- forecasts |> 
+  filter(!is.na(prediction), 
+         model %in% models_authors,
+         horizon == 2) |>
+  avail_forecasts(by = c("model", "target_type"))
+
+num_all <- all$`Number forecasts` |> sum()
+num_authors <- authors$`Number forecasts` |> sum()
+
+num_authors / num_all
 
@@
 
-# Figure 1: Number of forecasters across the study period
+# Figure 2: Number of forecasters across the study period
 ================================================================================
@@
 
-# Figure 2: Predictive performance across horizons (summary)
+# Figure 3: Predictive performance across horizons (summary)
 ================================================================================
@@
 
-# Figure 3: Forecasts and corresponding WIS
+# Figure 4: Forecasts and corresponding WIS
 ================================================================================
@@
 
-# Figure 4: Ranks for different models
+# Figure 5: Ranks for different models
 ================================================================================
@@
 
-# Figure 5: Comparison performance direct vs. Rt
+# Figure 6: Comparison performance direct vs. Rt
 ================================================================================
@@
 
-# Figure 6: Performance of "experts" and "non-experts"
+# Figure 7: Performance of "experts" and "non-experts"
 ================================================================================

### 2023-12-11 manuscript-first-submission/manuscript.tex
Respond to review comments

--- 2023-05-04 manuscript/manuscript.tex
@@
 \usepackage[numbers]{natbib}
-\bibliographystyle{custom}
 
@@
 
-\title{Human judgement forecasting of COVID-19 in the UK}
+\title{Human Judgement Forecasting of COVID-19 in the UK}
 %\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
@@
 
-From May 24th to August 16th, 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
+From May 24th to August 16th 2021, we elicited weekly one to four week ahead forecasts of cases and deaths from COVID-19 in the UK from a crowd of human forecasters.
 A median ensemble of all forecasts was submitted to the European Forecast Hub.
@@
 Human judgement forecasts and computational models can produce forecasts of similar quality for infectious disease such as COVID-19.
-The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a ``good" forecast is dependent on the forecast consumer.
+The results of forecast evaluations can change depending on what metrics are chosen and judgement on what does or doesn't constitute a "good" forecast is dependent on the forecast consumer.
 Combinations of human and computational forecasts hold potential but present real-world challenges that need to be solved.
@@
 Similarly, \citet{mcandrewChimericForecastingCombining2022} reported an ensemble of human forecasters to perform comparably to an ensemble of computational models when predicting incident COVID-19 cases, and worse when predicting incident deaths.
-\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit many participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
+\citet{farrowHumanJudgmentApproach2017} and in particular \citet{bosseComparingHumanModelbased2022} struggled to recruit a large number of participants (numbers of active forecasters ranged from 22 to 61 in \citet{mcandrewChimericForecastingCombining2022}, 7 to 24 in \citet{farrowHumanJudgmentApproach2017}, and 4 to 10 in \citet{bosseComparingHumanModelbased2022}).
 It is important to note that in previous studies (and also this one) human forecasters were free to use any resources, including computational models, in the process of creating a forecast, making it difficult to completely separate human judgement and computational modelling.
@@
 Forecasts were made for incident weekly reported numbers of cases of and deaths from COVID-19 on a national level for various European countries over a one to four week forecast horizon.
-While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday, and starting on Sunday.
+While forecasts were submitted on Mondays, weeks were defined as epidemiological weeks, ending on a Saturday and starting on Sunday.
 Forecast horizons were therefore in fact 5, 12, 19 and 26 days.
@@
 Every week, forecasts submitted to the hub were automatically checked for conformity with the required format and all eligible forecasts combined into different ensembles.
-Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
+Until the 12th of July 2021 the default Hub ensemble ("EuroCOVIDhub-ensemble") shown on all official Forecast Hub visualisations (https://covid19forecasthub.eu/) was a mean-ensemble (i.e., the $\alpha$-quantile of the ensemble is given by the mean of all submitted $\alpha$-quantiles).
 From the 29th of July onwards, the default Forecast Hub ensemble became a median ensemble.
@@
 The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
-The data and revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
+The data and revisions are displayed in Figure \ref{fig:plot-data}.
 All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).
@@
 Just as in the previous study, the default forecast shown was a repetition of the last known observation with constant uncertainty around it.
-The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e., as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
+The shown distribution was the exponential of a normal distribution with mean $\log (\text{last value})$ and uncertainty set to the standard deviation of the last four changes in weekly log observed forecasts (i.e.
+as $\sigma(\log(\text{value4}) - \log(\text{value3}), \, \log(\text{value3}) - \log(\text{value2}), \, \dots)$).
 In addition to information about past observations, participants could see various metrics and data such as the test positivity rate and vaccination rate sourced from Our World in Data \citep{owidcoronavirus}.
-Figure 3 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023} shows a screenshot of the forecast interface for direct forecasts.
+Figure \ref{fig:screenshot-classical} shows a screenshot of the forecast interface for direct forecasts.
 
@@
 This approach has been widely used for short-term forecasting \cite{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022} and used to produce reproduction number estimates \cite{abbottEstimatingIncreaseReproduction2021, abbottEstimatingTimevaryingReproduction2020a, sherrattExploringSurveillanceData2021}.
-Further details are given in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
+Further details are given in the SI.
 
-To obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
-Further details are given in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
+In order to obtain forecasts for deaths, we similarly fit a model that convolved observed and predicted reported cases as implied by the $R_t$ forecast over a  delay distributions \cite{sherrattExploringSurveillanceData2021, abbottEstimatingTimevaryingReproduction2020a} and scaled them by a fixed ratio to model the time between a case report and a reported death and the case fatality ratio using the \texttt{EpiNow2} \textsf{R} package \citep{epinow2}.
+Further details are given in the SI and in our previous work \citep{bosseComparingHumanModelbased2022}.
 
@@
 However, due to lack of development time, participants could not preview the death forecast implied by their current input for $R_t$ nor could they influence the estimated case fatality ratio or delay between reported cases and reported deaths.
-Figure 4 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023} shows a screenshot of the forecast interface for $R_t$ forecasts.
+Figure \ref{fig:screenshot-rt} shows a screenshot of the forecast interface for $R_t$ forecasts.
 
@@
 
-To boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
+In order to boost participation compared to our last crowd forecasting study in Germany and Poland \citep{bosseComparingHumanModelbased2022} which struggled in this regard, we announced an official tournament, the "UK Crowd Forecasting Challenge".
 Participants were asked to submit weekly predictions for reported cases and deaths linked to COVID-19 in the United Kingdom one to four weeks into the future.
@@
 averaged across forecast dates, horizons and forecast targets.
-For the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
+For the purpose of the tournament ranking, participants who did not submit a forecast in a given week were assigned the median score of all other participants who submitted a forecast that week.
 The UK crowd forecasting challenge was announced over Twitter and our networks.
@@
 
-We scored forecasts using the weighted interval score
-\cite{bracherEvaluatingEpidemicForecasts2021}. For a single interval, the interval score is computed as 
-  $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
-  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
-For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,
-  $$WIS = \frac{1}{K + 0.5} \cdot \left( w_0 \cdot |y - m| + \sum_{k = 1}^{K} w_k \cdot IS_{\alpha}(F, y) \right), $$
-  where $w_k$ is a weight for every interval. Usually, $w_k = \frac{\alpha_k}{2}$ and $w_0 = 0.5$.
-
+Our analysis of the data largely follows the one outlined in \citet{bosseComparingHumanModelbased2022}. We scored forecasts using the weighted interval score (WIS, formulas given in the SI) 
+\cite{bracherEvaluatingEpidemicForecasts2021}.
 The WIS is a strictly proper scoring rule yielding non-negative values, with smaller values implying better performance.
@@
 The WIS can be understood as an approximation of the continuous ranked probability score (CRPS) for forecasts in a quantile-based format, which itself represents a generalisation of the absolute error to predictive distributions.
-The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e., uncertainty of forecasts), overprediction and underprediction.
+The WIS can be decomposed into three separate penalty components: forecast dispersion (i.e.
+uncertainty of forecasts), overprediction and underprediction.
 
-\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS to better reflect the exponential nature of the underlying disease process.
+\citet{bosseTransformationForecastsEvaluating2023} recently suggested to transform forecasts and observations using the natural logarithm prior to applying the WIS in order to better reflect the exponential nature of the underlying disease process.
 We, therefore, also compute WIS values after transforming all forecasts and observations using the function $f\colon x \to \log (x + 1)$.
@@
 In order to measure probabilistic calibration \citep{gneitingProbabilisticForecastsCalibration2007}, we used the empirical coverage of all central 50\% and 90\% prediction intervals.
-Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g., the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
+Empirical coverage refers to the percentage of observations falling inside any given central prediction interval (e.g.
+the cumulative percentage of observed values that fall inside all central 50\% prediction intervals).
 
@@
 A total number of 90 participants submitted forecasts (more precisely, forecasts were submitted from 90 different accounts, some of them anonymous).
-Out of 90 participants, 21 self-identified as "experts", i.e., stated they had professional experience in infectious disease modelling or a related field.
+Out of 90 participants, 21 self-identified as "experts", i.e.
+stated they had professional experience in infectious disease modelling or a related field.
 
@@
 The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths) (see Figure \ref{fig:num-forecasters}B).
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/num-forecasters.png}
-\caption{\bf{Number of forecasts across the study period.} A: number of forecasts included in the Hub ensemble and the combined crowd ensemble.
-B: number of forecasts by "experts" and "non-experts".
-Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
-\label{fig:num-forecasters}
-\end{figure*}
 
@@
 B: Corresponding WIS values, decomposed into dispersion, overprediction and underprediction.
-C: 50\% prediction intervals on the log scale, i.e., after applying the natural logarithm to all forecasts and observations.
-D: Corresponding WIS on the log scale, i.e., the WIS applied to the log-transformed forecasts and observations.}
+C: 50\% prediction intervals on the log scale, i.e.
+after applying the natural logarithm to all forecasts and observations.
+D: Corresponding WIS on the log scale, i.e.
+the WIS applied to the log-transformed forecasts and observations.}
 \label{fig:forecasts-scores} 
@@
 WIS values the human forecasts of 1.02, 1.05, 1.06) and on the natural scale (rel.
-WIS values of 1.21, 1.25, 1.3) (compare Table 1 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}).
+WIS values of 1.21, 1.25, 1.3) (see Table \ref{tab:scores-4}).
 
@@
 
-\subsection*{Experts and non-experts}
+\subsection*{Experts and Non-Experts}
 
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.99\textwidth]{../output/figures/performance-expert.png}
-\caption{\bf{Predictive performance of self-reported "experts" and "non-experts" across forecast horizons.} Forecasts from "experts" and "non-experts" were combined to two separate median ensembles, including both direct and $R_t$ forecasts.
-A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
-E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance-experts}
-\end{figure*}
-
-\input{manuscript/performance-table-experts}
-
-When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}).
+When comparing two week ahead forecasts from "experts" and "non-experts", the ensemble of "experts" was better calibrated (see Figure \ref{fig:performance-experts}.
 For cases, "experts" achieved better scores than "non-experts" both on the log and on the natural scale.
@@
 %Calibration
-In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022}, we found almost all forecasts for cases to be underdispersed (i.e., too narrow/overconfident).
+In line with results from previous work \citep{bosseComparingHumanModelbased2022, sherrattPredictivePerformanceMultimodel2022}, we found almost all forecasts for cases to be underdispersed (i.e.
+too narrow/overconfident).
 Empirical coverage for death forecasts was higher than the corresponding coverage for cases for all forecasting approaches except the $R_t$ forecasts.
@@
 Sample sizes, however, were very low.
-Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
+Given that forecasters could simulate cases in the app, it is also possible that forecasters were in reality directly forecasting cases. 
 $R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths.
@@
 Forecasts for "experts" tended to be better calibrated than non-experts.
-However results should be taken with care considering relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths) and given that expert status was self-reported.
+However results should be taken with care in light of relatively low sample sizes (median of 11 "non-experts" for cases and 10 for deaths, median of 8 "experts" for cases and deaths), and given that expert status was self-reported.
 Furthermore, we only asked for professional involvement in a field related to infectious disease modelling, not specifically for familiarity with modelling of COVID-19 in the UK, and only offered participants a binary choice.
@@
 % \paragraph{Susceptibility of results}
-Overall, results of our study should be taken with caution due to several important limitations.
+Overall, results of our study should be taken with caution due to a number of important limitations.
 Firstly, our study was restricted to one location and to a relatively short period of thirteen weeks.
@@
 Highlighting this, prizes to the human forecasters were paid out based on the combined WIS on the log scale across all horizons and forecast targets.
-Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and pay-outs would have been different.
+Had we chosen to instead measure WIS on the natural scale, or to forecast only cases and continue to score on the log scale, rankings and payouts would have been different.
 
@@
 
-\subsection*{Data availability}
-All data and code are available publicy under a MIT license under \url{https://github.com/epiforecasts/uk-crowd-forecasting-challenge} and \url{https://doi.org/10.5281/zenodo.7897257}.
-The data has been published separately here: \url{https://doi.org/10.5281/zenodo.7897289}.
-Supplementary information \cite{bosseSupplementaryInformationHuman2023} to this manuscript is available at \url{https://doi.org/10.5281/zenodo.7897513}.
-
 % ================================= References =============================== %
@@
   $$IS_\alpha(F,y) = (u-l) + \frac{2}{\alpha} \cdot (l-y) \cdot 1(y \leq l) + \frac{2}{\alpha} \cdot (y-u) \cdot 1(y \geq u), $$ 
-  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e., the lower and upper bound of a single prediction interval.
+  where $1()$ is the indicator function, $y$ is the true value, and $l$ and $u$ are the $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$ quantiles of the predictive distribution $F$, i.e.
+the lower and upper bound of a single prediction interval.
 For a set of $K$ prediction intervals and the median $m$, the score is computed as a weighted sum,
@@
 Expert status was determined based on the participant's answer to the question whether they "worked in infectious disease modelling or had professional experience in any related field".}
-\label{fig:num-forecasters-SI}
+\label{fig:num-forecasters}
 \end{figure*}
@@
 E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
-\label{fig:performance-experts-SI}
+\label{fig:performance-experts}
 \end{figure*}
 
-%\input{manuscript/performance-table-experts}
+\input{manuscript/performance-table-experts}
 

### 2023-12-11 get-paper-data.R

@@ -1,5 +1,5 @@
 ## ========================================================================== ##
-##               Code to refetch data from                   ##
+##               Code to refetch data from Forecast Hub                       ##
 ## ========================================================================== ##
 
 library(covidHubUtils)

### 2024-02-01 Analysis.rmd
Update Figure 1 and correct some typos

@@
     mutate(status = str_to_title(status)) |>
-  ggplot(aes(x = target_end_date, y = true_value)) +
+  ggplot(aes(x = target_end_date, y = true_value, colour = "Weekly data")) +
   annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
@@
              mutate(true_value = true_value * 7), 
+           aes(fill = "7x daily cases"), 
+           colour = NA,
            stat = "identity",
-           fill = "skyblue3", alpha = 0.4) + 
+           alpha = 0.4) + 
   facet_wrap(~ target_type, 
@@
   scale_y_continuous(labels = label_fn) +
+  scale_fill_manual(values = c("7x daily cases" = "skyblue3")) +
+  scale_color_manual(values = c("Weekly data" = "black")) +
   theme_scoringutils() +
-  labs(y = "Observed values", x = "Date", color = "Data status")
-}
-
-p_data <- plot_data_simple(weekly_truth, daily_truth)
+  labs(y = "Observed values", x = "Date", color = " ", fill = " ")
+}
+
+(p_data <- plot_data_simple(weekly_truth, daily_truth))
 ggsave("output/figures/overview-data.png", plot = p_data, 

### 2024-02-01 manuscript/manuscript.tex
Update Figure 1 and correct some typos

@@
 \usepackage[numbers]{natbib}
-\bibliographystyle{custom}
+% \bibliographystyle{custom}
 
@@
 The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023.
-The data and revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
+Data revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}.
 All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).
@@
 
+\subsection*{Observed values}
+
+The study period (forecasts were made between May 24 and August 16, 2021, for targets between May 29 and September 11, 2021) was characterised by an increase in the number of cases and deaths in the United Kingdom.
+Reported cases in particular rose rapidly compared to pre-study levels, with a peak on July 17, 2021, followed by a trough and another subsequent increase in numbers.
+Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021.
+This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021} as well as the  European Football Championship \citep{dehningImpactEuro20202023}.
+An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.99\textwidth]{../output/figures/overview-data.png}
+\caption{\bf{Observed cases and deaths of COVID-19 in the UK}.
+A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021.
+Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.}
+\label{fig:overview-data}
+\end{figure*}
+
 \subsection*{Crowd forecast participation}
@@
 Abbott).
+Three other authors participated in the study (S.
+Funk, N.
+Bosse, and E.
+van Leuwwen).
+A total of 535 forecasts were submitted by human forecasters, 118 (22\%) of these were submitted by authors of this study.
 The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A).
@@
 
-\input performance-table.tex
+% \input performance-table.tex
 
@@
 \subsection*{Case forecasts}
+
+At the beginning of the study period, human forecasters as well as the Forecast Hub ensemble, consistently underpredicted case numbers (see Figure \ref{fig:forecasts-scores}A).
+All forecasting approaches overshot the peak in case numbers on July 17, 2021, overpredicting case numbers severely in the three weeks after, followed again by a small tendency to underpredict when case numbers rose once more in the 4th week after the peak.
 
@@
 \subsection*{Death forecasts}
+
+In the first part of the study period, most forecasting approaches (albeit not the direct crowd forecasts), showed a tendency to overpredict the increase in death numbers (see Figure \ref{fig:forecasts-scores}B).
+All forecasting approaches started to underpredict death numbers four weeks after the peak in case numbers on July 17, 2021, expecting a consequent drop in deaths that did not occur.
 
@@
 
-\input{manuscript/performance-table-experts}
+\input{performance-table-experts}
 
@@
 
+% Study period. 
+The study period was marked by an increase in both case and death numbers.
+Case numbers rose quickly compared to the pre-study period, peaking on July 17, 2021, followed by a trough and a subsequent further increase.
+Forecasts displayed a familiar pattern, where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak.
+Death numbers were increasing more slowly than during the previous peak in January 2021, in line with the beginning of vaccination efforts and a greater natural immunity in the population.
+The peak in case numbers did not correspond to a peak in death numbers, suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms.
+Forecasters tended to overpredict death numbers in the beginning, while underpredicting them in the end, expecting death numbers to fall after the peak in cases.
+The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.
+
 %Calibration
@@
 Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths.
-During the study period, the case fatality ratio (CFR) changed quite quickly.
-This was related to the rise of the Delta variant in the UK in the beginning of May with its higher case fatality rate \citep{bastIncreasedRiskHospitalisation2021}.
-In addition, the age distribution of cases changed over the course of the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
-One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update .
+During the study period, the case fatality ratio (CFR) likely changed quite quickly compared to the pre-study period.
+On the one hand, the rise of the Delta variant in the UK, which was first discovered in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022}.
+On the other hand, the ongoing COVID-19 vaccination and growing natural immunity in the population had decreasing effects on the CFR.
+In addition, the age distribution of cases changed throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
+Overall, the CFR was lower than during previous peaks of COVID-19.
+One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update.
+The present study only saw a steady increase in death numbers, which one could argue is relatively easy to predict, making it difficult to compare forecast performance with performance in other settings.
 A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.
@@
 The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon.
-However, as mentioned before, the CFR evolved during the study period.
-Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy.
-Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
+However, as mentioned before, the CFR likely evolved during the study period.
+Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting forecast accuracy.
+They also had no way to adjust the CFR manually.
+Giving human forecasters to see their implied death forecasts, as well as the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model.
 Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period.
@@
 However, as we used ad-hoc recruitment in our networks many of these self-identified experts are likely to be infectious disease modellers.
+
+It is plausible to hypothesise that the default baseline shown to forecasters in the app may influence their predictions.
+One could also interpret the $R_t$-forecast as a way of showing a different baseline forecast to the forecaster compared to the direct forecast.
+In our study, the default was a naive forecast with the median equal to the last value and uncertainty equal to the standard deviation of the last four changes in weekly log values.
+\cite{bosseComparingHumanModelbased2022} did not fnd conclusive evidence to that effect, but also did not analyse the question in detail.
+We suggest further research be done into potential priming effects that a default forecast can have on users.
 

### 2024-02-10 manuscript/manuscript.tex
Make small additions to the manuscript

@@
 Before creating the ensemble, we deleted forecasts that were clearly the result of a user or software error (such as forecasts that were zero everywhere).
+Our combined crowd ensemble, "epiforecasts-EpiExpert", but not the other two, entered the official European COVID-19 Forecast Hub ensemble ("EuroCOVIDhub-ensemble").
 
@@
 
+Importantly, in this study our combined crowd ensemble ("epiforecasts-EpiExpert") contributed to the European Forecast Hub ensemble.
+This in contrast to the study by \citet{bosseComparingHumanModelbased2022}, where they compared crowd forecasts against a hypothetical ensemble excluding the crowd forecasts.
+In the original study, including the crowd forecasts improved the Hub ensemble on average (however, the overall number of models included in the German and Polish Hub ensemble was smaller than the number of models in the European Forecast Hub ensemble).
+In our study, comparisons between our crowd ensembles and the Forecast Hub ensemble are therefore confounded by the fact the combined crowd ensemble was included in the Forecast Hub ensemble, possibly leading us to underestimate differences between the two.
+
 % \paragraph{Rt forecasting}
@@
 In our study, the default was a naive forecast with the median equal to the last value and uncertainty equal to the standard deviation of the last four changes in weekly log values.
-\cite{bosseComparingHumanModelbased2022} did not fnd conclusive evidence to that effect, but also did not analyse the question in detail.
+\cite{bosseComparingHumanModelbased2022} did not find conclusive evidence to that effect, but also did not analyse the question in detail.
 We suggest further research be done into potential priming effects that a default forecast can have on users.
@@
 All winners donated their prizes.
+We also thank Daniel J.
+McDonald for his kind and thoughtful review which has helped improve the manuscript substantially.
 

### 2024-02-14 manuscript/manuscript.tex
Update manuscript to incorporate feedback from Anne Cori

@@
 
-This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021.
+This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with one- to four-week ahead forecasts made over the course of thirteen weeks between May 24 and August 16, 2021.
+The study period is after the second wave of COVID-19 in the UK (which peaked in January 2021) and falls into a time when restrictions in the UK were gradually lifted as part of the roadmap out of lockdown (with final restrictions lifted on July 19, 2021).
+The forecast period also coincides with the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} starting April 2021.
+
 Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
@@
 Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021.
-This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021} as well as the  European Football Championship \citep{dehningImpactEuro20202023}.
+This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} as well as the  European Football Championship \citep{dehningImpactEuro20202023}.
 An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.
@@
 \caption{\bf{Observed cases and deaths of COVID-19 in the UK}.
-A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021.
+Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021.
+The green rectangle marks the study period from May 24 until September 11, 2021.
 Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.}
@@
 Case numbers rose quickly compared to the pre-study period, peaking on July 17, 2021, followed by a trough and a subsequent further increase.
-Forecasts displayed a familiar pattern, where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak.
-Death numbers were increasing more slowly than during the previous peak in January 2021, in line with the beginning of vaccination efforts and a greater natural immunity in the population.
-The peak in case numbers did not correspond to a peak in death numbers, suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms.
+Forecasts displayed a pattern where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak.
+Similar patterns have been observed previously in other short-term forecasts of COVID-19 (see e.g, \citep{sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021, bosseComparingHumanModelbased2022}).
+
+Death numbers during the study period were increasing more slowly than during the previous peak in January 2021, coinciding with the beginning of vaccination efforts and a growing immunity in the population \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023}.
+The peak in case numbers in July 2021 was not followed by a subsequent peak in death numbers (but rather a steady incline over several months), suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms.
 Forecasters tended to overpredict death numbers in the beginning, while underpredicting them in the end, expecting death numbers to fall after the peak in cases.
-The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.
+The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.
 
@@
 During the study period, the case fatality ratio (CFR) likely changed quite quickly compared to the pre-study period.
-On the one hand, the rise of the Delta variant in the UK, which was first discovered in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022}.
+On the one hand, the rise of the Delta variant in the UK, which was first detected in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022} (although \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023} estimated it to be lower than that of the Alpha variant).
 On the other hand, the ongoing COVID-19 vaccination and growing natural immunity in the population had decreasing effects on the CFR.
-In addition, the age distribution of cases changed throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
-Overall, the CFR was lower than during previous peaks of COVID-19.
+In addition, the age distribution of cases changed (hence modifying the overall CFR) throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
+Overall, the CFR was lower than during previous peaks of COVID-19 \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023}.
 One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update.
@@
 Importantly, in this study our combined crowd ensemble ("epiforecasts-EpiExpert") contributed to the European Forecast Hub ensemble.
-This in contrast to the study by \citet{bosseComparingHumanModelbased2022}, where they compared crowd forecasts against a hypothetical ensemble excluding the crowd forecasts.
+This is in contrast to the study by \citet{bosseComparingHumanModelbased2022}, where they compared crowd forecasts against a hypothetical ensemble excluding the crowd forecasts.
 In the original study, including the crowd forecasts improved the Hub ensemble on average (however, the overall number of models included in the German and Polish Hub ensemble was smaller than the number of models in the European Forecast Hub ensemble).
@@
 However, as mentioned before, the CFR likely evolved during the study period.
-Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting forecast accuracy.
+Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting predictive performance.
 They also had no way to adjust the CFR manually.

### 2024-02-22 manuscript/manuscript.tex
Update paper

@@
 \caption{\bf{Predictive performance across forecast horizons.} A-D: WIS stratified by forecast horizon for cases and deaths on the natural and log scale.
-E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.}
+E, F: Empirical coverage of the 50\% and 90\% prediction intervals stratified by forecast horizon and target type.
+Grey dashed lines denote the nominal coverage that a model should ideally achieve.}
 \label{fig:performance}
@@
 Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience.
+One important limitation of the approach is that $R_t$ values were estimated based on reported numbers of cases.
+This is susceptible to changes in testing and reporting and estimated $R_t$ values may not accurately reflect the true underlying transmission dynamics.
 In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more.
@@
 More research is required to obtain a better understanding of the role of subject matter expertise in infectious disease forecasting.
+Similarly, it would be interesting to explore the effects on predictive accuracy of providing forecasters with additional qualitative real-time information such as detailed descriptive reports that enhance the forecasters' understanding of the overall context beyond the numerical data that was visible in our application.
 Our results underline that it is difficult to evaluate forecast performance devoid of context that helps inform what a good or a bad forecast is.

### 2024-02-28 Analysis.rmd
Update date plot

@@
 plot_data_simple <- function(weekly_truth, daily_truth) {
+
+  date_range <- seq(min(weekly_truth$target_end_date), max(weekly_truth$target_end_date), by="2 week")
+
+  daily_truth <- daily_truth |> 
+    filter(target_end_date >= min(weekly_truth$target_end_date))
+
   weekly_truth |>
     mutate(status = str_to_title(status)) |>
-  ggplot(aes(x = target_end_date, y = true_value, colour = "Weekly data")) +
-  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
+    ggplot(aes(x = as.Date(target_end_date), y = true_value, colour = "Weekly data")) +
+    annotate("rect", xmin = as.Date("2021-05-24"), xmax = as.Date("2021-09-11"), ymin = -Inf, ymax = Inf,
              fill = "darkolivegreen3", alpha = .1) + 
+    geom_vline(xintercept = as.numeric(date_range), colour="grey90") + # Add vertical lines
     geom_line() + 
@@
     scale_color_manual(values = c("Weekly data" = "black")) +
+    labs(y = "Observed values", x = "Date", color = " ", fill = " ") + 
     theme_scoringutils() +
-  labs(y = "Observed values", x = "Date", color = " ", fill = " ")
+    scale_x_date(breaks = date_range) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
 }

### 2024-03-18 Analysis.rmd
small updates

@@
 ggsave("output/figures/overview-data.png", plot = p_data, 
-       width = 7, height = 5)
+       width = 7, height = 5, dpi = 300)
+ggsave("output/figures/Figure1.tif", plot = p_data, 
+       width = 7, height = 5, dpi = 300)
 

### 2024-03-18 manuscript/manuscript.tex
small updates

@@
 Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application.
-All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}.
+All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherPreregisteredShorttermForecasting2021a, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}.
 This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants.
@@
 
-If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021}.
+If not otherwise stated, we present results for two-week-ahead forecasts, following the practice adopted by the COVID-19 Forecast Hubs, which found predictive performance to be poor and unreliable beyond this horizon \cite{cramerEvaluationIndividualEnsemble2021, sherrattPredictivePerformanceMultimodel2022, bracherPreregisteredShorttermForecasting2021a}.
 We analysed all forecasts stratified by forecast target (cases or deaths), forecast horizon, and forecast approach.
@@
 Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021.
-This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} as well as the  European Football Championship \citep{dehningImpactEuro20202023}.
+This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} as well as the  European Football Championship \citep{dehningImpactEuro20202023}.Reported cases were likely influenced by an increased uptake of the NHS COVID-19 app in spring and summer 2021 \citep{kendallEpidemiologicalImpactsNHS2023}.
 An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.
@@
 Forecasts displayed a pattern where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak.
-Similar patterns have been observed previously in other short-term forecasts of COVID-19 (see e.g, \citep{sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021, bosseComparingHumanModelbased2022}).
+Similar patterns have been observed previously in other short-term forecasts of COVID-19 (see e.g, \citep{sherrattPredictivePerformanceMultimodel2022, bracherPreregisteredShorttermForecasting2021a, bosseComparingHumanModelbased2022}).
 
